# The Consensus Problem {#sec-consensus}

## Learning objectives

By the end of this chapter the reader should be able
to:

- State the double-spend problem precisely, and
  explain why a signature scheme alone cannot solve it.
- State the Byzantine Generals problem and derive the
  $n \geq 3f+1$ bound for classical Byzantine agreement.
- Describe state-machine replication as the abstraction
  that both permissioned BFT and Nakamoto consensus
  implement.
- Explain why open membership defeats naive
  identity-based voting (the Sybil attack), and why
  identity cannot be assumed as a primitive.
- Explain Nakamoto's substitution of probabilistic
  agreement, weighted by a scarce resource, for the
  deterministic agreement of classical BFT.
- Judge whether a given multi-party data problem calls
  for permissioned BFT, a permissionless chain, or
  neither.

## Orientation

We begin with a problem in the daily conduct of
multi-site clinical research. Consider a phase III
trial run jointly across several hospitals. Each site
enrols participants, assigns them to treatment arms by
randomisation, and records outcome events as they
occur. Each site keeps its own copy of a shared log of
these events, and no single site is trusted to serve as
the sole custodian of the record. Some time later a
regulator, or an independent data-monitoring board,
audits the trial and must be able to reconstruct one
ordered, tamper-evident history of what happened and
when. We shall refer to this arrangement throughout the
book as the trial ledger, and it is the running case
study against which the machinery of the following
chapters is to be measured.

The difficulty is not that any one hospital cannot keep
good records. The difficulty is that the several copies
must agree, and that they must be shown to agree to a
party who trusts none of the sites individually. To see
where the difficulty bites, consider three ways in which
the arrangement can fail.

The first failure is a conflict of order. The same
participant, or equivalently the same randomisation
slot, may be claimed at two sites at once: site 1
records that subject 0041 was enrolled and randomised
to the active arm at 09:14, while site 2 records the
same subject, or the same slot, enrolled one minute
later. Which entry is the true one? A digital signature
on each entry authenticates who wrote it, and proves the
entry was not altered after signing, but a signature
cannot order two conflicting entries in time, nor decide
which of the two the trial ought to honour. In the
vocabulary of digital cash, where the problem was first
studied and where we shall meet it below, this is the
double-spend problem. In the trial ledger it is the
problem of double enrolment.

The second failure is adversarial. Suppose one
hospital's data server is compromised, or merely
misconfigured, and sends one version of its log to the
regulator and a conflicting version to a second site.
The remaining participants are honest, yet they now hold
mutually inconsistent records, and no honest party can
tell, from its own copy alone, which sites have been
told what. Agreement among parties of whom some unknown
subset may send conflicting messages to different
recipients is the Byzantine Generals problem, posed by
Lamport, Shostak and Pease [@lamport1982byzantine], and
it is the abstract core of this chapter.

The third failure appears only once membership is
opened. Suppose that in place of a fixed consortium of
named hospitals we operate an open disease registry that
accepts case reports from anyone, and that we propose to
settle disputes by some form of majority vote among
reporters. Nothing then prevents a single actor from
registering under a thousand invented identities and
voting a thousand times, so that fabricated identities
swamp the honest reports. This is the Sybil attack, and
it is the reason that open-membership agreement is a
genuinely harder problem than the fixed-consortium case.

The three failures line up one to one with three
classical problems, and it is worth fixing that
correspondence before we develop each in turn.
@tbl-failure-modes gives the map, together with the
place in the book where each is treated in full.

| Failure | Trial-ledger instance | Abstract problem | Where treated |
|---|---|---|---|
| **Conflict of order** | Double enrolment on one randomisation slot | Double-spend | This chapter; @sec-pow |
| **Compromised site** | A site sends conflicting logs to different parties | Byzantine agreement | This chapter, the 3f+1 bound; @castro1999practical |
| **Fake identities** | Ballot stuffing in an open registry | Sybil attack | This chapter; @sec-pow |

: The three ways the trial ledger can fail, each an instance of a classical problem with a known treatment {#tbl-failure-modes}

These three failures motivate the chapter, but we should
state plainly at the outset where the argument is going
to land. When the parties are a fixed, named,
contractually accountable consortium, as the trial
hospitals are, the problem is the classical one of
agreement among known participants, solved by the
permissioned protocols we present below, and in a great
many such cases even that machinery is more than the
problem requires: a single signed, append-only database
with an audit log, maintained by one trusted operator,
would serve. The permissionless machinery of the later
chapters answers a different and harder trust problem,
that of the open registry with no accountable membership
at all. We shall be candid about this distinction
wherever it arises, since reaching for the permissionless
tool when the parties are in fact known and few is a
design error, and one this book is written in part to
help the reader avoid.

Digital cash and cryptocurrency furnish the historical
setting in which these problems were first isolated and
solved, and we shall draw on that setting freely, because
the mathematics was worked out there first. The setting,
however, is not the subject. The subject is agreement on
a single ordered history among mutually distrusting
parties, without a trusted referee, and the trial ledger
is our standing example of why a public-health data
system might have need of it. A second example, to which
we return in later chapters, is a vaccine cold-chain
provenance record, in which custody and temperature
readings for each lot are logged by many independent
handlers and must later be reconstructed as one
trustworthy history. This chapter isolates the goal of
agreement from its cryptographic and economic machinery
and studies it as a problem in its own right, because the
problem is old, is well posed mathematically, and was
understood to be hard for decades before Bitcoin existed.

We already possess the tools this chapter requires.
Byzantine agreement is a question about quorums and set
intersection, which is combinatorics we have done
before. The Sybil attack is a question about what a
voting rule assumes about its input space. Nakamoto's
contribution, previewed here and treated in full rigor
in @sec-pow, is a substitution of a deterministic
guarantee for a probabilistic one, precisely the kind of
trade a statistician is trained to evaluate rather than
to accept or reject on faith.

The chapter proceeds in three moves. We show first that
digital cash, and equally the trial ledger, has a
problem that signatures do not solve, namely agreement
on order. We then formalise the general problem of
agreement among possibly-lying parties and derive the
classical bound on how many liars a system can tolerate.
We show last why that classical solution collapses the
moment membership is opened to anyone, and what Nakamoto
substituted in its place.

## The statistician's contribution

1. **Naming the fault model.** Before asking whether a
   consensus protocol is 'secure', a statistician asks
   what it assumes about failures: how many parties may
   deviate, whether they may collude, and whether the set
   of parties is fixed or open. Every security claim in
   this book is conditional on such a model, and no tool
   states the model on our behalf.
2. **Trading determinism for probability.** Classical BFT
   gives a guarantee that holds with certainty once a
   fault threshold is respected. Nakamoto consensus gives
   a guarantee that holds with a probability approaching
   one as data accumulates. Recognizing which regime a
   system occupies, and demanding the relevant
   confidence statement rather than an absolute one, is
   a judgement no protocol makes on our behalf.
3. **Matching the trust model to the problem.** A fixed
   consortium of named, accountable institutions and an
   anonymous open population are different trust
   problems with different correct solutions. Reaching
   for a permissionless chain because it is the
   fashionable tool, when the parties are in fact known
   and few, is a design error a statistician is
   well placed to catch.

## The double-spend problem

Digital cash is a message-passing problem before it is a
cryptographic one, and the trial ledger, as we have
seen, is the same problem in another guise. Suppose party
$A$ holds a token and wishes to pay it to $B$. In a
physical-cash economy, handing over a coin is a single
physical act: $A$ cannot simultaneously hand the same
coin to $C$. A digital token, by contrast, is a string of
bits, and a string of bits may be copied and sent to $B$
and $C$ at once. This is the double-spend problem: absent
physical scarcity, preventing a party from spending the
same balance twice is a coordination problem, not a
copying problem. The reader will recognise the double
enrolment of the Orientation as the same difficulty, with
the coin replaced by a randomisation slot and the two
payees replaced by two trial sites.

A natural first instinct is that a digital signature
solves this. It does not, and seeing precisely why is the
first rigour check of the chapter. A signature scheme
gives $A$ a way to produce, for a message $m$, a value
$\sigma = \mathrm{Sign}(sk_A, m)$ that any holder of
$A$'s public key may verify against $m$. This proves two
things: that $m$ originated from whoever holds $sk_A$
(authentication), and that $m$ has not been altered since
signing (integrity). It proves nothing about *how many
times* $A$ has said something similar, nor about *which*
of two mutually exclusive statements by $A$ ought to be
believed.

Concretely, $A$ can produce two perfectly valid,
independently unforgeable signed messages:

$$
m_1 = \text{"pay coin } c \text{ to } B\text{"}, \qquad
m_2 = \text{"pay coin } c \text{ to } C\text{"}.
$$

Both $\sigma_1 = \mathrm{Sign}(sk_A, m_1)$ and
$\sigma_2 = \mathrm{Sign}(sk_A, m_2)$ verify. Signature
verification is a local, per-message predicate; it
cannot detect that $m_1$ and $m_2$ jointly spend the same
coin twice, because that fact is a relation between two
messages sent to two different recipients, possibly at
the same instant. Detecting it requires every party in
the system to observe the *same total order* over all
messages naming coin $c$, so that whichever of $m_1, m_2$
comes first is the one that is honoured and the other
rejected. That shared total order is consensus. In the
trial ledger the two messages are two enrolment claims on
the same slot, and the shared order is precisely what
allows the regulator to say which claim stands.
Signatures solve authentication; consensus solves order.
Confusing the two is, in our experience, the single most
common conceptual error made by newcomers to this field,
and it merits fixing permanently before we proceed.

::: {.callout-note icon=false collapse=true}
## Check your understanding: signatures and double-spend

Suppose every transaction were signed by a trusted
notary in addition to the payer, and the notary refused
to sign a message that conflicted with one it had
already notarized. Would this solve double-spending
without any further consensus mechanism? What has been
assumed?

The notary's local bookkeeping is exactly a total order
over transactions, so yes, this solves the problem, but
it has silently reintroduced a trusted, single point of
failure and authority: the notary. Distributed consensus
exists to deliver that same total-ordering service
without requiring one party to be trusted.
:::

## Byzantine agreement and the 3f+1 bound

The general problem behind 'get everyone to agree on an
order' was posed abstractly by Lamport, Shostak and Pease
[@lamport1982byzantine] as the Byzantine Generals
problem. We are given a set of $n$ generals, some unknown
subset of whom may be traitors, who must agree on a
common plan of action (attack or retreat) by exchanging
messages, in such a way that all *loyal* generals reach
the same decision, even though traitors may send
conflicting messages to different recipients, lie about
what others said, or say nothing at all. The compromised
hospital server of the Orientation is a traitor in
exactly this sense: it is not merely silent, it sends
inconsistent logs to different honest parties. A protocol
solving this is Byzantine fault tolerant (BFT) if it
guarantees two properties for the loyal generals:

- **Safety.** No two loyal parties ever decide
  differently.
- **Liveness.** Every loyal party eventually decides.

Faults here are *arbitrary* (Byzantine), not merely
crash faults: a faulty party may deviate from the
protocol in any way at all, including strategically, and
the protocol must still work. This is a much harder
requirement than tolerating parties that merely stop
responding, and it is the right model for a security
setting, where a faulty party may be an adversary rather
than a bug.

@fig-topology draws the failure concretely: one site
among five sends two conflicting versions of the shared
log to two honest neighbours, so that no honest party
can tell, from its own copy alone, which history is the
true one.

```{r}
#| label: fig-topology
#| fig-cap: 'A single compromised site breaks agreement by sending two conflicting versions of the shared log to different honest sites, which is Byzantine failure in the trial ledger.'
#| fig-width: 6
#| fig-height: 4
library(ggplot2)

PAL <- c('#2a78d6', '#1baf7a', '#eda100', '#008300',
         '#4a3aa7', '#e34948', '#e87ba4', '#eb6834')

book_theme <- ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(
    panel.grid.minor = ggplot2::element_blank(),
    panel.grid.major = ggplot2::element_line(
      linewidth = 0.25, colour = 'grey88'),
    axis.title = ggplot2::element_text(size = 10),
    legend.position = 'bottom'
  )

struct <- '#33356b'
attack <- '#e34948'

ang <- (seq(90, -198, length.out = 5)) * pi / 180
nodes <- tibble::tibble(
  site = paste('Site', 1:5),
  x = 2.6 * cos(ang),
  y = 2.6 * sin(ang),
  bad = c(FALSE, FALSE, TRUE, FALSE, FALSE)
)
comp <- nodes[nodes$bad, ]
tgt <- nodes[nodes$site %in% c('Site 2', 'Site 4'), ]
seg <- tibble::tibble(
  x = comp$x, y = comp$y,
  xend = tgt$x, yend = tgt$y,
  lab = c('log v1', 'log v2')
)
seg <- dplyr::mutate(seg,
  dx = xend - x, dy = yend - y,
  x = x + 0.30 * dx, y = y + 0.30 * dy,
  xend = xend - 0.32 * dx, yend = yend - 0.32 * dy,
  mx = (x + xend) / 2, my = (y + yend) / 2
)

ggplot() +
  geom_rect(data = nodes,
    aes(xmin = x - 0.62, xmax = x + 0.62,
        ymin = y - 0.32, ymax = y + 0.32, fill = bad),
    colour = NA) +
  geom_text(data = nodes, aes(x, y, label = site),
    colour = 'white', size = 3.3) +
  geom_segment(data = seg,
    aes(x = x, y = y, xend = xend, yend = yend),
    colour = attack, linewidth = 0.7,
    arrow = grid::arrow(length = grid::unit(0.18, 'cm'),
      type = 'closed')) +
  geom_label(data = seg, aes(mx, my, label = lab),
    colour = attack, fill = 'white', linewidth = 0, size = 2.9) +
  scale_fill_manual(
    values = c(`FALSE` = struct, `TRUE` = attack), guide = 'none') +
  coord_equal(clip = 'off') +
  theme_void()
```

### How many liars can be tolerated

Suppose $n$ parties run a quorum-based protocol: a
decision is made once some quorum of size $q$ agrees on
it, and up to $f$ of the $n$ parties are Byzantine. Two
requirements pin down the minimum $n$ as a function of
$f$.

**Safety requires overlapping quorums.** If two decisions
could be made by two disjoint-enough quorums, the
Byzantine parties could feed conflicting messages to each
quorum and cause two different loyal decisions. To rule
this out, any two quorums must share at least one *loyal*
member. Two quorums of size $q$ drawn from $n$ parties
overlap in at least $2q - n$ members; even if every
Byzantine party is placed in the overlap, a loyal member
survives in the overlap only if

$$
2q - n \; \geq \; f + 1.
$$

**Liveness requires an achievable quorum.** A quorum must
be reachable using loyal parties alone, since Byzantine
parties cannot be compelled to participate honestly. With
$f$ possible defectors, this requires

$$
q \; \leq \; n - f.
$$

Combining the two, the smallest workable $q$ satisfies
$q \geq \lceil (n+f+1)/2 \rceil$, and feasibility requires
this value not to exceed $n - f$. Substituting and
simplifying gives the classical bound:

$$
n \; \geq \; 3f + 1.
$$

With fewer than $3f+1$ parties, no protocol, however
clever, can guarantee both safety and liveness against
$f$ Byzantine parties: an adversary controlling $f$ of a
smaller system can always engineer either two
inconsistent decisions or a stall. The worked example
below checks this bound numerically by testing quorum
feasibility directly, rather than asking the reader to
take the algebra on trust.

@fig-quorum-intersection shows why the overlap argument
forces the bound. With n = 3f+1 sites, two quorums of
size 2f+1 must share f+1 members, and since at most f
of those shared members can be Byzantine, at least one
is guaranteed to be honest.

```{r}
#| label: fig-quorum-intersection
#| fig-cap: 'Two quorums of size 2f+1 drawn from 3f+1 sites must share at least f+1 members, so their overlap always contains at least one honest site.'
#| fig-width: 6.5
#| fig-height: 3.4
f <- 2
n <- 3 * f + 1
nd <- tibble::tibble(x = 1:n, y = 0)
bands <- tibble::tibble(
  xmin = c(0.5, 2.5, 5.5),
  xmax = c(2.5, 5.5, 7.5),
  fillc = c('grey88', '#d6e6f8', 'grey88')
)

ggplot() +
  geom_rect(data = bands,
    aes(xmin = xmin, xmax = xmax, ymin = -0.55, ymax = 0.55),
    fill = bands$fillc, colour = NA) +
  geom_point(data = nd, aes(x, y), size = 5, colour = struct) +
  annotate('text', x = 1.5, y = 0.88, label = 'A only',
    size = 2.9, colour = struct) +
  annotate('text', x = 4, y = 0.88,
    label = 'overlap = f+1 = 3\nat least one honest',
    size = 2.9, colour = '#1d5a9e', lineheight = 0.9) +
  annotate('text', x = 6.5, y = 0.88, label = 'B only',
    size = 2.9, colour = struct) +
  annotate('segment', x = 0.6, xend = 5.4, y = 1.6, yend = 1.6,
    colour = PAL[1], linewidth = 0.8) +
  annotate('text', x = 3, y = 1.95,
    label = 'Quorum A: 2f+1 = 5 sites', colour = PAL[1], size = 3.1) +
  annotate('segment', x = 2.6, xend = 7.4, y = -1.1, yend = -1.1,
    colour = PAL[2], linewidth = 0.8) +
  annotate('text', x = 5, y = -1.45,
    label = 'Quorum B: 2f+1 = 5 sites', colour = PAL[2], size = 3.1) +
  coord_cartesian(clip = 'off') +
  theme_void()
```

::: {.callout-note icon=false collapse=true}
## Check your understanding: why 3f, not 2f

A student argues that tolerating $f$ liars among $n$
parties should only require a strict majority, i.e.
$n \geq 2f+1$, since a majority already outvotes the
liars. Where does the argument fail?

A strict majority is enough to *outvote* liars in a
single vote, but Byzantine agreement requires that *any
two* quorums used across the protocol's messages still
intersect in a loyal party, and it requires that a quorum
be achievable without any cooperation from the (unknown)
faulty parties. A bare majority quorum can be entirely
disjoint from another bare majority quorum once one
removes the possibly-faulty parties from each,
breaking the safety property. The extra factor of $f$
is the price of never needing to trust which parties are
faulty.
:::

## State-machine replication as the abstraction

Byzantine agreement generalizes to a useful engineering
abstraction called state-machine replication (SMR): a
service is defined by a deterministic state machine (a
transition function applied to an ordered log of
inputs), and the log is replicated across $n$ parties
using a BFT agreement protocol so that every loyal
replica applies the same inputs in the same order and
therefore reaches the same state. A ledger is exactly
this: the 'inputs' are transactions, the 'state' is the
set of account balances or unspent outputs, and the
'transition function' is 'apply the transaction if it is
valid, else reject it.' The trial ledger is an instance
of the same abstraction, in which the inputs are
enrolment, randomisation and outcome events, the state is
the current enrolment and endpoint tally, and the
transition function admits an event if it is well formed
and consistent with the log so far. Every consensus
protocol in this book, from PBFT to Nakamoto's chain to
the proof-of-stake protocols of @sec-pos, is best
understood as an implementation of SMR under a particular
fault model and a particular membership assumption.

Castro and Liskov [@castro1999practical] gave the first
BFT SMR protocol efficient enough for real deployment:
Practical Byzantine Fault Tolerance (PBFT). PBFT runs
among a *fixed, known*
set of $n$ replicas, tolerates up to
$f = \lfloor (n-1)/3 \rfloor$ Byzantine replicas exactly
at the $3f+1$ bound, and reaches deterministic agreement
in a small, bounded number of message rounds per decision.
It is the template for every permissioned (consortium)
blockchain in production today.

The catch is in the phrase 'fixed, known set.' PBFT and
its relatives require that the identities of the $n$
participants be established in advance, typically by a
credentialing authority admitting named institutions.
This is a perfectly reasonable requirement in many real
settings, and a fatal one in others, which is exactly the
distinction the next section makes precise.

## Sybil attacks and the limits of majority vote

Suppose we relax PBFT's fixed-membership assumption, as
the open disease registry of the Orientation does, and
allow anyone to claim a seat in the consensus protocol
by, say, announcing a network address or registering as a
reporter. Nothing then prevents a single adversary from
announcing a thousand addresses. Douceur
[@douceur2002sybil] named this the Sybil attack and
established a stronger result: absent some mechanism that
makes acquiring an identity costly, no distributed
protocol can bound the number of identities a single
adversary controls, and therefore no identity-counting
rule (one address, one vote) can be made robust against a
resource-poor adversary who merely mints more identities.

This is not a narrow technical gap in PBFT; it is a
structural fact about open membership. The $3f+1$ bound
assumes $f$ is a bounded *fraction of a bounded set of
identities*. Once identity creation is free, $f$ is
unbounded relative to $n$ measured in identities, and the
bound is vacuous. A voting rule that counts identities is
the wrong object to defend, no matter how the counting is
implemented.

To illustrate, consider the open registry populated by
one hundred honest reporters, each of whom files a
truthful case report, and a single adversary who mints
one hundred and fifty free identities and files a
fraudulent report under each. The first code chunk below
makes the consequence concrete. Under a rule of one
identity, one vote, the fraudulent outcome commands a
majority while representing exactly one real,
resource-constrained actor.

```{r ch01-sybil-vote}
set.seed(20260702)
n_honest <- 100
n_sybil <- 150

# SYNTHETIC: identity-counted votes under free identity
# creation. Each 'A' is one honest participant; each 'B'
# is one Sybil identity minted by a single adversary.
votes <- c(rep('A', n_honest), rep('B', n_sybil))
table(votes)

prop_fraudulent <- mean(votes == 'B')
prop_fraudulent
```

A single actor with no special resources outvotes the
hundred honest reporters, because the ballot counted
identities and identities were free. The remedy is not a
better counting rule; it is to refuse to count identities
at all, and to weight influence instead by something the
adversary cannot mint for free. The second chunk repeats
the experiment, granting each honest reporter one unit of
a scarce resource and granting the adversary a *fixed
total* budget of that resource, which it may split across
as many Sybil identities as it likes.

```{r ch01-resource-weighted-vote}
set.seed(20260702)
r_honest <- rep(1, n_honest)

# SYNTHETIC: the adversary's total resource budget is
# fixed regardless of how many identities it mints.
# Splitting it more finely does not create more of it.
total_attacker_resource <- 30
r_sybil <- rep(total_attacker_resource / n_sybil, n_sybil)

weights <- c(r_honest, r_sybil)
side <- c(rep('A', n_honest), rep('B', n_sybil))
tapply(weights, side, sum)
```

The fraudulent side's total weight is fixed at 30 no
matter how many identities it is divided among; minting
more Sybils only slices the same budget more thinly. This
is the general principle: influence proportional to a
scarce, externally verifiable resource is Sybil-resistant
by construction, while influence proportional to
identity-count is not. Which resource to use, and how to
verify a claim on it without a trusted registrar, is
exactly the question @sec-pow answers for computation and
@sec-pos answers for stake.

::: {.callout-important}
## The sceptic's reflex

'Blockchain solves the Byzantine Generals problem' is a
narrative-tier claim as usually stated. Classical BFT,
solved by @castro1999practical, already solves Byzantine
Generals under a fixed-membership fault model. What
Nakamoto consensus adds is a solution to a harder,
different problem: Byzantine-style agreement under *open*
membership, where the number and identity of
participants is unknown and unbounded. Naming which
problem is being solved, fixed membership or open
membership, converts this from a marketing line into a
protocol fact.
:::

## Nakamoto's move: probabilistic agreement by scarce resource

Nakamoto's [@nakamoto2008bitcoin] contribution, in the
vocabulary now established, is to replace identity-counted
voting with resource-weighted influence, using
computational work as the scarce, externally verifiable
resource, and to accept a probabilistic form of agreement
in place of PBFT's deterministic one. Chapter @sec-pow
develops the full mechanism; here we note only the shape
of the trade.

Instead of $n$ known parties each casting one vote,
anyone may participate, and a participant's influence
over which history is canonical is proportional to the
computational work it has expended, verified cheaply by
everyone via a cryptographic puzzle. An adversary who
controls a minority of the total computational capacity
can, with probability that *decays exponentially* in the
number of confirmations, temporarily produce a longer
competing history, but cannot do so as a matter of
certainty, nor indefinitely. Safety is no longer a
theorem holding with probability one under a stated fault
bound; it is a tail probability that the reader,
possessing a Poisson-process and random-walk toolkit
already, is equipped to compute directly. @sec-pow does
exactly that computation.

This is not the first use of costly computation to deter
an adversary. Dwork and Naor [@dwork1993pricing] proposed
spending computation as a toll to deter denial-of-service
attacks and junk email, and Back [@back2002hashcash]
operationalised the idea as a practical proof-of-work
stamp. Nakamoto's innovation was to repurpose that toll,
not as an anti-spam fee but as the scarce resource that
determines *whose vote on transaction order counts*,
thereby converting a spam deterrent into a Sybil-resistant
consensus weight, and coupling it to a longest-valid-chain
rule that gives the system a mechanical way to resolve
competing histories without a vote in the classical sense
at all.

::: {.callout-note icon=false collapse=true}
## Check your understanding: what changed and what did not

Compare PBFT and Nakamoto consensus along three axes:
membership, the nature of the safety guarantee, and the
basis of a participant's influence. What is genuinely new
and what is a repackaging of an older idea?

Membership: fixed and known (PBFT) versus open and
unbounded (Nakamoto). Safety guarantee: deterministic
once $f$ respects $3f+1$ (PBFT) versus probabilistic,
improving with confirmations (Nakamoto). Basis of
influence: identity (PBFT's one-replica-one-vote among
admitted members) versus a scarce, externally verifiable
resource (Nakamoto's proof-of-work). The genuinely new
piece is Sybil-resistant weighting under open membership;
the toll-for-computation idea it is built from predates
Nakamoto by over a decade [@dwork1993pricing;
@back2002hashcash].
:::

## Permissioned versus permissionless: an engineering choice

None of the foregoing ranks one design above the other.
Permissioned BFT (PBFT and its modern descendants) gives
deterministic finality, high throughput, and low latency,
at the cost of requiring a credentialing process that
establishes who the $n$ parties are. Permissionless
Nakamoto-style consensus gives Sybil-resistant, open
membership with no credentialing authority at all, at the
cost of probabilistic finality, materially higher latency
before a transaction is safe to treat as settled, and (in
the proof-of-work instance) a real resource expenditure
that purchases the Sybil resistance.

The correct choice is a function of the trust
environment, not a matter of preference for
'decentralization' as a general good. When the set of
participants is small, known, accountable to a
regulator or contract, and stable over time,
permissioned BFT typically dominates on every operational
axis. When participation must be open to parties unknown
in advance, unaccountable to any shared authority, and
free to enter or leave, only a Sybil-resistant,
permissionless design is structurally viable, and its
extra cost is the price of that openness.

@tbl-three-designs sets the two consensus designs
beside the humbler option we keep returning to, a
single signed and audited database, along the axes
that decide between them.

| Property | **Permissioned BFT** | **Permissionless Nakamoto** | **Signed audited database** |
|---|---|---|---|
| **Membership** | Fixed, credentialed | Open, unbounded | Single operator |
| **Sybil resistance** | Not needed, set is known | By scarce resource | Not applicable |
| **Finality** | Deterministic, sub-second | Probabilistic, grows with confirmations | Immediate, one writer |
| **Throughput** | High | Low to moderate | Highest |
| **Right tool when** | Parties known, few, mutually distrusting | Membership open and unaccountable | One trusted operator suffices |

: Three designs for a shared record and the trust environment each one fits {#tbl-three-designs}

To illustrate the fixed-membership case with the trial
ledger in view, consider how the tolerable number of
faulty sites grows with the size of the consortium. Under
the classical bound a consortium of $n$ named
institutions can tolerate at most
$f = \lfloor (n-1)/3 \rfloor$ simultaneously Byzantine
members, those whose servers are compromised,
misconfigured, or otherwise misreporting, while still
guaranteeing agreement on the trial record.

```{r ch01-consortium-tolerance}
# SYNTHETIC: Byzantine fault tolerance of a fixed
# consortium ledger as a function of the number of
# named institutions, via f = floor((n - 1) / 3).
consortium <- tibble::tibble(n_sites = c(4, 7, 10, 13, 16))

consortium |>
  dplyr::mutate(
    max_faulty = (n_sites - 1L) %/% 3L,
    honest_needed = n_sites - max_faulty
  )
```

A consortium of thirteen parties, the twelve hospitals
and one regulator of the callout below, may thus tolerate
up to four compromised or misreporting members while
still reconstructing one agreed history, and it requires
nine honest members to do so. Should the consortium later
wish to tolerate more faults, the bound becomes a design
constraint on how many institutions must be enrolled, a
point we take up again in the exercises.

Plotting the same bound over a range of consortium
sizes, @fig-fault-tolerance shows that tolerance rises
as a step function: three further institutions buy
exactly one more tolerable fault, with the thirteen-party
consortium of the callout below sitting on the f = 4
tread.

```{r}
#| label: fig-fault-tolerance
#| fig-cap: 'Tolerance rises as a step function of consortium size, gaining one more tolerable Byzantine site only when three further institutions join.'
#| fig-width: 6
#| fig-height: 3.4
tol <- tibble::tibble(n = 1:19, f = (n - 1L) %/% 3L)
mark <- tibble::tibble(n = 13, f = 4L, lab = 'n = 13 tolerates f = 4')

ggplot(tol, aes(n, f)) +
  geom_step(colour = PAL[1], linewidth = 0.8) +
  geom_point(data = mark, aes(n, f), colour = PAL[3], size = 2.6) +
  geom_text(data = mark, aes(n, f, label = lab), colour = PAL[3],
    size = 3, hjust = 1.1, vjust = -0.8) +
  scale_x_continuous(breaks = seq(1, 19, 3)) +
  scale_y_continuous(breaks = 0:6) +
  labs(x = 'Consortium size n', y = 'Tolerable Byzantine sites f') +
  book_theme
```

::: {.callout-warning}
## Multi-hospital trial data: which tool, honestly

This is the trial ledger of the Orientation, now regarded
as an engineering choice. We have a phase III trial run
across a consortium of, say, twelve hospitals and a
regulator, none of whom fully trusts the others to
control the shared enrolment and endpoint log
unilaterally, but all of whom are named, contracted, and
accountable institutions. This is a textbook
fixed-membership setting: $n = 13$ is known, $f$ may be
bounded by contract (how many parties the consortium is
willing to assume might misreport or go offline), and a
PBFT-style consortium ledger delivers tamper-evident,
jointly-audited agreement on the trial record with
sub-second finality and no proof-of-work energy cost.
Reaching instead for a permissionless chain, with its
unbounded, anonymous validator set and probabilistic
finality, buys Sybil resistance the problem never needed,
at real cost in latency, throughput, and operational
complexity. The right engineering answer here is a
permissioned ledger, or, in a great many such consortia,
a conventional database with cryptographic audit logging
and no consensus protocol running at all, per the warning
in @sec-intro. We reserve the permissionless machinery
for the case that genuinely has open, unaccountable
membership, such as a public disease registry meant to be
independently verifiable by parties who will never be
enumerated in advance.
:::

## Permissioned ledgers and the health deployments

We have argued that the trial ledger, a fixed and named
consortium of accountable hospitals, is the permissioned
case, and we now take that case seriously. We do so for
two reasons: it is the honest setting for our running
example, and it is where very nearly every deployed
health-sector ledger actually lives. The open-membership
machinery that occupies the remainder of this book is,
for the health scientist, the exception and not the rule.

What does a ledger look like when the membership is
known in advance? The most widely deployed permissioned
design, Hyperledger Fabric, is instructive precisely
because it departs from the Bitcoin template. The public
chains of the preceding sections follow an order-execute
discipline: the consensus protocol orders the
transactions first, and every node then executes all of
them in that fixed order. Fabric, as Androulaki and
colleagues describe it [@androulaki2018fabric], rearranges
this into an execute-order-validate pipeline. A
transaction is first executed, or endorsed, by a
nominated subset of peers, who simulate it against their
copy of the state and sign the result; the signed
endorsements are then ordered by a separate ordering
service; and each peer finally validates the ordered
transactions against an endorsement policy before
committing them. The rearrangement is not cosmetic. It
lets mutually distrusting organisations agree on state
without requiring every node to run every computation in
lockstep, and it isolates the ordering service as the one
component that must be agreed upon.

The deeper point, for our purposes, is what a known
membership buys. When the writers are a fixed roster of
named institutions, the Sybil attack of the previous
section simply cannot arise, since no party can mint new
identities that the others will honour. The problem
reduces to classical Byzantine agreement among a known
set, which Castro and Liskov's practical protocol
[@castro1999practical] solves at the 3f+1 threshold we
derived, and proof-of-work, with its deliberate and
enormous energy cost, becomes unnecessary. We should be
precise, however, about how much fault tolerance a
deployed system actually buys, since the table below
names classical Byzantine agreement as the permissioned
answer and practice is often weaker. Fabric, for one,
has in its common configurations ordered transactions
with a crash-fault-tolerant protocol such as Raft, which
tolerates orderers that halt but assumes they do not lie;
full Byzantine ordering is available but is not the
default. The distinction is itself a trust decision:
Byzantine tolerance guards against a corrupted insider at
the ordering layer, crash tolerance only against one that
fails honestly, and a consortium must decide which threat
it is buying against. With that caveat noted, a
permissioned ledger recovers high throughput, low
latency, and negligible energy, at the price of the very
thing that makes public chains interesting: it
presupposes that the membership can be agreed upon in the
first place.

: The two architectures contrasted. The rearrangement of steps is what lets a permissioned ledger dispense with proof-of-work. {#tbl-fabric}

| Property | Public chain (order-execute) | Permissioned (execute-order-validate) |
|---|---|---|
| **Membership** | Open, unaccountable | Fixed, named, accountable |
| **Sybil resistance** | Costly scarce resource | Not needed |
| **Consensus** | Nakamoto or stake | Classical BFT [@castro1999practical], or crash-fault-tolerant in practice |
| **Energy** | Large by design | Negligible |
| **Throughput** | Low | High |
| **Health fit** | Rare | The usual case |

Where, then, has the health sector actually put these
ideas to work? The literature is by now sizeable.
Azaria and colleagues' MedRec [@azaria2016medrec] proposed
a ledger for medical-record access and permission
management, and it remains the most cited work in the
area. The informatics community's own assessments
[@kuo2017blockchain], the systematic review of Agbo and
colleagues [@agbo2019systematic], proposals for
clinical-trial provenance [@benchoufi2017trials], designs
for patient-driven interoperability [@gordon2018interoperability],
and integrations with the FHIR health-data standard
[@zhang2018fhirchain] together describe an active decade
of work. A public-health reader could be forgiven for
concluding that the matter is settled and the ledgers are
in the clinics.

Unfortunately, they are largely not, and it is worth
asking why. The systematic reviews are consistent on this
point: the great majority of these efforts are pilots and
proofs of concept, and few have reached production scale.
The explanation is not a failure of engineering but a
failure of fit, and Wust and Gervais give the decision
criterion that makes the failure legible
[@wust2018need]. A blockchain earns its considerable cost,
they argue, only when three conditions hold together:
there are multiple writers, no trusted third party is
available, and the writers' interests may conflict. Many
health proposals satisfy the first and third conditions
but not the second, because a trusted custodian, a
hospital's information-technology department or a trial's
data-coordinating centre, is in fact available and
acceptable. When such a custodian exists, a well-governed
database with cryptographic audit logging delivers
tamper-evidence at a fraction of the cost, and the ledger
is machinery in search of a problem.

::: {.callout-important}
## The sceptic's reflex

'Blockchain for health records' is, in the majority of
the proposals that carry the phrase, a tier-four
narrative wearing the costume of a tier-one protocol
fact. The reflex is to apply the Wust and Gervais test
before believing the pitch: name the writers, ask whether
a trusted custodian is genuinely unavailable, and ask
whether their interests actually conflict. Most proposals
fail at the second question.
:::

This returns us to the trial ledger, and to an honest
verdict. Of the health cases in the literature, a fixed
consortium of hospitals sharing a trial record is among
the strongest, since the distrust among named parties is
real and the regulatory demand for an independent audit
is genuine. Even here, however, the permissioned ledger
does not win by default. It competes against a simpler
arrangement in which one operator maintains a signed,
append-only log that every site replicates and checks, and
the choice between them turns on a single question: can
the consortium agree on who operates the log? If it can,
the database is the better tool; if it genuinely cannot,
the permissioned ledger earns its place. We shall keep
this question in view as the machinery grows more
elaborate in the chapters ahead.

::: {.callout-note icon=false collapse=true}
## Check your understanding: does the trial ledger need a chain?

A pharmaceutical sponsor proposes to run the trial ledger
on a public blockchain 'for maximum transparency and
immutability'. Applying the Wust and Gervais criterion,
what is the first question to ask, and what cheaper design
should be the comparator?

The first question is whether a trusted custodian is
genuinely unavailable, because if the sites will accept
any single operator, no blockchain of any kind is
warranted. A public chain is doubly wrong here: it exposes
enrolment and outcome data to the world, which the
hashing of @sec-intro shows is not anonymised by hashing
alone, and it pays proof-of-work costs to solve a Sybil
problem that a closed consortium does not have. The
correct comparator is a signed, replicated, append-only
audit log, and only if the sites cannot agree on its
operator does a permissioned consortium ledger become the
answer.
:::

## Worked example

The quorum-feasibility argument behind $n \geq 3f+1$ was
given algebraically above. This worked example checks it
computationally: for a range of fault counts $f$, is a
quorum of the required size simultaneously (a) large
enough to guarantee two quorums share a loyal member, and
(b) small enough to be reachable using loyal parties
alone? The function below implements exactly the two
inequalities derived earlier, and nothing else.

```{r ch01-quorum-bound}
quorum_exists <- function(n, f) {
  q_min <- ceiling((n + f + 1) / 2)
  q_min <= n - f
}

f_vals <- 1:6

# n = 3f: one short of the classical bound
purrr::map2_lgl(3 * f_vals, f_vals, quorum_exists)

# n = 3f + 1: exactly at the classical bound
purrr::map2_lgl(3 * f_vals + 1, f_vals, quorum_exists)
```

The first line is `FALSE` at every tested $f$: with only
$n = 3f$ parties, no quorum size satisfies both the
safety and the liveness inequality simultaneously, so no
protocol relying on quorum intersection can guarantee
both properties. The second line is `TRUE` at every
tested $f$: adding a single additional party, reaching
$n = 3f+1$, is both necessary and sufficient. The bound
derived by hand in the previous section is not an
approximation or a rule of thumb; it is exact, and the
simulation above is merely evaluating the same two
inequalities Lamport, Shostak, and Pease established, at
a handful of parameter values, as a check that the algebra
was not mis-transcribed.

## Collaborating with an LLM

**Prompt:** 'Explain why proof-of-work solves the
Byzantine Generals problem.'
**Watch for:** Most LLM answers to this prompt conflate
the *classical*, fixed-membership Byzantine Generals
problem, already solved by PBFT, with the *open-
membership* Sybil-resistance problem that proof-of-work
actually addresses. An answer that does not distinguish
the two, or that claims proof-of-work gives a
deterministic guarantee, is repeating a common but
imprecise account.
**Verification:** Ask the model to state explicitly
whether its claimed guarantee is deterministic or
probabilistic, and under what fault model. Cross-check
the stated fault model against @castro1999practical for
the permissioned case and against the tail-probability
argument of @sec-pow for the permissionless case.

**Prompt:** 'Derive the minimum number of nodes needed to
tolerate f Byzantine faults.'
**Watch for:** Models frequently state $n \geq 3f+1$
correctly as a fact but skip or garble the derivation,
particularly the step requiring quorum overlap to contain
a *loyal* member rather than merely a nonempty
intersection.
**Verification:** The reader may reproduce the two
inequalities, $2q - n \geq f+1$ and $q \leq n-f$,
independently, and check that the model's derivation
actually combines them rather than asserting the bound.
The `quorum_exists()` function
above is a fast independent check at specific values of
$n$ and $f$.

**Prompt:** 'Design a consensus mechanism for a
multi-hospital clinical trial consortium.'
**Watch for:** A default LLM answer often proposes a
permissionless, proof-of-work or proof-of-stake design by
association with 'blockchain' generally, without asking
whether membership is in fact fixed and known.
**Verification:** Before accepting any design, establish
whether the participant set is fixed, named, and
contractually accountable. If so, insist the model justify
any permissionless component against a plain PBFT-style
consortium ledger, or a conventional audited database, on
concrete operational grounds rather than by assertion.

## Exercises

1. A protocol uses quorums of size $q = \lceil (n+1)/2
   \rceil$ (a bare majority) among $n$ parties, of which
   $f$ may be Byzantine. Derive the largest $f$, as a
   function of $n$, for which this quorum size still
   satisfies both the overlap inequality and the
   achievability inequality from this chapter. Compare it
   to the $3f+1$ bound and explain the gap.
2. Extend the `quorum_exists()` function to also return the
   minimal feasible $q$ itself (not just a logical), and
   use it to tabulate, for $f = 1, \dots, 10$, the smallest
   $n$ at which a feasible quorum first appears. Confirm
   the pattern matches $n = 3f+1$ exactly.
3. Modify the Sybil-vote simulation so that the adversary's
   total resource budget, rather than being fixed at 30, is
   drawn as `total_attacker_resource <- n_honest * runif(1, 0, 1)`
   for a single simulated draw. At what fraction of
   `n_honest` does the resource-weighted vote flip in the
   adversary's favor? Is this threshold sensitive to
   `n_sybil`? Explain why or why not, referring to the
   mechanics of the simulation.
4. Prove that no protocol using a single, fixed quorum size
   $q$ can be both safe and live against $f \geq n/3$
   Byzantine parties, without assuming the specific
   inequalities of this chapter as given; construct the
   adversary strategy directly (i.e. exhibit two quorums an
   adversary can force to disagree).
5. A consortium of five named hospitals wants Byzantine
   fault tolerance against the possibility that any single
   member's system is compromised or misreports. Using the
   $3f+1$ bound, determine the largest $f$ the consortium
   can tolerate, and state what changes about the required
   design if the consortium later wants to tolerate two
   simultaneous compromised members instead.
6. Write one paragraph arguing, in the tier-language of
   @sec-intro, whether 'a permissionless blockchain is
   more secure than a permissioned consortium ledger' is a
   protocol fact, an empirical claim, an economic claim, or
   a narrative, as usually stated. Identify the missing
   qualifiers that would be needed to promote it to a
   tier-one or tier-two claim.

## Further reading

@lamport1982byzantine is the founding paper of this entire
chapter and should be read before anything else; the
generals allegory is a device, but the impossibility and
possibility results are the mathematical content. Read
@castro1999practical next for the first practical,
deployable BFT protocol, and note precisely which
assumptions (fixed membership, known $n$, partial
synchrony) make its efficiency possible.
@douceur2002sybil is short and should be read in full; it
is the paper that explains why every naive open-membership
voting scheme one might invent independently has already
been shown broken. @nakamoto2008bitcoin is the primary
source for the resolution this chapter previews and
@sec-pow develops in full; read it before any secondary
account. @narayanan2016bitcoin gives the fullest rigorous
textbook treatment of the material in this chapter and the
next several. @bonneau2015sok is the standard systematization
of the broader research literature this chapter draws its
first vocabulary from, and is the right next stop once the
primary sources are read. @dwork1993pricing and
@back2002hashcash are the pre-Bitcoin lineage of
'computation as a costly toll'; reading them clarifies
exactly what Nakamoto added and what was already known.

In conclusion, three points are to be emphasised. First,
the object every protocol in this book pursues is
agreement on one ordered history among mutually
distrusting parties, and for the trial ledger it is this
object, not any particular technology, that a regulator
ultimately audits. Second, the classical theory is sharp
and known: agreement among $n$ fixed, identified parties
tolerating $f$ Byzantine faults is possible precisely when
$n \geq 3f+1$, and no cleverness recovers the guarantee
below that threshold. Third, the permissionless machinery
of the later chapters earns its considerable extra cost
only under open, unaccountable membership, as in a public
disease registry, whereas when the participants are in
fact a fixed, named consortium, as the trial hospitals
are, a permissioned ledger, or even a well-governed
audited database, is the more responsible choice.
