# Cryptographic Primitives {#sec-crypto}

::: {.bookepigraph}
```{=latex}
\begin{flushright}\small
```
*We stand today on the brink of a revolution in
cryptography.*

[Diffie and Hellman]{.smallcaps}, *New Directions in
Cryptography* (1976)
```{=latex}
\end{flushright}
```
:::

## Learning objectives

By the end of this chapter the reader should be
able to:

- State the random-oracle idealisation of a
  cryptographic hash and name the five
  properties (determinism, preimage resistance,
  collision resistance, avalanche, uniformity)
  it is meant to capture.
- Derive the birthday bound and explain why
  collision resistance ($\sim 2^{n/2}$) is the
  harder property to guarantee than preimage
  resistance ($\sim 2^{n}$).
- Verify the avalanche and uniformity properties
  empirically in R and read the statistical
  tests as evidence for the random-oracle model.
- Construct a Merkle tree, produce a logarithmic
  inclusion proof, and verify membership without
  the underlying data.
- Explain, from the definitions of a finite field
  and a cyclic group, the elliptic-curve group law
  and why prime order and discrete-log hardness are
  what a signature scheme rests on.
- Write the ECDSA signing and verification
  equations, prove that an honest signature
  verifies, and show by direct algebra how nonce
  reuse recovers the private key.
- Explain why Schnorr's linearity permits signature
  aggregation and what Bitcoin's Taproot upgrade
  bought with it, and read the BLS pairing scheme as
  a cited interface.
- Judge when a published Merkle root is the right
  tool for dataset integrity and when it is not.

## Orientation

We begin with a problem from the conduct of a
clinical trial. A multi-institution trial, run
across several hospitals with no single trusted
custodian of the data, reaches its lock date. At
that moment the sites wish to commit irrevocably
to the dataset as it stands on that date: the
enrolments, the randomisation assignments, and the
outcome events recorded to date. Analysis and
unblinding will follow, and a regulator may, months
or years later, wish to verify that the dataset
examined at audit is exactly the one that was
locked, that no record was quietly altered, added,
or removed in the interval. We shall refer to this
shared, append-only record, maintained across
institutions with no single custodian, as the trial
ledger, and it recurs throughout the book. The
question before us is narrow and answerable: by what
mechanism can the sites commit to the locked data in
a way that a later auditor can check?

The machinery that answers it is the subject of this
chapter. The hash of the locked dataset is a short
fingerprint that changes if any byte of the data
changes. A Merkle tree built over the per-record
hashes lets an auditor verify a single record
without possessing the whole database. Digital
signatures attach to each entry the identity of the
coordinator who made it, so that attribution of an
eCRF record (an electronic case report form, the
record of one subject's data in the trial) becomes
itself cryptographically checkable. Digital cash and cryptocurrency are where
these primitives were sharpened, and we shall keep
that history in view; but for our purposes they are
the historical setting and not the frame.

The primitives are usually taught as engineering
artefacts, a matter of invoking a library and
obtaining a guarantee. For a reader with graduate
probability and algebra they are better understood
as mathematical objects with statistical behaviour
that can be measured. A
cryptographic hash is, to the precision that matters
here, a random function. An elliptic curve over a
finite field is a finite abelian group in which one
direction of arithmetic is easy and its inverse is
believed hard. Every security property we shall
invoke is a statement about the difficulty of a
search, and the difficulty of a search is a
probability calculation.

That reframing pays immediately. The reason six
confirmations suffice in @sec-pow, the reason a
light client can trust a block header without the
block body, and the reason a public key does not
leak its private key, are all consequences of two
numbers: the expected work to find a preimage and
the expected work to find a collision. We shall
derive both, and then confirm the underlying
random-oracle assumption by hashing tens of
thousands of strings in R and subjecting the output
to the same goodness-of-fit tests we would apply to
any claimed random-number generator.

Here the empirical stance deserves emphasis, and it
is worth stating in a form familiar to any reader
who has validated an assay before trusting its
readings. We do not take the random-oracle model on
faith. Cryptographers prove theorems about these
objects under stated hardness assumptions; a
statistician can, in addition, treat the deployed
hash as an unknown distribution and test its
observable consequences, exactly as one runs known
samples through a new assay to confirm that it
behaves as specified before admitting its readings
into evidence. The tests in this chapter do not
prove SHA-256 secure, and we shall be careful to say
what they do and do not establish. They verify that
its output is consistent with the random-oracle
model, which is the assumption every later
probability calculation silently uses.

::: {.callout-note}
## The vocabulary of this chapter

The argument turns on a handful of terms that a
first-year public-health reader is unlikely to
have met. We define them here and use them
consistently thereafter.

- **Random oracle.** An idealised hash function
  imagined as assigning each distinct input an
  independent, uniformly random output. It is the
  working model under which every probability in
  the chapter is computed.
- **Preimage resistance.** The property that,
  given a digest, no feasible search recovers an
  input producing it; the expected work is about
  $2^{n}$ for an $n$-bit digest.
- **Collision resistance.** The property that no
  feasible search finds two distinct inputs
  sharing a digest. The birthday bound makes this
  cost about $2^{n/2}$, far less than preimage
  resistance.
- **Birthday bound.** The result that a collision
  among random draws from a space of size $N$
  becomes likely after only about $\sqrt{N}$
  draws rather than $N$; it is the familiar
  birthday problem applied to digests.
- **Merkle tree.** A hash-based summary that
  commits to many records at once. Its top hash,
  the *Merkle root*, is a single fingerprint of
  the whole set, and an *inclusion proof* is the
  short list of hashes showing that one record
  belongs to it.
- **Discrete-logarithm problem.** The problem of
  recovering the scalar $d$ from the point $dG$ in
  an elliptic-curve group. Its presumed hardness
  is the wall between a public key and the private
  key behind it.
- **Commitment.** A short value binding a party to
  hidden data (*binding*) while revealing nothing
  about it until it is opened (*hiding*).
- **Zero-knowledge proof.** A protocol by which
  one party convinces another that a statement is
  true while conveying nothing beyond its truth.
:::

## The statistician's contribution

Three judgements in this chapter are not
automated by any cryptographic library, and are
exactly where quantitative training is decisive.

**1. Distinguishing the two resistance
properties and pricing each.** 'The hash is
secure' is a tier-four narrative until it is
resolved into two separate claims with two
separate costs. Preimage resistance costs an
attacker $\sim 2^{n}$ work; collision resistance
costs only $\sim 2^{n/2}$ by the birthday bound.
An analyst who conflates them will overstate the
security of any scheme whose safety rests on
collisions being hard, and the gap is not
marginal: for $n = 256$ it is the difference
between $2^{256}$ and $2^{128}$.

**2. Reading a statistical test as evidence for a
model, not proof of a property.** A
non-significant chi-square on the first output
byte is consistent with uniformity; it does not
establish that no structure exists, and it can
never rule out a cryptanalytic shortcut that
leaves the marginal distribution untouched.
Knowing the difference between 'I failed to reject
uniformity at the byte margin' and 'the function
is a random oracle' is the whole of the
statistician's discipline here.

**3. Deciding when the cryptography is the wrong
layer.** A Merkle root is a genuinely excellent
commitment to a locked dataset, and later in this
chapter we argue for it in exactly one
public-health setting, the trial ledger at database
lock. The same judgement requires saying, candidly,
that it needs no blockchain and that most integrity
problems are solved more cheaply by a signed hash in
a database. The tool is small; the discipline is
knowing its scope.

## Hash functions as random oracles

A cryptographic hash function
$H : \{0,1\}^{*} \to \{0,1\}^{n}$ maps an
arbitrary-length input to a fixed $n$-bit digest
(for SHA-256, $n = 256$). The random-oracle model
idealises $H$ as a function drawn uniformly at
random from all functions with that signature:
each distinct input is assigned an independent,
uniformly random $n$-bit output, and identical
inputs return identical outputs. No real function
is a random oracle (a random oracle has no short
description, whereas SHA-256 has a compact one),
but the model is the working assumption under
which the rest of the book's probability holds.
Five properties operationalise it.

- **Determinism.** $H(x)$ depends only on $x$.
  This is what makes a digest usable as an
  identifier: the same record always hashes to the
  same value.
- **Preimage resistance.** Given a target digest
  $y$, finding any $x$ with $H(x) = y$ should cost
  about $2^{n}$ evaluations. This is a brute-force
  search over the input space with success
  probability $2^{-n}$ per trial.
- **Collision resistance.** Finding any pair
  $x \neq x'$ with $H(x) = H(x')$ should be
  infeasible. As we derive below, the birthday
  bound makes this cost about $2^{n/2}$, far less
  than preimage resistance.
- **Avalanche.** A one-bit change in the input
  flips each output bit independently with
  probability one half. Under the random-oracle
  model this is automatic: the perturbed input is
  a distinct input and so gets an independent
  output.
- **Uniformity.** Digests are (marginally)
  uniformly distributed over $\{0,1\}^{n}$, so any
  fixed window of output bits looks like fair coin
  flips.

We begin with the atomic operation, reprised from
@sec-intro so this chapter stands alone.

```{r ch02-hash}
library(digest)
h <- function(x) digest(x, algo = 'sha256',
                        serialize = FALSE)

h('trial-record: patient 001, arm A')
h('trial-record: patient 001, arm B')
```

The inputs differ in one character and the digests
share no visible structure. Determinism and
avalanche are both on display: re-running the first
line always returns the same value, yet the
single-character change scrambles the output
completely.

### The birthday bound

Preimage resistance is the easy property to reason
about. Each evaluation of $H$ on a fresh input hits
a fixed target with probability $2^{-n}$, so the
number of trials to a first success is geometric
with mean $2^{n}$. More useful in practice: to find
an input whose digest begins with $k$ zero bits,
each trial succeeds with probability $2^{-k}$, so
the expected work is $2^{k}$ evaluations. That last
statement is the entire engine of proof-of-work in
@sec-pow, and it is worth stating precisely now: the
expected number of hash evaluations to find a digest
with $k$ leading zero bits is $2^{k}$.

Collision resistance is subtler. But how many
records must collide before we worry? For the trial
ledger the question is concrete: how many locked
records can share a digest space before two of them
receive the same fingerprint by accident, so that
the commitment would silently admit a substitution
of one for the other? The subtlety is the birthday
problem the reader already knows. Draw $m$ inputs
and hash them. Model the digests, under
the random oracle, as $m$ independent draws uniform
on a space of size $N = 2^{n}$. The probability that
all $m$ digests are distinct is

$$
\Pr(\text{no collision})
= \prod_{i=0}^{m-1}\Bigl(1 - \frac{i}{N}\Bigr)
\approx \exp\!\Bigl(-\frac{m(m-1)}{2N}\Bigr),
$$

using $1 - x \approx e^{-x}$ for small $x$. The
probability of at least one collision reaches one
half when the exponent is about $\ln 2$, that is
when

$$
m \approx \sqrt{2\ln 2}\,\sqrt{N}
\approx 1.177 \cdot 2^{n/2}.
$$

So a collision appears after roughly $2^{n/2}$
hashes, not $2^{n}$. For SHA-256 this is $2^{128}$
versus $2^{256}$: collision search is the square
root of preimage search in cost. This is why
collision resistance is the harder property to
deliver and the first to fall to cryptanalysis
(MD5 and SHA-1 were broken on collisions long
before any preimage attack), and why protocol
designers who need only preimage resistance can
sometimes tolerate a hash that others should not.

We can watch the birthday bound operate directly.
To make collisions findable on a laptop we truncate
SHA-256 to its first $b$ bits, so $N = 2^{b}$, and
count how many random inputs we hash before the
first repeat. The mean first-collision count should
land near $\sqrt{\pi N / 2}$, the birthday
expectation.

```{r ch02-birthday}
set.seed(202)

first_collision <- function(bits) {
  seen <- character(0)
  n <- 0L
  repeat {
    n <- n + 1L
    d <- substr(h(paste0('x-', n, '-',
                         sample(1e9, 1))), 1,
                bits %/% 4)
    if (d %in% seen) return(n)
    seen <- c(seen, d)
  }
}

reps <- 300
draws16 <- replicate(reps, first_collision(16))
c(mean_draws = mean(draws16),
  birthday_expectation = sqrt(pi * 2^16 / 2))
```

Truncating to 16 bits gives $N = 2^{16} = 65536$
and a predicted first-collision count of about
$\sqrt{\pi N/2} \approx 320$. The simulated mean
lands near that value, confirming the
$\sqrt{N} = 2^{n/2}$ scaling as a measured fact
rather than an asymptotic assertion. Doubling the
truncation width to 32 bits would multiply the
work by $2^{8} = 256$, the same square-root law
running in reverse.

We can see the square-root law across several
digest widths at once, not only the 16-bit case
the simulation used (@fig-birthday-prob).

```{r ch02-birthday-prob}
#| label: fig-birthday-prob
#| fig-cap: 'The birthday square-root law across digest widths: each added byte of width postpones the point where an accidental collision becomes likely by a factor of sixteen, since the crossover sits near two to the b over two records.'
library(ggplot2)
library(tibble)
library(dplyr)
library(tidyr)

PAL <- c('#2a78d6', '#1baf7a', '#eda100', '#008300',
         '#4a3aa7', '#e34948', '#e87ba4', '#eb6834')

book_theme <- theme_minimal(base_size = 11) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(
      linewidth = 0.25, colour = 'grey88'),
    axis.title = element_text(size = 10),
    legend.position = 'bottom')

coll_prob <- function(n, b) 1 - exp(-n * (n - 1) / 2^(b + 1))
n_grid <- 10^seq(0, 6, length.out = 400)
bday <- tidyr::expand_grid(n = n_grid,
                           b = c(16L, 24L, 32L)) |>
  dplyr::mutate(
    p = coll_prob(n, b),
    width = factor(paste0(b, '-bit digest'),
                   levels = c('16-bit digest',
                              '24-bit digest',
                              '32-bit digest')))

ggplot(bday, aes(n, p, colour = width)) +
  geom_hline(yintercept = 0.5, linewidth = 0.25,
             colour = 'grey60', linetype = 2) +
  geom_line(linewidth = 0.8) +
  scale_x_log10(breaks = 10^(0:6),
                labels = c('1', '10', '100', '1e3',
                           '1e4', '1e5', '1e6')) +
  scale_colour_manual(values = PAL[1:3], name = NULL) +
  labs(x = 'number of records n (log scale)',
       y = 'probability of at least one collision') +
  book_theme
```

::: {.callout-note icon=false collapse=true}
## Check your understanding: 128-bit collisions

A protocol identifies documents by the first 128
bits of their SHA-256 digest. Roughly how many
documents must exist before a collision becomes
likely, and how does that compare to forging a
document that matches one specific existing
identifier?

A collision becomes likely at about
$2^{128/2} = 2^{64}$ documents, by the birthday
bound. Matching one specific 128-bit identifier
(a preimage-style search) takes about $2^{128}$
attempts. The birthday attack is the square root
of the targeted attack, roughly $2^{64}$ versus
$2^{128}$, which is the difference between
'conceivable for a well-resourced adversary' and
'infeasible'. This is exactly why truncating a
hash to save space can quietly halve the security
exponent.
:::

The three resistance notions are worth setting
side by side, since the trial ledger leans on a
different one for each of its guarantees
(@tbl-resistance).

| Resistance property | Informal statement | Work factor | Ledger property it underwrites |
|:---|:---|:---|:---|
| **Preimage** | Given a digest, find any input that produces it | $\sim 2^{n}$ | One-wayness: a published root leaks nothing about the records, and the leading-zeros search of proof-of-work |
| **Second preimage** | Given one record, find a different record with the same digest | $\sim 2^{n}$ | No single committed record can be swapped for a chosen substitute |
| **Collision** | Find any two records that share a digest | $\sim 2^{n/2}$ | Binding of the Merkle root: no record altered, added, or removed without changing the root |

: Preimage, second-preimage, and collision resistance compared, with the work factor for an $n$-bit digest and the trial-ledger property each supports {#tbl-resistance}

### The avalanche effect, measured

The avalanche property is the claim that flipping
one input bit flips each of the 256 output bits
independently with probability one half. We do not
assume the claim; we validate it, in the sense a
laboratory validates an assay against a known
standard before trusting the readings it will later
return on unknown samples. If true, the Hamming
distance between the two digests (the number of
differing bits out of 256) is a sum of 256
independent fair coins, hence
$\mathrm{Binomial}(256, 0.5)$, with mean 128 and
standard deviation $\sqrt{256 \cdot 0.25} = 8$. We
test this directly: hash many random strings, flip
their last character, rehash, and record the
Hamming distance in bits.

```{r ch02-avalanche}
hex_to_bits <- function(hex) {
  bytes <- strtoi(
    substring(hex, seq(1, nchar(hex), 2),
              seq(2, nchar(hex), 2)),
    base = 16L)
  bits <- vapply(bytes,
    function(b) as.integer(rev(intToBits(b)[1:8])),
    integer(8))
  as.vector(bits)
}

flip_last_char <- function(s) {
  n <- nchar(s)
  ch <- substr(s, n, n)
  repl <- if (ch == 'a') 'b' else 'a'
  paste0(substr(s, 1, n - 1), repl)
}

set.seed(1)
N <- 4000
strs <- paste0('msg-', sprintf('%05d', sample(1e6, N)))
ham <- vapply(strs, function(s) {
  b1 <- hex_to_bits(h(s))
  b2 <- hex_to_bits(h(flip_last_char(s)))
  sum(b1 != b2)
}, numeric(1))

c(mean = mean(ham), sd = sd(ham),
  binom_sd = sqrt(256 * 0.25))
```

The empirical mean sits near 128 and the standard
deviation near the $\mathrm{Binomial}(256, 0.5)$
value of 8. The histogram, with the normal
approximation to that binomial overlaid, makes the
agreement visible.

```{r ch02-avalanche-plot}
#| label: fig-avalanche
#| fig-cap: 'Avalanche effect. Hamming distance between the SHA-256 digests of a string and its one-character perturbation, over 4000 strings, against the Normal approximation to Binomial(256, 0.5).'
library(ggplot2)
library(tibble)

ggplot(tibble(ham = ham), aes(ham)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 30, fill = 'grey70',
                 colour = 'white') +
  stat_function(
    fun = function(x) dnorm(x, 128, sqrt(64)),
    linewidth = 0.8) +
  labs(x = 'Hamming distance (bits changed of 256)',
       y = 'density',
       title = 'Avalanche: one input bit flips ~half the output bits')
```

A single input change scrambles the output as if it
were independently re-randomised. This is the
avalanche property stated as a measurable fact. We
should note the epistemic status carefully: the
observed $\mathrm{Binomial}(256, 0.5)$ shape is
consistent with the random-oracle model, and it is
what makes a one-character edit to a trial record
betray itself. It is not a proof that SHA-256 has
no exploitable structure, only that no structure
shows up in this margin.

### Uniformity of the output

Uniformity is the second observable consequence we
validate, and we test it as we would test any
claimed random-number generator before admitting it
into a simulation. If digests are uniform on
$\{0,1\}^{256}$, then any fixed byte of the output
is uniform on $\{0,1,\dots,255\}$. We hash 50000
distinct inputs, tabulate the first output byte into
its 256 cells, and apply a chi-square
goodness-of-fit test against the uniform
expectation.

```{r ch02-uniformity}
set.seed(2)
M <- 50000
first_byte <- vapply(seq_len(M), function(i) {
  strtoi(substr(h(paste0('u-', i)), 1, 2),
         base = 16L)
}, numeric(1))

obs <- tabulate(first_byte + 1, nbins = 256)
chisq <- chisq.test(obs, p = rep(1 / 256, 256))
c(statistic = unname(chisq$statistic),
  df = unname(chisq$parameter),
  p_value = chisq$p.value)
```

The chi-square statistic sits near its degrees of
freedom (255, the mean of a $\chi^{2}_{255}$) and
the p-value is unremarkable. There is no evidence
against uniformity of the first byte, consistent
with a hash that behaves as a random oracle. The
same caveat applies: this is a test of one marginal,
and passing it is necessary but nowhere near
sufficient for cryptographic security. A backdoored
hash could pass every marginal test we can afford to
run while failing catastrophically on a structured
input the designer chose in advance.

::: {.callout-important}
## The sceptic's reflex

Marketing copy calls encryption 'military-grade'
and systems 'unhackable'. Both phrases are tier-four
narrative wearing the cadence of a protocol fact.
'Military-grade' names no property: AES-256 is
excellent and freely available, and the adjective
adds nothing measurable. 'Unhackable' is stronger
and worse, because the security of every primitive
in this chapter is conditional. Collision resistance
holds only until someone finds a
better-than-birthday attack, exactly as they did for
MD5 and SHA-1. The correct claim is always
quantitative and conditional: 'finding a collision
costs about $2^{128}$ evaluations under current
cryptanalysis'. When a vendor omits the number and
the condition, demote the claim and ask for both.
:::

## Merkle trees

A Merkle tree, named for Ralph Merkle who
introduced the hash-tree construction, turns a list
of records into a single digest, the root, such
that membership of any one record can be proven
with a number of hashes logarithmic in the list
length. Leaves are the
hashes of the records; each internal node is the
hash of its two children concatenated; the root is
the hash at the top. Two facts make the structure
useful, and both follow from the hash properties
above. First, the root is a binding commitment:
changing any record changes its leaf, which by
collision resistance changes every ancestor up to
the root. Second, proving that a record is in the
tree requires revealing only the sibling hash at
each level, $\lceil \log_2 m \rceil$ hashes for $m$
leaves, not the other $m - 1$ records.

We implement the tree with the same `h()` from
above. When a level has an odd number of nodes we
duplicate the last one, the common convention that
keeps the tree binary. This code is self-contained
and additional to the verified companion blocks.

```{r ch02-merkle-build}
hash_pair <- function(a, b) h(paste0(a, b))

merkle_root <- function(leaves) {
  level <- vapply(leaves, h, character(1))
  while (length(level) > 1) {
    if (length(level) %% 2 == 1) {
      level <- c(level, level[length(level)])
    }
    idx <- seq(1, length(level), by = 2)
    level <- vapply(idx,
      function(i) hash_pair(level[i], level[i + 1]),
      character(1))
  }
  unname(level)
}

set.seed(101)
records <- paste0('trial-record-',
                  sprintf('%03d', 1:8))
root <- merkle_root(records)
root
```

The eight synthetic records collapse to one 256-bit
root. But how would a regulator check a single
record without the whole database? This is what the
inclusion proof provides. To prove that
`records[k]` is committed, we walk from its leaf to
the root, recording at each level the sibling hash
and whether it sits to the left or the right. The
verifier replays that walk from the leaf hash and
checks that it reconstructs the published root.

```{r ch02-merkle-proof}
merkle_proof <- function(leaves, index) {
  level <- vapply(leaves, h, character(1))
  proof <- list()
  i <- index
  while (length(level) > 1) {
    if (length(level) %% 2 == 1) {
      level <- c(level, level[length(level)])
    }
    # sibling is to the right of an odd (left) node,
    # to the left of an even (right) node
    if (i %% 2 == 1) {
      proof[[length(proof) + 1]] <-
        list(hash = unname(level[i + 1]),
             side = 'right')
    } else {
      proof[[length(proof) + 1]] <-
        list(hash = unname(level[i - 1]),
             side = 'left')
    }
    idx <- seq(1, length(level), by = 2)
    level <- vapply(idx,
      function(j) hash_pair(level[j], level[j + 1]),
      character(1))
    i <- (i + 1) %/% 2
  }
  proof
}

merkle_verify <- function(leaf, proof, root) {
  acc <- h(leaf)
  for (step in proof) {
    acc <- if (step$side == 'right')
             hash_pair(acc, step$hash)
           else hash_pair(step$hash, acc)
  }
  acc == root
}
```

We now demonstrate the three properties that make
the structure worth its complexity: every genuine
record verifies, a proof is logarithmic in size,
and a forged record fails.

```{r ch02-merkle-demo}
proof_3 <- merkle_proof(records, 3)

all_verify <- vapply(seq_along(records),
  function(k) merkle_verify(records[k],
    merkle_proof(records, k), root),
  logical(1))

c(all_leaves_verify = all(all_verify),
  proof_length = length(proof_3),
  log2_leaves = log2(length(records)),
  forged_leaf_verifies =
    merkle_verify('trial-record-forged',
                  proof_3, root))
```

Every one of the eight records verifies against the
root; the proof for a leaf is three hashes long,
matching $\log_2 8 = 3$; and a forged record
presented with a genuine proof fails, because its
leaf hash does not reconstruct the root. A light
client holding only the 256-bit root can therefore
confirm that a specific record belongs to a
committed set of any size, receiving only
$\log_2 m$ hashes and none of the other records.
That is the whole basis of simplified payment
verification in Bitcoin, where a wallet verifies
that a transaction is in a block using the block
header's Merkle root without downloading the block.

The proof path is easier seen than described. In
@fig-merkle-proof we draw the eight-leaf tree of
the code above and highlight the inclusion proof
for the third leaf.

```{r ch02-merkle-diagram}
#| label: fig-merkle-proof
#| fig-cap: 'Inclusion proof for the third of eight leaves. The verifier holds only the target leaf in dark blue and the three sibling hashes in amber, and rehashes up the blue path to reconstruct the root; the six uninvolved records in grey are never revealed.'
library(ggplot2)
library(tibble)

nodes <- tibble::tribble(
  ~x,  ~y, ~label,   ~role,
  1,   0, 'L1',     'other',
  2,   0, 'L2',     'other',
  3,   0, 'L3',     'target',
  4,   0, 'L4',     'sibling',
  5,   0, 'L5',     'other',
  6,   0, 'L6',     'other',
  7,   0, 'L7',     'other',
  8,   0, 'L8',     'other',
  1.5, 1, 'H(1:2)', 'sibling',
  3.5, 1, 'H(3:4)', 'path',
  5.5, 1, 'H(5:6)', 'other',
  7.5, 1, 'H(7:8)', 'other',
  2.5, 2, 'H(1:4)', 'path',
  6.5, 2, 'H(5:8)', 'sibling',
  4.5, 3, 'root',   'path')

edges <- tibble::tribble(
  ~x,  ~y, ~xend, ~yend, ~erole,
  1,   0, 1.5, 1, 'other',
  2,   0, 1.5, 1, 'other',
  3,   0, 3.5, 1, 'path',
  4,   0, 3.5, 1, 'sibling',
  5,   0, 5.5, 1, 'other',
  6,   0, 5.5, 1, 'other',
  7,   0, 7.5, 1, 'other',
  8,   0, 7.5, 1, 'other',
  1.5, 1, 2.5, 2, 'sibling',
  3.5, 1, 2.5, 2, 'path',
  5.5, 1, 6.5, 2, 'other',
  7.5, 1, 6.5, 2, 'other',
  2.5, 2, 4.5, 3, 'path',
  6.5, 2, 4.5, 3, 'sibling')

txt_col <- ifelse(nodes$role %in% c('target', 'path'),
                  'white', 'grey15')
nodes$role <- factor(nodes$role,
  levels = c('target', 'path', 'sibling', 'other'),
  labels = c('target leaf', 'recomputed path',
             'proof sibling', 'uninvolved'))

fill_named <- c('target leaf' = '#1d5a9e',
                'recomputed path' = '#2a78d6',
                'proof sibling' = '#eda100',
                'uninvolved' = '#9a9a92')
edge_cols <- c(path = '#2a78d6', sibling = '#eda100',
               other = 'grey80')

ggplot() +
  geom_segment(data = edges,
    aes(x = x, y = y, xend = xend, yend = yend,
        colour = erole),
    linewidth = 0.7, show.legend = FALSE) +
  scale_colour_manual(values = edge_cols) +
  geom_tile(data = nodes, aes(x = x, y = y, fill = role),
            width = 0.82, height = 0.55, colour = 'white') +
  geom_text(data = nodes, aes(x = x, y = y, label = label),
            colour = txt_col, size = 2.5) +
  scale_fill_manual(values = fill_named, name = NULL) +
  expand_limits(x = c(0.4, 8.6), y = c(-0.4, 3.5)) +
  theme_void(base_size = 11) +
  theme(legend.position = 'bottom')
```

::: {.callout-note icon=false collapse=true}
## Check your understanding: proof size and privacy

A Merkle tree commits to one million records. How
many hashes does an inclusion proof contain, and
which of the other records does the verifier learn
about?

The proof contains
$\lceil \log_2 10^{6} \rceil = 20$ hashes. The
verifier learns nothing about the contents of the
other records: each proof step reveals only a
sibling digest, and by preimage resistance a digest
does not disclose the record that produced it. This
combination, membership provable in 20 hashes and
no leakage of the rest, is precisely what lets a
data custodian publish a root as a commitment while
keeping the underlying records confidential.
:::

### Public-health application: committing to a locked trial dataset

We return now to the problem with which the chapter
opened, for here the cryptography is genuinely the
right tool. At database lock the trial ledger fixes
its dataset before unblinding and analysis. The
recurring integrity question is whether any record
was altered after lock, for example to remove an
inconvenient adverse event or to adjust an outcome
in a direction favourable to the sponsor. We build a
Merkle tree over the hashes of the locked records
and publish the root, whether in a journal, a
registry, a timestamping service, or simply a
widely-witnessed public post. The root is a 256-bit
commitment to the entire dataset. Later, any party,
and in particular the regulator of the opening
scenario, may be handed a single record together
with a logarithmic inclusion proof and verify that
the record was in the locked set; and by collision
resistance no record can be altered, added, or
removed without changing the published root and
thereby exposing the tampering. The records
themselves never need to be disclosed in order to
make the commitment, so patient confidentiality is
preserved, since the published root reveals nothing
about the data.

@fig-tamper traces what the regulator sees when a
single record is altered after lock.

```{r ch02-tamper-diagram}
#| label: fig-tamper
#| fig-cap: 'Tamper propagation. A single post-lock edit to one record changes its leaf hash, and by collision resistance the change cascades up every ancestor in red until the root no longer matches the value the regulator holds.'
library(ggplot2)
library(tibble)

tnodes <- tibble::tribble(
  ~x,  ~y, ~label,   ~role,
  1,   0, 'L1',     'other',
  2,   0, 'L2',     'other',
  3,   0, 'L3',     'other',
  4,   0, 'L4',     'other',
  5,   0, 'L5',     'other',
  6,   0, 'L6*',    'tampered',
  7,   0, 'L7',     'other',
  8,   0, 'L8',     'other',
  1.5, 1, 'H(1:2)', 'other',
  3.5, 1, 'H(3:4)', 'other',
  5.5, 1, 'H(5:6)', 'prop',
  7.5, 1, 'H(7:8)', 'other',
  2.5, 2, 'H(1:4)', 'other',
  6.5, 2, 'H(5:8)', 'prop',
  4.5, 3, 'root',   'prop')

tedges <- tibble::tribble(
  ~x,  ~y, ~xend, ~yend, ~erole,
  1,   0, 1.5, 1, 'other',
  2,   0, 1.5, 1, 'other',
  3,   0, 3.5, 1, 'other',
  4,   0, 3.5, 1, 'other',
  5,   0, 5.5, 1, 'other',
  6,   0, 5.5, 1, 'prop',
  7,   0, 7.5, 1, 'other',
  8,   0, 7.5, 1, 'other',
  1.5, 1, 2.5, 2, 'other',
  3.5, 1, 2.5, 2, 'other',
  5.5, 1, 6.5, 2, 'prop',
  7.5, 1, 6.5, 2, 'other',
  2.5, 2, 4.5, 3, 'other',
  6.5, 2, 4.5, 3, 'prop')

ttxt <- ifelse(tnodes$role %in% c('tampered', 'prop'),
               'white', 'grey15')
tnodes$role <- factor(tnodes$role,
  levels = c('tampered', 'prop', 'other'),
  labels = c('altered record', 'changed ancestry',
             'unchanged'))

tfill <- c('altered record' = '#e34948',
           'changed ancestry' = '#e34948',
           'unchanged' = '#9a9a92')
tedge <- c(prop = '#e34948', other = 'grey80')

ggplot() +
  geom_segment(data = tedges,
    aes(x = x, y = y, xend = xend, yend = yend,
        colour = erole),
    linewidth = 0.7, show.legend = FALSE) +
  scale_colour_manual(values = tedge) +
  geom_tile(data = tnodes, aes(x = x, y = y, fill = role),
            width = 0.82, height = 0.55, colour = 'white') +
  geom_text(data = tnodes, aes(x = x, y = y, label = label),
            colour = ttxt, size = 2.5) +
  annotate('text', x = 4.5, y = 3.6,
           label = 'root no longer matches the regulator copy',
           colour = '#e34948', size = 2.8) +
  scale_fill_manual(values = tfill, name = NULL) +
  expand_limits(x = c(0.4, 8.6), y = c(-0.4, 3.9)) +
  theme_void(base_size = 11) +
  theme(legend.position = 'bottom')
```

::: {.callout-warning}
## This needs no blockchain

The mechanism above is a hash tree and a published
root. It requires no consensus protocol, no tokens,
no mining, and no chain. A single trustworthy
publication of the root at lock is sufficient,
because the security rests entirely on collision
resistance, not on distributed agreement. A
blockchain would add value only if no party could be
trusted even to publish the root honestly and
promptly, which is rarely the situation in a
regulated trial with an established registry. This
is the falsifiability discipline of @sec-intro in
action: the useful primitive here is the Merkle
commitment, and attaching 'blockchain' to it is
tier-four narrative unless a specific trust failure
is named that only consensus can repair. The
timestamping lineage that predates blockchains
[@haber1991timestamp] made exactly this point:
tamper-evident chronology is a hashing problem, and
was solved as one.

It was also, and this is the part the promotional
literature omits, *standardised* as one. The
Internet time-stamping protocol of 2001 accepts a
hash, never the underlying record, and returns a
signed token establishing that the datum existed
before a stated moment [@rfc3161]; the certificate
transparency standard supplies an append-only Merkle
log with logarithmic inclusion and consistency
proofs, and has run at internet scale for over a
decade [@rfc6962; @trillian]. Between them these
deliver every property our locked-dataset example
requires: proof of existence, proof of inclusion,
and proof that the log was only ever appended to.
Bitcoin's own data structure, as its academic
genealogy makes plain, was borrowed from this
lineage with little modification
[@narayanan2017pedigree], which means the
hash-linked log is separable from the consensus
machinery and can be had without it. A trial ledger
that names no adversary the notary cannot handle is
therefore not buying tamper-evidence with its
blockchain; it already had tamper-evidence, and is
buying only expense. Readers who wish to see the
regulatory question taken seriously, control by
control against the electronic-records requirements
of 21 CFR Part 11, should consult @oakley2023scrybe,
which is nearly alone in the literature in doing so,
and which concludes that the public chains fail
those controls.
:::

## The algebra we have been assuming {#sec-algebra}

The preface asked for comfort with groups and finite
fields, and we now cash that cheque, because the rest of
this chapter is unintelligible without them and because
the reader who takes el The Merkle
commitment settles whether a record was altered, but
it says nothing about who entered it. In the trial
ledger we also wish to know which site coordinator
recorded a given eCRF entry, and to have that
attribution be checkable by a party, such as the
regulator, who trusts none of the coordinators in
advance. The public-key idea
[@diffie1976new; @rivest1978method] supplies exactly
this: a signer holds a private scalar and publishes
a corresponding public key, such that anyone can
verify a signature but only the holder of the
private key can produce one. A coordinator who signs
each entry with a site-held private key makes
authorship of that entry cryptographically checkable,
and forging an attribution is as hard as the
discrete-logarithm problem we come to below.

To make that sentence precise rather than
atmospheric, we need three algebraic objects, and we
build them from the ground up because they are the
load-bearing structure beneath every signature in
this book, and because a reader who takes them on
faith is in no position to judge a claim made about
them.

A **finite field** $\mathbb{F}_{p}$, for $p$ prime,
is the set $\{0, 1, \dots, p-1\}$ with addition and
multiplication carried out modulo $p$. What earns it
the name *field* is that every non-zero element has a
multiplicative inverse: for each $a \neq 0$ there is
a unique $a^{-1}$ with $a\,a^{-1} \equiv 1 \pmod p$.
That inverse exists precisely because $p$ is prime,
so $\gcd(a, p) = 1$ and Bézout's identity supplies
integers with $a x + p y = 1$, whence $x \equiv
a^{-1}$. Division is therefore total, exactly as over
the rationals, and this is the single property that
makes the curve arithmetic below work: the slope of a
line is a ratio, and a ratio needs a reciprocal. Over
$\mathbb{Z}$ modulo a *composite* $n$, some non-zero
elements have no inverse, division breaks, and the
construction collapses. Primality is not decoration.

A **group** $(\mathcal{G}, +)$ is a set with one
associative operation, an identity element, and an
inverse for every element. It is **abelian** if the
operation commutes and **cyclic** if some single
element $G$, the *generator*, produces every element
by repeated application: $\mathcal{G} = \{0\cdot G,
1\cdot G, 2\cdot G, \dots\}$, where $k \cdot G$ means
$G$ combined with itself $k$ times. In a cyclic group
of order $n$ (the number of elements), $n \cdot G$
returns to the identity, so the exponents live in
$\mathbb{Z}_{n}$, the integers modulo $n$. When $n$
is prime, a theorem of Lagrange guarantees that
*every* non-identity element is a generator and that
the group has no non-trivial subgroups, which is why
cryptographic curves are chosen to have prime order:
it forecloses an entire family of attacks that pick
off a small subgroup.

Blockchains realise the public-key idea over a group
whose elements are the points of an elliptic curve.
The group that matters for Bitcoin and much of
Ethereum is
secp256k1: the points $(x, y)$ satisfying
$y^{2} = x^{3} + 7$ over the prime field
$\mathbb{F}_{p}$ for a specific 256-bit prime $p$,
together with a point at infinity $\mathcal{O}$ as
the identity. These points form a cyclic group of
prime order $n$, generated by a fixed base point $G$.
The order $n$ is itself a 256-bit prime, close to but
not equal to $p$; the distinction between the *field*
modulus $p$ (which bounds the coordinates) and the
*group* order $n$ (which bounds the scalars) is a
persistent source of confusion, and we will keep the
two moduli visibly separate in the code.

The group law is geometric: to add two points, draw
the line through them, find its third intersection
with the curve, and reflect across the $x$-axis. The
reflection is what makes the operation a *group*
rather than a mere "third point" map, because it
supplies inverses: the inverse of $(x, y)$ is its
mirror image $(x, -y)$, the line through the two is
vertical, meets the curve only at those two points
and at the point at infinity, and so their sum is
$\mathcal{O}$, the identity, exactly as a group
demands. Turning the geometry into arithmetic gives
the formulas the code below uses. For $P \neq Q$ the
chord has slope
$$\lambda = \frac{y_{Q} - y_{P}}{x_{Q} - x_{P}},$$
and for $P = Q$ the chord becomes the tangent, whose
slope is obtained by implicitly differentiating
$y^{2} = x^{3} + ax + b$ to get $2y\,\mathrm{d}y =
(3x^{2}+a)\,\mathrm{d}x$, hence
$$\lambda = \frac{3x_{P}^{2} + a}{2 y_{P}}.$$
In either case the third point, reflected, is
$$x_{R} = \lambda^{2} - x_{P} - x_{Q}, \qquad
y_{R} = \lambda(x_{P} - x_{R}) - y_{P}.$$
Every operation in these formulas, including the
division hidden inside $\lambda$, is carried out in
$\mathbb{F}_{p}$, which is why we needed a field and
not merely a ring: the slope is a reciprocal, and
only a field guarantees it exists.

A private key is a scalar $d$; the public key is the
point $Q = dG$, meaning $G$ added to itself $d$
times. This is computed not by $d$ literal additions
but by **double-and-add**: reading the bits of $d$
from the top, double the running point at each step
and add $G$ where the bit is one, so a 256-bit scalar
costs about 256 doublings rather than $2^{256}$
additions. The asymmetry this creates is the whole
game. Computing $Q = dG$ takes a few hundred group
operations; recovering $d$ from $Q$ and $G$ is the
**elliptic-curve discrete logarithm problem**, for
which the best known algorithms (Pollard's rho and
its parallel variants) run in time on the order of
$\sqrt{n}$. For secp256k1, $n \approx 2^{256}$, so
$\sqrt{n} \approx 2^{128}$: this is the source of the
"128-bit security" figure, and the reason it is
half the key length rather than the full length. It
is the same square-root gap the birthday bound gave
us for hashes, arising here for the same reason, that
the attacker's best strategy is a collision search
rather than exhaustion. That $2^{128}$ wall, easy to
multiply the base point by a scalar, infeasible to
invert, is what separates a public key from its
private scalar.

To make the group concrete we verify the one axiom
that is not obvious from the geometry, associativity,
on a small curve. We take $y^{2} = x^{3} + 7$ over
$\mathbb{F}_{97}$, which is secp256k1's own curve
equation shrunk to a field we can enumerate by hand.
This curve has exactly 79 points including the point
at infinity, and 79 is prime, so by Lagrange every
non-identity point generates the whole group and the
toy group is a faithful structural miniature of the
real one.

```{r ch02-ecc}
mod_inv <- function(a, p) {
  a <- a %% p
  for (x in 1:(p - 1)) if ((a * x) %% p == 1) return(x)
  stop('no inverse')
}

# Curve: y^2 = x^3 + a x + b over F_p,
# with the point at infinity as NA.
ec_add <- function(P, Q, a, p) {
  if (all(is.na(P))) return(Q)
  if (all(is.na(Q))) return(P)
  if (P[1] == Q[1] && (P[2] + Q[2]) %% p == 0)
    return(c(NA, NA))
  if (all(P == Q)) {
    m <- ((3 * P[1]^2 + a) * mod_inv(2 * P[2], p)) %% p
  } else {
    m <- ((Q[2] - P[2]) *
          mod_inv((Q[1] - P[1]) %% p, p)) %% p
  }
  xr <- (m^2 - P[1] - Q[1]) %% p
  yr <- (m * (P[1] - xr) - P[2]) %% p
  c(xr %% p, yr %% p)
}

p <- 97; a <- 0; b <- 7   # secp256k1's equation, tiny field
pts <- do.call(rbind, Filter(Negate(is.null),
  lapply(0:(p - 1), function(x) {
    y2 <- (x^3 + a * x + b) %% p
    ys <- which(((0:(p - 1))^2 %% p) == y2) - 1
    if (length(ys)) cbind(x, ys) else NULL
  })))

P <- pts[3, ]; Q <- pts[5, ]; R <- pts[9, ]
lhs <- ec_add(ec_add(P, Q, a, p), R, a, p)
rhs <- ec_add(P, ec_add(Q, R, a, p), a, p)
list(n_points = nrow(pts) + 1L,   # +1 for the point at infinity
     P = P, Q = Q, R = R,
     associative = isTRUE(all.equal(lhs, rhs)),
     result = lhs)
```

Associativity holds: $(P + Q) + R$ equals
$P + (Q + R)$. With associativity, an identity, and
inverses (the reflection of a point), the points form
an abelian group, and scalar multiplication $dG$ is
well-defined. Inverting it, the discrete-log problem,
is the hardness the signature scheme rests on.

### ECDSA from the ground up

ECDSA is the scheme Bitcoin launched with, and we now
give it in full, because a reader who cannot write the
verification equation cannot judge a claim that a
signature scheme is sound, and judging such claims is
the entire purpose of this book. Fix the group of
prime order $n$ with generator $G$. Scalars, the
private keys and nonces, live in $\mathbb{Z}_{n}$;
points live on the curve over $\mathbb{F}_{p}$. Write
$x(\,\cdot\,)$ for the $x$-coordinate of a point,
reduced modulo $n$.

To **sign** a message digest $m \in \mathbb{Z}_{n}$
under private key $d$:

1. draw a per-signature nonce $k$ uniformly from
   $\{1, \dots, n-1\}$;
2. compute the point $R = kG$ and set $r = x(R) \bmod
   n$ (if $r = 0$, redraw);
3. compute
   $$s = k^{-1}\,(m + r\,d) \bmod n$$
   (if $s = 0$, redraw).

The signature is the pair $(r, s)$. Note that both
$k^{-1}$ and the arithmetic are taken modulo the group
order $n$, not the field prime $p$: the scalars and
the coordinates live in different rings, and conflating
them is the most common implementation error in this
scheme.

To **verify** $(r, s)$ on digest $m$ against public
key $Q = dG$, compute $w = s^{-1} \bmod n$, then
$$u_{1} = m\,w \bmod n, \qquad u_{2} = r\,w \bmod n,$$
form the point $X = u_{1} G + u_{2} Q$, and **accept
if and only if** $x(X) \bmod n = r$.

That the honest signer is accepted is a short
algebraic **completeness** argument, and it is worth
seeing in full because it is the whole scheme. Starting
from the definition of $s$,
$$s = k^{-1}(m + rd) \;\Longrightarrow\; k = s^{-1}(m
+ rd) = s^{-1}m + s^{-1}rd = u_{1} + u_{2}\,d
\pmod n.$$
Multiply both sides by $G$ and use $Q = dG$:
$$kG = (u_{1} + u_{2} d)G = u_{1}G + u_{2}(dG) = u_{1}G
+ u_{2}Q = X.$$
So $X = kG = R$, their $x$-coordinates agree, and
$x(X) \bmod n = r$ as required. The verifier
reconstructs the signer's nonce point $R$ without ever
learning $k$, and a forger who cannot solve the
discrete logarithm cannot produce an $s$ that makes
the two sides meet.

We implement all of it on the toy curve. The scalar
multiplication routine is the double-and-add described
above; everything else is a transcription of the three
displayed equations.

```{r ch02-ecdsa}
# Toy curve y^2 = x^3 + 7 over F_97; prime group order n = 79.
n <- 79
ec_mul <- function(k, P, a, p) {      # double-and-add
  Rr <- c(NA, NA); Acc <- P; k <- k %% n
  while (k > 0) {
    if (k %% 2 == 1) Rr <- ec_add(Rr, Acc, a, p)
    Acc <- ec_add(Acc, Acc, a, p); k <- k %/% 2
  }
  Rr
}
G <- pts[1, ]                          # a generator (order 79)

sign_ecdsa <- function(m, d, k) {
  R <- ec_mul(k, G, a, p); r <- R[1] %% n
  s <- (mod_inv(k, n) * (m + r * d)) %% n
  c(r = r, s = s)
}
verify_ecdsa <- function(m, sig, Q) {
  w  <- mod_inv(sig[['s']], n)
  u1 <- (m * w) %% n; u2 <- (sig[['r']] * w) %% n
  X  <- ec_add(ec_mul(u1, G, a, p), ec_mul(u2, Q, a, p), a, p)
  !all(is.na(X)) && (X[1] %% n) == sig[['r']]
}

d <- 42; Q <- ec_mul(d, G, a, p)       # key pair
sig <- sign_ecdsa(m = 17, d = d, k = 13)
list(public_key = Q, signature = sig,
     verifies       = verify_ecdsa(17, sig, Q),
     wrong_message  = verify_ecdsa(18, sig, Q))
```

The genuine signature verifies; the same signature
checked against a different digest does not. The
scheme works, and now we can see precisely how it
breaks.

### Why the nonce is the whole ballgame

ECDSA's fragility is not folklore, and it is not
abstract. It is one line of algebra, and it has cost
real custodians real coins. Suppose a signer reuses a
nonce $k$ on two different digests $m_{1} \neq m_{2}$.
Because $r = x(kG) \bmod n$ depends only on $k$, both
signatures carry the *same* $r$, which is visible on
the chain and announces the mistake to the world. The
two $s$ values are
$$s_{1} = k^{-1}(m_{1} + rd), \qquad s_{2} =
k^{-1}(m_{2} + rd).$$
Subtract: $s_{1} - s_{2} = k^{-1}(m_{1} - m_{2})$, so
the nonce falls out immediately,
$$k = \frac{m_{1} - m_{2}}{s_{1} - s_{2}} \bmod n,$$
and with $k$ in hand the private key follows from
either signature by solving its defining equation,
$$d = \frac{s_{1} k - m_{1}}{r} \bmod n.$$
There is no brute force here and no discrete logarithm
to solve. Two signatures sharing a nonce hand over the
private key by arithmetic a student can do by hand.

```{r ch02-nonce-reuse}
k <- 13                                # the fatal reuse
sA <- sign_ecdsa(17, d, k)
sB <- sign_ecdsa(55, d, k)             # same k, new message
k_rec <- ((17 - 55) *
          mod_inv((sA[['s']] - sB[['s']]) %% n, n)) %% n
d_rec <- ((sA[['s']] * k_rec - 17) *
          mod_inv(sA[['r']], n)) %% n
c(same_r = unname(sA[['r']] == sB[['r']]),
  recovered_key = d_rec, true_key = d)
```

The recovered key equals the true key exactly. This is
the failure that broke a well-known games console's
code-signing in 2010, and that has drained wallets
whose software drew nonces from a weak generator. It is
also the reason the uniformity and unpredictability of
the nonce source matter as much as the secrecy of the
key, and why the random-oracle discipline earlier in
this chapter is load-bearing rather than ornamental: a
nonce that is merely *biased*, not repeated, leaks the
key too, more slowly, through a lattice attack on many
signatures. The defensive response adopted across the
ecosystem is to stop trusting randomness altogether and
derive the nonce *deterministically* from the message
and the key (the construction of RFC 6979), so that the
same message never produces two nonces and a broken
generator cannot produce a predictable one.

One further wrinkle Bitcoin had to patch is
**malleability**. If $(r, s)$ verifies, so does
$(r,\,n - s)$: negating $s$ leaves the verification
equation satisfied, because it flips $X$ to its mirror
image, which shares the same $x$-coordinate. An
attacker who cannot forge a signature can still flip
its low bit and change the transaction's identifier
without the signer's consent. Bitcoin's rule requiring
the *low* value of $s$ (BIP 62, and later enforced as
part of the SegWit upgrade) closes this by declaring
one of the two equivalent signatures canonical and
rejecting the other.

### Schnorr, and why Bitcoin moved to it

Schnorr signatures, named for Claus Schnorr who
proposed them [@schnorr1991efficient], authenticate
the same way but with a cleaner equation, and the
cleanliness is not aesthetic: it is what makes
aggregation possible. We met the Schnorr identification
protocol earlier in this chapter; the signature is its
Fiat--Shamir form. The signer with key $d$ and public
key $Q = dG$ draws a nonce $k$, forms $R = kG$, sets
the challenge $e = H(R \,\|\, Q \,\|\, m)$ by hashing
rather than asking a verifier, and publishes $(R, s)$
with
$$s = k + e\,d \pmod n.$$
Verification checks the single linear equation
$$sG \;\stackrel{?}{=}\; R + e\,Q,$$
which holds for an honest signer because $sG = (k +
ed)G = kG + e(dG) = R + eQ$. Contrast this with ECDSA,
where the nonce enters through an inverse and an
$x$-coordinate, both non-linear. The Schnorr equation
is linear in the secret $d$ and the nonce $k$, and
linear equations add.

That is the property Bitcoin's 2021 Taproot upgrade was
after. Because the equation is linear, $n$ signers with
keys $Q_{1}, \dots, Q_{n}$ can each contribute a share
and produce a *single* signature valid for the combined
key $\sum_i Q_{i}$, so a multi-signature wallet or a
payment channel closes with one signature that is
indistinguishable on chain from an ordinary single-party
one. This shrinks both the data written to the chain and
the verification cost, and it improves privacy by making
complex spending policies look ordinary. The standard
that specifies Bitcoin's variant, including the
deterministic nonce and a fixed rule for the sign of
$R$, is BIP 340. Naive aggregation invites a
**key-cancellation** attack, in which a malicious signer
chooses its public key as a function of the others' to
forge on their behalf; production multi-signature
protocols such as MuSig2 defeat it by committing to all
keys before any nonce is revealed, which is the same
commit-then-reveal discipline that recurs throughout
this book.

### BLS and pairings, as an interface

BLS signatures, named for Boneh, Lynn and Shacham
[@boneh2001short], push aggregation further still, and
we treat them honestly as an interface rather than an
implementation, because the object they rely on, a
cryptographic pairing, is genuinely beyond the algebra
this book develops. A **pairing** is a map $e(\cdot,
\cdot)$ taking two curve points to an element of a
field that is *bilinear*: $e(aP, bQ) = e(P, Q)^{ab}$.
That one identity is all we need to read the scheme. A
BLS signature on message $m$ under key $d$ is the
single point $\sigma = d\,H(m)$, where $H$ maps the
message to a curve point, and verification checks
$$e(\sigma, G) \;\stackrel{?}{=}\; e\bigl(H(m),\,
Q\bigr),$$
which holds because bilinearity turns the left side
into $e(H(m), G)^{d}$ and the right side into
$e(H(m), G)^{d}$ as well. The payoff is that signatures
from many signers on the same message *add* as curve
points into one short signature verifiable against the
summed public key. This is why BLS is the signature of
proof-of-stake consensus in @sec-pos, where a single
block may carry attestations from tens of thousands of
validators and compressing them into one object is what
keeps verification tractable. We will use this property;
we will not build the pairing, and the reader should
treat its security as an assumption imported from a
literature we cite rather than a result we prove
[@boneh2001short].

::: {.callout-note icon=false collapse=true}
## Check your understanding: why aggregation matters

A proof-of-stake block collects attestations from
10000 validators. Contrast the verification burden
if each attestation carried an independent ECDSA
signature versus a single BLS-aggregated signature,
and name the property that makes aggregation
possible.

With independent ECDSA signatures the verifier
performs 10000 separate verifications and the block
carries 10000 signatures, roughly 640 KB at 64 bytes
each. With BLS aggregation the 10000 signatures on
the shared attestation collapse to one group element
and one aggregate public key, verified in a single
pairing check. The enabling property is linearity of
the signature scheme in the key and message: BLS (and
Schnorr) verification equations are linear, so sums of
keys and signatures verify together, whereas ECDSA's
equation is not, so its signatures do not aggregate.
This is the difference between a consensus protocol
that scales to thousands of validators and one that
does not.
:::

### A note on quantum threats

We should not close the discussion of signatures
without asking how durable their hardness is, for a
question that once seemed academic has become a
genuine design consideration. What would a
large-scale quantum computer do to the primitives of
this chapter? The answer divides sharply, and the
division is the practical point. Shor's algorithm
solves the discrete-logarithm and integer-
factorisation problems in polynomial time, and these
are precisely the problems on which ECDSA and RSA
rest. A sufficiently large quantum computer would
therefore recover a private key from its public key
and forge signatures at will, which is to say that
the public-key signatures of this chapter would fail
completely rather than merely weaken. Hash functions
fare far better. Grover's algorithm speeds an
unstructured search only quadratically, so inverting
an $n$-bit hash falls from about $2^{n}$ to about
$2^{n/2}$ work; a hash loses roughly a square-root
factor of its security and no more, which for
SHA-256 still leaves a comfortable $2^{128}$. The
analysis of quantum attacks on Bitcoin by Aggarwal
and colleagues [@aggarwal2018quantum] works these
asymmetries through in detail and reaches the same
qualitative conclusion: signatures are the exposed
surface, and hashes are relatively robust.

For a health ledger the implication is not
hypothetical, and it turns on the longevity of the
record. A trial ledger is a long-lived object: a
signature made today to attribute an eCRF entry may
need to remain unforgeable for decades, through the
archival life of the trial and any later re-analysis
or litigation. This is the 'harvest now, forge later'
horizon, in which an adversary records today's public
keys and signatures and waits for the cryptographic
era to turn, at which point yesterday's unforgeable
attributions become forgeable. The horizon is genuine
for any ledger meant to outlive the current generation
of public-key cryptography, and it argues, on the one
hand, for preferring hash-based commitments wherever a
hash will do the work, since these are the resilient
primitive, and, on the other, for designing the
signature layer to be replaceable, so that a migration
to a post-quantum scheme can be carried out without
discarding the ledger. We do not attempt to recommend
a specific post-quantum signature here; the field is
still settling, and the prudent posture, in our view,
is to keep the option open rather than to commit early.

## Commitments and zero-knowledge

The Merkle root is one instance of a commitment: a
short value that binds a party to hidden data
(binding), while revealing nothing about it until
opened (hiding). The same idea generalises. A
commitment scheme lets a party publish a commitment
now and reveal the committed value later, with the
scheme guaranteeing the value cannot be changed in
between. The binding property rests on collision
resistance, exactly as the Merkle root does, and the
hiding property on the one-wayness of the hash, so a
commitment asks of the reader nothing that the
present chapter has not already supplied.

A commitment fixes a value, but it does not, by
itself, let one demonstrate anything about that value
while it remains hidden. That further step, proving a
property of committed data without opening it, is the
province of the zero-knowledge proof, to which we now
turn.

## Zero-knowledge proofs

We begin with a claim that sounds paradoxical and is
not. Can one party convince another that a statement
is true while conveying nothing whatever beyond its
truth, not even the reason it is true? The
affirmative answer, formalised by Goldwasser, Micali
and Rackoff [@goldwasser1989knowledge], whose study
of the knowledge complexity of interactive proof
systems founded the subject, is among the deeper
results of modern cryptography, and it rests on
three properties that a
reader trained in statistics will find unusually
congenial. Consider a prover, who claims to hold a
witness to some statement, and a verifier, who will
accept or reject after an exchange of messages. An
interactive proof system is required to satisfy three
conditions.

- **Completeness.** If the statement is true and the
  prover is honest, the verifier accepts with
  overwhelming probability. An honest prover in
  possession of the witness is essentially never
  turned away.
- **Soundness.** If the statement is false, then no
  prover, however devious, can make the verifier
  accept except with some small probability, the
  soundness error. This error is not a defect to be
  apologised for; it is the very quantity the
  construction controls, and repetition drives it
  toward zero.
- **Zero-knowledge.** The verifier learns nothing
  from the exchange beyond the truth of the
  statement, in the precise sense that whatever the
  verifier sees could have been simulated, without
  the prover at all, from the statement alone.

The property that repays statistical attention is
soundness, and it is worth dwelling on why. A
soundness error of $\varepsilon$ means that a cheating
prover, one who does not hold the witness, slips a
false statement past the verifier with probability at
most $\varepsilon$ on a single run. But how small must
$\varepsilon$ be? Here the analogy to hypothesis
testing is exact rather than decorative. The verifier
is running a test whose null hypothesis is 'the prover
does not know the witness', and the soundness error is
precisely the type-one error rate, the probability of
accepting a false claim. A single round of a protocol
whose challenge is one bit has soundness error one
half, which no statistician would accept as a test.
There are then two routes to a small error, and they
are interchangeable. One may repeat the protocol:
because the rounds are independent, $t$ repetitions
drive the soundness error to $2^{-t}$, the geometric
decay that @fig-zk-soundness traces against the
conventional 0.05 level. Or one may widen the
challenge within a single round, so that a prover who
can answer only one of the $\lvert C \rvert$ possible
challenges for a given commitment succeeds with
probability at most $1 / \lvert C \rvert$. The Schnorr
protocol below takes the second route, drawing its
challenge from the whole scalar field, and so attains
a negligible soundness error in one round with no
repetition at all; the repetition picture belongs to
its small-challenge cousins. One chooses whichever
route the setting favours to meet the tolerance
demanded.

```{r ch02-zk-soundness}
#| label: fig-zk-soundness
#| fig-cap: 'Soundness error under repetition. A sigma-protocol with per-round soundness error one half sees that error fall geometrically as two to the minus t over t independent rounds, crossing the conventional 0.05 level between the fourth and fifth round.'
library(ggplot2)
library(tibble)

PAL <- c('#2a78d6', '#1baf7a', '#eda100', '#008300',
         '#4a3aa7', '#e34948', '#e87ba4', '#eb6834')

book_theme <- theme_minimal(base_size = 11) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(
      linewidth = 0.25, colour = 'grey88'),
    axis.title = element_text(size = 10),
    legend.position = 'bottom')

rounds <- 0:12
sound <- tibble::tibble(t = rounds, err = (1 / 2)^rounds)

ggplot(sound, aes(t, err)) +
  geom_hline(yintercept = 0.05, linewidth = 0.25,
             colour = 'grey60', linetype = 2) +
  geom_line(linewidth = 0.8, colour = PAL[1]) +
  geom_point(size = 1.6, colour = PAL[1]) +
  annotate('text', x = 9, y = 0.075,
           label = 'conventional 0.05 level',
           colour = 'grey40', size = 2.8) +
  scale_y_log10(breaks = 10^(0:-4),
                labels = c('1', '0.1', '0.01',
                           '0.001', '0.0001')) +
  scale_x_continuous(breaks = seq(0, 12, 2)) +
  labs(x = 'number of independent rounds t',
       y = 'soundness error (log scale)') +
  book_theme
```

The abstraction becomes concrete in the oldest and
cleanest example, the identification protocol of
Schnorr [@schnorr1991efficient]. Recall the
elliptic-curve group of the preceding section, with
base point $G$ of prime order and the
discrete-logarithm problem, recovering $d$ from
$Q = dG$, believed hard. A prover wishes to convince a
verifier that she knows the private scalar $d$ behind
a public key $Q = dG$, without revealing $d$. The
exchange has the three-move shape that names this
family of protocols, the sigma-protocol, after the
shape of its message flow.

- **Commitment.** The prover draws a fresh random
  scalar $r$, computes the point $R = rG$, and sends
  $R$. The scalar $r$ is a secret nonce, and its role
  is exactly that of a commitment: it binds the prover
  to a value she cannot afterward change.
- **Challenge.** The verifier draws a random scalar
  $c$ and returns it. This is the public coin, chosen
  after the commitment is fixed.
- **Response.** The prover sends $s = r + c\,d$,
  computed with the secret scalars, and the verifier
  accepts if $sG$ equals $R + cQ$.

The algebra closes because
$sG = (r + c\,d)G = rG + c(dG) = R + cQ$, so an honest
prover always passes, which is completeness. Soundness
is the interesting direction: a prover who does not
know $d$ cannot in general answer a challenge she could
not predict, and one can show that a prover able to
answer two different challenges for the same
commitment $R$ would thereby reveal $d$ outright, which
is the knowledge-extraction argument that gives the
soundness its teeth. The protocol is zero-knowledge
because the triple $(R, c, s)$ can be simulated without
$d$ at all: choose $s$ and $c$ at random and set
$R = sG - cQ$, and the resulting transcript is
distributed exactly as a genuine run, carrying no
information about the secret. That last observation is
the whole content of the zero-knowledge property in
miniature.

One inconvenience remains. The protocol as described
is interactive, in that the verifier must be present to
issue the challenge $c$. But what if the prover wishes
to leave behind a proof that anyone can check later,
with no live verifier at all? The device that removes
the verifier is the Fiat-Shamir heuristic, and its
intuition is simple. The challenge is meant
to be unpredictable to the prover before she commits,
and a cryptographic hash, which the earlier part of
this chapter established behaves as a random oracle,
supplies exactly such unpredictability. We replace the
verifier's coin with $c = H(R, \text{statement})$, the
hash of the commitment together with the statement
being proved. The prover cannot steer the output of the
hash, and so cannot choose $R$ to suit a favourable
$c$; the interactive protocol collapses to a single
non-interactive proof, the pair $(R, s)$, that any
party can verify by recomputing the challenge. This is,
incidentally, how the interactive identification scheme
becomes the Schnorr signature met earlier, and it is
one more reason the random-oracle tests of this chapter
are load-bearing rather than ornamental.

The Schnorr proof is small and quick, but its statement
is fixed and simple, namely knowledge of one discrete
logarithm. The ambition of the last decade has been to
prove, just as compactly, that an arbitrary computation
was carried out correctly, that a whole batch of
transactions is valid, or that a committed record
satisfies an elaborate predicate. The proofs that
achieve this go by the name of succinct non-interactive
arguments of knowledge, or SNARKs. What does 'succinct'
buy? It is the property that the proof is far smaller
than the computation it attests to and can be checked
in far less time than re-executing that computation
would take, so that a verifier who would need hours to
redo the work can instead check a proof of a few
hundred bytes in milliseconds. Ben-Sasson, Chiesa,
Tromer and Virza [@bensasson2014snark] gave a
construction for a general von Neumann computer
architecture, bringing arbitrary programs within reach;
Groth [@groth2016size] produced the compact
pairing-based scheme, with proofs of constant size,
that is now widely deployed; and Ben-Sasson and
colleagues [@bensasson2018stark] later introduced
transparent STARKs, which dispense with the trusted
setup that pairing-based SNARKs require and rest their
security on hashes alone, at the cost of somewhat larger
proofs. The application that turned these ideas from
theory into consequence was Zerocash
[@bensasson2014zerocash], which used succinct proofs to
build a payment system in which a transaction proves its
own validity, that its inputs and outputs balance and
the spender is authorised, while revealing neither
sender, receiver, nor amount, and so turned
zero-knowledge into private payments.

We should be candid about the boundary of our
competence here. The internal machinery of these
constructions, the arithmetic circuits, the polynomial
commitments, the elliptic-curve pairings, is a
substantial body of cryptographic engineering that lies
well beyond the scope of this book. We cite these
systems as an interface and not as an implementation:
what matters for our purposes is the guarantee they
expose, a short proof that a stated computation was
performed correctly, and not the mechanism by which
they attain it.

Why should a book concerned with public-health data
attend to any of this? Because zero-knowledge is the
natural language for privacy-preserving health
analytics, and the connection is not merely
metaphorical. To illustrate, consider a multi-site
trial with an eligibility protocol under which a
prospective participant must fall within an age band
and carry a laboratory value in a specified range. A
site can, in principle, prove to the coordinating
centre that a given individual satisfies these
criteria, that the age lies in one interval and the
lab value in another, without transmitting the date of
birth or the laboratory result itself. The statement
'this record satisfies the predicate' is established;
the record stays home. Broadened, this is the same
aspiration as secure multiparty computation and
federated analysis, in which a statistic, a regression
coefficient or a pooled prevalence, is computed jointly
across institutions while each retains custody of its
raw data and surrenders none of it.

The analogy must be drawn with care, for it is easily
overstated. A zero-knowledge proof establishes the
truth of a statement about data that has already been
committed; it does not, on its own, compute a joint
statistic across parties. Proving that each site's
record satisfies a predicate is a different operation
from combining those records into a pooled estimate,
and the latter is the province of secure multiparty
computation proper, of which zero-knowledge proofs are
a component rather than a synonym. We flag the
distinction because conflating the two is a common
route to overstating what a privacy-preserving pipeline
actually delivers. The constructions themselves we take
no further here, and the reader who wishes to pursue
them is referred to the primary sources named above.

::: {.callout-note icon=false collapse=true}
## Check your understanding: soundness as type-one error

A verifier runs a single round of an interactive proof
whose soundness error is one half. A statistician
objects that a test which wrongly accepts a false claim
half the time is worthless. How is the objection
answered, and how many independent rounds are needed to
drive the probability of accepting a false claim below
one in a million?

The objection is answered by repetition. The soundness
error is exactly a type-one error rate, and because the
rounds are independent, $t$ of them multiply the
per-round error to $(1/2)^{t}$. To fall below
$10^{-6}$ we need $(1/2)^{t} < 10^{-6}$, that is
$t > 6 / \log_{10} 2 \approx 19.9$, so twenty rounds
suffice. This is the same logic by which one lowers a
significance threshold, and it is why a single round
with soundness error one half is not an embarrassment
but a starting point: one buys as much soundness as the
setting requires by paying in rounds.
:::

Before the worked example we gather the four
primitives of the chapter into one view, each with
the guarantee it provides and, as important, the
guarantee it does not (@tbl-primitives).

| Primitive | Guarantee | Role in the trial ledger | What it cannot do |
|:---|:---|:---|:---|
| **Hash** | Deterministic one-way fingerprint | Detects any change to a record | Attribute authorship, or reveal what changed beyond the digest |
| **Digital signature** | Authenticity and non-repudiation | Attributes each eCRF entry to a site coordinator | Prove integrity of anything beyond what was signed, or establish time |
| **Merkle tree** | Logarithmic membership proof over a binding root | Lets an auditor verify one record against the root without the dataset | Say who entered a record, or when |
| **Commitment** | Binding and hiding | Publishes a value now to be revealed later, fixed in between | Prove a property of the hidden value, which needs zero-knowledge |

: The four primitives of this chapter, the guarantee each provides, its role in the trial ledger, and its limits {#tbl-primitives}

## Worked example

To illustrate, consider the trial ledger at its lock
date, with the pieces now assembled into one
end-to-end analysis of the problem with which we
began. We commit to the locked dataset, prove one
record's membership to an auditor, and confirm that
any tampering is caught by the published root. This
is the trial-ledger application of the previous
section, run as code on a SYNTHETIC locked dataset
standing in for the real trial records.

```{r ch02-worked}
set.seed(303)

# SYNTHETIC locked trial dataset: one row per
# subject, serialised to a canonical string.
locked <- data.frame(
  subject = sprintf('S%04d', 1:16),
  arm = sample(c('A', 'B'), 16, replace = TRUE),
  outcome = rbinom(16, 1, 0.3),
  stringsAsFactors = FALSE
)
record_strings <- with(locked,
  paste(subject, arm, outcome, sep = '|'))

published_root <- merkle_root(record_strings)

# An auditor is handed subject 7's record plus a
# proof, and verifies membership against the
# published root alone.
k <- 7
audit_ok <- merkle_verify(
  record_strings[k],
  merkle_proof(record_strings, k),
  published_root)

# Someone alters subject 12's outcome after lock.
tampered <- record_strings
tampered[12] <- paste0(tampered[12], '-edited')

c(auditor_confirms_membership = audit_ok,
  proof_hashes = length(merkle_proof(record_strings, k)),
  tamper_changes_root =
    merkle_root(tampered) != published_root)
```

The auditor confirms subject 7's record with a
four-hash proof (the tree over 16 leaves has depth
four) and never sees the other fifteen records. The
post-lock edit to subject 12 changes the root, so
anyone holding the originally published root detects
that the dataset is no longer the one committed to.
The guarantee is exactly as strong as collision
resistance, which the birthday analysis priced at
about $2^{128}$ work, and no stronger. It is also, as
the warning above insisted, achieved with no
blockchain: a hash tree and one honest publication.

## Collaborating with an LLM

Language models are fluent about cryptography and
confidently wrong at its edges, which is precisely
where security lives. They are of use in
accelerating routine work, provided the parts that
matter are verified against a primary source.

**Prompt.** 'Write an R function to compute a Merkle
root and an inclusion proof using the digest
package.'

*Watch for:* the odd-node convention (duplicate the
last node, or promote it unchanged?), off-by-one
errors in sibling indexing, and whether the proof
records left/right order at all. A verifier that
ignores order will accept proofs for the wrong
position.

*Verification:* the structural checks in this
chapter's demo (every leaf verifies, a forged leaf
fails, tampering changes the root) are the minimum.
These are correctness checks on the implementation,
not a security proof of the scheme.

**Prompt.** 'Explain why collision resistance is
weaker than preimage resistance for a 256-bit hash.'

*Watch for:* confident but reversed statements. Models
sometimes assert collisions cost $2^{256}$, missing
the birthday factor entirely, or garble which
property protects which use case.

*Verification:* the reader should re-derive the
birthday bound independently ($m \approx 2^{n/2}$)
and check the model's numbers against it. The
truncated-hash simulation in this chapter is a fast
empirical cross-check.

**Prompt.** 'Is it safe to reuse the same nonce
across two ECDSA signatures?'

*Watch for:* any answer that does not immediately say
this leaks the private key. Nonce reuse is a
catastrophic, well-documented failure, and a model
that hedges is a warning about the model.

*Verification:* consult the scheme's specification,
not the model, for anything a key depends on. Treat
LLM cryptography as a tier-two claim requiring a
primary source.

## Exercises

1. **Birthday bound by simulation.** Extend the
   `first_collision()` function to truncation widths
   of 8, 12, 16, and 20 bits. For each, estimate the
   mean first-collision count over several hundred
   replications and plot it against $\sqrt{\pi N/2}$
   on log axes. Confirm the slope corresponds to the
   $2^{n/2}$ law and report the fitted exponent.

2. **Leading-zeros work.** Verify empirically that
   the expected number of SHA-256 evaluations to find
   a digest with $k$ leading zero bits is $2^{k}$.
   Estimate the mean trial count for $k = 8, 12, 16$
   and compare to $2^{k}$. Explain how this
   experiment is the kernel of proof-of-work in
   @sec-pow.

3. **Avalanche at the byte level.** The chapter tested
   avalanche in bits. Repeat the analysis counting
   differing bytes out of 32 instead of bits out of
   256, derive the distribution differing bytes should
   follow under the random oracle, and test the fit.
   (Hint: a byte differs with probability
   $1 - 2^{-8}$.)

4. **Merkle proof forgery.** Attempt to construct a
   valid inclusion proof for a record not in the tree,
   without finding a hash collision. Argue from the
   `merkle_verify()` logic why this reduces to finding
   a preimage or collision, and state which resistance
   property each forgery strategy would have to defeat.

5. **Second-preimage on a truncated hash.** Using
   truncation to make it feasible, empirically estimate
   the work to find a second input hashing to the same
   truncated digest as a fixed target, and confirm it
   scales as $2^{b}$ rather than $2^{b/2}$. Explain why
   this differs from the collision experiment in
   exercise 1.

6. **Group order.** For the toy curve
   $y^{2} = x^{3} + 7$ over $\mathbb{F}_{97}$,
   enumerate all points (the chapter's `pts` matrix,
   plus the point at infinity) and count them. Confirm
   the count is 79, and since 79 is prime, verify
   Lagrange's theorem in its strongest form here: the
   order of *every* non-identity point equals 79, so
   every point is a generator. Then take the composite-
   order curve $y^{2} = x^{3} + 2x + 3$ over the same
   field, show its order is 100, and exhibit a point
   whose order is a proper divisor of 100. This is the
   small-subgroup structure that prime order forecloses.

7. **ECDSA end to end.** Using the chapter's
   `sign_ecdsa()` and `verify_ecdsa()`, pick your own
   private key $d$ and nonce $k$ on the toy curve, sign
   two different digests, and confirm both verify.
   Then tamper with one component of a signature (the
   $r$, then the $s$) and confirm verification fails in
   each case. Finally demonstrate malleability: show
   that $(r,\,n - s)$ verifies whenever $(r, s)$ does,
   and explain from the verification equation why.

8. **Recover a key from a reused nonce.** Given only
   the two signatures produced with a shared nonce in
   the chapter (their $r$, $s_{1}$, $s_{2}$ and the two
   digests), and *without* using the variable `d`,
   recover the private key by the two-line derivation
   in the text. Confirm it matches. Then argue why a
   nonce that is not reused but merely predictable
   (say, a counter) is equally fatal, and name the
   deterministic construction the ecosystem adopted in
   response.

## Further reading

The Merkle tree originates with @merkle1988digital,
which introduces the hash tree as a way to sign many
messages from one public value; read it for the
construction in its original framing. The
timestamping papers of @haber1991timestamp are the
direct ancestors of the tamper-evident-log idea and
make explicit that chronology and integrity are
hashing problems, a useful corrective to
blockchain-first accounts. For public-key
cryptography, @diffie1976new is the conceptual origin
and @rivest1978method the first concrete realisation;
both are short and worth reading as primary sources.

On signatures, @schnorr1991efficient introduces the
scheme whose linearity underlies modern aggregation,
and @boneh2001short introduces the short pairing-based
signatures that make proof-of-stake consensus
tractable at @sec-pos scale. The zero-knowledge
framework is @goldwasser1989knowledge, foundational
for the privacy constructions later chapters build on.
For the primitives in blockchain context,
@narayanan2016bitcoin (chapters 1 and 4) gives the
rigorous treatment this chapter matches to a
statistical audience, and @antonopoulos2017mastering
gives the concrete Bitcoin realisations of hashing,
Merkle trees, and ECDSA. The proof-of-work use of the
leading-zeros bound traces to @back2002hashcash, which
@sec-pow develops in full.

## In conclusion

In conclusion, three points are to be emphasised.
First, the security of every mechanism in this
chapter is quantitative and conditional, resolved
into a specific cost under current cryptanalysis
(preimage resistance at about $2^{n}$, collision
resistance at about $2^{n/2}$ by the birthday
bound), and never the flat 'unbreakable' of
marketing copy; an analyst who conflates the two
resistance properties will overstate by the square
root of the security exponent. Second, the
statistical tests we ran on the hash validate the
random-oracle model as one validates an assay,
establishing that the output is consistent with the
model at the margins we were able to examine while
establishing nothing about structure we did not
test, and the distinction between failing to reject
uniformity and proving security is the whole of the
discipline. Third, for the trial ledger the right
primitive is the Merkle commitment to the locked
dataset, which needs no blockchain, and the
statistician's contribution lies as much in
declining to reach for consensus machinery that a
single honest publication renders unnecessary as in
deploying the cryptography where a genuine trust
failure warrants it.
