# Proof-of-Work and Nakamoto Consensus {#sec-pow}

::: {.bookepigraph}
```{=latex}
\begin{flushright}\small
```
*The race between the honest chain and an attacker chain
can be characterized as a Binomial Random Walk.*

[Satoshi Nakamoto]{.smallcaps}, *Bitcoin: A Peer-to-Peer
Electronic Cash System* (2008)
```{=latex}
\end{flushright}
```
:::

## Learning objectives

By the end of this chapter the reader
should be able to:

- State the proof-of-work puzzle as a
  Bernoulli search over nonces and explain
  why it is progress-free (memoryless).
- Describe Nakamoto consensus, the
  longest-chain (most-work) rule, and how
  cost substitutes for identity to resist
  Sybil attacks.
- State the three Bitcoin Backbone
  properties (common prefix, chain quality,
  chain growth) precisely and say what each
  guarantees.
- Derive the simple gambler's-ruin bound
  $(q/p)^z$ for a double-spend and explain
  how Satoshi's Poisson mixture refines it.
- Compute an attacker-success probability,
  verify the six-confirmations claim, and
  cross-check the closed form by simulation.
- Treat confirmation depth as a
  tail-probability threshold, the same
  decision as choosing a critical value or a
  group-sequential stopping boundary.

## Orientation

We begin with a data problem. Consider a
consortium of hospitals running a
multi-site clinical trial. Enrollment,
randomization, and outcome events are
appended to a shared, append-only log that
no single institution owns, and a regulator
will audit that log after the fact. We shall
refer to this arrangement throughout as the
trial ledger. Suppose an outcome event, the
death of a subject, say, or a confirmed
endpoint, is appended at one site. Here is
the question that governs everything
downstream: how long must the other sites
wait before they may treat that event as
settled, so that an interim analysis, or a
report to the data and safety monitoring
board, may rely upon it? Act too soon and
the record may yet be revised beneath the
analysis; wait too long and the monitoring
that safeguards the trial's subjects is
delayed.

The previous two chapters assembled the
parts. @sec-consensus posed open-membership
Byzantine agreement as the problem, and
@sec-crypto supplied the hash function that
behaves as a random oracle. This chapter
puts them together into the mechanism that
made Bitcoin work, and although digital cash
is the historical setting in which that
mechanism was born, our interest here is in
what it offers the trial ledger. The
vocabulary is one the reader already owns.
The proof-of-work puzzle is a sequence of
independent Bernoulli trials, that is,
repeated yes-or-no experiments, each with a
minuscule success probability. The security of a
confirmed record is the tail of a biased
random walk. The choice of how long to wait
before treating a record as settled is the
choice of a critical value: a threshold on a
quantified error probability.

The single most important reframing in the
chapter is this. A blockchain does not give
certainty that a record is permanent. It
gives a probability that the record will be
reversed, and that probability decays
geometrically in the number of blocks mined
on top of it. 'Finality' is therefore not a
binary state the ledger enters. It is a
confidence statement, and the reader who has
ever chosen a significance level, or set a
stopping boundary for a group-sequential
trial, already knows how to reason about it.

We shall build every claim twice: once as a
piece of mathematics, and once as runnable R
that the reader may re-execute. The
attacker-success calculation reproduced here
is that of Satoshi Nakamoto, the pseudonymous
author of the 2008 Bitcoin whitepaper, given
in its Section 11 [@nakamoto2008bitcoin], and
we verify the six-confirmations claim rather
than repeat it on faith.

Before proceeding we fix the terminology on
which the remainder of the chapter depends, so
that the reader need not carry these terms as
mysteries.

::: {.callout-note}
## The vocabulary of this chapter

- **Proof-of-work.** A scheme in which
  appending a block requires solving a
  computational puzzle that is costly to
  perform but cheap to check, so that influence
  over the ledger is bought with computation
  rather than with identity.
- **Nonce.** A number placed in a block header
  and varied by the miner in search of a header
  whose hash meets the target. It is the one
  field the miner is free to change at will.
- **Target and difficulty.** The target is the
  threshold a block hash must fall below to be
  valid; the difficulty is the reciprocal
  measure of how demanding that threshold is. A
  smaller target means a higher difficulty and
  more expected work per block.
- **Nakamoto consensus.** The procedure,
  introduced with Bitcoin, by which open and
  mutually distrusting parties agree on one
  history: they extend and accept the chain
  carrying the most accumulated proof-of-work.
- **Longest-chain (heaviest-chain) rule.** The
  fork-choice component of Nakamoto consensus:
  honest nodes treat the chain with the greatest
  total work as authoritative, which reduces to
  the longest chain when difficulty is held
  constant.
- **Fork and orphan block.** A fork is a
  temporary split in which two valid blocks
  extend the same parent. When one branch is
  abandoned, the discarded block is termed an
  orphan, and the transactions it held return to
  the pending pool.
- **Double-spend.** The canonical attack, in
  which a party spends a record, waits for it to
  be accepted, and then publishes a competing
  chain that erases it, thereby reversing a
  transaction already treated as settled.
- **Probabilistic finality and confirmation
  depth.** A confirmed record is never permanent
  with certainty; the probability of its
  reversal merely shrinks as further blocks
  accumulate above it. The confirmation depth is
  the number of such blocks, written $z$, at
  which one agrees to treat the record as
  settled.
:::

## The statistician's contribution

Three judgments in this chapter fall to the
analyst, and no protocol automates them.

**1. Finality is a confidence statement, not
a fact.** The protocol emits a reversal
probability; a human decides how small it
must be before acting. That is a critical
value chosen against a quantified tail risk,
and setting it well requires knowing the cost
of a wrong decision, which the chain cannot
know on our behalf. A merchant selling
coffee, a clearing house settling a bond, and
a monitoring board acting on an interim
endpoint should not use the same depth.

**2. The model's assumptions are where the
real risk resides.** Satoshi's bound assumes a
fixed attacker hash share $q$, honest blocks
arriving as a Poisson process, and no network
partition. Each assumption can fail:
hash-rate can be rented in bursts, propagation
delay creates natural forks, and a
sufficiently resourced adversary changes $q$
mid-attack. Naming which assumption a given
deployment violates is the analyst's job, not
the whitepaper's.

**3. 'Immutable' is a claim to be demoted.**
The correct statement is 'exponentially
expensive to revise, with the exponent set by
depth and the base set by the honest majority
margin'. Reporting the number, not the
adjective, is the contribution.

## The proof-of-work puzzle

### The puzzle as a Bernoulli search

A block header commits to the previous
block's hash, a Merkle root of the
transactions (a hash-based summary that commits
to every transaction in the block), a
timestamp, and a nonce. To
mine a block a miner searches for a nonce
such that the header hash, read as a 256-bit
integer, falls below a target $T$:

$$
H(\text{header}) < T .
$$

Because $H$ behaves as a random oracle, an
idealized function whose every output is, in
effect, an independent uniform draw
(@sec-crypto), each distinct nonce yields an
output that is, for all practical purposes,
uniform on $\{0, 1, \dots, 2^{256}-1\}$.
A single attempt therefore succeeds with
probability

$$
p = \frac{T}{2^{256}} ,
$$

independently across nonces. Mining is a
sequence of i.i.d. Bernoulli$(p)$ trials with
$p$ extremely small.

It is often more intuitive to state the
puzzle in leading zero bits. If the target
requires the top $d$ bits of the hash to be
zero, then

$$
p = 2^{-d} ,
$$

and the expected number of attempts to find a
block is $1/p = 2^{d}$. Each additional
required zero bit halves the success
probability and doubles the expected work.
Stated in hexadecimal digits, each leading
hex zero divides the valid target set by 16,
a fact we will measure directly in
@sec-mining.

### Progress-free means memoryless

The property that makes proof-of-work a fair
lottery rather than a race is that it is
progress-free. A miner who has failed a
million times is in exactly the same position
as a miner who just started: the next attempt
still succeeds with probability $p$. There is
no partial credit, no accumulated progress
toward the solution. Formally, the number of
attempts $N$ until the first success is
geometric,

$$
\Pr(N = n) = (1-p)^{n-1} p ,
\qquad \mathbb{E}[N] = 1/p ,
$$

and the geometric distribution is the unique
memoryless distribution on the positive
integers:

$$
\Pr(N > m + n \mid N > m) = \Pr(N > n) .
$$

Memorylessness has a consequence the reader
will recognize. When many independent miners
each
run their own tiny-$p$ Bernoulli search in
parallel, the time until some miner succeeds
is, in the continuous limit, exponential, and
the stream of blocks is a Poisson process (the
standard model for events that arrive
independently at a steady average rate).
That is the entire statistical foundation of
the next chapter, so we flag it here and
develop it in @sec-mining.

Progress-freedom is also what makes hash-power
the sole determinant of success. One cannot be
clever about finding a nonce; one can only
attempt more of them per second. Mining reduces one
resource, computation, to one number, and that
reduction is what lets cost stand in for
identity.

::: {.callout-note icon=false collapse=true}
## Check your understanding: doubling the difficulty

A chain currently requires $d$ leading zero
bits, so a block takes an expected $2^{d}$
hash attempts. The difficulty is raised to
require $d+3$ zero bits. By what factor does
the expected number of attempts per block
change, and does a miner who has already tried
$2^{d}$ nonces at the new difficulty have any
advantage on the next attempt?

The expected attempts rise by a factor of
$2^{3} = 8$, since $\mathbb{E}[N] = 2^{d+3} =
8 \cdot 2^{d}$. The miner has no advantage:
the search is progress-free, so having tried
$2^{d}$ nonces leaves the success probability
of the next attempt at exactly $2^{-(d+3)}$,
unchanged. Past failures carry no information
about future success. This is the memoryless
property, and it is why a large miner's
advantage is strictly proportional to
hash-rate, with no economies of scale in the
search itself.
:::

## Nakamoto consensus

### The longest-chain rule

Proof-of-work resolves the question of which
party may append the next block. Nakamoto
consensus resolves the question of which chain
the parties agree upon. The rule is simple to
state:
honest nodes always extend the chain with the
most accumulated proof-of-work, and treat that
chain as the truth. In the common case where
difficulty is constant, most-work reduces to
longest, so the mechanism is usually called
the longest-chain rule.

The rule is a decentralized tie-break. When
two miners find a block at nearly the same
time, the network temporarily holds two
competing chains of equal length. The tie
resolves when the next block extends one of
them; that chain now has more work, honest
nodes switch to it, and the orphaned block's
transactions return to the pool. No vote is
taken and no coordinator is consulted. The
heaviest chain wins by construction.

We can picture the tie-break directly.
@fig-pow-fork shows two branches leaving a
shared block: the honest branch pulls ahead,
and the attacker branch, once it falls behind,
is abandoned.

```{r ch03-fork}
#| label: fig-pow-fork
#| fig-cap: 'A fork resolves by accumulated work: honest nodes follow the heavier branch, so the honest chain that pulls ahead wins and the attacker branch that falls behind is abandoned.'
library(ggplot2)
library(tibble)

PAL <- c('#2a78d6', '#1baf7a', '#eda100', '#008300',
         '#4a3aa7', '#e34948', '#e87ba4', '#eb6834')

book_theme <- ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(
    panel.grid.minor = ggplot2::element_blank(),
    panel.grid.major = ggplot2::element_line(
      linewidth = 0.25, colour = 'grey88'),
    axis.title = ggplot2::element_text(size = 10),
    legend.position = 'bottom'
  )

blocks <- tibble::tribble(
  ~x, ~y,   ~branch,    ~label,
  0,  0,    'common',   'shared',
  1,  0.7,  'honest',   'H1',
  2,  0.7,  'honest',   'H2',
  3,  0.7,  'honest',   'H3',
  4,  0.7,  'honest',   'H4',
  1, -0.7,  'attacker', 'A1',
  2, -0.7,  'attacker', 'A2'
)
w <- 0.6; h <- 0.5
blocks <- transform(blocks,
  xmin = x - w / 2, xmax = x + w / 2,
  ymin = y - h / 2, ymax = y + h / 2)
cols <- c(common = '#33356b', honest = PAL[1], attacker = '#9a9a92')
links <- tibble::tribble(
  ~x,  ~y,    ~xend, ~yend, ~branch,
  0.3, 0.05,  0.7,   0.55,  'honest',
  1.3, 0.7,   1.7,   0.7,   'honest',
  2.3, 0.7,   2.7,   0.7,   'honest',
  3.3, 0.7,   3.7,   0.7,   'honest',
  0.3, -0.05, 0.7,  -0.55,  'attacker',
  1.3, -0.7,  1.7,  -0.7,   'attacker'
)

ggplot() +
  geom_segment(data = links,
    aes(x = x, y = y, xend = xend, yend = yend, colour = branch),
    arrow = arrow(length = unit(0.12, 'cm'), type = 'closed'),
    linewidth = 0.5) +
  geom_rect(data = blocks,
    aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax,
        fill = branch), colour = 'white') +
  geom_text(data = blocks, aes(x = x, y = y, label = label),
    colour = 'white', size = 3) +
  annotate('text', x = 4, y = 1.2, label = 'honest branch wins',
    colour = PAL[1], size = 3.2) +
  annotate('text', x = 2, y = -1.2, label = 'attacker branch abandoned',
    colour = '#9a9a92', size = 3.2) +
  scale_fill_manual(values = cols) +
  scale_colour_manual(values = cols) +
  coord_equal(xlim = c(-0.5, 4.8), ylim = c(-1.5, 1.5)) +
  theme_void() +
  theme(legend.position = 'none')
```

### Cost substitutes for identity

The reason this survives an open-membership
setting is that appending a block costs real
computation, so votes are weighted by
hash-power rather than by identity. In a
classical Byzantine agreement protocol
(@sec-consensus) an adversary who can forge
identities can manufacture a majority: the
Sybil attack [@douceur2002sybil]. Nakamoto
consensus makes identities free but makes
influence expensive. Creating a thousand fake
nodes confers no advantage, because each block still
requires the same expected $1/p$ hash
attempts. To outweigh the honest network an
attacker must out-compute it, which is the
content of the phrase 'one CPU, one vote' in
the whitepaper, correctly read as 'one unit of
hash-power, one unit of vote'.

This is the pivotal design move. The Sybil
problem is not solved by authenticating
participants; it is dissolved by making the
scarce resource external and physical.
Proof-of-work is, in the language of
@dwork1993pricing who first proposed pricing
via processing, a cost function that gates
participation.

### Probabilistic finality

Because an attacker with a minority of
hash-power can still, by chance, find several
blocks in a row, no transaction is ever
absolutely final. What the longest-chain rule
delivers is that the probability of reversing
a transaction buried under $z$ subsequent
blocks decays geometrically in $z$. Finality
is asymptotic, approached but never reached.

Here the public-health reader has a ready
analogy. In group-sequential monitoring of a
trial we do not wait for the study to end
before acting; we examine the accumulating
data at interim looks and act when the
evidence crosses a stopping boundary chosen
to control the overall error probability.
Waiting for confirmations is the same shape
of decision. Each new block is a further look
at the ledger, and we treat a record as
settled once the reversal probability has
fallen below a boundary we have set in
advance. The analogy is honest, but it has a
limit, and we should mark it plainly. In
group-sequential monitoring the uncertainty
is sampling error, and the data, once
recorded, do not conspire against us. In the
ledger the uncertainty is adversarial: a
reorganization is not the luck of the draw
but the product of an attacker spending
hash-power to rewrite history. The tail we
compute is therefore a bound against a
strategic opponent, not a sampling
distribution, and that distinction warrants a
more conservative boundary than sampling error
alone would demand.

This is the deepest contrast with the
classical Byzantine-fault-tolerant protocols
studied in @sec-consensus and revisited for
proof-of-stake in @sec-pos. A protocol such as
PBFT, or Practical Byzantine Fault Tolerance,
the protocol of Castro and Liskov
[@castro1999practical], delivers deterministic
finality within a known number of rounds, at
the cost of a fixed, known validator set.
Nakamoto consensus buys
open membership at the price of finality that
is only ever probabilistic. Neither is
strictly better; they price the same guarantee
in different currencies, and knowing which one
is buying is the point.

::: {.callout-important}
## The skeptic's reflex

The reader will encounter the claim that a
blockchain is 'immutable' and that confirmed
transactions are 'permanent'. The tier should
be named. 'Immutable' is
delivered in the cadence of a protocol fact,
but the protocol fact is weaker and more
precise: a confirmed transaction is
exponentially expensive to revise, with a
reversal probability that is small but never
zero and that depends on the attacker's
hash share. 'Immutable' should be demoted to
'exponentially expensive to revise, and here
is the exponent'. The adjective hides exactly
the quantity a statistician is equipped to
report.
:::

## The Bitcoin Backbone

Satoshi's whitepaper argued security by a
single tail-probability calculation. A decade
of formal work, initiated by the Bitcoin
Backbone analysis of Garay, Kiayias, and
Leonardos [@garay2015backbone], which recast
Nakamoto's informal security argument as a set
of provable guarantees, and extended to the
asynchronous setting by @pass2017asynchronous,
distilled the protocol into three properties
that hold with
overwhelming probability provided the honest
parties control a sufficient majority of
hash-power. The properties are stated over the
chains held by honest parties as the protocol
runs.

**Common prefix.** For any two honest parties
at any two times, the shorter chain, after
removing its last $k$ blocks, is a prefix of
the longer chain, except with probability
dropping exponentially in $k$. In words:
honest parties agree on all but the most recent
few blocks. This is precisely what licenses
'wait $z$ confirmations': the common-prefix
parameter $k$ is the confirmation depth, and
the exponential drop is the reversal
probability. Common prefix is the formal home
of probabilistic finality.

**Chain quality.** In any sufficiently long
window of an honest party's chain, the fraction
of blocks contributed by honest parties is
bounded below (roughly by the honest hash
share, degraded by an adversarial factor). In
words: the adversary cannot monopolize the
ledger's content. Chain quality guarantees that
honest transactions actually get included, not
merely that parties agree on some ordering.
Without it a majority-but-not-total attacker
could agree with everyone while censoring
chosen transactions.

**Chain growth.** The honest parties' chains
grow by at least a guaranteed number of blocks
in any sufficiently long window. In words: the
ledger makes progress, it does not stall.
Chain growth is what lets us assume that
waiting real time produces confirmations at a
predictable rate, which is the premise of every
latency calculation in @sec-mining.

The three properties are complementary. Common
prefix is safety (agreement), chain growth is
liveness (progress), and chain quality is a
fairness or censorship-resistance guarantee
sitting between them. Stated compactly: common
prefix says honest parties agree, chain growth
says
they keep moving, chain quality says the
adversary cannot fill the chain with its own
blocks. All three degrade gracefully rather
than failing catastrophically as the adversary
approaches the honest-majority threshold.

@tbl-pow-backbone collects the three
properties beside what each guarantees and how
we read it for the trial ledger.

| Property | Informal statement | What it guarantees | Trial-ledger reading |
|----------|--------------------|--------------------|----------------------|
| **Common prefix** | Honest parties agree on all but the last few blocks | Safety: no reorganization below depth $k$ | The $z$ in 'wait $z$ confirmations' is this $k$ |
| **Chain quality** | The adversary cannot monopolize the block content | Censorship resistance: honest events get in | A settled outcome event cannot be quietly excluded |
| **Chain growth** | The chain gains blocks at a guaranteed rate | Liveness: the ledger does not stall | Confirmations accrue in predictable real time |

: The three Bitcoin Backbone properties, what each guarantees, and how the trial ledger reads it {#tbl-pow-backbone}

### GHOST, an alternative fork-choice

The longest-chain rule discards the work in
orphaned blocks. When block propagation is
slow relative to the block interval, forks are
frequent, orphaned work is large, and security
suffers because the effective honest hash-power
is fragmented. The GHOST rule (Greedy Heaviest
Observed Subtree) of Sompolinsky and Zohar
[@sompolinsky2015ghost]
changes the fork-choice: rather than following
the longest chain, a node at each fork follows
the subtree containing the most total work,
counting orphaned blocks toward the weight of
their branch. This lets a chain tolerate faster
block times without the security loss that
naive longest-chain would suffer, and a variant
of it informed Ethereum's early fork-choice.
The point for our purposes is that longest-chain
is one fork-choice rule among several, each with
a different tradeoff between block interval,
orphan rate, and security, and each analysable
by the same probabilistic tools.

## Double-spend as gambler's ruin

### The random-walk model

We now come to the chapter's mathematical
core, and we shall frame it from the outset as
the computation of the very error probability
the trial ledger requires. The canonical
attack on Nakamoto consensus is the
double-spend. An attacker pays a merchant,
waits for the transaction to be buried under
$z$ confirmations, takes delivery of the goods,
and then attempts to publish a longer private
chain that omits the payment, reversing it. The
attack succeeds if and only if the attacker's
private chain overtakes the honest chain. The
same adversary threatens the trial ledger: a
site that wishes to un-record a settled outcome
event must, by exactly this mechanism, outpace
the honest sites, and the probability that it
succeeds is the probability that our
'settled' verdict was premature.

Model the race by the lead of the honest chain
over the attacker. Let the attacker command a
fraction $q$ of total hash-power and the honest
network $p = 1 - q$, with $q < p$. Each new
block is honest with probability $p$ and
attacker's with probability $q$. The gap
between the chains is a random walk that steps
$+1$ (honest extends its lead) with probability
$p$ and $-1$ (attacker closes the gap) with
probability $q$. The attacker wins if this walk
ever reaches zero starting from a lead of $z$.

This is the classical gambler's ruin, the
problem of a bettor with finite capital playing
against a richer adversary until one side is
exhausted, and its solution is well known. For
a walk biased away from the
origin ($q < p$), the probability of ever
reaching zero starting from distance $z$ is

$$
\Pr(\text{catch up from } z)
= \left( \frac{q}{p} \right)^{z} .
$$

The derivation is the standard one. Let $a_z$
be the catch-up probability from a gap of $z$.
First-step analysis gives $a_z = q\,a_{z-1} +
p\,a_{z+1}$ with $a_0 = 1$, whose bounded
solution for $q < p$ is $a_z = (q/p)^{z}$. The
security of $z$ confirmations, in this simplest
model, is that reversal probability falls
geometrically with ratio $q/p < 1$. If the
attacker holds ten percent of hash-power,
$q/p = 1/9$, and each confirmation cuts the
reversal odds by a factor of nine.

### Satoshi's Poisson refinement

The simple bound $(q/p)^{z}$ is pessimistic in
one respect and optimistic in another, and
Satoshi's Section 11 calculation corrects it.
The subtlety is that during the time the honest
network builds its $z$ confirmations, the
attacker is not idle: it is mining its private
chain in parallel. So at the moment the merchant
accepts, the attacker has already made some
private progress.

Satoshi models this as follows. While the honest
network finds $z$ blocks, the number of blocks
the attacker finds is Poisson with mean

$$
\lambda = z \, \frac{q}{p} ,
$$

since the attacker's blocks arrive at rate
$q/p$ relative to the honest network's. Condition
on the attacker having found $k$ private blocks.
If $k \ge z$ the attacker is already ahead and
succeeds with probability one. If $k < z$ the
attacker is behind by $z - k$ and must close that
gap, which by gambler's ruin happens with
probability $(q/p)^{z-k}$. Mixing over the
Poisson distribution of $k$,

$$
\Pr(\text{success})
= 1 - \sum_{k=0}^{z}
\frac{\lambda^{k} e^{-\lambda}}{k!}
\left[ 1 - \left(\tfrac{q}{p}\right)^{z-k}
\right] .
$$

This is the formula we now evaluate. It
refines the crude $(q/p)^z$ by accounting for
the attacker's head start, and it is precisely
the reversal probability the trial ledger
needs: the chance that a record treated as
settled at depth $z$ is later undone. But how
deep is deep enough? The question has no answer
in the abstract, for the depth that suffices
depends on the attacker share we are willing to
assume and the tolerance we are willing to
accept. We begin with the case Satoshi treats,
an attacker holding ten percent of hash-power
who is watched to a depth of six.

```{r ch03-attack}
p_attack <- function(q, z) {
  p <- 1 - q
  lambda <- z * (q / p)
  k <- 0:z
  1 - sum(dpois(k, lambda) * (1 - (q / p)^(z - k)))
}

# Satoshi's claim: q = 0.10, z = 6 gives well under 0.1 percent.
p_attack(0.10, 6)
```

The value is well below one in a thousand,
confirming the whitepaper's assertion that six
confirmations suffice against an attacker with
ten percent of hash-power. This is the origin of
the folk rule 'wait for six confirmations': it
is not an arbitrary constant but the depth at
which the
Section 11 tail probability drops below a
tolerance a merchant, or a monitoring board, is
willing to accept. A different tolerance implies
a different depth.

It is worth seeing the whole tail at once.
@tbl-pow-depth tabulates the reversal
probability across depths and attacker shares,
the table a merchant or a monitoring board
reads down a column to the depth at which the
probability first falls below its tolerance.

```{r ch03-depth-table}
#| label: tbl-pow-depth
library(knitr)

qs <- c(0.05, 0.10, 0.25, 0.40)
zs <- 0:12
tab <- sapply(qs, function(q) sapply(zs, function(z) p_attack(q, z)))
tab_df <- data.frame(z = zs, formatC(tab, format = 'e', digits = 2),
                     check.names = FALSE)
names(tab_df) <- c('z', paste0('q = ', formatC(qs, format = 'f', digits = 2)))
knitr::kable(tab_df, align = 'r',
  caption = 'Reversal probability P(success) by confirmation depth z (rows) and attacker hash share q (columns), from the Section 11 formula. Reading down a column shows the geometric decay for a fixed attacker; comparing columns shows how the depth needed for a target tolerance climbs as q grows.')
```

### The family of attack curves

Plotting the attack probability against
confirmation depth for several attacker shares
makes the geometric decay, and its sensitivity to
$q$, visible at once.

```{r ch03-plot}
#| label: fig-pow-family
#| fig-cap: 'Attacker success probability by confirmation depth z, for hash shares q in {0.10, 0.25, 0.40}, on a log scale. Deeper confirmations buy exponentially more security, but the slope flattens as q approaches one half.'
library(dplyr)
library(tidyr)
library(ggplot2)

grid <- expand_grid(q = c(0.10, 0.25, 0.40), z = 0:12) |>
  rowwise() |>
  mutate(p = p_attack(q, z)) |>
  ungroup()

ggplot(grid, aes(z, p, colour = factor(q))) +
  geom_line() + geom_point(size = 1) +
  scale_y_log10() +
  labs(x = 'Confirmations behind (z)',
       y = 'P(attacker overtakes), log scale',
       colour = 'q (hash share)',
       title = 'Section 11 attacker success by confirmations')
```

Read @fig-pow-family as a statistician reads a
power curve. For $q = 0.10$ the reversal
probability falls steeply, and six confirmations
already place it out of reach. For $q = 0.40$ the
curve is nearly flat: an attacker with forty
percent of hash-power is so close to parity that
no realistic confirmation depth secures the
transaction, because the random walk's drift
away from zero is weak. The confirmation depth
that buys a target security level is a function
of the attacker share we are willing to assume,
and assuming $q$ is a modeling decision, not a
protocol output.

### Cross-checking the closed form

A closed-form probability should always be
checked against a simulation of the same model,
if only to catch transcription errors. We draw
the attacker's private block count as Poisson,
resolve the residual gap by the gambler's-ruin
ratio, and compare.

```{r ch03-mc}
mc_attack <- function(q, z, sims = 2e5) {
  p <- 1 - q
  lambda <- z * (q / p)
  k <- rpois(sims, lambda)
  gap <- z - k
  catch <- ifelse(gap <= 0, 1, (q / p)^gap)
  mean(runif(sims) < catch)
}

set.seed(21)
tibble::tibble(
  z = 1:8,
  closed_form = sapply(1:8, function(zz) p_attack(0.10, zz)),
  monte_carlo = sapply(1:8, function(zz) mc_attack(0.10, zz))
)
```

The two columns agree to simulation noise. Note
the honest epistemic claim: agreement validates
the transcription of the model into code, not
the model itself. The model is Satoshi's, and
its assumptions (fixed $q$, Poisson honest
blocks, no partition) are the things a deployment
can violate. A simulation cannot vindicate an
assumption it also makes.

::: {.callout-note icon=false collapse=true}
## Check your understanding: choosing a depth is choosing a critical value

A merchant will accept a transaction once the
reversal probability falls below a tolerance
$\alpha$. Explain in what precise sense choosing
the confirmation depth $z$ is the same kind of
decision as choosing a critical value in a
hypothesis test, and what plays the role of the
'cost of a Type I error'.

The reversal probability $\Pr(\text{success})$
is a monotone decreasing function of $z$.
Choosing the smallest $z$ with
$\Pr(\text{success}) \le \alpha$ is exactly
choosing a rejection threshold so that a tail
probability sits below a preset level, the same
operation as picking a critical value so that
$\Pr(\text{Type I error}) \le \alpha$. The tail
here is the tail of the biased random walk rather
than of a null sampling distribution, but the
logic is identical: fix the tolerable error
probability, then read off the threshold. The
'cost of a wrong decision' is the value of the
goods released against an unconfirmed payment;
higher stakes justify a smaller $\alpha$ and
hence a deeper $z$, just as a costlier false
positive justifies a stricter significance level.
The chain supplies the tail; the human supplies
$\alpha$.
:::

### Setting the depth in practice

We may now return to the question with which
the chapter opened. How many confirmations
should elapse before the trial ledger treats an
appended outcome event as settled, so that an
interim analysis or a report to the monitoring
board may rely upon it? The machinery of the
preceding sections answers it directly. The
reversal probability is $\Pr(\text{success})$
as a function of the assumed attacker share $q$
and the depth $z$, and the required depth is
the smallest $z$ at which that probability
falls below the tolerance we are prepared to
accept.

To illustrate, consider the trial ledger with
an assumed adversary share $q = 0.10$ and
ten-minute blocks. A monitoring board content
with a reversal probability below one in a
thousand may act at a depth of six, which by
`p_attack(0.10, 6)` is roughly one chance in
four thousand and corresponds to about an hour
of waiting. A board that judges the stakes
higher, and demands one in a million, must wait
to a depth of eleven, since $\Pr(\text{success})$
first crosses $10^{-6}$ between $z = 10$ and
$z = 11$, which is under two hours. The extra
security is bought with roughly fifty additional
minutes of latency, and whether that trade is
worth making is a judgment about the trial, not
about the chain.

The tradeoff is exactly the latency-versus-risk
tradeoff of the double-spend calculation. Acting
on a shallow confirmation is fast but exposes the
workflow to a reversal: if the block is later
orphaned, the outcome event it recorded is
un-recorded, and any interim decision taken in
reliance on it, an early-stopping recommendation,
say, rests on a record that no longer exists. The
same logic governs a vaccine cold-chain
provenance system, our secondary example, in
which each custody hand-off of a refrigerated
consignment is anchored to the chain: a premature
reliance on a shallow confirmation risks
certifying a hand-off that is subsequently undone.
Waiting for a deep confirmation is safe but slow,
and in either setting the latency itself carries a
cost. The decision is a threshold on a quantified
reversal probability, and it should be set by the
consequence of a reversal, not by a default.

We pause to note what has just been done, because
the habit is one this book means to instill. We
named a tolerance, derived the depth that meets it,
and reported the latency it costs. The estimate was
accompanied by the assumption it rests on, namely
$q$, and a reader who disputes the assumption can
recompute the answer. Contrast this with the
standard to which the public-health literature holds
itself when it evaluates a digital intervention in
earnest. The English exposure-notification
application of the early 2020s was assessed by two
independent estimators, one built from an
epidemiological model and one from a matched
statistical comparison of neighboring districts.
They disagreed. The model-based estimate put the
cases averted in the first months at roughly
284,000; the design-based comparison put them at
roughly 594,000, better than twice as many
[@wymant2021nhsapp]. The authors published both,
with intervals, and declined to choose. Over the
first year the same program was credited with
averting on the order of a million cases and nine
thousand deaths [@kendall2023nhsapp]. Two features
of that work should be held in mind for the whole of
this book. The first is that a two-fold disagreement
between estimators, honestly reported, is worth more
than a single confident number, and the reader who
becomes uncomfortable at such a spread has
misunderstood what an estimate is. The second is
that the intervention thus evaluated ran on Bluetooth
radios and public-key signatures, and contained no
ledger of any kind. It is, so far, the only digital
public-health intervention of the decade with a
credible estimate of the deaths it prevented.

Here candor is owed, and we state the limit
plainly. For most health-data integrity problems
the parties share a regulator or a sponsor who can
serve as a custodian, and a well-governed database
with a cryptographic audit log delivers
tamper-evidence without any of this
probabilistic-finality machinery, as @sec-intro
argued. On the one hand the confirmation-depth
calculation is the right tool when the log must be
trust-minimized across parties with no common
authority; on the other, when a common authority
does exist, the calculation is machinery in search
of a problem. The analyst's first contribution is
therefore not to compute the depth but to ask
whether the condition that makes the depth
meaningful actually holds.

## Worked example: a tamper-evident toy chain

To make the linkage concrete we build a small
proof-of-work chain from scratch, then
demonstrate that altering a buried block is
detectable. To fix ideas, let each block hold a
batch of trial events, a day's enrollments,
randomizations, and outcome records from one
site, so that the `data` field stands for that
batch. The function `validate_chain` is then the
regulator's audit: months after the fact, the
regulator recomputes every hash and checks that
the recorded history is internally consistent
and carries the required proof-of-work. This is
the mechanism behind 'tamper-evident' stated as
running code rather than as an adjective. The
data here are SYNTHETIC, standing in for the
event batches only as the labels 'block-1'
through 'block-6'.

```{r ch03-chain}
library(digest)

make_block <- function(prev_hash, data, nonce = 0) {
  list(prev_hash = prev_hash, data = data, nonce = nonce, hash = NA)
}

mine_block <- function(block, difficulty = 2) {
  target <- strrep('0', difficulty)
  attempts <- 0L
  repeat {
    attempts <- attempts + 1L
    block$nonce <- block$nonce + 1
    block$hash <- digest(
      paste0(block$prev_hash, block$data, block$nonce),
      algo = 'sha256', serialize = FALSE)
    if (startsWith(block$hash, target)) {
      block$attempts <- attempts
      return(block)
    }
  }
}

build_chain <- function(n, difficulty = 2) {
  chain <- vector('list', n)
  prev <- strrep('0', 64)
  for (i in seq_len(n)) {
    blk <- mine_block(make_block(prev, paste0('block-', i)), difficulty)
    chain[[i]] <- blk
    prev <- blk$hash
  }
  chain
}

validate_chain <- function(chain, difficulty = 2) {
  target <- strrep('0', difficulty)
  for (i in seq_along(chain)) {
    b <- chain[[i]]
    recomputed <- digest(paste0(b$prev_hash, b$data, b$nonce),
                         algo = 'sha256', serialize = FALSE)
    if (recomputed != b$hash) return(FALSE)
    if (!startsWith(b$hash, target)) return(FALSE)
    if (i > 1 && b$prev_hash != chain[[i - 1]]$hash) return(FALSE)
  }
  TRUE
}

set.seed(3)
chain <- build_chain(6, difficulty = 2)
tampered <- chain
tampered[[3]]$data <- 'tampered!'
c(valid_original = validate_chain(chain, 2),
  valid_tampered = validate_chain(tampered, 2))
```

The original six-block chain validates: the
regulator's audit passes. Altering the data in
block 3, as if a site had quietly revised a
day's outcome records after the fact, breaks
validation for two independent reasons, both of
which `validate_chain` checks. The recomputed
hash of block 3 no longer matches its stored
hash, and even had the attacker recomputed that
hash, it would no longer begin with the required
zero digits, nor would it match the `prev_hash`
recorded in block 4. To repair the chain the
attacker would have to re-mine block 3 and every
block after it, each at full proof-of-work cost.
This is the microcosm of the double-spend
analysis, and the point for the trial ledger is
the same one seen from the auditor's side: a
buried record cannot be revised without redoing
all the work stacked upon it, and the deeper the
record, the more work required. That is the
gambler's-ruin bound once more, read from the
attacker's side of the ledger.

@fig-pow-hashchain renders the cascade the
audit detects: the tamper at block 3, and the
prev-hash links above it that no longer close.

```{r ch03-hashchain}
#| label: fig-pow-hashchain
#| fig-cap: 'Each block commits to its predecessor by a prev-hash pointer, so tampering with block 3 breaks the links above it: block 3 no longer matches the pointer carried by block 4, and the invalidation cascades to block 5.'
library(ggplot2)
library(tibble)

n <- 5
bx <- seq_len(n)
wc <- 0.7; hc <- 0.9
fill_cols <- c(ok = '#33356b', tampered = '#e34948', broken = '#ffffff')
line_cols <- c(ok = '#33356b', tampered = '#e34948', broken = '#e34948')
text_cols <- c(ok = '#ffffff', tampered = '#ffffff', broken = '#e34948')
chain_blocks <- tibble::tibble(
  x = bx, label = paste0('block ', bx),
  state = c('ok', 'ok', 'tampered', 'broken', 'broken'),
  xmin = bx - wc / 2, xmax = bx + wc / 2, ymin = -hc / 2, ymax = hc / 2)
chain_blocks$fillc <- fill_cols[chain_blocks$state]
chain_blocks$outc <- line_cols[chain_blocks$state]
chain_blocks$txtc <- text_cols[chain_blocks$state]
arrows <- tibble::tibble(
  x = bx[-1] - wc / 2, xend = bx[-n] + wc / 2,
  broken = c(FALSE, FALSE, TRUE, FALSE))
arrows$col <- ifelse(arrows$broken, '#e34948', '#33356b')

ggplot() +
  geom_segment(data = arrows,
    aes(x = x, xend = xend, y = 0, yend = 0, colour = col),
    arrow = arrow(length = unit(0.15, 'cm'), type = 'closed'),
    linewidth = 0.5) +
  geom_rect(data = chain_blocks,
    aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax,
        fill = fillc, colour = outc), linewidth = 0.7) +
  geom_text(data = chain_blocks,
    aes(x = x, y = 0, label = label, colour = txtc), size = 2.6) +
  annotate('text', x = 3, y = 0.72, label = 'tampered',
    colour = '#e34948', size = 3) +
  annotate('text', x = 4.5, y = 0.72, label = 'invalidated',
    colour = '#e34948', size = 3) +
  scale_fill_identity() +
  scale_colour_identity() +
  coord_equal(xlim = c(0.4, 5.6), ylim = c(-0.9, 1.1)) +
  theme_void() +
  theme(legend.position = 'none')
```

## Collaborating with an LLM

An AI assistant is useful for this material and
also a frequent source of confidently stated
errors. We consider three patterns, each as a
Prompt, a Watch-for, and a Verification.

**Prompt.** 'Derive the probability that a
proof-of-work double-spend attacker with hash
share q overtakes the honest chain from z blocks
behind, and give the simple gambler's-ruin
bound.'

- *Watch for.* Assistants frequently return the
  crude $(q/p)^z$ bound and present it as the
  full answer, silently dropping Satoshi's Poisson
  correction for the attacker's head start. They
  also sometimes invert $q$ and $p$, giving a
  bound greater than one.
- *Verification.* Check the boundary cases by
  hand: the bound must equal one at $z = 0$ and
  must lie in $[0, 1]$ for all $q < p$. Compare
  the closed form against the `mc_attack`
  simulation above; they must agree to simulation
  noise.

**Prompt.** 'State the three Bitcoin Backbone
properties and what each guarantees.'

- *Watch for.* Conflation of common prefix (a
  safety property) with chain quality (a
  censorship-resistance property), and omission of
  the honest-majority precondition under which all
  three hold. Some assistants invent a fourth
  property or attribute the framework to the wrong
  paper.
- *Verification.* Map each stated property to
  safety, liveness, or fairness and confirm the
  three roles are covered exactly once. Check the
  attribution against @garay2015backbone and the
  asynchronous extension against
  @pass2017asynchronous.

**Prompt.** 'Write R to compute the
six-confirmations reversal probability and plot
it against depth.'

- *Watch for.* Off-by-one errors in the Poisson
  sum's upper limit, use of `=` for assignment or
  the `%>%` pipe against house style, and code
  that silently assumes $q < p$ without flagging
  that the formula is invalid for $q \ge 0.5$.
- *Verification.* Re-run the code and reproduce
  `p_attack(0.10, 6)` below one in a thousand.
  Confirm the function is monotone decreasing in
  $z$ and that it returns 1 at $z = 0$.

## Conclusion

In conclusion, three points are to be
emphasized. First, finality on a proof-of-work
ledger is a confidence statement rather than a
fact, and the confirmation depth is a critical
value chosen against a quantified reversal
probability, the same kind of decision as
setting a significance level or a stopping
boundary in group-sequential monitoring; for the
trial ledger, the depth at which an outcome event
may be treated as settled is ours to set by the
consequence of a reversal, and no protocol sets
it on our behalf. Second, the gambler's-ruin
analysis, refined by Satoshi's Poisson correction
for the attacker's head start, supplies the
number, and we have verified the folk 'six
confirmations' claim against both the closed form
and a simulation rather than repeating it on
faith. Third, the machinery earns its place only
where the log must be trust-minimized across
parties with no common authority; where a
regulator or sponsor can serve as a custodian, a
well-governed database with a cryptographic audit
log is the more honest choice, and saying so is
itself part of the analyst's contribution.

## Exercises

1. **Leading zeros and expected work.** A chain
   targets a hash below $T = 2^{256 - d}$. Derive
   the per-attempt success probability and the
   expected number of attempts per block as
   functions of $d$. Show that adding one required
   hexadecimal zero (four bits) multiplies the
   expected work by sixteen, and connect this to
   the memoryless property.

2. **Derive the gambler's-ruin bound.** Starting
   from the first-step relation $a_z = q\,a_{z-1}
   + p\,a_{z+1}$ with $a_0 = 1$, solve for the
   bounded catch-up probability when $q < p$ and
   obtain $a_z = (q/p)^z$. Explain why the
   solution is unbounded, hence rejected, when
   $q > p$, and what that says about a
   majority attacker.

3. **Verify six confirmations across shares.**
   Using `p_attack`, tabulate the smallest
   confirmation depth $z$ that drives the reversal
   probability below $10^{-3}$ for $q \in \{0.05,
   0.10, 0.20, 0.30, 0.40\}$. Plot required depth
   against $q$ and comment on the behavior as $q$
   approaches one half. Cite where your reasoning
   depends on the model assumptions.

4. **Simulate the full random walk.** Rather than
   the Poisson-plus-ruin shortcut of `mc_attack`,
   simulate the honest-versus-attacker race
   directly as a biased $\pm 1$ random walk run
   until the attacker either falls hopelessly
   behind or overtakes. Estimate the reversal
   probability for $q = 0.25$, $z = 6$ and compare
   to `p_attack(0.25, 6)`. Discuss the variance of
   your estimator and how many simulations you
   need for a given precision.

5. **A confirmation-depth decision rule.** A
   supply-chain custody system will treat a
   hand-off as settled once the reversal
   probability falls below a tolerance $\alpha$.
   Write a function that, given $q$ and $\alpha$,
   returns the required depth $z$ and the expected
   wall-clock latency assuming ten-minute blocks.
   Tabulate the latency-versus-$\alpha$ tradeoff
   and frame the choice of $\alpha$ as a decision
   under a quantified error probability.

6. **Chain quality by simulation.** Simulate a
   sequence of blocks where each is honest with
   probability $p$ and adversarial with
   probability $q$. Over sliding windows, estimate
   the fraction of honest blocks and compare its
   lower tail to the honest share $p$. Discuss how
   your empirical bound relates to the chain-quality
   guarantee of @garay2015backbone and why the
   adversary's realized share can exceed $q$ in
   short windows.

## Further reading

- @nakamoto2008bitcoin is the primary source and
  the origin of the Section 11 attacker-success
  calculation reproduced in this chapter. The
  reader should consult Section 11 before any
  secondary account; it is two pages and it is
  the ground truth.

- @garay2015backbone is the foundational formal
  treatment, defining common prefix, chain
  quality, and chain growth and proving they hold
  under an honest majority in a synchronous
  network. @pass2017asynchronous extends the
  analysis to the more realistic asynchronous
  setting with bounded delay, and is the reference
  for how network latency degrades the guarantees.

- @sompolinsky2015ghost introduces the GHOST
  fork-choice rule and analyses the security cost
  of high orphan rates, the natural next step once
  one understands why longest-chain discards work.

- @rosenfeld2014analysis gives a careful and
  more general analysis of the double-spend
  attack, including the attacker's optimal
  strategy and expected profit, going beyond the
  whitepaper's bound. @grunspan2018double
  provides a rigorous closed-form treatment of the
  double-spend probability and corrects several
  approximations in circulation; it repays
  reading, for it places the Poisson mixture on
  firm footing.

- @dwork1993pricing is the original proposal to
  price access via a computational cost function,
  the intellectual ancestor of proof-of-work as a
  Sybil-resistance mechanism.

- @narayanan2016bitcoin remains the reference
  textbook; its treatment of consensus and the
  longest-chain rule is the rigorous complement to
  this chapter's probabilistic emphasis.
