# The Statistics of Mining {#sec-mining}

## Learning objectives

By the end of this chapter the reader should be
able to:

- Derive block discovery as a Poisson process
  from tiny-probability Bernoulli hashing, and
  state precisely why block inter-arrival times
  are exponential.
- Fit the geometric model for attempts-to-success
  and confirm that mean work scales as
  $16^{\text{difficulty}}$.
- Analyse the difficulty retarget rule as a
  discrete-time feedback controller acting on a
  stochastic input, and predict its transient
  response to a hash-power shock.
- Estimate the orphan (fork) rate from the mean
  block time and a propagation window, and explain
  why shorter target intervals mechanically raise
  it.
- Estimate current hash rate from observed block
  timestamps, with an interval that reflects the
  exponential sampling variability.
- Name the failure modes of the simple exponential
  model on real chains.

## Orientation

We begin with an observation that should reassure
the reader. A biostatistics student who has fit an
exponential survival model, and an epidemiologist
who has watched Poisson counts accumulate on a
disease-surveillance system, already command every
tool this chapter requires. The reason is simple to
state: block arrivals are arrival data. A block
discovered on a chain is an event in continuous
time, the interval between two consecutive blocks
is a waiting time, and the stream of such events is
a point process of precisely the kind one meets in
survival analysis and in the surveillance of
incident cases. Nothing in the machinery below is
foreign to a reader trained on time-to-event
outcomes; it is a familiar subject presented in
unfamiliar terms.

The public-health problem that motivates the
chapter is one of settlement timing. The running
case study of this book is a multi-institution
clinical-trial data-integrity ledger, which we
shall refer to henceforth as the trial ledger.
Several hospitals share an append-only record of
enrolment, randomisation, and outcome events, no
single institution is trusted to hold the master
copy, and a regulator audits the record after the
fact. The record advances only as fast as blocks
are appended to it, so the arrival stream that this
chapter studies is exactly what governs how quickly
a randomisation becomes irrevocably committed and
how long an auditor must wait before a record may be
treated as settled. A vaccine cold-chain provenance
system, in which each custody handoff of a
temperature-sensitive consignment is written to a
shared ledger, furnishes a secondary example with
the same arrival-time character. In both, the
question 'how long until this record is safe to
rely upon?' is a question about block inter-arrival
times, and it is a statistical question.

Seen this way, mining is not, at its core, a
cryptographic puzzle. It is a stochastic process
observed through a control system. A miner performs
an enormous number of near-independent trials, each
with a vanishingly small success probability, and
the resulting arrival stream of blocks is a Poisson
process whose rate the protocol holds fixed by
feedback. Every object in this description is one
the reader already owns: Bernoulli trials, the
geometric and exponential waiting-time laws,
superposition of Poisson processes, memorylessness,
and a proportional controller correcting a noisy
plant.

::: {.callout-note}
## The vocabulary of this chapter

The statistical terms below will be familiar to a
reader trained in survival analysis; we state them
here in the arrival-time language this chapter uses,
alongside the few terms specific to mining.

- **Bernoulli trial and the geometric distribution.**
  A Bernoulli trial is a single success-or-failure
  experiment with fixed success probability $p$; one
  hashing attempt is such a trial. The geometric
  distribution is the law of the number of trials
  needed to obtain the first success, with mean
  $1/p$; here it counts the hashes a block costs.
- **Exponential distribution and memorylessness.**
  The exponential is the continuous waiting-time law
  with a constant hazard, meaning that the chance of
  an event in the next instant does not depend on how
  long one has already waited. This 'memoryless'
  property is the survival-analysis constant-hazard
  assumption, and it makes the wait for the next
  block indifferent to past failures.
- **Poisson process and superposition.** A Poisson
  process is a stream of events in continuous time
  whose counts in disjoint windows are independent
  Poisson variables and whose gaps are exponential;
  it is the standard model for events arriving at
  random, as in disease surveillance. Superposition
  is the fact that merging several independent
  Poisson processes yields one Poisson process whose
  rate is the sum of the parts, which is how many
  independent miners combine into a single arrival
  stream.
- **Hash rate.** The number of hashing attempts a
  miner, or the whole network, performs per second.
  It sets the block-arrival rate and is not published
  by the chain, so it must be estimated.
- **Difficulty retarget.** The rule by which the
  protocol periodically rescales the mining
  difficulty, and hence the per-attempt success
  probability, so as to hold the mean interval
  between blocks at a fixed target.
- **Feedback controller and set point.** A feedback
  controller is a rule that measures a system's
  output, compares it with a desired value called
  the set point, and adjusts an input to reduce the
  discrepancy. The retarget is such a controller,
  with the ten-minute mean interval as its set point.
- **Orphan (or fork) rate.** The fraction of mined
  blocks that are discarded because two miners found
  a block at nearly the same time and the network
  eventually kept only one. A discarded block is
  called an orphan or stale block.
- **Quantile-quantile (QQ) plot.** A diagnostic that
  plots the ordered observed values against the
  quantiles a fitted distribution predicts; points on
  the diagonal indicate agreement in distribution,
  and systematic departure signals that the assumed
  law is wrong.
:::

The identification with event and arrival data is
exact, and we shall exploit it in line rather than
reserve it for a closing remark. A block
inter-arrival time is an exponential waiting time,
so the memoryless property is the constant-hazard
assumption, a drifting arrival rate is a
time-varying hazard, and the quantile-quantile plots
and goodness-of-fit tests we shall apply are the
very diagnostics one uses to check an exponential
survival fit. The transfer runs in both directions:
the mining chain is a concrete, fully observed point
process on which to rehearse methods whose
biomedical applications, in surveillance and in
trials, are noisier and harder to validate.

We proceed constructively. We first establish the
Poisson structure, then measure it in R on a toy
chain whose hashing is genuine. We fit the geometric
law of attempts and read its diagnostics as survival
diagnostics; we develop a worked surveillance
analogy and note carefully where it ends; we treat
difficulty adjustment as a controller and watch it
absorb a doubling of hash power; then we derive a
fork-rate estimate and a hash-rate estimator. We
close by asking where the tidy exponential model
breaks on real chains, because that boundary,
whether on a chain or in a surveillance stream, is
where the interesting science resides.

## The statistician's contribution

Three judgements sit at the centre of this chapter,
and it merits emphasis that none of them is
automated by any mining tool or block explorer. Each
is a modelling act of the sort a biostatistician
performs when deciding how to treat a stream of
adverse-event or incidence times.

1. **Recognising the renewal and Poisson
   structure.** A renewal process is one in which
   the waiting-time clock restarts afresh after each
   event, so the gaps are independent and
   identically distributed. Seeing that independent
   tiny-probability hashing produces a Poisson
   arrival process, and therefore that inter-arrival
   times are exponential and memoryless, is what
   licenses every downstream calculation: the
   waiting-time law, the superposition of many
   miners into one aggregate process, and the fork
   rate. Nothing about the source code announces
   this; it is a modelling act.

2. **Treating difficulty as a controller, not a
   constant.** The retarget rule is a
   discrete-time feedback controller acting on a
   stochastic input. Framing it that way turns
   vague questions ('will block times recover after
   a hash-power surge?') into precise ones about
   transient response, stability, and oscillation,
   answerable with the tools of control theory
   rather than intuition, following Kraft, who
   analysed the retarget rule as a control problem
   [@kraft2016difficulty].

3. **Knowing the exponential model's failure
   modes.** The memoryless model is exact under
   constant hash rate and instantaneous
   propagation, and it frays predictably when those
   assumptions fail: propagation delay, non-constant
   hash rate across a retarget epoch, and
   deviations documented on real chains by Bowden
   and colleagues, who modelled Bitcoin
   block-arrival times and identified where the
   simple Poisson model fails [@bowden2020modeling].
   A statistician's value is
   as much in stating where the model stops holding
   as in fitting it.

## Block discovery as a Poisson process

We begin with the generating mechanism. Fix a
difficulty and let $H(\cdot)$ be the hash function.
A single hashing attempt succeeds when its digest
falls below the target, an event of probability $p$.
In our toy model the target is a prefix of $d$
hexadecimal zeros, so $p = 16^{-d}$, because each
hex digit takes $16$ equally likely values and $d$
of them must be zero. Attempts are, for all
practical purposes, independent Bernoulli$(p)$
trials with $p$ minuscule.

From this modest premise three classical facts
follow immediately.

- The number of attempts until the first success is
  Geometric$(p)$, with mean $1/p = 16^{d}$. This is
  the number of hashes one block costs.
- A miner performing hashes at rate $r$ per second
  therefore finds blocks at rate $\lambda = r p$
  per second. Because $p$ is tiny and $r$ is huge,
  the discrete geometric count is indistinguishable
  from a continuous exponential waiting time: the
  time to the next block is Exponential$(\lambda)$,
  with mean $1/\lambda$.
- If many miners hash independently at rates
  $r_1, r_2, \dots$, their block discoveries
  superpose into a single Poisson process of rate
  $\lambda = p \sum_i r_i$, proportional to total
  hash power. The identity of the finder is an
  independent multinomial draw with probabilities
  $r_i / \sum_j r_j$.

It helps to hold the whole correspondence in view
at once. Every object in the mining process carries
a distribution and an exact public-health twin,
together with the diagnostic one reaches for to
check it (@tbl-distribution-map).

| Object | Distribution | Public-health twin | Diagnostic |
|---|---|---|---|
| **Single hash attempt** | Bernoulli$(p)$ | Event indicator | n/a |
| **Attempts to success** | Geometric$(p)$ | Trials to first event | Geometric QQ |
| **Waiting time to next block** | Exponential$(\lambda)$ | Survival time under constant hazard | Exponential QQ, KS |
| **Arrivals in a window** | Poisson$(\lambda w)$ | Surveillance case counts | Dispersion check |

: The mining distributions and their public-health twins {#tbl-distribution-map}

The decisive property is memorylessness. Having
hashed for an hour without success leaves the
distribution of the remaining wait unchanged: the
process has no notion of being 'due'. A miner who
has found nothing all day is in exactly the position
of one who just started. This is the mining folklore
'hashing has no memory' stated as the defining
property of the exponential law,
$P(T > s + t \mid T > s) = P(T > t)$. The reader
will recognise it at once as the constant-hazard
assumption of survival analysis: a memoryless
waiting time is precisely one whose hazard does not
depend on how long we have already waited, and the
exponential is the unique continuous law with that
property.

The claim is worth seeing rather than only stating.
If we plot the survival curve of a fresh wait and,
on the same axes, the survival curve of the
remaining wait given that a period $s$ has already
elapsed without success, the two lie exactly on top
of one another (@fig-memoryless). Waiting confers no
advantage.

```{r}
#| label: fig-memoryless
#| fig-cap: 'The survival curve of a fresh exponential
#|   wait and the survival curve of the remaining wait,
#|   given that a period has already elapsed without
#|   success, coincide exactly; this is the memoryless
#|   property that makes mining a constant-hazard
#|   process.'
library(ggplot2)
PAL <- c('#2a78d6', '#1baf7a', '#eda100', '#008300',
         '#4a3aa7', '#e34948', '#e87ba4', '#eb6834')
book_theme <- theme_minimal(base_size = 11) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(
      linewidth = 0.25, colour = 'grey88'),
    axis.title = element_text(size = 10),
    legend.position = 'bottom')

lam <- 1
tt <- seq(0, 4, by = 0.05)
lab_u <- 'Unconditional P(T > t)'
lab_c <- 'Conditional P(T > s + t | T > s)'
mem <- tibble::tibble(
  t = rep(tt, 2),
  survival = c(exp(-lam * tt), exp(-lam * tt)),
  curve = factor(rep(c(lab_u, lab_c), each = length(tt)),
                 levels = c(lab_u, lab_c)))

ggplot(mem, aes(t, survival, colour = curve,
                linetype = curve)) +
  geom_line(linewidth = 0.8) +
  scale_colour_manual(
    values = setNames(PAL[1:2], c(lab_u, lab_c))) +
  scale_linetype_manual(
    values = setNames(c(1, 2), c(lab_u, lab_c))) +
  book_theme +
  labs(x = 'additional wait t', y = 'survival probability',
       colour = NULL, linetype = NULL)
```

We measure this on a toy chain whose hashing is
genuine SHA-256. The mining primitive below is the
one from @sec-pow, reused verbatim: it increments a
nonce (a counter varied to change the hashed input)
until the digest (the hash output) carries the
required prefix, and records the attempt count.

```{r ch04-mine-primitive}
library(digest)

make_block <- function(prev_hash, data, nonce = 0) {
  list(prev_hash = prev_hash, data = data,
       nonce = nonce, hash = NA)
}

mine_block <- function(block, difficulty = 2) {
  target <- strrep('0', difficulty)
  attempts <- 0L
  repeat {
    attempts <- attempts + 1L
    block$nonce <- block$nonce + 1
    block$hash <- digest(
      paste0(block$prev_hash, block$data, block$nonce),
      algo = 'sha256', serialize = FALSE)
    if (startsWith(block$hash, target)) {
      block$attempts <- attempts
      return(block)
    }
  }
}
```

## The geometric law of attempts-to-success

We now put the geometric prediction to the test. We
harvest per-block attempt counts at two difficulties
and compare them with theory. Attempts to mine one
block are Geometric with success probability
$p = 16^{-d}$, so the mean is $16^{d}$.

```{r ch04-geometric}
set.seed(4)
mine_attempts <- function(n, difficulty) {
  prev <- strrep('0', 64)
  out <- numeric(n)
  for (i in seq_len(n)) {
    blk <- mine_block(
      make_block(prev, paste0('b', i)), difficulty)
    out[i] <- blk$attempts
    prev <- blk$hash
  }
  out
}

a2 <- mine_attempts(250, 2)
a3 <- mine_attempts(150, 3)
tibble::tibble(
  difficulty = c(2, 3),
  mean_attempts = c(mean(a2), mean(a3)),
  theoretical = c(16^2, 16^3),
  ratio = c(NA, mean(a3) / mean(a2))
)
```

The observed mean attempts track $16^{d}$, and the
ratio between difficulties 3 and 2 lands near 16:
the quantitative content of 'one more hex zero
costs sixteen times the work'. But agreement in the
mean is a weak test, and a natural question arises.
Does the whole distribution match, or have we
confirmed only its first moment? A quantile-quantile
plot against the geometric law settles the point,
confirming the distributional form and not merely
the mean.

We can look even more directly first. Overlaying the
geometric mass function on the difficulty-2 attempt
counts already in hand shows the whole shape, and
not only the mean, before we turn to the more
sensitive quantile-quantile view (@fig-geom-pmf).

```{r}
#| label: fig-geom-pmf
#| fig-cap: 'The distribution of attempts to mine one
#|   block at difficulty 2, overlaid with the geometric
#|   mass function it should follow; agreement across
#|   the whole shape, not merely at the mean, is what
#|   licenses the geometric model.'
p2 <- 16^(-2)
kk <- seq_len(max(a2))
pmf <- tibble::tibble(k = kk, d = dgeom(kk - 1, prob = p2))
maxd <- dgeom(0, p2)

ggplot() +
  geom_histogram(
    data = tibble::tibble(a = a2),
    aes(a, after_stat(density)),
    bins = 30, fill = '#9a9a92',
    colour = 'white', alpha = 0.7) +
  geom_line(data = pmf, aes(k, d),
            colour = PAL[1], linewidth = 0.8) +
  annotate('text', x = 0.42 * max(a2), y = 0.92 * maxd,
           label = 'empirical attempts', colour = 'grey25',
           hjust = 0, size = 3.2) +
  annotate('text', x = 0.42 * max(a2), y = 0.78 * maxd,
           label = 'geometric pmf', colour = 'grey25',
           hjust = 0, size = 3.2) +
  book_theme +
  labs(x = 'attempts to mine one block (difficulty 2)',
       y = 'density')
```

```{r ch04-qq}
#| label: fig-geom-qq
#| fig-cap: 'Attempts-to-mine at difficulty 3 against
#|   the geometric model. Points on the diagonal
#|   indicate agreement in distribution, not only in
#|   mean.'
library(ggplot2)
q_theory <- qgeom(ppoints(length(a3)),
                  prob = 16^(-3))
ggplot(tibble::tibble(sample = sort(a3),
                      theory = sort(q_theory)),
       aes(theory, sample)) +
  geom_point(size = 0.8) +
  geom_abline(linetype = 2) +
  labs(x = 'Geometric quantiles',
       y = 'Observed attempts',
       title = paste('Attempts to mine at difficulty',
                     '3 vs geometric model'))
```

It merits emphasis that the diagnostic is not an
analogue of a survival-analysis tool; it is the same
tool. The quantile-quantile plot a biostatistician
draws to check whether censored survival times
follow an assumed law is doing the work here, applied
to a mining process in place of a time-to-event
outcome. Points on the diagonal license the
distributional assumption; systematic departure from
it, in either setting, is the signal that the assumed
law is wrong.

::: {.callout-note icon=false collapse=true}
## Check your understanding: memorylessness and the 16-fold cost

A miner has hashed for what would be, on average,
half a block's worth of work at difficulty 3
without finding a block. Is the expected remaining
work now $16^3 / 2$? Separately, why does adding one
hex zero to the target multiply the expected work
by exactly 16?

The expected remaining work is still the full
$16^3$, not half. Attempts are independent
Bernoulli trials, so the count to the next success
is memoryless (geometric); past failures carry no
information about how much longer the wait will be.
The 16-fold cost follows because requiring one more
leading hex zero divides the set of valid targets
by 16, so $p$ falls from $16^{-d}$ to
$16^{-(d+1)}$ and the mean $1/p$ rises by a factor
of 16.
:::

## Difficulty adjustment as feedback control

Block intervals are exponential with mean
difficulty divided by hash rate. If total hash rate
changes, the mean interval changes in inverse
proportion, and the protocol must move difficulty
to hold the mean at its set-point (ten minutes on
Bitcoin). But how does a protocol with no central
operator hold a mean arrival rate fixed against a
fluctuating and adversarial input? The retarget rule
is the answer: every epoch (2016 blocks on Bitcoin)
it scales difficulty by the ratio of observed mean
interval to target,

$$
d_{\text{new}} = d_{\text{old}} \cdot
  \frac{\text{target interval}}
       {\text{observed mean interval}} .
$$

Kraft [@kraft2016difficulty] analyses precisely this
rule as a control problem, and it is the framing we
adopt: the retarget is a discrete-time feedback
controller acting on a stochastic plant. The control
action is applied once per epoch, not continuously,
so the loop has a built-in dead time (a delay before
any correction can take effect): a change in hash
rate is uncorrected until the next retarget.

It is worth drawing the loop before analysing it.
Read as a block diagram, the retarget is an ordinary
negative-feedback controller with five parts: a set
point, a comparator, an update rule, a stochastic
plant, and a measured output fed back to the
comparator (@fig-control-loop).

```{r}
#| label: fig-control-loop
#| fig-cap: 'Difficulty adjustment read as a
#|   discrete-time feedback loop: the observed mean
#|   interval is compared with the ten-minute set
#|   point, the difficulty update responds once per
#|   epoch, and the mining process turns hash rate and
#|   difficulty into the next interval, so a hash-rate
#|   change is corrected only at the following retarget.'
box <- tibble::tibble(
  label = c('Set point\n10 min', 'Comparator',
            'Difficulty\nupdate',
            'Mining process\nhash rate x difficulty',
            'Observed mean\ninterval'),
  xc = c(1.2, 3.4, 5.7, 8.6, 11.6),
  hw = c(0.9, 0.75, 0.9, 1.35, 1.1))
box$xmin <- box$xc - box$hw
box$xmax <- box$xc + box$hw
box$ymin <- 2.4
box$ymax <- 3.2
box$border <- c('#33356b', '#33356b', '#33356b',
                PAL[1], '#33356b')

fwd <- tibble::tibble(
  x = box$xmax[1:4], xend = box$xmin[2:5])

ggplot() +
  geom_rect(data = box,
            aes(xmin = xmin, xmax = xmax,
                ymin = ymin, ymax = ymax),
            fill = 'white', colour = box$border,
            linewidth = 0.5) +
  geom_text(data = box, aes(xc, (ymin + ymax) / 2,
            label = label), size = 2.9, colour = '#33356b') +
  geom_segment(data = fwd, aes(x = x, xend = xend),
               y = 2.8, yend = 2.8, colour = '#33356b',
               linewidth = 0.4,
               arrow = arrow(length = unit(0.16, 'cm'),
                             type = 'closed')) +
  geom_segment(aes(x = box$xc[5], y = 2.4,
                   xend = box$xc[5], yend = 1.5),
               colour = '#33356b', linewidth = 0.4) +
  geom_segment(aes(x = box$xc[5], y = 1.5,
                   xend = box$xc[2], yend = 1.5),
               colour = '#33356b', linewidth = 0.4) +
  geom_segment(aes(x = box$xc[2], y = 1.5,
                   xend = box$xc[2], yend = 2.4),
               colour = '#33356b', linewidth = 0.4,
               arrow = arrow(length = unit(0.16, 'cm'),
                             type = 'closed')) +
  geom_segment(aes(x = box$xc[4], y = 3.9,
                   xend = box$xc[4], yend = 3.2),
               colour = PAL[3], linewidth = 0.4,
               arrow = arrow(length = unit(0.16, 'cm'),
                             type = 'closed')) +
  annotate('text', x = box$xc[4], y = 4.05,
           label = 'hash-rate change', size = 2.7,
           colour = PAL[3]) +
  annotate('text', x = (box$xc[5] + box$xc[2]) / 2,
           y = 1.32, label = 'feedback once per epoch',
           size = 2.7, colour = 'grey35') +
  coord_cartesian(xlim = c(0, 13), ylim = c(1.1, 4.2),
                  clip = 'off') +
  theme_void()
```

The reader with a trials background will find the
mechanism intuitive, for it is an adaptive procedure
of a familiar kind. A retarget re-estimates a
nuisance quantity, the effective hash rate, from the
data accumulated over an epoch, and adjusts a design
parameter, the difficulty, so as to hold a target
quantity, the mean arrival rate, at its set-point.
The construction has the same shape as an adaptive
clinical-trial design that re-estimates the sample
size at an interim analysis when the observed
nuisance variance departs from the value assumed at
planning. In both cases one looks at accumulating
data, updates an estimate of something one does not
control, and moves a lever to keep an operating
characteristic on target.

We should be careful, however, to mark where the
analogy ends, for it ends at an important place. The
retarget acts on the generating rate itself: raising
the difficulty makes blocks physically harder to
find and so slows their true arrival. Adaptive
sample-size re-estimation, by contrast, acts only on
the sampling plan, that is, on how many patients are
enrolled or how long they are followed; it does not,
and could not, alter the underlying event rate of
the disease. The mining controller reaches into the
data-generating process; the trial design reaches
only into the observation process. This distinction
is not a quibble. It is the reason a difficulty
retarget can erase a rate change entirely, whereas an
adaptive trial design must live with whatever event
rate nature supplies.

Before simulating the transient, we can tabulate
what a doubling of hash power moves, and what the
retarget subsequently restores (@tbl-hash-doubling).

| Quantity | Before retarget | After retarget |
|---|---|---|
| **Per-hash success probability** | Unchanged; set by difficulty, not power | Halved, as difficulty doubles |
| **Expected block interval** | Halved, near five minutes | Restored to ten minutes |
| **Interval variance** | Quartered; varies as the squared mean | Restored |
| **Orphan rate ($\approx \Delta / \mu$)** | Doubled | Restored to baseline |

: What a doubling of hash power moves, before and after the retarget {#tbl-hash-doubling}

We now simulate an instantaneous doubling of hash
power and watch the controller correct. But how
quickly, and how smoothly, does the mean interval
return to its set-point? The epoch is shortened to 40
blocks for illustration; the real Bitcoin epoch is
2016.

```{r ch04-retarget}
#| label: fig-retarget
#| fig-cap: 'The difficulty controller absorbs a
#|   doubling of hash power at block 250. Blocks
#|   briefly arrive twice as fast (trailing mean near
#|   five), then the next retarget restores the
#|   ten-block set-point.'
trailing_mean <- function(x, k) {
  vapply(seq_along(x),
         function(i) mean(x[max(1, i - k + 1):i]),
         numeric(1))
}

simulate_retarget <- function(n = 600, target = 10,
                              epoch = 40,
                              shock_at = 250) {
  hashrate <- 1
  difficulty <- target
  interval <- numeric(n)
  for (i in seq_len(n)) {
    if (i == shock_at) hashrate <- 2 * hashrate
    interval[i] <- rexp(1, rate = hashrate / difficulty)
    if (i %% epoch == 0) {
      recent <- mean(interval[(i - epoch + 1):i])
      difficulty <- difficulty * target / recent
    }
  }
  tibble::tibble(block = seq_len(n),
                 interval = interval)
}

set.seed(5)
sim <- simulate_retarget()
sim$roll <- trailing_mean(sim$interval, 20)

ggplot(sim, aes(block, roll)) +
  geom_line() +
  geom_hline(yintercept = 10, linetype = 2) +
  geom_vline(xintercept = 250, linetype = 3) +
  labs(x = 'block',
       y = 'trailing-20 mean interval',
       title = paste('Hash power doubles at block',
                     '250; controller re-targets'))
```

Immediately after the shock the trailing mean drops
toward five, because blocks arrive twice as fast at
the old difficulty, then the next retarget roughly
doubles difficulty and pulls the mean back to ten.
This is the negative-feedback dynamics of any
proportional controller, and it explains why a
permanent change in global hash rate produces only
a transient change in block time.

The controller is stable but not free of
oscillation. Because it corrects on the mean of a
noisy epoch and acts with a one-epoch lag, an
over-correction can overshoot in the following
epoch, and the standard rule's damping (Bitcoin
clamps the per-epoch adjustment to the interval
$[1/4, 4]$) exists precisely to bound this
oscillation. A statistician recognises the classic
tension between a controller aggressive enough to
track real changes and cautious enough not to chase
sampling noise.

::: {.callout-important}
## The sceptic's reflex

A recurring claim is that mining difficulty 'always
rises'. Name the tier before believing it. This is
an empirical claim about the past, not a protocol
fact about the future. The retarget rule is
symmetric: it lowers difficulty exactly as readily
as it raises it, and difficulty has fallen
materially in every sustained hash-rate decline
(for instance after large regional miner
shutdowns). The monotone-rise story is a narrative
extrapolated from a bull-market regime, and the
controller's own definition falsifies it. What the
protocol guarantees is regulation toward the target
interval, not a monotone difficulty path.
:::

## A surveillance reading of the same mathematics

To make the epidemiologic identification concrete,
rather than merely asserted, we develop it as a
worked example. Consider a disease-surveillance
system that records the time of each incident case
reported to a health department. Under a stable
incidence the reports arrive as a Poisson process,
so the inter-arrival gaps between successive cases
are exponential, and the analyst's first task is
exactly the one performed above for blocks: fit the
rate and check the exponential assumption with a
quantile-quantile plot and a goodness-of-fit test.
An outbreak is an increase in the arrival rate, and
detecting it is the surveillance counterpart of
detecting a hash-rate shock. To illustrate, consider
a synthetic record of six hundred cases in which the
incidence doubles midway through, at case three
hundred.

```{r}
#| label: fig-surveillance
#| fig-cap: 'SYNTHETIC disease-surveillance case
#|   arrivals. A doubling of incidence at case 300
#|   halves the trailing-mean inter-arrival gap. Unlike
#|   the mining chain, no controller restores the
#|   set-point: in surveillance the rate change is the
#|   signal we wish to detect, not a disturbance to
#|   reject.'
set.seed(72)
# SYNTHETIC: case inter-arrival gaps in days. Baseline
# incidence gives a mean gap of 7 days; an outbreak at
# case 300 doubles the rate, halving the mean gap.
baseline <- rexp(300, rate = 1 / 7)
outbreak <- rexp(300, rate = 2 / 7)
case_gap <- c(baseline, outbreak)

surveil <- tibble::tibble(
  case = seq_along(case_gap),
  roll = trailing_mean(case_gap, 30))

ggplot(surveil, aes(case, roll)) +
  geom_line() +
  geom_hline(yintercept = 7, linetype = 2) +
  geom_vline(xintercept = 300, linetype = 3) +
  labs(x = 'case number',
       y = 'trailing-30 mean inter-arrival (days)',
       title = paste('Synthetic surveillance stream:',
                     'incidence doubles at case 300'))
```

The trailing-mean gap halves at the change-point, in
the same manner as the trailing-mean block interval
dropped after the hash-power shock. It merits notice
that a marginal test on the pooled record is
relatively insensitive here, because a fifty-fifty
mixture of two exponentials remains roughly
exponential in its unordered distribution, so a
quantile-quantile plot or a Kolmogorov-Smirnov test
(a goodness-of-fit test comparing an empirical
distribution with a hypothesised one) applied to all
six hundred gaps at once may fail to reject. The change is a sequential phenomenon, not a
marginal one, and it is visible only when the data
are read in arrival order. This is the same lesson a
survival analyst learns from an unmodelled
change-point in the hazard: pooling across the change
hides it, and a time-ordered or change-point analysis
is required to recover it.

Here the two settings part company, and the parting
is instructive. The mining protocol treats a rate
change as a disturbance and deploys the retarget
controller to erase it, returning the mean interval
to its set-point. The surveillance analyst treats the
identical rate change as the quantity of interest and
wishes above all to detect it, quickly and with
controlled false-alarm rate. The mathematics of the
arrival process is common to both; what differs is
whether the change is a nuisance to be regulated away
or a finding to be reported. The reader designing an
alarm rule for the trial ledger, in which an
unexpected surge of outcome events written across
several hospitals might signal a safety problem, sits
on the surveillance side of this divide and not the
controller side.

## Fork and orphan rate

Two miners can find a block at nearly the same
height before either has heard of the other. When
the network converges on one, the other becomes an
orphan (or stale) block. This is not an academic
concern for the trial ledger: a randomisation event
written into a block that is later orphaned appears,
for a short interval, to be committed and then is
reorganised out of the accepted history, which is
precisely why an auditor must wait a number of
confirmations before treating a record as settled.
How often, then, does such a competing block occur?
Under the Poisson model the rate is a short
calculation, and Decker and Wattenhofer, who
measured block propagation in the live Bitcoin
network, ground it in observed delays
[@decker2013information]. Let $\mu$ be the mean
block time, so blocks arrive as a Poisson process
of rate $\lambda = 1/\mu$, and let $\Delta$ be the
propagation window: the time for a newly found
block to reach the rest of the network. A fork
occurs when a competing block is found within
$\Delta$ of a given block. By the memoryless
property, the wait to the next block is
Exponential$(\lambda)$, so

$$
P(\text{fork}) = P(T < \Delta)
  = 1 - e^{-\lambda \Delta}
  = 1 - e^{-\Delta / \mu}
  \approx \frac{\Delta}{\mu}
$$

for $\Delta \ll \mu$. The approximation is the
statement that matters: the orphan rate is, to
first order, the propagation window measured in
units of the mean block time. Below we compute it
across regimes and confirm the closed form by
simulating exponential inter-arrival gaps and
counting those shorter than the window.

```{r ch04-fork}
orphan_prob <- function(mean_block_time, prop_window) {
  lambda <- 1 / mean_block_time
  1 - exp(-lambda * prop_window)
}

simulate_orphans <- function(n_blocks, mean_block_time,
                             prop_window) {
  gaps <- rexp(n_blocks, rate = 1 / mean_block_time)
  mean(gaps < prop_window)
}

set.seed(40)
grid <- expand.grid(
  mean_block_time = c(600, 150, 15),
  prop_window = c(2, 6)
)
grid$closed_form <- with(grid,
  orphan_prob(mean_block_time, prop_window))
grid$simulated <- with(grid,
  mapply(simulate_orphans, 50000, mean_block_time,
         prop_window))
grid
```

The simulated column tracks the closed form to
Monte Carlo noise. Read down the table: holding the
propagation window fixed, shrinking the mean block
time from 600 seconds to 15 raises the orphan rate
by roughly the same factor, because the rate is
approximately $\Delta / \mu$. This is the
mechanical cost of fast blocks. A chain that targets
short intervals to appear 'fast' spends a larger
fraction of its work on blocks that lose the fork
race, which both wastes hash power and weakens
security, since orphaned work does not defend the
chain. The propagation window $\Delta$ is a network
property that no protocol parameter can eliminate,
so it sets a floor on the block time below which
orphan rates become costly.

::: {.callout-note icon=false collapse=true}
## Check your understanding: why fast blocks fork more

A chain proposes to cut its target block time from
ten minutes to one minute to improve user
experience, keeping the same peer-to-peer network
and therefore the same propagation window $\Delta$.
Roughly what happens to the orphan rate, and why is
this a security concern and not merely a bandwidth
one?

The orphan rate scales approximately as
$\Delta / \mu$, so cutting $\mu$ by a factor of ten
raises the orphan rate by roughly ten. It is a
security concern because work spent on orphaned
blocks does not contribute to the longest chain: an
honest majority whose effective (non-orphaned) hash
rate is reduced by forking defends the chain with
less than its nominal hash power, which lowers the
cost of a given-depth reorganisation for an
attacker whose blocks, being self-referential, do
not fork against themselves.
:::

## Estimating hash rate from block timestamps

The chain does not publish the global hash rate; it
must be inferred. The estimator falls out of the
same model. Expected work per block is $16^{d}$
hashes (in Bitcoin, the analogue is
$D \cdot 2^{32}$, where $D$ is the difficulty). If
blocks arrive with mean interval $\bar{t}$ seconds,
then the hash rate is expected work per block
divided by mean interval,
$\hat{R} = 16^{d} / \bar{t}$. The sampling
variability is inherited from $\bar{t}$: for $n$
exponential intervals the mean has relative standard
error $1/\sqrt{n}$, and by the delta method (the
first-order propagation of variance through a smooth
transformation) $\hat{R}$ has the same relative
standard error,
because $\hat{R}$ is a smooth reciprocal of
$\bar{t}$.

```{r ch04-hashrate}
# Hash-rate estimate from observed block intervals.
# rel SE of mean of n exponentials is 1/sqrt(n),
# and hat R is a reciprocal of that mean, so it
# inherits the same relative SE (delta method).
estimate_hashrate <- function(intervals, difficulty) {
  work <- 16^difficulty
  rate_hat <- work / mean(intervals)
  n <- length(intervals)
  se <- rate_hat / sqrt(n)
  c(rate_hat = rate_hat,
    lower = rate_hat - 1.96 * se,
    upper = rate_hat + 1.96 * se)
}

set.seed(41)
difficulty <- 3
true_rate <- 500                      # hashes/sec, SYNTHETIC
mean_interval <- 16^difficulty / true_rate
intervals <- rexp(2016, rate = 1 / mean_interval)
estimate_hashrate(intervals, difficulty)
```

The SYNTHETIC intervals were generated at a true
rate of 500 hashes per second, and the estimate
brackets it with a 95 percent interval whose width
reflects the $1/\sqrt{2016} \approx 2.2$ percent
relative standard error of an epoch's worth of
blocks. This is why hash-rate figures quoted from a
single day of blocks are noisy, and why practitioners
average over a difficulty epoch or longer. The
estimator is nothing more exotic than a
method-of-moments fit of an exponential rate (one
that equates the sample mean to its theoretical
value and solves for the rate), rescaled by the
known per-block work.

## When the exponential model frays

The exponential model is exact only under two
assumptions: constant hash rate and instantaneous
propagation. Real chains violate both, and knowing
where is the deliverable [@bowden2020modeling].

- **Propagation delay** makes the process not quite
  Poisson at short time scales, and produces the
  orphan events the pure model treats as
  measure-zero. It also induces weak negative
  dependence between consecutive very-short
  intervals.
- **Non-constant hash rate** within a retarget
  epoch means the rate parameter $\lambda$ drifts,
  so the intervals are exponential with a slowly
  varying rate rather than i.i.d. Across a retarget
  the interval distribution is a mixture, and the
  mean interval before a large upward hash-rate move
  differs from the mean after it. @bowden2020modeling
  document exactly these departures on Bitcoin data
  and propose non-homogeneous and doubly-stochastic
  refinements.
- **Timestamp noise.** Block timestamps are set by
  miners within a tolerance and are not exact
  arrival times, so empirical inter-arrival
  distributions carry measurement error, including
  occasional negative apparent intervals.

The honest workflow is the biostatistician's, and it
is the same whether the data are blocks or cases:
propose the memoryless model as the null, fit it, and
use quantile-quantile plots and formal tests to
locate the epochs and events where it is rejected.
Those rejections are not failures of the analysis;
they are where the physics of propagation and the
economics of hash-rate migration become visible, just
as the rejection of a constant-hazard model in a
survival analysis is where the biology of a
time-varying risk becomes visible. That the mining
chain is a fully observed point process, with every
event timed and no censoring, is precisely what makes
it a convenient rehearsal ground for methods whose
biomedical applications, in surveillance streams and
in adverse-event monitoring, are noisier and harder
to validate.

## Worked example

We tie the chapter together on a single synthetic
chain. We generate an epoch of block intervals from
the exponential model, fit the rate as if the data
had been downloaded, test the fit, then read off the
implied hash rate and the orphan rate. This is the
end-to-end analysis one would run on a real
block-explorer extract, or, in the register of this
book, on an export of the trial ledger's block
timestamps when an auditor asks how quickly records
have been settling. The generator stands in for the
download.

```{r ch04-worked}
#| label: fig-exp-fit
#| fig-cap: 'Block intervals against the fitted
#|   exponential (SYNTHETIC). On real data the
#|   interesting departures cluster around retargets
#|   and hash-rate shocks.'
set.seed(13)
# SYNTHETIC: intervals from Exp(mean 600s), the very
# model under test. Swap in a real CSV to test a chain.
intervals <- rexp(2016, rate = 1 / 600)

fit_rate <- 1 / mean(intervals)
ks <- ks.test(intervals, 'pexp', rate = fit_rate)
c(mean_interval = mean(intervals),
  fitted_rate = fit_rate,
  ks_p_value = ks$p.value)

ggplot(tibble::tibble(
    x = sort(intervals),
    theo = qexp(ppoints(length(intervals)), fit_rate)),
    aes(theo, x)) +
  geom_point(size = 0.6) +
  geom_abline(linetype = 2) +
  labs(x = 'exponential quantiles',
       y = 'observed intervals',
       title = paste('Block intervals vs fitted',
                     'exponential (SYNTHETIC)'))
```

The QQ plot lies on the diagonal and the
Kolmogorov-Smirnov test does not reject, because we
drew from the model. Now we exploit the same
intervals for the two estimates the chapter
developed. Taking difficulty 3 as the toy analogue,
the implied hash rate and the orphan rate for a
five-second propagation window are read off
directly.

```{r ch04-worked-estimates}
hr <- estimate_hashrate(intervals, difficulty = 3)
orphan <- orphan_prob(
  mean_block_time = mean(intervals),
  prop_window = 5)
c(rate_hat = unname(hr['rate_hat']),
  orphan_rate = orphan)
```

On genuine data the science is exactly where this
tidy picture breaks. Around difficulty retargets, or
during hash-rate shocks such as the one the
controller absorbed above, the memoryless assumption
frays, the QQ plot bends away from the diagonal, and
the KS test begins to reject. Reporting that boundary
honestly, rather than declaring a clean fit, is the
statistician's contribution to on-chain measurement.

In conclusion, three points are to be emphasized.
First, block arrivals are arrival data, and the
exponential and Poisson machinery a biostatistician
already owns, together with the quantile-quantile and
goodness-of-fit diagnostics of survival analysis, is
the correct and sufficient toolkit for on-chain
timing; it should be the first instrument reached
for, not a borrowed metaphor. Second, difficulty
adjustment warrants reading as a discrete-time
feedback controller acting on the generating rate,
and questions about recovery after a hash-rate change
are then answerable by its transient response rather
than by folklore about difficulty always rising or by
appeal to a continuous equilibrium force that does
not exist. Third, the value of the statistician lies
as much in stating where the memoryless model fails,
through propagation delay, non-constant hash rate,
and timestamp noise, as in fitting it, for those
rejections, whether encountered on a chain or in a
disease-surveillance stream, are where the science
resides. The reader who carries these three points
away has, we believe, grasped the essential argument
of the chapter.

## Collaborating with an LLM

An AI assistant is a capable partner for the algebra
and the plotting here, and an unreliable one for the
modelling judgements. It should be used with the
following discipline.

**Prompt.** 'Derive the orphan-rate formula for a
Poisson block process with mean interval mu and
propagation window Delta, and give the small-Delta
approximation.'

- *Watch for:* a confident derivation that silently
  assumes propagation is instantaneous, or that
  conflates the probability of a fork with the
  long-run fraction of orphaned blocks (they differ
  once forks can themselves fork).
- *Verification:* re-derive from
  $P(T < \Delta) = 1 - e^{-\Delta/\mu}$
  independently, and confirm the $\Delta/\mu$ limit
  against the
  simulation in the fork-rate section. The model
  presupposes a single competing miner within the
  window; state that assumption explicitly.

**Prompt.** 'Write R to estimate Bitcoin hash rate
from a vector of block timestamps and difficulty.'

- *Watch for:* an estimator that omits the
  $2^{32}$ (or $16^{d}$) work factor, that reports a
  point estimate with no uncertainty, or that treats
  raw timestamps as exact arrival times despite their
  known miner-set tolerance.
- *Verification:* check the units by dimensional
  analysis (work per block divided by seconds per
  block yields hashes per second), and confirm the
  relative standard error scales as $1/\sqrt{n}$ by
  resampling.

**Prompt.** 'Explain why block times return to ten
minutes after hash rate doubles.'

- *Watch for:* an answer that credits the return to
  an equilibrium force acting continuously, rather
  than to a once-per-epoch discrete controller with
  a dead-time lag.
- *Verification:* run the retarget simulation and
  confirm the recovery is stepwise at epoch
  boundaries, not smooth, and that the transient
  persists for up to a full epoch.

## Exercises

1. **Superposition.** Two mining pools hash at rates
   $r_1$ and $r_2$. Prove that the combined block
   process is Poisson with rate $p(r_1 + r_2)$ and
   that, given a block is found, the probability it
   came from pool 1 is $r_1 / (r_1 + r_2)$,
   independent of the arrival time. Verify by
   simulation.

2. **Geometric to exponential.** Show analytically
   that as $p \to 0$ with $rp = \lambda$ held fixed,
   the Geometric$(p)$ attempt count, rescaled by the
   hash rate $r$ into a time, converges in
   distribution to Exponential$(\lambda)$. Then
   reproduce the convergence numerically by driving
   `mine_attempts()` at increasing difficulty and
   overlaying the exponential limit.

3. **Controller stability.** Modify
   `simulate_retarget()` to clamp the per-epoch
   adjustment factor to $[1/4, 4]$, as Bitcoin does.
   Apply a four-fold hash-rate jump and compare the
   transient and any overshoot with and without the
   clamp. Characterise the trade-off between
   tracking speed and oscillation.

4. **Fork rate versus block time.** Using
   `orphan_prob()`, plot the orphan rate against
   mean block time for a fixed propagation window,
   on a log-log scale. Confirm the near-unit slope
   predicted by the $\Delta/\mu$ approximation and
   identify where the exact expression departs from
   it.

5. **Hash-rate estimator coverage.** By Monte Carlo,
   verify that the 95 percent interval in
   `estimate_hashrate()` attains nominal coverage for
   $n = 2016$, and investigate its coverage for small
   $n$ (say $n = 20$), where the normal approximation
   to the exponential-mean sampling distribution is
   poor. Propose and test a gamma-based interval.

6. **Model failure.** Generate intervals whose rate
   changes partway through an epoch (a hash-rate
   shock), fit a single exponential, and show through
   the QQ plot and KS test that the pooled fit is
   rejected even though each segment is exponential.
   Discuss how this maps onto a survival analysis
   with an unmodelled change-point in the hazard.

## Further reading

- @nakamoto2008bitcoin introduces proof-of-work and
  the difficulty-adjustment target; read Section 4
  and the difficulty discussion as the primary
  source for the retarget rule.
- @kraft2016difficulty analyses difficulty
  adjustment explicitly as a control problem and
  studies its stability, the framing this chapter
  adopts.
- @bowden2020modeling is the essential empirical
  companion: it models Bitcoin block-arrival times
  and documents precisely where the homogeneous
  Poisson model fails, motivating non-homogeneous
  and doubly-stochastic refinements.
- @decker2013information measures block propagation
  in the real Bitcoin network and connects
  propagation delay to the fork rate, grounding the
  orphan-rate calculation in observed data.
- @goffard2023blockastics, the BLOCKASTICS course
  materials, are the closest existing
  statistics-course precedent, presenting blockchain
  as a subject for a probability and statistics
  curriculum; they are the natural next text for a
  reader who wants this chapter's viewpoint extended.
- @narayanan2016bitcoin remains the rigorous
  reference textbook; its treatment of mining and
  difficulty complements the process-oriented account
  given here.
