# Security Economics and Attacks {#sec-security}

::: {.bookepigraph}
```{=latex}
\begin{flushright}\small
```
*We show that the Bitcoin protocol is not
incentive-compatible.*

[Eyal and Sirer]{.smallcaps}, *Majority Is Not Enough:
Bitcoin Mining Is Vulnerable* (2014)
```{=latex}
\end{flushright}
```
:::

## Learning objectives

By the end of this chapter the reader should be
able to:

- Distinguish what cryptography guarantees in
  Bitcoin from what economic incentives
  guarantee, and locate the boundary precisely.
- Compute a rental-based cost-of-attack for a
  51 percent attack and explain why the attacker
  recovers most of that cost.
- State the Eyal-Sirer selfish-mining result and
  reproduce their closed-form relative-revenue
  curve, including the sub-50-percent
  profitability threshold.
- Explain incentive compatibility as an
  equilibrium property of a mechanism, and audit
  the assumptions a compatibility claim rests on.
- Assess when a low cost-of-attack makes a small
  permissionless chain the wrong custodian for a
  health-data ledger.

## Orientation

We begin with a data-integrity problem to which
the book will return throughout. Several hospitals
participating in a multi-institution clinical trial
share a single append-only record of who was
enrolled, how each participant was randomized, and
which outcome and adverse events occurred, with no
single institution trusted to hold the master copy
and a regulator who audits the record only after
the fact. We shall call this the trial ledger.
Suppose now that a sponsor, a participating site,
or an outside party wished to alter the shared
record: to suppress an adverse event, say, or to
move an enrollment date so that a protocol deviation
disappears from view. Two questions decide whether
such a ledger deserves anyone's trust. What would
that alteration cost the party who attempted it,
and who would bear the loss were it to succeed?

These are not, in the first instance, cryptographic
questions. A digital signature can make each entry
attributable and each subsequent edit detectable,
and we quantified that guarantee in @sec-crypto.
What cryptography cannot tell us is whether a
coalition with enough of the system's resources
will find it profitable to rewrite recent history,
nor what an honest custodian stands to lose if it
does. Those are questions about payoffs, and they
are answered, if they are answered at all, by
incentives, audit, and consequences. The habit the
statistician already possesses, of asking who
benefits from a given data pattern and whether they
could plausibly have produced it, is precisely the
habit this chapter formalizes for a shared ledger.

The mechanics for such a ledger were built in the
previous two chapters, which constructed
proof-of-work and studied mining as a renewal
process (a sequence of block discoveries separated
by independent, identically distributed waiting
times). This chapter asks the question those
mechanics defer: why does anyone follow the rules?
A miner, or a custodian of the trial ledger, is
free software running on hardware its owner
controls. Nothing in the protocol compels a
participant to extend the chain everyone else
extends, to publish a block promptly, or to refrain
from rewriting recent history. The protocol is a
proposal, and the network's integrity is the claim
that following the proposal is the
profit-maximizing strategy.

That reframing is the chapter's thesis. Bitcoin's
security, and by extension the trial ledger's, is
economic, not merely cryptographic. Cryptography
does one job, and does it to a standard we
quantified in @sec-crypto: it makes signatures
unforgeable and the ledger tamper-evident, so no
participant can spend another's coins or silently
rewrite a committed block. Cryptography says nothing
about whether a majority coalition of miners will
choose to reorganize the chain, or whether a large
miner will withhold blocks so as to orphan its
rivals' work, that is, to cause validly mined
blocks to be discarded because a competing chain
has outpaced them. Those are not cryptographic
questions. The
reader trained on clinical data will recognize the
shape of the argument at once, for research
integrity too is maintained not by cryptography but
by the prospect that a fabricated pattern will be
detected, questioned, and punished. We shall see the
analogy sharpen, and also where it breaks, as the
chapter proceeds.

A reader with masters-level game theory and
probability already owns the tools. The
cost-of-attack is a comparison of two revenue
streams. Selfish mining is a Markov chain (a random
process whose next state depends only on the
current one) on the lead a miner holds over the
public chain, and its profitability is a rational
function of two parameters. Incentive compatibility
is the statement that honest mining is a Nash
equilibrium (a strategy profile from which no
participant gains by deviating alone), or close
enough to one that deviation does not pay. The work of the chapter is to write
these down and check them, because the field states
them as slogans ('51 percent is the security
threshold') that the arithmetic does not support.
We proceed as follows. We first price a majority
attack and examine what its cost recovers. We then
derive the selfish-mining threshold and read it as
a security parameter. We close by returning to the
trial ledger and asking, candidly, whether the
permissionless machine, meaning the open system
that any party may join without approval, is the
right custodian for it at all.

The chapter turns on a compact vocabulary, which we
fix here before proceeding, so that the reader need
not carry these terms as unknowns through the
argument that follows.

::: {.callout-note}
## The vocabulary of this chapter

- **The 51 percent (majority) attack.** An attack
  in which a single party commands more than half of
  the network's mining power and uses it to rewrite
  recent history, to reorder or exclude entries, and
  to spend the same coins twice.
- **Cost of attack.** The expenditure a party must
  incur to mount such an attack. Here it is priced
  as the cost of renting a majority of the mining
  power for the duration of the attack window.
- **Selfish mining.** A strategy in which a miner
  keeps the blocks it finds secret and releases them
  at moments of its own choosing, so as to waste the
  effort of honest miners and capture more than its
  fair share of the reward.
- **Hash-power fraction (alpha).** The share of the
  network's total mining power that a given miner
  controls, written $\alpha$.
- **Communication (propagation) advantage (gamma).**
  The fraction of honest miners that side with a
  selfish miner's block, rather than the competing
  honest block, when the two are released at once.
  Written $\gamma$, it measures how effectively the
  selfish miner wins the race to propagate its block
  first.
- **Relative revenue.** The share of the total block
  reward a miner earns in the long run. Honest mining
  earns a relative revenue equal to $\alpha$; an
  attack pays only when it earns more.
- **Incentive compatibility.** The property that
  following the protocol honestly is the strategy
  that maximizes a participant's own payoff, so that
  no participant gains by deviating from it.
- **Eclipse attack.** An attack that surrounds a
  node's network connections with the attacker's own
  peers, so that the victim is fed a false view of
  the ledger or has its own entries suppressed.
:::

Before pricing any single attack, we lay out the
four we shall meet and the layer that stops each,
so the cryptography-versus-economics boundary is
visible from the outset (@tbl-attack-taxonomy). Only
the first is a cryptographic matter; the rest are
questions of payoffs and network structure.

| Attack | Prevented by | Cost driver | Trial-ledger reading |
|--------|--------------|-------------|----------------------|
| **Signature forgery** | Cryptography | Breaking ECDSA, computationally infeasible | An outsider cannot author an entry in a site's name |
| **51 percent reorg** | Economics | Rental of a hash-power majority for the window | A funded party rewrites recent enrollments and outcomes |
| **Selfish mining** | Economics | Hash share above $(1-\gamma)/(3-2\gamma)$ | A large site farms orphans, biasing whose entries survive |
| **Eclipse and censorship** | Network design | Surrounding a node's peer connections | A site is fed a false ledger view or has its entries suppressed |

: The chapter's four attacks and the layer that stops each {#tbl-attack-taxonomy}

## The statistician's contribution

Three judgments structure the chapter, and none
is automated by any tool.

**1. Separate the cryptographic guarantee from
the economic guarantee, and never let one borrow
the other's certainty.** A signature scheme's
security is a statement about computational
hardness that holds against any adversary below a
work bound. An incentive-compatibility claim is a
statement about a rational adversary's payoffs
that holds only under a stated behavioral and
distributional model. The first is close to a
protocol fact; the second is an economic claim in
the sense of @sec-intro, conditional on
assumptions. Treating 'the chain is secure' as
one undifferentiated guarantee is the field's
most common category error, and the statistician's
first contribution is to refuse it. The whole
chapter can be read off a single partition: what
the mathematics guarantees, what the economics
guarantees only under a model, and what neither
underwrites at all (@tbl-guarantees). The rightmost
column is where every failure in this chapter lives.

| Cryptography guarantees | Incentives guarantee (under a model) | Neither guarantees |
|-------------------------|--------------------------------------|--------------------|
| Entries are attributable to a signer | Rewriting recent history costs more than it returns, on a large chain | That a small chain's security budget exceeds an edit's off-chain value |
| Any edit to a committed block is detectable | Honest extension is an equilibrium below the threshold | That a miner values disruption less than reward |
| No one spends coins they cannot sign for | Withholding does not pay below $(1-\gamma)/(3-2\gamma)$ | That the propagation parameter $\gamma$ is known and fixed |
| The issuance schedule cannot be rewritten | Pooling cuts variance without changing mean reward | That pooling will not concentrate share past the threshold |

: What the two guarantees cover, and what falls between them {#tbl-guarantees}

**2. Audit the assumptions behind an
incentive-compatibility claim before believing the
conclusion.** Every equilibrium result in this
chapter rests on premises: that miners are
myopically revenue-maximizing, that hash power is
divisible and rentable, that block propagation is
described by a single communication-advantage
parameter, that no miner values disruption for its
own sake. The selfish-mining result is a theorem
about a model, and the model can fail. Naming the
premises and asking whether they hold for a
specific chain is analytic work, not a literature
citation.

**3. Recognize when a threshold model's
assumptions fail so its threshold stops meaning
what it says.** '51 percent' is exact under a
model in which the only attack is a longest-chain
race by a myopic majority. Selfish mining moves
the relevant threshold below 50 percent; a chain
whose miners can be bribed moves it again;
economic irrationality (an attacker who will pay
to destroy) removes it. The statistician's job is
to know which threshold governs which threat, and
to say plainly when the headline number is the
wrong one.

## The 51 percent attack and its cost

### What a majority actually buys

Recall the double-spend analysis of @sec-pow, where
a double-spend is the act of spending the same coins
twice by rewriting the ledger after a payment has
been accepted. An attacker with hash-power fraction
$q$ who trails
the honest chain by $z$ confirmations is running a
biased random walk, and the probability of ever
catching up is $(q/p)^{z}$ when $q < p = 1 - q$,
which decays geometrically in $z$. When
$q \ge 1/2$ that ratio reaches one: catch-up
becomes certain, the geometric safety in $z$
collapses, and no number of confirmations
protects a transaction. A majority of hash power
therefore buys the ability to rewrite recent
history at will, to exclude or reorder
transactions, and to double-spend against any
confirmation depth.

It is worth stating precisely what a majority does
not buy, because the folklore overreaches. It
does not let the attacker forge signatures, spend
coins it does not hold keys for, or alter the
issuance schedule; those are enforced by every
node's validation rules, which a majority of hash
power does not override. The 51 percent attack is
a liveness-and-ordering attack on recent history,
that is, an attack on whether and in what order
entries appear rather than on their authenticity,
not a theft of arbitrary funds. That distinction
is exactly the cryptography-versus-incentives
boundary this chapter is about.

### A rental cost-of-attack

The security question is not whether a majority
attack is possible, for above 50 percent it plainly
is, but what it costs. But how much does a majority
cost to assemble? The relevant model is rental: an
attacker need not buy hardware, only command a
majority of hash power for the duration of the
attack window. We reuse the proof-of-work cost
helper from the mining chapter, which prices the
attack as the rental of just over half the
network's hash rate for the window.

```{r ch05-pow-cost}
pow_attack_cost <- function(network_hashrate_ths,
                            hours,
                            rent_per_ths_hour) {
  0.51 * network_hashrate_ths * hours *
    rent_per_ths_hour
}

# Illustrative figures, not a live measurement.
c(cost_6h  = pow_attack_cost(4e8, 6,  0.02),
  cost_24h = pow_attack_cost(4e8, 24, 0.02))
```

The arithmetic is deliberately transparent: cost
scales linearly in the hash rate commanded, in the
attack window, and in the unit rental price. The
figures are illustrative rather than a live
measurement of any chain. The lesson is
structural. Doubling the confirmation depth a
merchant waits for roughly doubles the window an
attacker must sustain, and so roughly doubles the
cost; this is the economic reading of the
confirmation-depth policy derived in @sec-pow.

### Cost recovery and the proof-of-stake contrast

A subtlety the headline cost hides is what happens
to the attacker's capital afterward. Under
proof-of-work the attacker rents or owns
general-purpose hashing hardware. When the attack
ends, the hardware retains almost all its value:
it can be resold, or redirected to honest mining,
or rented back out. The economic loss is the
rental premium and the opportunity cost of the
window, not the capital itself. Proof-of-work
attack cost is, in this sense, a recurring flow
that the attacker largely recovers.

Proof-of-stake inverts this, and we develop the
contrast fully in @sec-pos. There the analogue of
'commanding a majority' is holding a large
fraction of the staked capital, and an attack that
violates the protocol triggers slashing: the
offending stake is destroyed by the protocol
itself. The attacker's own capital sits in the
blast radius. The cost is not a recoverable rental
flow but a stock of capital placed at risk of
outright loss. Whether that makes proof-of-stake
more or less secure is a genuine and contested
question; the point here is only that the two
cost structures differ in kind, flow-recoverable
versus stock-at-risk, and conflating them is a
common error.

::: {.callout-note icon=false collapse=true}
## Check your understanding: what a majority cannot do

A commentator writes that an entity controlling
55 percent of Bitcoin's hash power 'could drain
every wallet on the network'. Using the
cryptography-versus-incentives distinction, state
precisely what is wrong with this claim and what a
55 percent coalition can in fact do.

Draining arbitrary wallets requires spending coins
whose private keys the attacker does not hold,
which is a forgery of ECDSA signatures, ECDSA being
the elliptic-curve digital-signature scheme Bitcoin
uses. That is prevented cryptographically and is not affected by
hash-power share; every node rejects a transaction
whose signatures do not verify, regardless of who
mined the block containing it. A 55 percent
coalition can rewrite recent history: exclude or
reorder transactions, censor addresses, and
double-spend its own coins against any
confirmation depth by out-racing the honest chain.
It cannot move funds it cannot sign for. The claim
commits the category error of borrowing
cryptographic certainty (unforgeability) to
describe an economic-ordering attack.
:::

## Selfish mining

### The result and why it is striking

The 51 percent analysis suggests a comforting
threshold: stay below half the hash power and the
chain is safe. Eyal and Sirer, whose selfish-mining
result first showed that a miner can profit by
withholding the blocks it finds, overturned that
comfort [@eyal2014majority]. They exhibit a
strategy, selfish mining, under which a miner with
strictly less than half the hash power earns
revenue exceeding its fair share, and, once its
share passes a threshold that can be well below 50
percent, earns more by deviating than by mining
honestly. The striking consequence is that the
honest-majority assumption is not enough. A
rational miner with, say, a third of the hash power
may be incentivized to attack, so the coalition
that threatens the chain can be far smaller than a
majority.

The mechanism is withholding, and it will repay a
moment's translation into terms the applied reader
knows well. A selfish miner who finds a block does
not publish it. It mines privately on its secret
extension, building a lead. When the honest network
finds and publishes a competing block, the selfish
miner releases its withheld block to create a fork,
and, crucially, races to have its version adopted.
Honest miners who had begun extending the public
block find their work orphaned. The selfish miner
has converted its private lead into wasted honest
effort, raising its share of the blocks that
ultimately survive above its share of the hash
power.

We can picture the maneuver as a Markov chain on
the lead the selfish miner holds, with its withheld
branch riding above the public chain
(@fig-selfish-states). The state is that lead: zero
when nothing is withheld, one when a single block
is held back, two or more when the private branch
has pulled ahead.

```{r}
#| label: fig-selfish-states
#| fig-cap: 'The selfish miner modeled as a Markov chain on the lead it holds over the public chain. Finding a block extends a withheld private branch, drawn in red above the public chain, while an honest block prompts the miner to publish and race to have its branch adopted. The state records how many blocks the private branch leads by.'
library(ggplot2)

PAL <- c('#2a78d6', '#1baf7a', '#eda100', '#008300',
         '#4a3aa7', '#e34948', '#e87ba4', '#eb6834')

book_theme <- ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(
    panel.grid.minor = ggplot2::element_blank(),
    panel.grid.major = ggplot2::element_line(
      linewidth = 0.25, colour = 'grey88'),
    axis.title = ggplot2::element_text(size = 10),
    legend.position = 'bottom')

indigo <- '#33356b'
attack_col <- '#e34948'

nodes <- data.frame(
  x = c(1, 3, 5), y = 0, lab = c('0', '1', '2+'))
pub <- data.frame(
  xmin = c(0.6, 1.6, 2.6), xmax = c(1.4, 2.4, 3.4),
  ymin = 1.2, ymax = 1.55)
wh <- data.frame(
  xmin = c(2.6, 3.6), xmax = c(3.4, 4.4),
  ymin = 1.95, ymax = 2.3)

ggplot() +
  geom_rect(data = pub, aes(xmin = xmin, xmax = xmax,
            ymin = ymin, ymax = ymax),
            fill = 'white', colour = indigo,
            linewidth = 0.6) +
  geom_rect(data = wh, aes(xmin = xmin, xmax = xmax,
            ymin = ymin, ymax = ymax),
            fill = 'white', colour = attack_col,
            linewidth = 0.6) +
  annotate('segment', x = 3, xend = 3,
           y = 1.55, yend = 1.95,
           colour = attack_col, linewidth = 0.5) +
  annotate('text', x = 0.6, y = 1.72,
           label = 'public chain', colour = indigo,
           size = 3.1, hjust = 0) +
  annotate('text', x = 4.4, y = 2.45,
           label = 'withheld private branch',
           colour = attack_col, size = 3.1, hjust = 1) +
  geom_point(data = nodes, aes(x, y), shape = 21,
             size = 16, fill = 'white', colour = indigo,
             stroke = 0.7) +
  geom_text(data = nodes, aes(x, y, label = lab),
            colour = indigo, size = 4) +
  annotate('curve', x = 1.35, xend = 2.65,
           y = 0.28, yend = 0.28, curvature = -0.4,
           colour = attack_col, linewidth = 0.5,
           arrow = arrow(length = unit(0.02, 'npc'),
                         type = 'closed')) +
  annotate('curve', x = 3.35, xend = 4.65,
           y = 0.28, yend = 0.28, curvature = -0.4,
           colour = attack_col, linewidth = 0.5,
           arrow = arrow(length = unit(0.02, 'npc'),
                         type = 'closed')) +
  annotate('text', x = 3, y = 0.72, colour = attack_col,
           size = 3.1,
           label = 'selfish finds a block: extend the secret lead') +
  annotate('curve', x = 2.65, xend = 1.35,
           y = -0.28, yend = -0.28, curvature = -0.4,
           colour = indigo, linewidth = 0.5,
           arrow = arrow(length = unit(0.02, 'npc'),
                         type = 'closed')) +
  annotate('curve', x = 4.65, xend = 3.35,
           y = -0.28, yend = -0.28, curvature = -0.4,
           colour = indigo, linewidth = 0.5,
           arrow = arrow(length = unit(0.02, 'npc'),
                         type = 'closed')) +
  annotate('text', x = 3, y = -0.72, colour = indigo,
           size = 3.1,
           label = 'honest finds a block: publish and race to be adopted') +
  annotate('text', x = 1, y = -1.05, colour = 'grey45',
           size = 2.9,
           label = 'lead 0: honest block adopted') +
  coord_cartesian(xlim = c(0.2, 5.8),
                  ylim = c(-1.2, 2.6)) +
  theme_void()
```

The essential move here is not a forgery but a
strategic misreport. The selfish miner never breaks
the letter of the protocol: every block it releases
is valid, every signature verifies. It cheats by
controlling the timing of disclosure, withholding
information at moments chosen to disadvantage its
rivals. The reader who has thought about
publication bias and selective outcome reporting
will find this familiar. An investigator who runs
an analysis, dislikes the result, and delays or
suppresses its disclosure has broken no rule of
arithmetic; the reported estimates are each
internally valid, yet the aggregate record is
distorted by what was withheld and when. Selfish
mining is that maneuver made mechanical, with a
propagation race in place of a file drawer. We
should note carefully where the analogy ends, for
it ends at an important place. Mining rewards are
zero-sum: a block the selfish miner steals is a
block some honest miner does not earn, and the
total is fixed by the issuance schedule. Scientific
evidence is not zero-sum in this way. A suppressed
trial does not transfer its truth to a competitor;
it removes information from the common pool, and the
loss is borne diffusely by future patients rather
than captured by any rival. The incentive to
withhold is real in both settings, but the
accounting of who gains and who loses differs, and
we shall not press the analogy past that boundary.

### The closed form

@eyal2014majority model the strategy as a Markov
chain on the selfish miner's lead over the public
chain and derive the miner's long-run relative
revenue in closed form. Two parameters govern it.
Let $\alpha$ be the selfish miner's fraction of
total hash power, and let $\gamma$ be the fraction
of the honest hash power that, in a fork, ends up
mining on the selfish miner's block rather than
the honest one. The parameter $\gamma$ is a
communication advantage: it measures how
effectively the selfish miner wins the propagation
race when it releases a withheld block to answer
an honest one. The relative revenue is

$$
R(\alpha, \gamma) =
\frac{\alpha(1-\alpha)^2\bigl(4\alpha +
\gamma(1-2\alpha)\bigr) - \alpha^3}
{1 - \alpha\bigl(1 + (2-\alpha)\alpha\bigr)}.
$$

Honest mining earns relative revenue equal to
$\alpha$: a miner with fraction $\alpha$ of the
hash power wins fraction $\alpha$ of the blocks in
the long run. Selfish mining pays whenever
$R(\alpha, \gamma) > \alpha$. The algebra of that
inequality yields a clean profitability threshold:

$$
\alpha > \frac{1 - \gamma}{3 - 2\gamma}.
$$

At $\gamma = 0$, the selfish miner has no
propagation advantage and the threshold is
$1/3$: withholding pays only above a third of the
hash power. At $\gamma = 1$, the selfish miner
always wins the race and the threshold falls to
$0$: any positive share benefits from deviating.
Realistic $\gamma$ sits between, so the operative
threshold sits between $0$ and $1/3$, in every
case below the naive $1/2$.

### Reproducing the curve

We implement the closed form directly and plot the
relative-revenue family against the honest
baseline $y = \alpha$.

```{r ch05-selfish-fun}
selfish_revenue <- function(alpha, gamma) {
  num <- alpha * (1 - alpha)^2 *
    (4 * alpha + gamma * (1 - 2 * alpha)) -
    alpha^3
  den <- 1 - alpha * (1 + (2 - alpha) * alpha)
  num / den
}

selfish_threshold <- function(gamma) {
  (1 - gamma) / (3 - 2 * gamma)
}

# Thresholds at the three communication regimes.
c(gamma_0   = selfish_threshold(0),
  gamma_0.5 = selfish_threshold(0.5),
  gamma_1   = selfish_threshold(1))
```

The printed thresholds are $1/3$, $1/4$, and $0$,
matching the formula. Now the curves.

```{r ch05-selfish-plot}
#| label: fig-selfish
#| fig-cap: 'Selfish-mining relative revenue against honest mining. Each curve is R(alpha, gamma) for one communication advantage; the dashed line is the honest baseline y = alpha. Where a curve rises above the dashed line, withholding pays. Vertical marks show the profitability thresholds (1-gamma)/(3-2gamma).'
library(dplyr)
library(tidyr)
library(ggplot2)

gammas <- c(0, 0.5, 1)
alpha_grid <- seq(0.001, 0.499, length.out = 400)

curves <- expand_grid(gamma = gammas,
                      alpha = alpha_grid) |>
  mutate(revenue = selfish_revenue(alpha, gamma),
         gamma = factor(gamma))

thresholds <- tibble(
  gamma = factor(gammas),
  alpha_star = selfish_threshold(gammas))

ggplot(curves,
       aes(alpha, revenue, colour = gamma)) +
  geom_line(linewidth = 0.8) +
  geom_abline(slope = 1, intercept = 0,
              linetype = 2) +
  geom_vline(data = thresholds,
             aes(xintercept = alpha_star,
                 colour = gamma),
             linetype = 3) +
  coord_cartesian(xlim = c(0, 0.5),
                  ylim = c(0, 0.75)) +
  labs(x = expression('hash-power fraction ' *
                        alpha),
       y = 'relative revenue',
       colour = expression(gamma),
       title = paste('Selfish mining beats honest',
                     'below 50% hash power'))
```

The dashed diagonal is honest revenue. For each
$\gamma$ the solid curve lies below the diagonal
until $\alpha$ crosses the dotted vertical (the
threshold), after which selfish mining earns
strictly more than the miner's fair share. The
$\gamma = 1$ curve is above the diagonal for every
positive $\alpha$, so with a decisive propagation
advantage there is no safe minority share at all.
The takeaway is the one @eyal2014majority
emphasized: the security threshold is not 50
percent, and under favorable propagation it can
be pushed arbitrarily close to zero.

The revenue curves fix a value of $\gamma$ and vary
$\alpha$; it is worth also seeing the threshold
itself as a function of $\gamma$, since that is the
security parameter a chain operator can influence
(@fig-threshold). The safe share is largest, one
third, only when the attacker has no propagation
edge, and it slides to zero as that edge becomes
decisive.

```{r}
#| label: fig-threshold
#| fig-cap: 'The selfish-mining profitability threshold falls as the communication advantage grows. The curve is the safe-share boundary (1-gamma)/(3-2gamma); a miner whose share lands in the shaded region above the curve earns more by withholding, while a miner below it does better mining honestly. The safe share is one third when the attacker has no propagation edge and vanishes as gamma approaches one.'
gamma_grid <- seq(0, 1, length.out = 200)
thr_df <- tibble::tibble(
  gamma = gamma_grid,
  alpha_star = selfish_threshold(gamma_grid))

ggplot(thr_df, aes(gamma, alpha_star)) +
  geom_ribbon(aes(ymin = alpha_star, ymax = 0.5),
              fill = PAL[1], alpha = 0.12) +
  geom_line(colour = PAL[1], linewidth = 0.9) +
  annotate('text', x = 0.48, y = 0.42,
           label = 'selfish mining pays',
           colour = PAL[1], size = 3.4) +
  annotate('text', x = 0.66, y = 0.08,
           label = 'honest mining safe',
           colour = 'grey40', size = 3.4) +
  annotate('point', x = 0, y = 1 / 3,
           colour = PAL[1], size = 1.6) +
  annotate('text', x = 0.04, y = 1 / 3 + 0.035,
           hjust = 0, label = 'gamma = 0: one third',
           colour = 'grey30', size = 3.1) +
  annotate('point', x = 1, y = 0,
           colour = PAL[1], size = 1.6) +
  annotate('text', x = 0.96, y = 0.035, hjust = 1,
           label = 'gamma = 1: zero',
           colour = 'grey30', size = 3.1) +
  coord_cartesian(ylim = c(0, 0.5)) +
  labs(x = expression('communication advantage ' *
                        gamma),
       y = expression('threshold share ' *
                        alpha^'*')) +
  book_theme
```

### What the model assumes

The result is a theorem about a model, and the
statistician's contribution is to inventory the
premises. The Markov-chain derivation assumes a
single selfish coalition (not several competing
ones), a fixed and known $\gamma$, myopic
revenue-maximization over block rewards with no
regard for exchange-rate consequences, and no
difficulty-adjustment feedback within the analysis
window. Sapirshtein and colleagues tightened the
picture by solving for the optimal withholding
policy as a Markov decision process (a Markov chain
in which a decision-maker chooses actions so as to
maximize long-run reward), showing that Eyal and
Sirer's strategy is not always optimal and
that the true profitability threshold can be lower
still [@sapirshtein2016optimal]. That is a
strengthening of the result, not a refutation:
optimizing the attacker only lowers the threshold
further. The honest reading is that
$1/3$ at $\gamma = 0$ is an upper bound on the
safe share under this family of attacks, and the
safe share is smaller once the attacker plays
optimally or enjoys any propagation edge.

::: {.callout-important}
## The skeptic's reflex

'Proof-of-work is 51-percent-attack-proof below a
majority' is a narrative wearing the cadence of a
protocol fact. Demote it. It is an economic claim,
conditional on a model in which the only deviation
is an honest-strategy longest-chain race. Selfish
mining is a deviation outside that model, and under
it the operative threshold is $(1-\gamma)/(3-2
\gamma)$, which is at most $1/3$ and can approach
$0$. The correct statement is conditional: honest
mining is incentive-compatible only for shares
below a threshold that depends on propagation, and
that threshold is below one half. Anyone quoting
'51 percent' as the security margin is quoting the
wrong number.
:::

## Incentive compatibility as mechanism design

### Why honest miners extend the longest chain

Nothing in the protocol forces a miner to build on
the chain everyone else builds on. A miner could
extend a stale block, or start a private fork, at
no cryptographic cost. Yet honest miners
overwhelmingly extend the longest (more precisely,
the highest-total-work) chain. The reason is
expected reward, and it is worth making the
argument explicitly because it is the template for
every incentive claim in the field.

A block earns its reward only if it ends up on the
chain the network eventually accepts, which is the
one with the most accumulated work. A miner who
extends a minority branch is betting that its
branch will win the work race against the rest of
the network. Unless that miner commands a majority
of hash power, the bet is a biased random walk
running against it, and the branch is orphaned with
probability approaching one. The expected reward
from extending the majority chain therefore
dominates the expected reward from any minority
branch, for any sub-majority miner. Orphan risk is
the enforcement mechanism: the protocol does not
forbid deviation, it makes deviation lose money in
expectation. This is precisely the logic that
selfish mining exploits and inverts, by
manufacturing orphans for honest miners rather than
suffering them.

### Mining pools and centralization pressure

The variance of solo mining is punishing. A small
miner finds blocks as a Poisson process (events
arriving independently at a constant average rate)
with a minute rate, so its reward is a high-variance
lottery with long dry spells. Pooling hash power,
meaning that many miners combine their effort in a
mining pool and share the rewards in proportion to
the work each contributes, converts the lottery
into a near-deterministic income stream, reducing
variance without changing expected reward. That is
a rational risk-management response, and it is why
pools dominate. It also concentrates the decision
of what to mine into a few pool operators, which is
a centralization pressure the base protocol does
nothing to resist. A statistician recognizes this
as a variance-reduction incentive with a
governance externality: individually rational
pooling aggregates into a small number of entities
whose combined share can approach the selfish-
mining and majority thresholds discussed above.
The security model assumes diffuse hash power; the
economics of variance pushes toward its
concentration.

### The security-performance frontier

Protocol parameters that improve performance often
degrade security, and the trade-off is
quantifiable. Gervais and colleagues built a
framework that, for given block interval, block
size, and network propagation, computes the cost of
attacks such as double-spending and selfish mining,
mapping out a security-performance frontier, the
boundary describing how much security must be
surrendered to gain throughput and conversely
[@gervais2016security]. Shorter block intervals and
larger blocks raise throughput but also raise the
stale-block rate, meaning the frequency with which
validly mined blocks are discarded as the network
converges on one chain, which improves an attacker's
relative position: more natural forks mean a
withholding
attacker's manufactured forks are cheaper to win.
The frontier makes explicit that 'faster' and
'more secure' are competing objectives on a
measurable curve, not free improvements. Bonneau
and colleagues, in their systematization-of-
knowledge survey, place these results in a common
framework, cataloguing Bitcoin's stability and
incentive assumptions and the attacks that probe
them [@bonneau2015sok], and it is the right map for
orienting among the individual results.

### The game-theoretic frame

Assembling the pieces, the security argument for
proof-of-work is a claim in mechanism design. The
protocol designer chooses rules (the reward
schedule, the longest-chain rule, the difficulty
adjustment) so that the behavior the system needs,
honest extension of the canonical chain, is an
equilibrium of the game the miners play. The
system is secure to the extent that honest mining
is a Nash equilibrium, or close enough to one that
no profitable unilateral deviation exists. Selfish
mining is precisely the demonstration that honest
mining is not an equilibrium for a sufficiently
large or well-connected miner, which is why the
result matters beyond its arithmetic: it is a proof
that the mechanism is not incentive-compatible over
the whole parameter range its designers implicitly
assumed. Reading Bitcoin as a mechanism, and asking
whether honest behavior is an equilibrium under
stated assumptions, is the frame that turns a pile
of attacks into a single question.

::: {.callout-note icon=false collapse=true}
## Check your understanding: threshold under a propagation edge

A mining pool controls a well-connected relay
network, so that in a fork roughly 60 percent of
honest hash power adopts its block. It holds 28
percent of total hash power. Is honest mining
incentive-compatible for this pool, and what is the
operative threshold?

The communication advantage is $\gamma = 0.6$, so
the selfish-mining threshold is
$(1 - 0.6)/(3 - 2 \cdot 0.6) = 0.4 / 1.8 \approx
0.222$. The pool's share of $0.28$ exceeds this
threshold, so honest mining is not
incentive-compatible: the pool earns more by
withholding than by mining honestly. Note that a
naive '51 percent' analysis would have declared the
pool harmless. The propagation edge, not just the
raw share, is what moves the threshold, which is
why $\gamma$ must be part of any serious assessment.
:::

## A public-health reading: threat-modeling the trial ledger

We now return to the trial ledger with which the
chapter opened, and to the two questions we posed
of it: what an alteration would cost, and who would
bear the loss. A recurring proposal is to put such
a ledger, or a vaccine cold-chain provenance system
of the same character, on a permissionless
blockchain, so that its integrity does not depend
on trusting a single custodian. The security
analysis of this chapter is exactly the threat
model that proposal needs, and we shall find that it
usually undermines the proposal rather than
supporting it.

The integrity of a permissionless chain is only as
strong as the cost of attacking it. That cost, as
the rental calculation showed, is a function of the
chain's total hash rate. A large chain like Bitcoin
has an attack cost in the millions of dollars per
day, but a small or new chain, exactly the kind a
niche health registry might spin up or select for
low fees, can have a 51 percent rental cost of a
few thousand dollars, which is trivial relative to
the stakes of a trial record. Worse, the value an
attacker extracts from corrupting the ledger is not
the on-chain token value to which the security
budget is scaled; it is the off-chain value of
altering a record, which can be far larger. A chain
secured by a small mining reward is defending
high-stakes data with a low-value security budget,
an economic mismatch the cost-of-attack framing
makes explicit.

There is a further failure, and it is the one that
actually occurs. A threat model is a claim about
where trust resides, and a system can satisfy every
consensus property we have derived while placing its
trust somewhere the analysis never looked. The
clearest documented instance is a health credential:
an international travel pass, built on a permissioned
identity ledger, deployed at scale, and audited
independently. The audit found that the laboratories
issuing the credentials did not hold their own
signing keys. A single commercial intermediary held
them [@lin2022iata]. Whatever the ledger's consensus
guarantees may have been, an adversary who compromised
that one firm could mint whatever credentials it
pleased, and the decentralization the architecture
advertised was, in the auditors' assessment,
nullified in practice. The ledger appeared in the
architecture diagram and was absent from the threat
model. We commend this as the first question to ask
of any health-ledger proposal put before a committee,
in preference to any question about consensus: *who
holds the keys, and what happens if that party is
compromised?* The chain is rarely the weakest link.
It is merely the most conspicuous one, and the
literature on data quality in these systems makes the
complementary point that a ledger records faithfully
whatever it is given, including what is false
[@abuhalimeh2023quality].

To illustrate, consider a consortium of six
hospitals that proposes to host the trial ledger on
a small proof-of-work chain, chosen for its low
transaction fees, whose total hash rate is a
thousandth of a large network's. Suppose a site
under enrollment pressure wishes to erase a single
serious adverse event, and that doing so would
protect a product whose off-chain value, in avoided
liability and preserved approval prospects, runs to
several million dollars. We price the two sides of
the ledger explicitly.

```{r ch05-ledger-threat}
# SYNTHETIC illustration, not a live measurement.
small_chain_hashrate_ths <- 4e5   # 1/1000 of large
attack_window_hours      <- 6     # confirmations
rent_per_ths_hour        <- 0.02

attack_cost <- pow_attack_cost(
  small_chain_hashrate_ths,
  attack_window_hours,
  rent_per_ths_hour)

offchain_value_of_edit <- 3e6     # avoided liability

c(attack_cost_usd  = attack_cost,
  offchain_value   = offchain_value_of_edit,
  ratio_value_cost = offchain_value_of_edit /
    attack_cost)
```

Laid out as a tidy ledger, the two sides of the
threat model sit side by side and the mismatch is
plain (@tbl-ledger-threat).

```{r}
#| label: tbl-ledger-threat
knitr::kable(
  tibble::tibble(
    Quantity = c(
      'Small-chain hash rate (TH/s)',
      'Attack window (hours)',
      'Rental price (USD per TH/s-hour)',
      'Rental cost of the reorg (USD)',
      'Off-chain value of the edit (USD)',
      'Value-to-cost ratio'),
    Value = c(
      format(small_chain_hashrate_ths, big.mark = ',',
             scientific = FALSE),
      format(attack_window_hours),
      format(rent_per_ths_hour),
      format(round(attack_cost), big.mark = ',',
             scientific = FALSE),
      format(offchain_value_of_edit, big.mark = ',',
             scientific = FALSE),
      paste0(round(offchain_value_of_edit /
                     attack_cost), ' to 1'))),
  caption = paste('The hospital-consortium threat',
                  'model, SYNTHETIC, priced on both',
                  'sides of the ledger.'),
  align = c('l', 'r'))
```

The rental cost of rewriting six hours of the small
chain's history is a few tens of thousands of
dollars, while the off-chain value of the edit is
some two orders of magnitude larger. The ratio is
the whole argument: when the reward for corrupting
the ledger dwarfs the cost of doing so, the
economic security the chain was supposed to provide
has evaporated, and the append-only record protects
nothing that a determined and moderately funded
party cannot undo. The signatures still verify; the
integrity does not.

For high-stakes, low-on-chain-value health data, a
permissioned design with legal accountability often
dominates. If the participating institutions are
known, credentialed, and legally liable (the
hospitals, the trial sponsor, the regulator), the
threat is not an anonymous majority coalition but a
nameable party who can be audited, sued, and
prosecuted. A permissioned ledger, or a well-
governed replicated database with cryptographic
audit logging, delivers tamper-evidence, the
property actually wanted, without staking integrity
on a hash-rate market the health system does not
control. This is the negative result @sec-intro
promised the book would state candidly: for
high-stakes health data, legal accountability in a
permissioned design frequently dominates economic
security in a small permissionless one, and the
cost-of-attack calculation is how a statistician
demonstrates the point rather than merely asserting
it. We do not claim the permissioned design is free
of its own difficulties, for concentrating
custody re-introduces the trusted party the
blockchain was meant to eliminate; we claim only
that the economic accounting must be done before,
not after, the architecture is chosen.

## Worked example

We tie the chapter together by placing the two
attack economies side by side and reading the
selfish-mining threshold as a security parameter.
The first computation contrasts a rental
proof-of-work attack with the stock-at-risk
proof-of-stake structure developed in @sec-pos, so
the flow-versus-stock distinction is concrete.

```{r ch05-cost-contrast}
library(tibble)

pos_attack_cost <- function(total_staked_usd,
                            slash_fraction = 1) {
  0.34 * total_staked_usd * slash_fraction
}

# Illustrative, not a live measurement.
tibble(
  mechanism = c('PoW rental (6h)',
                'PoW rental (24h)',
                'PoS stake-at-risk'),
  cost_usd = c(
    pow_attack_cost(4e8, 6,  0.02),
    pow_attack_cost(4e8, 24, 0.02),
    pos_attack_cost(1e11, 1)),
  recovered = c('most (hardware retains value)',
                'most (hardware retains value)',
                'little (slashed)'))
```

The proof-of-work rows are a recoverable rental
flow; the proof-of-stake row is capital largely
destroyed on use. The numbers are illustrative, but
the two columns encode the structural lesson of the
chapter: a headline cost means little without
knowing how much of it the attacker gets back.

The second computation reads the selfish-mining
threshold as the security parameter it is, tabling
the minimum safe hash-power share against the
communication advantage.

```{r ch05-threshold-table}
gamma_seq <- c(0, 0.25, 0.5, 0.75, 1)

tibble(
  gamma = gamma_seq,
  safe_share_upper = selfish_threshold(gamma_seq),
  revenue_at_share =
    selfish_revenue(selfish_threshold(gamma_seq) +
                      1e-6, gamma_seq))
```

At each $\gamma$, `safe_share_upper` is the largest
share for which honest mining remains
incentive-compatible, and it falls monotonically
from $1/3$ to $0$ as the attacker's propagation
advantage grows. The revenue evaluated just above
each threshold sits essentially at the honest
baseline, confirming the threshold is where the two
strategies cross. The single number a security
assessment should quote is not 50 percent but this
$\gamma$-dependent share, and producing it is the
chapter's deliverable.

## Collaborating with an LLM

Language models are fluent on this material and
also reliably wrong in specific ways. Three
prompt patterns, each with what to watch for and
how to verify.

**Prompt.** 'Derive the selfish-mining
profitability threshold and explain each step.'

*Watch for.* Models frequently quote the threshold
as a flat $1/3$ or $25$ percent, dropping the
dependence on the communication advantage $\gamma$
entirely, and some conflate the Eyal-Sirer
strategy with the later optimal policy of
@sapirshtein2016optimal without noting the
difference.

*Verification.* Check the stated threshold reduces
to $(1-\gamma)/(3-2\gamma)$ and reproduces $1/3$
at $\gamma = 0$ and $0$ at $\gamma = 1$. Evaluate
the closed form at a share just above the claimed
threshold and confirm it exceeds $\alpha$, as the
worked example does.

**Prompt.** 'Estimate the cost of a 51 percent
attack on a given chain.'

*Watch for.* A confident single dollar figure that
silently fixes a rental price, an attack window, and
a hash rate without stating them, and that omits
the cost-recovery point entirely, so the number
reads as a pure loss when most of it is recovered.

*Verification.* Require the three inputs
(hash rate, window, unit rental price) as explicit
arguments, recompute with the transparent helper,
and insist the answer separates the gross cost from
the net unrecovered cost.

**Prompt.** 'Is putting our patient registry on a
blockchain more secure than a database?'

*Watch for.* An answer that treats 'blockchain' as
uniformly high-integrity and skips the cost-of-
attack scaling, missing that a small chain's
security budget may be trivial relative to the
value of corrupting a health record.

*Verification.* Ask the model to state the chain's
approximate 51 percent rental cost and compare it
to the off-chain value of altering a record. If it
cannot, treat its security claim as a narrative and
apply the permissioned-alternative analysis above.

## Exercises

1. **Boundary of the majority attack.** Starting
   from the catch-up probability $(q/p)^{z}$ for
   $q < p$, show that as $q \to 1/2$ the
   probability tends to $1$ for every fixed $z$,
   and interpret why no finite confirmation depth
   protects a transaction against a majority.

2. **Cost scaling.** Using `pow_attack_cost`,
   derive how the attack cost scales if a merchant
   doubles its required confirmation depth (and
   hence the attack window). Confirm the linear
   scaling numerically and explain the connection
   to the confirmation-depth policy of @sec-pow.

3. **Reproduce the threshold algebraically.**
   Starting from $R(\alpha, \gamma) > \alpha$ with
   the closed form given, derive the profitability
   threshold $(1-\gamma)/(3-2\gamma)$. Verify your
   derivation against `selfish_threshold` at
   $\gamma \in \{0, 0.5, 1\}$.

4. **Selfish mining by simulation.** Implement the
   selfish-mining Markov chain on the miner's lead
   over the public chain, simulate it for a few
   thousand blocks at $\alpha = 0.35$ and several
   $\gamma$, and estimate the miner's realized
   relative revenue. Compare the estimates to
   `selfish_revenue` and report the agreement to
   simulation noise.

5. **Pooling and variance.** Model solo mining
   reward over a fixed horizon as a compound
   Poisson process and quantify the reduction in
   the coefficient of variation of reward when a
   miner joins a pool holding $k$ times its hash
   power. Discuss how this variance incentive
   drives the centralization pressure of the
   chapter.

6. **Health-ledger threat model.** For a
   hypothetical patient-consent registry, estimate
   the 51 percent rental cost of hosting it on a
   small chain with a stated hash rate, estimate the
   off-chain value of corrupting a consent record,
   and write a short recommendation comparing the
   permissionless option to a permissioned ledger
   with legal accountability. Be explicit about
   which assumptions your recommendation is most
   sensitive to.

## Further reading

@eyal2014majority is the primary source for
selfish mining and should be read first; the
Markov-chain derivation of the relative-revenue
formula reproduced here is worth working through in
full, because the modeling choices are where the
result's assumptions live.
@sapirshtein2016optimal recasts the problem as a
Markov decision process and solves for the optimal
withholding policy, showing the Eyal-Sirer strategy
is not always optimal and lowering the
profitability threshold further; read it for how
much the honest-majority assumption actually buys.
@gervais2016security is the quantitative
security-performance framework, and the source to
consult before claiming any parameter change is a
free improvement. @bonneau2015sok is the
systematization-of-knowledge survey that organizes
Bitcoin's stability and incentive results into a
common framework, and is the best single map of the
territory. The reference textbook throughout,
@narayanan2016bitcoin, treats mining incentives and
the 51 percent attack with the rigor this chapter
aims to extend, and its treatment of pool economics
and centralization pressure complements the game-
theoretic framing developed here.

## In conclusion

In conclusion, three points are to be emphasized.
First, the security of a shared ledger, whether
Bitcoin's or the trial ledger's, is an economic
guarantee resting on a stated behavioral and
distributional model, and it must never be allowed
to borrow the near-certainty of the cryptographic
guarantee that sits beside it; the two answer
different questions, and conflating them is the
field's most common error. Second, the headline
'51 percent' is the wrong security parameter for
proof-of-work: selfish mining moves the operative
threshold to $(1-\gamma)/(3-2\gamma)$, at most a
third and possibly near zero, and a serious
assessment must quote this propagation-dependent
share rather than the folklore number. Third, for
high-stakes health data whose off-chain value
exceeds the on-chain security budget, legal
accountability in a permissioned design frequently
dominates economic security in a small
permissionless one; the cost-of-attack calculation
developed here is the instrument by which the
statistician decides the question honestly, and it
should be carried out before an architecture is
chosen, not after.
