# Ethereum and Programmable State {#sec-ethereum}

::: {.bookepigraph}
```{=latex}
\begin{flushright}\small
```
*What Ethereum intends to provide is a blockchain with a
built-in fully fledged Turing-complete programming
language that can be used to create 'contracts' that can
be used to encode arbitrary state transition functions.*

[Vitalik Buterin]{.smallcaps}, *Ethereum Whitepaper*
(2014)
```{=latex}
\end{flushright}
```
:::

## Learning objectives

By the end of this chapter the reader should be
able to:

- Contrast the UTXO model with the account model
  as two representations of a record that changes
  over time, the event-log form versus the
  current-state form, and derive the consequences of
  each for replay protection and parallel validation.
- Explain why the account nonce is a counter, not a
  proof-of-work nonce, and state precisely what
  problem it solves.
- Explain gas as a metered-computation solution to
  the halting problem, and convert gas and gwei into
  ether costs.
- Derive and simulate the EIP-1559 base-fee update
  rule, and read it as a discrete-time feedback
  controller analogous to the difficulty adjustment
  of @sec-mining.
- State what EIP-4844 blobs price, and why that
  pricing is a data-availability problem rather than
  a computation problem.
- Evaluate a 'health records on Ethereum' proposal
  against the constraint that on-chain data is
  public.

## Orientation

We begin with a problem in the governance of
research data. Consider a longitudinal cohort study
intended to follow its participants for a decade or
more. The consent each participant grants is
anything but monolithic: a given variable may be
used for one purpose and not another, a given
specimen may be shared with one laboratory and not a
second, a permission may expire on a stated date,
and any of it may be withdrawn at any time. In a
study whose data are handled by dozens of analysts
across several institutions over many years,
enforcing that consent correctly is a substantial
administrative burden, and the characteristic
failure is familiar: a permission is revoked in one
system and never propagated to the others, or a
data-use condition is honoured by a conscientious
analyst and quietly overlooked by a hurried one. We
should prefer, if we could arrange it, that the
permission rules execute themselves, so that access
is granted or refused by machinery rather than by
each downstream analyst's individual diligence.

The device that makes a rule execute itself, in a
setting where no single custodian is trusted to run
it faithfully, is the smart contract, and it is the
principal subject of this chapter. The idea did not
originate in public health; its historical setting
is digital cash, and the question Ethereum was built
to answer was first posed in those terms. Bitcoin's
ledger answers one narrow question, namely which
unspent outputs exist and who may spend them, and
that question has a compact and beautifully parallel
representation, the UTXO set, which needs nothing
more to support a payment system. Ethereum was
designed to answer a broader question: given an
arbitrary program and an arbitrary starting state,
what is the state after applying an arbitrary
sequence of transactions? A consent policy that
enforces itself is one instance of such a program,
and answering the general question in a
decentralised, adversarial setting is what the
remainder of this chapter develops.

The reader already holds the right frame. A
blockchain that supports general computation is a
replicated deterministic state machine: every
validator applies the same transition function to
the same starting state and must reach the same
ending state, or consensus fails. That is a
transition function in the Markov-chain sense we
have used for years, except that the state space is
astronomically large (every account balance and
every contract's storage, all at once) and the
transition function is written by whoever deploys a
contract. The engineering problem is making that
computation both expressive and safe to run on a
machine we do not control, and the solution, gas, is
a pricing mechanism before it is anything else.

This chapter also introduces the first genuine
feedback controller we shall simulate end to end,
namely the base fee of EIP-1559, the standard
adopted in 2021 under which the protocol itself
computes the per-gas price and adjusts it block by
block rather than leaving it to an open auction. It
is a smaller and more transparent system than the
mining difficulty
adjustment of @sec-mining, and studying it here, with
the earlier controller already in hand, makes the
general pattern, an error signal driving a
multiplicative update toward a set point, impossible
to miss the next time it appears.

We proceed as follows. We begin by contrasting the
UTXO and account representations of a ledger, a
choice we shall recognise as the event-log versus
current-state distinction already met in
representing a longitudinal record. We then describe
the EVM as the shared deterministic machine on which
contracts run, and gas as the pricing mechanism that
makes running untrusted code on that machine safe.
With those pieces in hand we develop EIP-1559 as a
feedback controller and simulate its transient
response, return to the programmable-consent problem
with which we opened and scope precisely what a
contract can and cannot be trusted to hold, and
close with the data-availability pricing of
EIP-4844, a 2024 change that adds a cheap,
short-lived form of bulk on-chain storage.

Before proceeding we fix the vocabulary on which
the chapter depends, so that the reader need not
carry these terms as mysteries. Each is given in
plain form here and treated more fully, with its
statistical reading, in the section or table that
follows.

::: {.callout-note}
## The vocabulary of this chapter

- **The UTXO and account models.** Two ways a
  ledger may represent balances. In the UTXO
  (unspent-transaction-output) model, used by
  Bitcoin, the ledger stores discrete unspent
  outputs, and a balance is obtained by summing the
  ones an owner may still spend. In the account
  model, used by Ethereum, the ledger stores one
  running balance per address and overwrites it as
  transactions arrive.
- **Ethereum Virtual Machine (EVM).** The shared,
  rule-bound computer that every Ethereum node runs
  identically, so that a given program applied to
  given inputs yields one and only one result on
  every machine.
- **Smart contract.** A program stored on the
  ledger that runs automatically and identically on
  every node when it is called, enforcing its rules
  without any custodian trusted to execute them
  faithfully.
- **Gas and gwei.** Gas is the unit in which the
  EVM meters computational work, each operation
  costing a fixed amount. Gwei is the small unit of
  ether, $10^{-9}$ ether, in which the price paid
  per unit of gas is quoted.
- **Account nonce.** A per-account counter,
  advanced by one with each transaction the account
  sends, that orders that account's transactions
  and prevents a signed transaction from being
  replayed. It is unrelated to the mining nonce of
  proof-of-work met in @sec-pow.
- **Externally owned and contract accounts.** An
  externally owned account is controlled by a
  private key held by a person or institution and
  may originate a transaction; a contract account
  is controlled by its own stored code and acts
  only when it is called.
:::

## The statistician's contribution

1. **Reading a fee market as a control system.**
   EIP-1559 is not merely 'a fee'; it is a
   proportional feedback loop with a known gain and a
   known set point. Recognising the update rule as a
   discrete-time controller lets us predict its
   transient behaviour under a demand shock and reason
   about stability, rather than treating the base fee
   as an opaque market outcome.
2. **Recognising gas as a price on computational
   complexity.** Gas costs are a per-opcode tariff
   schedule. Auditing a contract's expected gas
   consumption is a complexity analysis in disguise,
   and the statistician's habit of decomposing a total
   into per-unit costs applies directly.
3. **Auditing determinism and its limits.** The EVM
   is deterministic by specification, but a system
   'wired to' the chain (an oracle feeding it a
   price, or a multisig, a wallet that requires
   several keys to authorise an action, holding an
   admin key) reintroduces trust and single points
   of failure that the deterministic core does not
   have. Distinguishing
   what the protocol guarantees from what a particular
   deployment adds is an exercise in scoping a claim
   correctly, which is exactly the discipline of
   @sec-intro's falsifiability tiers.

## From UTXO to the account model

We have already, in earlier chapters, met two ways
of representing a record that changes over time. One
may store the current state, a single row per entity
that is overwritten as facts change, or one may
store an event log, an append-only sequence of the
changes themselves from which the current state is
reconstructed by replay. The epidemiologist knows
both forms intimately: a registry table holding each
subject's present status is a current-state
representation, whereas a sequence of timestamped
enrolment, visit, and outcome events is an event
log. The contrast between Bitcoin's UTXO model and
Ethereum's account model is precisely this
data-modelling choice, transposed to a ledger, and
the trade-offs are the ones the reader would
anticipate. We set the two representations of the
same ledger side by side in @fig-event-vs-state
before drawing out their consequences.

```{r ch06-event-vs-state}
#| label: fig-event-vs-state
#| fig-cap: >
#|   The same ledger under two representations. On
#|   the left the UTXO model keeps an append-only log
#|   of immutable outputs from which a balance is
#|   derived by replay; on the right the account
#|   model keeps one mutable row per address, plus a
#|   nonce that is overwritten as each transaction
#|   arrives.
library(ggplot2)
library(tibble)

PAL <- c('#2a78d6', '#1baf7a', '#eda100', '#008300',
         '#4a3aa7', '#e34948', '#e87ba4', '#eb6834')

book_theme <- ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(
    panel.grid.minor = ggplot2::element_blank(),
    panel.grid.major = ggplot2::element_line(
      linewidth = 0.25, colour = 'grey88'),
    axis.title = ggplot2::element_text(size = 10),
    legend.position = 'bottom'
  )

structure_col <- '#33356b'

events <- tibble(
  y = c(6.0, 4.5, 3.0, 1.5),
  label = c('t1  new output: +2 to A',
            't2  new output: +1 to B',
            't3  A spends output to B',
            't4  new output: +1 to A')
)

ggplot() +
  geom_rect(data = events,
    aes(xmin = 0.3, xmax = 4.3, ymin = y - 0.55,
        ymax = y + 0.55),
    fill = '#eef4fc', colour = PAL[1], linewidth = 0.4) +
  geom_text(data = events, aes(x = 0.55, y = y, label = label),
    hjust = 0, size = 2.9, colour = structure_col) +
  annotate('text', x = 0.3, y = 7.3, hjust = 0,
    label = 'event log (UTXO reading)', fontface = 'bold',
    size = 3.3, colour = structure_col) +
  annotate('segment', x = 5.0, xend = 5.0, y = 6.4, yend = 1.0,
    arrow = arrow(length = unit(0.15, 'cm'), ends = 'first'),
    colour = 'grey55', linewidth = 0.4) +
  annotate('text', x = 5.15, y = 3.7, hjust = 0.5, angle = 90,
    label = 'append only, nothing overwritten',
    size = 2.6, colour = 'grey45') +
  annotate('text', x = 6.6, y = 7.3, hjust = 0,
    label = 'current state (account reading)',
    fontface = 'bold', size = 3.3, colour = structure_col) +
  geom_rect(aes(xmin = 6.6, xmax = 9.9, ymin = 1.5, ymax = 6.55),
    fill = '#eafaf3', colour = PAL[2], linewidth = 0.4) +
  annotate('segment', x = 6.6, xend = 9.9, y = 5.6, yend = 5.6,
    colour = PAL[2], linewidth = 0.3) +
  annotate('text', x = 6.85, y = 6.05, hjust = 0,
    label = 'address    balance', fontface = 'italic',
    size = 2.8, colour = structure_col) +
  annotate('text', x = 6.85, y = 5.1, hjust = 0,
    label = 'A              3', size = 2.9, colour = structure_col) +
  annotate('text', x = 6.85, y = 4.3, hjust = 0,
    label = 'B              1', size = 2.9, colour = structure_col) +
  annotate('segment', x = 6.6, xend = 9.9, y = 3.6, yend = 3.6,
    colour = PAL[2], linewidth = 0.3, linetype = 'dashed') +
  annotate('text', x = 6.85, y = 3.1, hjust = 0,
    label = 'nonce (version)', fontface = 'italic',
    size = 2.7, colour = structure_col) +
  annotate('text', x = 6.85, y = 2.3, hjust = 0,
    label = '4  (overwritten each tx)', size = 2.8,
    colour = structure_col) +
  xlim(0, 10) + ylim(0.5, 7.7) +
  theme_void()
```

To make the choice concrete, consider the running
case study of this book, a data-integrity ledger for
a multi-institution clinical trial in which several
hospitals share a single append-only record of
enrolment, randomisation, and outcome events, with
no single trusted custodian and a regulator auditing
after the fact. We shall refer to it as the trial
ledger. The trial ledger is, by its nature, an event
log: no hospital overwrites a prior entry, and the
state of the trial at any moment (how many subjects
are enrolled, to which arm each was randomised) is
reconstructed by replaying the recorded events.
Whether one then chooses to maintain a current-state
summary alongside the log, and who is trusted to
compute it, is precisely the UTXO-versus-account
question in applied dress.

Bitcoin's state is a set of unspent outputs, each
locked by a script; a transaction consumes some
outputs entirely and creates new ones. There is no
notion of an account balance anywhere in the
protocol: a balance is a client-side convenience
computed by summing the outputs one can spend. This
has two consequences worth stating precisely.

**Replay protection is implicit.** An output can be
spent exactly once, because spending it removes it
from the UTXO set; there is no separate counter to
maintain, because consumption itself is the ordering
mechanism.

**Validation is embarrassingly parallel.** Two
transactions that spend disjoint sets of outputs
touch disjoint state and can be validated
concurrently; the only shared resource is the UTXO
set membership test, which is a hash-table lookup.

Ethereum instead stores a mapping from account
address to account state: a balance, a nonce, and
(for a contract account) a code hash and a storage
root. A transaction debits one account and credits
another by directly mutating the balance fields,
the way a bank's ledger works, not the way a
till full of physical banknotes works. In the
vocabulary just introduced, the account model is a
current-state representation: it stores each
account's present balance as a single mutable field
and overwrites it as transactions arrive. The UTXO
model, by contrast, is closer to an event log, in
that the protocol stores the discrete outputs
themselves and a balance is a derived quantity,
reconstructed by summing the outputs an owner may
still spend. The current-state form is the more
natural representation for a programmable system,
because a contract routinely needs to reason about
the current balance of an address as a single
persistent number (a token contract's ledger, an
escrow's holdings) rather than reconstruct it from a
history of discrete outputs every time.

The convenience of the current-state form is not
free. The account model gives up the two UTXO
advantages above and must manufacture substitutes.

**Replay protection becomes explicit: the nonce.**
Because an account balance is a mutable field, two
identical transactions ('send 1 ETH from A to B')
are literally indistinguishable at the level of
their content. Ethereum therefore attaches to every
account a monotonically increasing counter, the
account nonce, incremented by exactly one each time
the account originates a transaction. A transaction
is valid only if its declared nonce equals the
sender account's current nonce; once processed, the
account's nonce advances, so the identical
transaction cannot be replayed, and transactions
from one sender are strictly ordered by this
counter.

::: {.callout-warning}
## Two different things named 'nonce'

Do not conflate the account nonce with the
proof-of-work mining nonce of @sec-pow. The mining
nonce is a search variable in a discrete-uniform
brute-force search over a hash's range, re-chosen at
random or by increment until a difficulty target is
met; it exists per attempted block header and has no
relationship to any specific account. The account
nonce is a per-account counter with no cryptographic
search attached to it at all; it exists to give a
total order to one account's transactions and to
make replay of a signed message impossible. Ethereum
uses the same English word for two unrelated
objects, one from the consensus layer, one from the
account layer, and the confusion this produces in
casual writing is common enough to be worth naming
explicitly.
:::

**Parallelisability is no longer free.** Because two
transactions from different accounts can still touch
the same piece of shared state (both call the same
contract, both trade against the same liquidity
pool), validating and executing transactions in the
account model cannot in general be parallelised
without first inspecting which storage slots each
transaction reads and writes. This is a genuine cost
of the account model relative to UTXO's disjoint-set
structure, and it is one reason Bitcoin-style chains
remain attractive for pure payment throughput even
though they cannot express contracts.

@tbl-utxo-account collects the contrast in one
place, with the two consequences just derived, replay
protection and parallelism, set beside the audit and
use characteristics that follow from them.

| Property | UTXO model (Bitcoin) | Account model (Ethereum) |
|---|---|---|
| **State representation** | Set of unspent outputs; balance is derived by replay | Mutable row per address (balance, nonce) overwritten in place |
| **Replay protection** | Implicit: spending removes the output | Explicit: per-account nonce counter |
| **Parallel validation** | Embarrassingly parallel over disjoint outputs | Not free: must inspect storage slots touched |
| **Audit trail** | Native event log of discrete outputs | Current-state snapshot; history reconstructed from blocks |
| **Natural use** | High-throughput payments | Programmable stateful contracts |

: The UTXO and account models as event-log and current-state representations of the same ledger {#tbl-utxo-account}

::: {.callout-note icon=false collapse=true}
## Check your understanding: why not just hash the account state directly

**Q.** Why does Ethereum need an explicit nonce at
all? Why not forbid resending a transaction
whose exact byte content has already been seen,
the way one might deduplicate by hashing?

**A.** Deduplicating by content hash prevents replay
of the identical bytes, but it does not establish
*order*. Suppose account A sends two different
valid transactions in the same block, one to pay
rent and one to buy groceries; both are distinct in
content, so content-hash deduplication would accept
either first. The chain still needs a rule for which
one A intended to happen first, because A's balance
after the first affects whether the second is
affordable. The nonce solves both problems at once:
it forbids replay (nonce reused) and it fixes total
order per sender (nonce 5 must be processed before
nonce 6), which a content hash alone cannot express.
:::

## The EVM as a shared deterministic state machine

The Ethereum Virtual Machine is a stack-based
virtual machine specified precisely enough that any
two independent, correct implementations execute
any transaction to bit-identical resulting state.
That specification is Wood's Ethereum Yellow Paper
[@wood2014yellow], the formal mathematical
definition of the machine and of its
state-transition rules. That determinism is the
entire
point: thousands of independently operated nodes
must agree on the result of running arbitrary code
without communicating during the computation, and
they can only do so if 'run this code on this
input' has exactly one possible answer.

Two kinds of accounts populate this state machine.
An **externally owned account** (EOA) is controlled
by a private key and has no code; it is the analogue
of a Bitcoin address, the thing a human or an
institution actually holds keys to. A **contract
account** has code and, on receiving a transaction
(or a message call from another contract), executes
that code deterministically against its own
persistent storage. Only an EOA can originate a
transaction, meaning the chain always starts a state
transition from a genuine act of key-holder
authorization: this is the anchor that ties the
'world computer' picture back to the same signature
scheme studied in @sec-crypto. @tbl-eoa-contract
sets the two account kinds side by side, read once
more against the trial ledger, where a hospital holds
an EOA and the shared record's logic lives in a
contract account.

| Property | Externally owned account | Contract account |
|---|---|---|
| **Controlled by** | A private key | Its own deployed code |
| **Can initiate a transaction** | Yes | No; only reacts to a call |
| **Holds code** | No | Yes, executed deterministically |
| **Trial-ledger reading** | A hospital or auditor holding keys | The shared enrolment-logic and permission rules |

: Externally owned accounts and contract accounts contrasted {#tbl-eoa-contract}

Buterin, who proposed Ethereum in 2013, in the
original Ethereum whitepaper [@buterin2014ethereum]
describes this design as a single shared computer
that anyone may submit programs to and whose entire
history of execution is
public and auditable. The framing is useful and also
easy to over-read: it is one *logical* computer with
thousands of *redundant* physical executions, which
is why every unit of computation is expensive
relative to a conventional server, a point the next
section makes precise.

## Gas as a solution to the halting problem

A Turing-complete language admits programs whose
halting cannot be decided in general (the halting
problem). A blockchain that executes arbitrary
submitted code therefore faces a genuine hazard: a
contract could contain a loop that never terminates,
and every validating node would be obligated to run
it forever, or forever fail to reach the state that
depends on it. But how is a decentralised network to
bound the cost of running code it did not write,
without first solving a problem known to be
undecidable?

Ethereum does not solve the halting problem. No one
can. It sidesteps it economically: every EVM
opcode, that is, every single primitive machine
instruction, has a fixed gas cost, a sender must
pre-purchase a
gas limit for a transaction by paying up front, and
execution halts the instant the gas supplied is
exhausted, reverting all state changes made by that
transaction (though the gas already consumed is not
refunded). A transaction is therefore always
guaranteed to terminate in bounded wall-clock time,
because it is guaranteed to terminate in bounded
*opcode count*, and opcode count times gas cost per
opcode is bounded by the gas limit the sender agreed
to pay for. This converts an undecidable question
('will this program halt?') into a decidable
resource-accounting one ('has this program consumed
its budget?'), at the cost that some programs that
would eventually halt on their own are killed early
because their author underpriced them.

Gas is denominated in an abstract unit; what the
sender actually pays is
$$
\text{fee} = \text{gas used} \times \text{gas price},
$$
where gas price is quoted in **gwei**, $10^{-9}$
ether. A simple transfer costs 21{,}000 gas; a
contract call can cost anywhere from tens of
thousands to millions of gas depending on the storage
writes and computation it performs
[@antonopoulos2018ethereum]. Because gas price is
itself set by a market (studied next), the ether cost
of the same computation varies with network demand
even though the gas quantity for a given opcode
sequence does not.

The vocabulary of the fee market is compact but easy
to conflate, so we gather it in @tbl-gas-glossary,
each term paired with a statistical reading that
recasts it as a familiar quantity: a per-unit cost, a
price, a unit, a controller state variable, a
premium, or a budget cap.

| Term | Meaning | Statistical reading |
|---|---|---|
| **Gas** | Abstract unit metering work done by opcodes | Cost per instruction |
| **Gas price** | Ether paid per unit of gas | Price per unit |
| **Gwei** | $10^{-9}$ ether, the unit gas price is quoted in | Unit of the price |
| **Base fee** | Protocol-set price per gas, burned | Controller state variable |
| **Tip (priority fee)** | Extra per-gas bid for ordering | Priority premium |
| **Gas limit** | Maximum gas a transaction may consume | Budget cap |

: A gas glossary, each term paired with its statistical reading {#tbl-gas-glossary}

::: {.callout-note icon=false collapse=true}
## Check your understanding: gas versus a CPU quota

**Q.** How does gas differ from an ordinary
compute-time quota, such as a cloud function's
execution-time limit?

**A.** A time quota measures wall-clock seconds on
one machine and is not portable across hardware of
different speeds. Gas instead prices each EVM
*opcode* individually, at a cost fixed by the
protocol specification and independent of any single
executor's hardware. This is essential in a system
where thousands of heterogeneous machines must all
agree on exactly how much 'work' a given piece of
code represents: the accounting has to be part of
the protocol, not an artefact of whichever machine
happens to run it, or nodes with faster hardware
could disagree with slower ones about whether a
transaction should have been killed.
:::

## EIP-1559: the fee market as a feedback controller

Prior to EIP-1559, Ethereum used a first-price
auction for block space: senders bid a gas price,
miners included the highest bidders, and the market
cleared with high variance and a poor user
experience, in that bidders routinely over- or
under-paid relative to the true clearing price. EIP-1559 [@eip1559] replaced this
with a protocol-computed **base fee** that adjusts
block by block toward a target utilisation, plus an
optional **priority fee** (tip) that senders set to
compete for ordering within a block. The base fee
itself is burned, not paid to the block proposer,
which is a monetary-policy design choice this chapter
does not pursue further; the object of interest here
is the update rule.

Each block has a gas limit and, since the London
upgrade, a gas target set at half the gas limit. Let
$g_n$ be the gas used in block $n$, let
$\bar g$ be the gas target, and let $b_n$ be the
base fee in force for block $n$. The protocol sets

$$
b_{n+1} = b_n \left(1 + \frac{1}{8}\cdot
\frac{g_n - \bar g}{\bar g}\right).
$$

Read as a control system: $g_n - \bar g$ is the
error signal (how far the block's gas usage deviated
from the target), $\bar g$ normalises it to a
fraction, $1/8$ is the proportional gain, and the
update multiplies the previous base fee by
$(1 + \text{gain} \times \text{error fraction})$.
This is a discrete-time, first-order, multiplicative
proportional controller: if blocks run consistently
full ($g_n > \bar g$), the base fee rises
geometrically each block until demand is priced out
and usage falls back toward the target; if blocks run
consistently empty, the base fee decays geometrically
toward zero (in practice, toward a protocol floor).
The controller has no integral or derivative term, so
it corrects proportionally to the current error only,
the same structural family as the Bitcoin difficulty
adjustment of @sec-mining, which also multiplies the
previous target by a ratio of observed to desired
throughput. The two differ in what they control (a
price, versus a proof-of-work difficulty) and in
their update cadence (every block, versus every 2016
blocks), but the underlying idea, a set point, an
observed deviation, and a bounded multiplicative
correction, is identical, and recognising the pattern
here is the payoff of having derived it once already.

For the reader whose intuition is clinical rather
than industrial, an honest analogy from our own
field is adaptive dose titration. In titrating a
drug toward a target response, one raises the dose
when the measured response falls short of the target
and lowers it when the response overshoots, adjusting
in bounded increments at each visit rather than
leaping to a putative optimum. The base-fee rule has
the same shape: the target utilisation plays the role
of the target response, block fullness plays the role
of the measured response, and the fixed 1/8 gain
plays the role of the bounded per-visit dose step. We
should be candid, however, about where the analogy
ends, for it ends rather quickly. A titration
protocol governs a single patient whose response
follows a pharmacological process that is, at least
in principle, stable and non-strategic; the base fee
governs an aggregate of anonymous and frankly
adversarial bidders who will reorganise their
behaviour in anticipation of the rule itself. A
clinician, moreover, may exercise judgment, impose
safety bounds, and abandon a titration that is going
wrong, whereas the base-fee controller is a fixed
formula with no such discretion. The analogy is
therefore useful for locating the base fee in a
family of mechanisms the reader already understands,
and misleading if pressed into a claim that a fee
market will behave like a well-characterised
dose-response curve.

The following simulation is entirely synthetic: gas
usage per block is not drawn from any real chain. It
is constructed in three phases so that the
controller's full transient response is visible: a
baseline near the target, a sustained demand shock
that pushes blocks toward the gas limit, and a
cooldown in which usage falls *below* target, the
realistic consequence of a base fee that has become
expensive enough to price out marginal demand, before
the system settles back to a target-hugging baseline.
A demand shock followed immediately by a return to
exactly-target usage would leave the base fee
permanently elevated with no force pulling it back
down, because the controller only reacts to the
*current* deviation from target; the cooldown phase
is what actually drives the fee back down, and is a
realistic feature of fee markets, not an artefact
introduced for cosmetic effect.

```{r ch06-eip1559-sim}
#| label: fig-basefee
#| fig-cap: >
#|   SYNTHETIC gas usage across a demand shock and
#|   its cooldown (top) and the resulting EIP-1559
#|   base fee path (bottom). The base fee rises
#|   geometrically while blocks run above target,
#|   then falls once the shock's own expense pushes
#|   usage below target during the cooldown.
library(ggplot2)
library(dplyr)
library(tibble)

set.seed(20260703)

gas_limit <- 30e6
gas_target <- gas_limit / 2
n_blocks <- 220
shock_start <- 60
shock_end <- 100
cooldown_end <- 160

# SYNTHETIC gas usage in three phases: near-target
# noise, a sustained demand shock near the gas
# limit, a cooldown below target (demand priced out
# by the now-elevated fee), then a return to
# near-target noise.
baseline_frac <- rnorm(n_blocks, mean = 0.5, sd = 0.03)
shock_frac <- rnorm(n_blocks, mean = 0.97, sd = 0.02)
cooldown_frac <- rnorm(n_blocks, mean = 0.15, sd = 0.03)
phase <- ifelse(
  seq_len(n_blocks) < shock_start, 'baseline',
  ifelse(seq_len(n_blocks) <= shock_end, 'shock',
    ifelse(seq_len(n_blocks) <= cooldown_end,
      'cooldown', 'recovery'))
)
gas_used_frac <- pmin(pmax(
  case_when(
    phase == 'shock' ~ shock_frac,
    phase == 'cooldown' ~ cooldown_frac,
    TRUE ~ baseline_frac
  ), 0
), 1)
gas_used <- gas_used_frac * gas_limit

# base fee update rule, EIP-1559, applied
# sequentially; base_fee[1] is an arbitrary starting
# level in gwei.
base_fee <- numeric(n_blocks)
base_fee[1] <- 20
for (n in seq_len(n_blocks - 1)) {
  error_frac <- (gas_used[n] - gas_target) / gas_target
  base_fee[n + 1] <- base_fee[n] * (1 + error_frac / 8)
}

sim <- tibble(
  block = seq_len(n_blocks),
  phase = phase,
  gas_used_frac = gas_used_frac,
  base_fee_gwei = base_fee
)

p_gas <- sim |>
  ggplot(aes(block, gas_used_frac)) +
  geom_line(colour = 'grey40') +
  geom_hline(yintercept = 0.5, linetype = 'dashed') +
  labs(
    x = NULL, y = 'gas used / gas limit',
    title = 'SYNTHETIC block fullness (target = 0.5)'
  ) +
  theme_minimal()

p_fee <- sim |>
  ggplot(aes(block, base_fee_gwei)) +
  geom_line(colour = 'steelblue') +
  labs(
    x = 'block number', y = 'base fee (gwei)',
    title = 'EIP-1559 base fee response'
  ) +
  theme_minimal()

p_gas
p_fee
```

```{r ch06-eip1559-summary}
sim |>
  filter(block %in% c(
    shock_start - 1, shock_start + 10,
    shock_end, shock_end + 30, cooldown_end + 30
  )) |>
  select(block, phase, gas_used_frac, base_fee_gwei)
```

The base fee is essentially flat while blocks
hover near the 50 percent target, accelerates upward
once the shock pushes usage toward the gas limit
(each block above target compounds the previous
block's fee by up to 12.5 percent, the maximum
single-block change under this rule), and then falls
back once the cooldown phase pushes usage below
target, settling near the end of the run *below* the
pre-shock base fee: the controller has no memory of
where it started, only of the current error, so the
extended below-target cooldown pulls the fee past its
original level before the final recovery phase holds
it near target again. The 1/8 gain bounds the
per-block change to at most $\pm 12.5\%$, which is a
deliberate design choice: it makes the fee
predictable enough that a wallet can reliably compute
'the base fee for the next block' from public
information (it is a formula, not an auction
outcome), at the cost of responding to a genuine
demand shock only gradually, block by block, rather
than instantaneously.

::: {.callout-important}
## The sceptic's reflex

'EIP-1559 makes fees predictable' is often stated as
if it means fees became low or stable in level. It
does neither. It is a tier-one protocol fact that the
*update rule* is deterministic and computable from
public information; it is a tier-three economic claim
that this makes the ether cost of a transaction
low or that base fee volatility disappeared. During a
genuine demand shock the base fee still compounds
geometrically block by block, and a user submitting a
transaction still faces uncertainty about the tip
required to be included promptly. What EIP-1559
actually delivers is a *known functional form* for
the base fee's evolution, which is valuable for
wallets and for the statistician modelling the
series, not a promise about its level.
:::

## EIP-4844: pricing data availability

EIP-4844 [@eip4844] introduces **blob-carrying
transactions**: a transaction type that attaches a
large binary object (a 'blob', roughly 128 KB) which
is stored by the consensus layer for a bounded
retention window (currently on the order of eighteen
days) rather than being kept permanently in state or
processed by the EVM at all. Blobs are priced by
their own independent EIP-1559-style mechanism, with
their own target and limit per block, entirely
separate from the gas market for ordinary execution.

The reason this needs its own market is that a blob
is not purchasing *computation*, it is purchasing
**data availability**: a rollup (a layer-2 chain that
batches many transactions and periodically posts a
compressed summary back to Ethereum) needs the
underlying transaction data to be publicly available
long enough for anyone to reconstruct and challenge
the rollup's state, but has no need for the EVM to
execute anything against that data directly. Charging
for this in ordinary gas conflated two different
scarce resources, computation and storage-for-a-
retention-window, that have different cost
structures and different demand curves; EIP-4844
separates them into two independently clearing
markets, each obeying the same controller family
introduced above. This is best understood as the
first concrete step of a broader scaling roadmap in
which the base chain increasingly specialises in
ordering and data availability, and execution moves
to rollups; the details of that architecture are
beyond this chapter's scope, but the pricing logic
is already in hand.

## Scaling and the layer-two stack

To teach EIP-4844 and its blobs without first
teaching rollups is to hand the reader an answer
while withholding the question. Blobs exist because
rollups need somewhere cheap to put their data, and
rollups exist because the base chain, on its own,
cannot carry the transaction volume that a
general-purpose settlement layer is asked to bear. We
take up that question here, and only with it in view
does the data-availability pricing of the previous
section acquire its motivation.

The difficulty is usually stated as a trilemma. A
blockchain is asked to be decentralised (runnable by
many independent, modestly resourced nodes), secure
(costly to attack), and scalable (able to process
many transactions per second), and the informal
observation, which is engineering folklore rather
than a theorem, is that a single base layer can
comfortably supply any two of the three but not all
three at once. The reason is already in hand from the
account and EVM material of this chapter. Every full
node re-executes every transaction against the shared
state, because that is precisely what makes the
result trustless: consensus holds only if each node
can independently reach the same ending state. But
how, then, is such a chain to process more
transactions per second? Raising the throughput of a
chain that every node must fully re-execute raises
the hardware demand on every node in lockstep, and a
chain that only a few well-resourced machines can
keep up with has surrendered the decentralisation
that motivated it in the first place. The
re-execution that buys trustlessness is exactly what
caps throughput.

The dominant response, and the one the EIP-4844
material was quietly preparing, is to stop asking the
base chain to execute everything. A rollup executes
transactions off-chain, on a separate system with
its own sequencer, the party that decides the order
in which the incoming transactions are processed,
and posts back to the base chain only a compressed
record of what it did, together with
enough information for anyone to check the result.
Crucially, the rollup does not run its own consensus
over who may write history; it borrows the base
chain's. Its data, or a proof of its correctness, is
anchored on Ethereum, so that the security of the
batched transactions reduces to the security of the
layer they settle on. In the standard vocabulary the
base chain is called layer one and the rollup built
upon it layer two. The throughput gain comes from
doing the expensive work, execution, once and
off-chain, while the base chain is asked only to
store the data and check a summary. We set the
arrangement out in @fig-rollup-flow.

```{r ch06-rollup-flow}
#| label: fig-rollup-flow
#| fig-cap: 'A rollup executes transactions off-chain and posts compressed data plus a state commitment back to the base chain, so that data availability and settlement are inherited from layer one while execution is not.'
library(ggplot2)

PAL <- c('#2a78d6', '#1baf7a', '#eda100', '#008300',
         '#4a3aa7', '#e34948', '#e87ba4', '#eb6834')
structure_col <- '#33356b'

box <- function(xmin, xmax, ymin, ymax, fill, col) {
  annotate('rect', xmin = xmin, xmax = xmax, ymin = ymin,
    ymax = ymax, fill = fill, colour = col, linewidth = 0.4)
}

ggplot() +
  box(0.2, 2.2, 5.0, 6.4, '#eef4fc', PAL[1]) +
  annotate('text', x = 1.2, y = 5.7, label = 'many users',
    size = 3.0, colour = structure_col) +
  box(3.3, 7.0, 4.2, 7.2, '#f2f0fb', PAL[5]) +
  annotate('text', x = 5.15, y = 6.7, label = 'rollup (layer two)',
    fontface = 'bold', size = 3.1, colour = structure_col) +
  annotate('text', x = 5.15, y = 6.0, label = 'off-chain execution',
    size = 2.7, colour = structure_col) +
  annotate('text', x = 5.15, y = 5.4, label = '+ sequencer orders txs',
    size = 2.7, colour = structure_col) +
  annotate('text', x = 5.15, y = 4.7, label = 'batches many txs',
    size = 2.6, colour = 'grey45') +
  box(8.1, 11.6, 3.4, 7.2, '#eafaf3', PAL[2]) +
  annotate('text', x = 9.85, y = 6.7, label = 'base chain (layer one)',
    fontface = 'bold', size = 3.1, colour = structure_col) +
  annotate('text', x = 9.85, y = 6.0, label = 'blob data (EIP-4844)',
    size = 2.7, colour = structure_col) +
  annotate('text', x = 9.85, y = 5.4, label = '+ state commitment',
    size = 2.7, colour = structure_col) +
  annotate('text', x = 9.85, y = 4.6,
    label = 'fraud proof or validity proof',
    size = 2.5, colour = 'grey45') +
  annotate('text', x = 9.85, y = 4.0, label = 'checked here',
    size = 2.5, colour = 'grey45') +
  annotate('segment', x = 2.25, xend = 3.25, y = 5.7, yend = 5.7,
    arrow = arrow(length = unit(0.14, 'cm')), colour = 'grey50') +
  annotate('text', x = 2.75, y = 6.05, label = 'submit',
    size = 2.4, colour = 'grey45') +
  annotate('segment', x = 7.05, xend = 8.05, y = 5.7, yend = 5.7,
    arrow = arrow(length = unit(0.14, 'cm')), colour = 'grey50') +
  annotate('text', x = 7.55, y = 6.05, label = 'post data',
    size = 2.4, colour = 'grey45') +
  annotate('text', x = 7.55, y = 5.35, label = '+ proof',
    size = 2.4, colour = 'grey45') +
  annotate('segment', x = 9.85, xend = 5.15, y = 3.3, yend = 4.15,
    arrow = arrow(length = unit(0.12, 'cm')), colour = structure_col,
    linetype = 'dashed', linewidth = 0.35) +
  annotate('text', x = 7.5, y = 3.0,
    label = 'security inherited from layer one',
    size = 2.6, colour = structure_col) +
  xlim(0, 11.9) + ylim(2.6, 7.4) +
  theme_void()
```

How the summary is checked divides rollups into two
families, and the distinction merits precise
statement, because it is a distinction in trust
assumption and not merely in engineering. An
OPTIMISTIC rollup posts its transaction data and
simply asserts that the resulting state is correct,
without proof; the assertion is accepted by default,
subject to a challenge window during which any honest
party who has recomputed the batch may submit a fraud
proof demonstrating an invalid transition and reverse
the result. Arbitrum [@kalodner2018arbitrum], one of
the most widely used optimistic rollups in
production, is the canonical instance of this
design. Its security rests
on the assumption that at least one honest and
attentive verifier exists and will act within the
window, and on the underlying data being available
for that verifier to check, which is exactly why
blobs [@eip4844] matter: the fraud proof is possible
only if the data was published in the first place. A
VALIDITY rollup, commonly a zero-knowledge rollup
(one whose proof establishes that the computation
was performed correctly while disclosing nothing
about its inputs), instead posts a succinct
cryptographic proof that the state transition was
computed correctly, which the
base chain verifies directly; there is no challenge
window and no honest-verifier assumption, because an
invalid batch cannot produce a valid proof. The proof
machinery this relies upon is the zero-knowledge
material we shall take up in @sec-crypto, and we defer
its mechanics to that discussion. Gudgeon et al.
[@gudgeon2020layertwo] offer a systematisation of
knowledge across the whole layer-two design space, and
the reader who wishes to place these two families in
their full taxonomy is referred to it. We set the two
side by side in @tbl-rollup-families.

| Property | Optimistic rollup | Validity (zero-knowledge) rollup |
|---|---|---|
| **What is posted** | Transaction data plus an asserted new state | Transaction data plus a succinct validity proof |
| **How correctness is enforced** | Fraud proof during a challenge window | Cryptographic proof verified on-chain immediately |
| **Trust assumption** | At least one honest, attentive verifier acts in time | None beyond the soundness of the proof system |
| **Withdrawal finality** | Delayed by the challenge window | As fast as the base chain confirms the proof |
| **Canonical example** | Arbitrum [@kalodner2018arbitrum] | See the zero-knowledge material of @sec-crypto |

: Optimistic and validity rollups contrasted {#tbl-rollup-families}

Rollups are not the only route off the base layer.
The older idea is the payment or state channel, in
which two parties (or a small group) open a channel
by committing funds on-chain, then transact between
themselves off-chain by exchanging signed balance
updates, and settle back on-chain only when they
close the channel; the intervening transactions,
possibly very many of them, never touch the base
layer at all. The Lightning Network
[@poon2016lightning] is the best-known realisation of
this idea, layered on Bitcoin, and it routes payments
across a mesh of such channels so that two parties
without a direct channel may still transact through
intermediaries. Channels excel at high-frequency
payments between stable sets of counterparties and
are correspondingly poor at one-off interactions with
strangers, to whom one would not open a funded
channel in the first place; rollups, by contrast,
serve open participation at the cost of the heavier
machinery just described.

We should be candid about what layer two costs, for
the throughput and the lower per-transaction fee are
not obtained for nothing. Croman et al.
[@croman2016scaling], in an early and careful analysis
of Bitcoin's scaling limits, make the general point
that scaling a decentralised ledger forces trade-offs
rather than dissolving them, and the point transfers
directly to layer two. On the one hand, a rollup
genuinely raises throughput and lowers cost, by
amortising execution over a batch and by paying only
for data availability rather than for base-layer
execution. On the other hand, it introduces trust
assumptions the base layer did not carry: most rollups
today run a single sequencer that orders transactions,
and a centralised sequencer can censor or reorder,
which is a weaker guarantee than the base chain's; an
optimistic rollup imposes a challenge-window latency,
of the order of a week for withdrawals, during which a
posted result is not yet final; and a validity rollup
substitutes the considerable complexity of a proving
system, whose own soundness must then be trusted. None
of these is disqualifying, but each is a real cost,
and a proposal that presents layer two as free
throughput has simply not looked at the other side of
the ledger.

The bearing on our own subject is direct. Consider a
national-scale health-provenance system of the kind
this book keeps returning to, one that records every
vaccine vial and every custody transfer as it moves
from manufacture to administration. The event volume
such a system generates, tens or hundreds of millions
of discrete provenance events, is far beyond what a
base chain that every node re-executes could settle
directly, and if one genuinely intends to operate at
national scale, a rollup is the architecture by which
the necessary throughput is reached. It is worth
noticing that the sequencer-trust question this raises
is not a new question at all. Whom do we trust to
order and to publish the batched events, and what
happens if that party misbehaves, is precisely the
custodian question of @sec-consensus returning in a
new guise, one layer up. The technology has moved the
trust rather than abolished it, and locating exactly
where it now sits is the whole of the analytic task.

We should be equally candid in the other direction.
For the single-consortium trial ledger that has served
as our running example, a handful of hospitals sharing
one append-only record, the base layer's throughput is
already ample, and reaching for a rollup would add
sequencer trust, proof-system complexity, or a
challenge window in exchange for a throughput headroom
the application does not need. The question of whether
an added layer is warranted is a special case of the
more basic question Wüst and Gervais [@wust2018need]
pose, namely whether one needs a blockchain at all;
the discipline of asking it honestly does not stop at
layer one. Layer two earns its complexity only when
the transaction volume genuinely exceeds what the base
layer can settle, and below that threshold it is
complexity for its own sake.

::: {.callout-note icon=false collapse=true}
## Check your understanding: what does a rollup actually inherit

**Q.** A rollup is routinely described as inheriting
the security of the base chain it settles on.
Inheriting it in exactly what respect, and in what
respect not?

**A.** It inherits *data availability* and
*settlement*: once the rollup's data and its
commitment are posted to the base chain, they are as
available and as immutable as any other base-chain
data, and correctness is anchored on-chain by a fraud
proof (optimistic) or a validity proof
(zero-knowledge). What it does *not* inherit is
censorship resistance in ordering. A single sequencer
decides, off-chain, which transactions are included
and in what order, and it can delay or exclude a user
even though it cannot forge an invalid state
transition. An optimistic rollup, moreover, does not
inherit the base chain's near-immediate finality,
because a posted result is contestable until its
challenge window closes. So 'inherits security'
means the base chain guarantees that anyone can
reconstruct and verify the rollup's state, not that
the sequencer cannot make a particular user wait.
:::

## When code is law: smart-contract security

The phrase most often used to commend the smart
contract is that with it, code is law: the agreement
executes
itself exactly as written, and no clerk, no court, and
no counterparty may intervene to alter the outcome.
The determinism this chapter has emphasised is exactly
what makes the promise credible. A contract does what
its bytecode says and nothing else, every node agrees
on the result, and no privileged party can quietly
reinterpret the terms. What is said less often, and
matters more, is the price of the same property. A
contract executes exactly as written including where
what was written is wrong, and on an immutable chain a
discovered bug cannot be patched in place the way a
flaw in a conventional service is corrected by
deploying a fix. The code is law even when the code is
mistaken, and the mistake is then as immutable as
everything else on the chain.

The sharpest test this promise has faced is the attack
on 'The DAO' in 2016. The DAO was a large
investment fund encoded entirely as an Ethereum
contract, holding what was at the time a substantial
fraction of all circulating ether. An attacker
discovered that its withdrawal logic was vulnerable to
reentrancy and used the flaw to drain roughly a third
of the fund before the community could react. The
mechanism merits concrete statement, because it is
both subtle and, once seen, unmistakable. Reentrancy
arises when a contract makes an external call, handing
control to another contract, before it has updated its
own internal state to record the effect of that call.
The DAO sent ether to the caller before decrementing
the caller's recorded balance; the receiving
contract's code, invoked automatically on receipt,
simply called back into the withdrawal function again,
and because the balance had not yet been reduced, the
second call passed the same check and paid out a
second time, and a third, looping until the funds were
gone.

The response is what makes the episode the sharpest
test 'code is law' has faced. If code truly is law,
then the attacker merely executed the contract as
written and was entitled to keep the proceeds. The
Ethereum community, confronted with a loss of that
magnitude, chose otherwise, and enacted a hard fork
that rewrote the chain's history to return the funds,
a decision so contentious that a portion of the
community rejected it and continued the original,
unforked chain as Ethereum Classic, precisely on the
principle that the code's outcome, however unwelcome,
ought to stand. That split is the standing reminder
that 'code is law' is an aspiration held by fallible
people, not a property the technology can guarantee on
its own.

Reentrancy is only one of a family of vulnerability
classes whose incidence is measurable rather than
anecdotal. Atzei et al. [@atzei2017survey] provide a
systematic survey of attacks on Ethereum contracts,
organising them into recurring categories, and Luu et
al. [@luu2016oyente], with the Oyente analysis tool,
showed that a large fraction of then-deployed
contracts were flagged for at least one such class by
automated symbolic analysis. The point for our
purposes is not the precise percentages, which have
shifted as tooling and language practice improved, but
that these are structured, recurring, and to a
considerable degree detectable defects rather than
one-off misfortunes. @tbl-vuln-classes collects four
of the most common classes with their mechanism and
the standard mitigation.

| Vulnerability class | Mechanism | Standard mitigation |
|---|---|---|
| **Reentrancy** | An external call hands control to an untrusted contract before state is updated, letting the callee re-enter and repeat the operation | Update state before the external call (checks-effects-interactions), or use a reentrancy guard |
| **Integer overflow/underflow** | Arithmetic wraps past the fixed integer width, so a balance or counter silently jumps to a spurious value | Checked arithmetic (default in modern compilers) or a safe-math library |
| **Access control** | A privileged function omits or mis-scopes its authorisation check, letting any caller invoke it | Explicit caller checks on every privileged entry point, and least-privilege role design |
| **Oracle manipulation** | Logic depends on an external price or data feed an attacker can distort, for example via a flash-loan-driven price move | Time-weighted or multi-source feeds, sanity bounds, and treating any single oracle as untrusted |

: Four common smart-contract vulnerability classes {#tbl-vuln-classes}

The bearing on the programmable-consent contract of
this chapter is uncomfortable and merits facing
squarely. That contract is exactly the kind of code
whose bugs are irreversible. A consent contract that
mishandles a revocation, that continues to adjudicate
'allow' after a participant has withdrawn permission
because of an off-by-one in its logic or a reentrancy
in its update path, has not committed a spreadsheet
typo that a later correction quietly overwrites. It
has recorded a permanent, on-chain error whose effects
may already have propagated to every party that
consulted it. This is a direct argument, from the
security properties rather than from the privacy
properties, for the same architecture the earlier
public-health discussion reaches on other grounds:
keep the on-chain logic minimal and auditable, and
keep the sensitive data and the consequential
decisions off-chain, where a mistake can be corrected
before it becomes irreversible. The smaller the
on-chain surface, the smaller the immutable attack
surface.

::: {.callout-important}
## The sceptic's reflex

'The contract has been audited, so it is safe' is
the reassurance to distrust most. An audit is a finite
inspection, by fallible reviewers, of a program whose
behaviour is in general undecidable; it can establish
the presence of the bugs it finds and never the
absence of the bugs it does not. The DAO's code had
been read by many capable people before it was
drained. Automated tools such as Oyente
[@luu2016oyente] raise the floor by flagging known
classes mechanically, and they merit running, but a
clean report is evidence of the absence of those
particular classes, not a proof of correctness. We
should treat 'audited' as a tier-two claim about
what a reviewer looked for and did not find, never as
a tier-one guarantee that the code is safe, and we
should scope the irreversible on-chain logic small
enough that an audit has a chance of being complete.
:::

## Public health connection: programmable consent, precisely scoped

We return now to the cohort study with which we
opened. The proposal there was that a smart contract
might encode a consent policy directly, so that the
rule 'this variable may be released to this
researcher if this condition holds' is enforced by
code rather than by a data-use agreement that
depends on institutional follow-through. But how much
of a consent arrangement can a contract safely be
made to carry? The computational content of the
proposal is genuine and useful. A contract can hold
a permission state (who is authorised, under what
condition, until when) and can enforce access rules
against *anything routed through it* with the same
determinism this chapter has emphasised throughout,
thereby removing a class of administrative failure in
which a revoked permission is not actually propagated
to every system that ought to honour it. This is the
self-executing rule the opening asked for, and to
that extent the intuitive appeal of the proposal is
warranted.

It has also, unusually for this literature, been
tried on human beings. A dynamic-consent ledger was
deployed across two centres in a sixty-subject
decentralised trial, with protocol amendments
deliberately introduced so that re-consent behaviour
could be measured rather than assumed
[@huh2022metory]. The consent mechanism itself
performed well: participants agreed to amendments in
some ninety-six per cent of instances, at a median
response time of a fifth of an hour, which is a
turnaround no paper-based re-consent process can
approach. What degraded was everything the ledger did
not control. Adherence to the *schedule* of
measurements ran at sixty-nine per cent against
ninety-one per cent adherence to the intervention
itself, and it fell further after each major
amendment. The study is worth dwelling on precisely
because its authors report the disappointing number
alongside the flattering one, and because the pattern
it reveals is the pattern of the whole field: the
cryptographic mechanism did what it promised, and the
human process around it remained the binding
constraint. A biobank consent portal in Malta
[@mamo2020dwarna] reaches the same accommodation from
the engineering side, and it is the accommodation
every workable design in this literature converges on:
the personal data live off the chain, the consent
*events* live on it, and erasure is effected by
destroying keys. That is a sound architecture. It is
also a quiet admission that the chain is not doing
the interesting work.

The proposal breaks, however, the moment it is
extended to storing the protected health information
itself on chain. Every full node holds a complete copy of
chain state, in perpetuity, and no consensus
mechanism studied in this book (nor any studied
later) makes that state confidential; encryption of
a field does not fix this, because the ciphertext
persists forever and any weakness discovered in the
encryption scheme, or any future key compromise,
retroactively de-anonymises every record ever posted.
A blockchain is therefore a defensible place to store
a **commitment** to a permission (a hash of a consent
document, an access-control state machine, a
revocation flag) and an indefensible place to store
the health data that permission governs. The correct
architecture keeps PHI off-chain, in a system subject
to ordinary access controls and regulatory
obligations (HIPAA-equivalent controls, in a US
setting), and puts on-chain only the tamper-evident
metadata: who is permitted, since when, subject to
what condition, and a hash committing to the
off-chain record so that any party can verify a copy
has not been altered without ever learning its
content. @fig-consent-flow draws the arrangement:
the contract holds only a commitment and the access
rules, adjudicates each request to allow or deny, and
never touches the protected data, which stays in the
off-chain store.

```{r ch06-consent-flow}
#| label: fig-consent-flow
#| fig-cap: >
#|   Programmable consent, scoped correctly. The
#|   participant grants a permission whose commitment
#|   and rules live in an on-chain contract; a data
#|   user request is adjudicated allow or deny by that
#|   contract; and the protected health information
#|   itself stays in an off-chain store, with only a
#|   hash committed on chain.
library(ggplot2)

structure_col <- '#33356b'
PAL <- c('#2a78d6', '#1baf7a', '#eda100', '#008300',
         '#4a3aa7', '#e34948', '#e87ba4', '#eb6834')

box <- function(xmin, xmax, ymin, ymax, fill, col) {
  annotate('rect', xmin = xmin, xmax = xmax, ymin = ymin,
    ymax = ymax, fill = fill, colour = col, linewidth = 0.4)
}

ggplot() +
  box(0.2, 2.4, 6.2, 7.6, '#eef4fc', PAL[1]) +
  annotate('text', x = 1.3, y = 6.9, label = 'participant',
    size = 3.1, colour = structure_col) +
  box(3.9, 7.1, 5.4, 8.4, '#f2f0fb', PAL[5]) +
  annotate('text', x = 5.5, y = 7.9, label = 'consent contract',
    fontface = 'bold', size = 3.1, colour = structure_col) +
  annotate('text', x = 5.5, y = 7.2, label = '(on-chain)',
    size = 2.6, colour = 'grey45') +
  annotate('text', x = 5.5, y = 6.5, label = 'commitment (hash)',
    size = 2.7, colour = structure_col) +
  annotate('text', x = 5.5, y = 5.9, label = '+ access rules',
    size = 2.7, colour = structure_col) +
  box(8.6, 10.7, 6.2, 7.6, '#eef4fc', PAL[1]) +
  annotate('text', x = 9.65, y = 6.9, label = 'data user',
    size = 3.1, colour = structure_col) +
  annotate('segment', x = 2.4, xend = 3.85, y = 6.9, yend = 6.9,
    arrow = arrow(length = unit(0.14, 'cm')), colour = 'grey50') +
  annotate('text', x = 3.1, y = 7.3, label = 'grants',
    size = 2.5, colour = 'grey45') +
  annotate('segment', x = 8.55, xend = 7.15, y = 6.9, yend = 6.9,
    arrow = arrow(length = unit(0.14, 'cm')), colour = 'grey50') +
  annotate('text', x = 7.85, y = 7.3, label = 'request',
    size = 2.5, colour = 'grey45') +
  box(3.9, 5.4, 3.0, 4.2, '#eafaf3', PAL[2]) +
  annotate('text', x = 4.65, y = 3.6, label = 'allow',
    size = 3.0, colour = '#0f7a52') +
  box(5.7, 7.1, 3.0, 4.2, '#f1f1ef', '#9a9a92') +
  annotate('text', x = 6.4, y = 3.6, label = 'deny',
    size = 3.0, colour = 'grey40') +
  annotate('segment', x = 5.0, xend = 4.65, y = 5.35, yend = 4.25,
    arrow = arrow(length = unit(0.12, 'cm')), colour = PAL[2]) +
  annotate('segment', x = 6.0, xend = 6.4, y = 5.35, yend = 4.25,
    arrow = arrow(length = unit(0.12, 'cm')), colour = '#9a9a92') +
  annotate('rect', xmin = 3.9, xmax = 7.1, ymin = 0.5, ymax = 1.8,
    fill = 'white', colour = structure_col, linewidth = 0.4,
    linetype = 'dashed') +
  annotate('text', x = 5.5, y = 1.15,
    label = 'PHI in off-chain store',
    size = 2.9, colour = structure_col) +
  annotate('segment', x = 5.5, xend = 5.5, y = 5.35, yend = 1.85,
    arrow = arrow(length = unit(0.12, 'cm'), ends = 'both'),
    colour = structure_col, linetype = 'dashed',
    linewidth = 0.35) +
  annotate('text', x = 7.35, y = 3.3, hjust = 0, label = 'hash only',
    size = 2.5, colour = 'grey45') +
  annotate('text', x = 5.5, y = 0.05, label = 'PHI stays off-chain',
    fontface = 'bold', size = 2.9, colour = structure_col) +
  xlim(0, 10.9) + ylim(-0.3, 8.6) +
  theme_void()
```

To illustrate the commitment idea with something
concrete, consider a single SYNTHETIC consent
record, held in an off-chain store, of which we place
on chain nothing but a cryptographic hash. The hash
serves two purposes at once. It reveals nothing about
the participant, since a good hash is not invertible,
and yet it detects any later alteration of the
off-chain record, because changing even one field
changes the hash and the change no longer matches the
value already committed. The following computes the
commitment for an original record and for a silently
widened one, in which a permission that named a
single laboratory has been extended to a second.

```{r ch06-consent-commitment}
library(digest)
library(tibble)
library(dplyr)

# SYNTHETIC consent record. In a real deployment
# this off-chain record lives in a HIPAA-governed
# store; only the hash commitment goes on chain.
consent <- tibble(
  participant_id = 'SYNTHETIC-0007',
  permitted_use = 'genomic-analysis',
  shared_with = 'lab-B',
  expires = '2030-01-01',
  revoked = FALSE
)

commit <- function(record) {
  digest(record, algo = 'sha256')
}

original_hash <- commit(consent)

# a single silent edit: the permission is widened
# from one laboratory to two.
tampered <- consent |>
  mutate(shared_with = 'lab-B;lab-C')

tampered_hash <- commit(tampered)

tibble(
  version = c('original', 'tampered'),
  sha256_prefix = c(
    substr(original_hash, 1, 16),
    substr(tampered_hash, 1, 16)
  ),
  matches_commitment = c(
    original_hash == original_hash,
    tampered_hash == original_hash
  )
)
```

The tampered record fails to match the on-chain
commitment, so the alteration is detectable by any
party holding the committed hash, and yet that hash,
placed on a public and permanent ledger, discloses
none of the participant's information. This is the
whole of what a blockchain contributes to the
arrangement: tamper-evidence for a record whose
contents remain elsewhere.

The secondary example the reader should keep in mind
is vaccine cold-chain provenance. A distribution
network would like an auditable guarantee that each
vial remained within its permitted temperature range
from manufacture to administration, across carriers
and jurisdictions that do not fully trust one
another. The structure of a defensible design is
identical to the consent case: the sensor readings
and custody transfers live off-chain, and the ledger
holds only periodic hash commitments to them, so that
a regulator can later verify that a presented
temperature log has not been retrofitted to conceal
an excursion, without the ledger itself becoming a
public repository of the underlying operational data.

::: {.callout-warning}
## Most 'health records on Ethereum' proposals fail this test

The reader should examine any concrete proposal for
its precise placement of the boundary between what
is committed on-chain and what is stored off-chain.
A proposal that puts
diagnosis codes, lab values, or free-text clinical
notes on a public chain, encrypted or not, has
misunderstood the guarantee a blockchain provides.
The tier-one protocol fact is that on-chain state is
public and permanent; a proposal that treats
encryption as a substitute for that fact is making a
tier-four narrative claim about future
cryptographic durability, not a tier-one one about
the system as built.
:::

## Worked example: base fee response to a rollup-driven demand shock

To illustrate, suppose gas usage on the base chain
follows the SYNTHETIC pattern already simulated, and
suppose we wish to quantify two things a network
operator or an application developer genuinely cares
about: how many blocks it takes the base fee to
double from its pre-shock level, and how far it
overshoots below the pre-shock level during the
recovery. Both are directly readable from the
simulated series already computed above.

```{r ch06-worked-example}
pre_shock_fee <- sim |>
  filter(block == shock_start - 1) |>
  pull(base_fee_gwei)

doubling_block <- sim |>
  filter(block >= shock_start, base_fee_gwei >= 2 * pre_shock_fee) |>
  slice(1) |>
  pull(block)

trough_after_shock <- sim |>
  filter(block > shock_end) |>
  slice_min(base_fee_gwei, n = 1) |>
  select(block, base_fee_gwei)

tibble(
  pre_shock_fee_gwei = pre_shock_fee,
  doubles_by_block = doubling_block,
  blocks_to_double = doubling_block - shock_start
) |>
  bind_cols(trough_after_shock |> rename(
    trough_block = block, trough_fee_gwei = base_fee_gwei
  ))
```

Two features generalise beyond this particular
synthetic run. First, doubling time is itself
computable directly from the gain: a base fee that
compounds by a constant fraction $\epsilon$ per
block for $k$ blocks satisfies
$b_{n+k} = b_n(1+\epsilon)^k$, so if the shock holds
usage at a constant fraction above target the
doubling time is $\log 2 / \log(1+\epsilon)$ blocks,
which is a direct application of the compound-growth
reasoning the reader already owns from continuous-
time interest and population models, just discretised
to one block per period. Second, the post-shock
undershoot exists because the controller has no
memory of the pre-shock level, only of the current
error; a controller with an integral term would
return exactly to a target level, and the base fee
rule deliberately has none, trading long-run level
stability for a simple, publicly computable per-block
rule.

## Collaborating with an LLM

**Prompt:** 'Explain why Ethereum's account model
allows re-entrancy attacks and Bitcoin's UTXO model
does not.'
**Watch for:** a plausible-sounding answer that
conflates re-entrancy (a contract-execution ordering
bug, specific to stateful contract calls) with the
account-versus-UTXO distinction itself; an LLM may
imply the account model *causes* re-entrancy when in
fact re-entrancy requires a contract that makes an
external call before updating its own state, a coding
pattern, not an unavoidable feature of the account
model.
**Verification:** ask for the specific EVM opcode
sequence (a `CALL` before a storage write) that
creates the vulnerability, and confirm against the
Yellow Paper's description of message calls
[@wood2014yellow] that re-entrancy is a consequence
of contract code, not of the account-balance state
representation per se.

**Prompt:** 'Derive the maximum possible base fee
after $k$ consecutive full blocks under EIP-1559.'
**Watch for:** an LLM silently assuming the gain
$1/8$ acts additively rather than
multiplicatively-compounding, giving a linear rather
than a geometric bound.
**Verification:** derive it yourself as
$b_0 (1 + 1/8)^k$ and check the LLM's closed form
reduces to the same expression; then check it
numerically against the simulation code in this
chapter by forcing `gas_used_frac` to 1 for $k$
consecutive blocks.

**Prompt:** 'Is it safe to store patient consent
records on Ethereum?'
**Watch for:** an answer that treats 'consent
records' as a single object and glosses over which
fields are commitments versus which are the
underlying protected data; this is exactly the
tier-slippage the public-health section above warns
against.
**Verification:** insist the answer separately name,
for each field of the proposed record, whether it is
stored on-chain in plaintext, on-chain as a
commitment (hash), or off-chain entirely, and reject
any answer that does not make this partition
explicit.

## In conclusion

In conclusion, three points are to be emphasised.
First, the account model is a current-state
representation of a ledger, and the account nonce is
the price of that choice: replay protection and a
total order per sender, which the UTXO event log
obtains for nothing, must in the account model be
maintained explicitly, and the reader who keeps the
event-log versus current-state distinction in view
will not be surprised by where each cost falls.
Second, gas and the EIP-1559 base fee are,
respectively, a price on computation and a feedback
controller on block utilisation; the base fee shares
its structural family with the difficulty adjustment
of @sec-mining, and it appears to us that recognising
the controller pattern, a set point, an error signal,
and a bounded multiplicative correction, is a more
durable acquisition than any particular fee level.
Third, and most consequentially for our own field,
the defensible use of a blockchain in public health
is to hold a tamper-evident commitment to a
permission or a provenance record, and never the
protected data itself; a proposal that places
protected health information on chain, encrypted or
otherwise, has misread the guarantee the technology
provides, and the reader is now equipped to say
precisely why.

## Exercises

1. Derive the exact per-block bound on base fee
   change implied by the EIP-1559 update rule (that
   is, show the maximum possible ratio
   $b_{n+1}/b_n$ and the minimum possible ratio, and
   state under what gas-usage condition each is
   attained).
2. Modify `ch06-eip1559-sim` so that instead of a
   single demand shock, gas usage follows a
   sinusoidal daily cycle around the target (SYNTHETIC
   data, your choice of amplitude and period in
   blocks). Plot the resulting base fee path and
   comment on whether the controller tracks a
   periodic input with a phase lag, and explain why
   a purely proportional controller cannot track a
   periodic input without lag.
3. A transaction's total gas limit is 100{,}000 and
   the network base fee is 30 gwei with a 2 gwei
   priority fee. Compute the maximum ether cost of
   the transaction, and explain under what condition
   the sender pays strictly less than this maximum.
4. Explain, in terms of the account nonce, why a
   transaction with nonce 7 submitted before a
   pending transaction with nonce 6 from the same
   account cannot be included first, and contrast this
   with how Bitcoin would handle two transactions
   from the same address with no ordering dependency
   between them.
5. Using the falsifiability tiers of @sec-intro,
   classify each of the following claims and justify
   the tier: (a) 'the EVM executes deterministically
   given identical inputs'; (b) 'the average base fee
   fell after EIP-4844 shipped'; (c) 'rollups will
   make on-chain health data storage economically
   viable'.
6. A contract's admin key can pause the contract or
   redirect funds. Write two or three sentences
   evaluating the claim 'this smart contract is
   trustless' against that fact, using the sceptic's
   reflex framework.

## Further reading

@wood2014yellow is the Ethereum Yellow Paper, the
formal specification of the EVM and the account-state
transition function; read it as the ground truth for
every claim this chapter makes about determinism and
gas accounting. @buterin2014ethereum is the original
Ethereum whitepaper and is the right source for the
'world computer' framing and the motivating
comparison with Bitcoin's more limited scripting
language. @eip1559 is the EIP specifying the base-fee
mechanism in full, including edge cases (block-limit
elasticity, the initial base fee at the London
upgrade) this chapter's simulation abstracts away.
@eip4844 specifies blob transactions and the
independent blob-gas market; read it alongside the
main EIP-1559 specification to see how directly the
second market reuses the first's control law.
@antonopoulos2018ethereum is the standard practitioner
reference for account types, gas mechanics, and
contract deployment, useful for the operational detail
this chapter deliberately omits in favour of the
underlying control-theoretic and economic structure.
@narayanan2016bitcoin remains the standard comparison
point for the UTXO model this chapter contrasts
Ethereum against throughout.
