# Decentralised Finance as Mathematics {#sec-defi}

## Learning objectives

By the end of this chapter you should be able
to:

- Derive the constant-product invariant, show
  that the marginal price of a pool equals its
  reserve ratio, and explain slippage as a
  consequence of the curvature of the bonding
  curve.
- Prove that impermanent loss equals
  $2\sqrt{r}/(1+r) - 1$ in the price ratio $r$
  and recognise it as the arithmetic-geometric
  mean (Jensen) inequality made financial.
- Analyse an over-collateralised lending
  position, compute its liquidation price, and
  explain how forced selling cascades.
- Explain why oracles are the trust boundary of
  decentralised finance and how a flash loan
  weaponises a single-block spot price.
- Classify stablecoin designs and diagnose a
  deleveraging spiral of the kind that destroyed
  TerraUSD.
- Distinguish yield that is a real cash flow
  from yield that is inflationary token
  emission.
- Explain why a parametric health-financing
  instrument, of the kind the World Bank's 2017
  pandemic bonds attempted, inherits the oracle
  and depeg risks of the DeFi plumbing on which
  it would be built.

## Orientation

We begin not with a cryptocurrency but with a
problem in public-health finance. When an outbreak
is declared, money moves slowly. Donor pledges are
announced, then wend their way through
appropriations, disbursement conditions, and the
accounts of intermediary agencies, so that the cash
arrives weeks or months after the epidemic curve
has already turned. The epidemiology is unforgiving
on this point: a case count that doubles every few
days does not wait for a committee. We are led to
ask whether the financing might be made to move at
the speed of the disease.

But how might money be made to arrive on time? One
answer, much discussed since the West African Ebola
epidemic of 2014, is an instrument that pays out
automatically the moment a verifiable trigger
fires: reported cases crossing a threshold, a
declaration by the World Health Organization, a
mortality index exceeding a cutoff. The premium is
posted in advance, the trigger is written down, and
when the index crosses it the capital is released
with no committee in the loop. The intuitive appeal
is plain, since the delay is, at least in
principle, engineered out.

The honest cautionary case is the World Bank's
Pandemic Emergency Financing Facility, whose
catastrophe bonds were issued in 2017 with exactly
this structure. When the Ebola epidemic of 2018 and
2019 spread through the eastern Democratic Republic
of the Congo, the bonds' parametric triggers, which
required among other conditions a threshold of
confirmed deaths together with a minimum outbreak
growth rate sustained over a fixed window, had been
engineered so conservatively that payout was
delayed for months and, when it came, was partial
and late relative to the need. The lesson is
precise, and it is not that the plumbing failed.
The capital transfer, once triggered, worked as
written. What was problematic was the trigger
itself: the choice of index, the thresholds, the
data feed that measured them, and the lag those
choices imposed. We shall take this as the
organising moral of the chapter. The hard part of a
parametric instrument is the trigger design and the
oracle that reports the index, not the payment
mechanism.

That is the frame in which the machinery of this
chapter merits a public-health student's attention.
Decentralised finance (DeFi) is the collection of
financial primitives, exchange, lending,
derivatives, and stable value, rebuilt as smart
contracts running on a programmable chain such as
the Ethereum machine of @sec-ethereum. Digital cash
and cryptocurrency are its historical setting, and
we shall use their vocabulary throughout, but they
are not our reason for studying it. Our reason is
that these primitives are precisely the plumbing a
parametric health instrument would require: an
escrow to hold the premium capital, an oracle to
deliver the triggering index on-chain, a stable
unit of account so the payout is denominated in
something that does not itself lurch, and, should
anyone propose to invest the float in the interim,
an exchange and a lending market in which to do so.
The failure modes of that plumbing, oracle
manipulation, the depeg of a supposedly stable
token, and the reflexive spiral in which a defence
mechanism accelerates the very collapse it was
meant to arrest, are exactly the risks that a
parametric health instrument would inherit. To
audit the instrument, one must first be able to
audit the plumbing.

It is worth tabulating that plumbing before we
build any of it, so the correspondence between
each primitive and the risk it carries is in view
from the start (@tbl-defi-primitives).

| Primitive | Role in a parametric health instrument | Inherited risk |
|---|---|---|
| **Escrow / conditional payout** | Holds the premium capital and releases it when the trigger fires | Smart-contract bug; a trigger that fires late or not at all |
| **Oracle** | Delivers the triggering index on-chain | Manipulation, reporting lag, single-reporter trust |
| **Stablecoin** | Denominates the payout in a stable unit | Depeg; reserve or custodian failure |
| **Exchange (AMM)** | Converts or invests the idle float | Impermanent loss and slippage |
| **Lending market** | Earns interest on the float | Liquidation cascade; dependence on a price oracle |

: Each decentralised-finance primitive maps to a role a parametric health instrument would ask of it and to a distinct risk that role imports {#tbl-defi-primitives}

Stripped of the promotional language, almost all of
that plumbing is elementary applied mathematics: a
swap is a move along a level set of a two-variable
function, a loan is an inequality on a collateral
ratio, and a stablecoin is a control law with a
feedback loop that sometimes goes unstable. The
reader who has taken a course in convex analysis
and one in stochastic processes already owns the
tools to audit the entire field, and, we shall
argue, to audit a parametric instrument built on
top of it.

That is the stance of this chapter. We treat each
mechanism as an object with a closed form, derive
the closed form, and then ask the question a
statistician is trained to ask: under what
assumptions is the stated behaviour true, and how
would we measure a departure from it? The automated
market maker yields a clean invariant and an exact
expression for the trader's execution cost. The
liquidity provider's dreaded 'impermanent loss'
turns out to be nothing more exotic than Jensen's
inequality, computable in one line, and it is
exactly the convexity cost that a proposal to
'stake the fund's float to earn a yield for the
endowment' quietly conceals. Lending and
liquidation are threshold crossings of a ratio.
Each result is small; the value is in seeing that
it is small, because the surrounding discourse
insists that it is deep.

Two ideas recur and deserve flagging at the outset.
The first is convexity: the market maker holds a
portfolio whose value is a concave function of
price, and concavity relative to a linear
buy-and-hold benchmark is exactly the source of
impermanent loss. The second is the oracle: every
contract that references an off-chain quantity, a
price, a case count, a cold-chain temperature, has
a single point at which an external number enters
the system, and that point is where the trust the
chain claims to remove quietly re-enters. Most DeFi
exploits are, at root, attacks on the oracle, and
the pandemic-bond episode is, at root, a failure of
the same boundary: the index and its measurement,
not the transfer. A quantitative reader who keeps
these two ideas in view will not be surprised by
much of what follows.

Two applications thread the whole book and we shall
return to them here. The first, which we shall call
the trial ledger, is a multi-institution
clinical-trial data-integrity ledger: several
hospitals maintaining an append-only shared record
of enrolment, randomisation, and outcome events,
with no single trusted custodian, which a regulator
audits after the fact. The second, our secondary
example, is a vaccine cold-chain provenance system,
in which each custody handoff and temperature
reading of a shipment is recorded immutably. Both
matter to the present chapter for one reason: each
is a candidate source of the verifiable trigger a
parametric instrument consults. An outcome-event
feed drawn from the trial ledger, or a breach count
drawn from the cold-chain record, is an oracle in
exactly the sense we shall shortly make precise, and
it inherits exactly the trust and manipulation
questions we shall raise about price oracles.

## The statistician's contribution

Three judgements recur in what follows, and each is
one that no contract and no dashboard will make on
our behalf.

1. **Impermanent loss is Jensen's inequality,
   not a fee or a bug.** The liquidity provider
   holds a portfolio that is rebalanced, by
   construction, to sell whatever rose and buy
   whatever fell. Against a static hold that is a
   guaranteed drag whenever prices move, and its
   magnitude is the gap between an arithmetic and
   a geometric mean. Recognising the convexity
   tells us that no amount of 'the loss is
   impermanent' reassurance alters the
   expectation: the provider is short volatility,
   and must be paid for it in fees, or the
   position loses. This is the reckoning any
   proposal to invest a fund's float in a
   liquidity pool must confront honestly.

2. **The oracle is the only assumption that
   matters.** A lending market, a derivative, and
   an algorithmic stablecoin are each only as
   sound as the price feed they consult, just as a
   parametric health bond is only as sound as its
   case-count feed. Auditing a DeFi protocol is,
   to first order, auditing its oracle: is the
   price a manipulable single-block spot, a
   time-weighted average, or a signed feed from a
   quorum of reporters, and what does an attacker
   with a flash loan pay to move it? We read this
   as a question about the variance and the
   adversarial robustness of an estimator, which
   is familiar ground.

3. **Separate cash-flow yield from emission.**
   An advertised annual percentage yield is a sum
   of two very different quantities: fees or
   interest paid by real counterparties, which is
   income, and freshly minted governance tokens,
   which is dilution wearing the costume of
   income. The first is sustainable; the second is
   a wealth transfer from later entrants to
   earlier ones and decays as emission schedules
   taper. Decomposing a headline number into these
   two parts is elementary and, unfortunately,
   almost never done in the marketing.

## Constant-function market makers

A constant-function market maker (CFMM) replaces
the order book of a traditional exchange with a
pool of two assets and a rule: any trade is
permitted so long as it leaves a fixed function
of the reserves unchanged. Let the pool hold
reserves $x$ of asset $X$ and $y$ of asset $Y$.
The dominant design, which Adams and colleagues
introduced at scale in the Uniswap protocol
[@adams2020uniswapv2], is the constant-product
market maker (CPMM), whose invariant is

$$
x \, y = k .
$$

A trader who deposits $\mathrm{d}x$ of $X$
withdraws the amount $\mathrm{d}y$ of $Y$ that
keeps the product at $k$:

$$
(x + \mathrm{d}x)(y - \mathrm{d}y) = k
\quad\Longrightarrow\quad
\mathrm{d}y = \frac{y \, \mathrm{d}x}{x + \mathrm{d}x}.
$$

The marginal (spot) price of $Y$ in units of $X$
is the limiting exchange rate for an infinitesimal
trade,

$$
p \;=\; \lim_{\mathrm{d}x \to 0}
\frac{\mathrm{d}x}{\mathrm{d}y}
\;=\; \frac{x}{y},
$$

the reserve ratio. This is the whole pricing
mechanism: there is no order book and no
counterparty quote, only the ratio of what the
pool currently holds. The question of when such a
function admits sensible prices and bounded losses
is taken up rigorously by Angeris and colleagues
[@angeris2020improved; @angeris2021analysis], and
the constant product is the canonical case they
treat.

Execution price degrades with trade size. For a
finite purchase the trader pays, on average,
$\mathrm{d}x / \mathrm{d}y = (x + \mathrm{d}x)/y$
per unit of $Y$, which exceeds the spot price $p$
by $\mathrm{d}x / y$. The excess is slippage. It
arises because the marginal price along the
invariant, $m(x) = x/y = x^2/k$, is convex in the
reserve, so that the average price paid while
traversing a finite interval sits strictly above
the price at the starting point. That is Jensen's
inequality appearing for the first time in this
chapter, and it returns below in a more
consequential role.

We can pause on a single picture that makes the
whole mechanism legible. The invariant is a
hyperbola, a trade is a step along it, and the
chord and the tangent at the starting point are
the two prices we have just distinguished
(@fig-cpmm-hyperbola).

```{r}
#| label: fig-cpmm-hyperbola
#| fig-cap: 'The constant product x times y equals k
#|   is a hyperbola, and a finite trade steps the pool
#|   from the before point to the after point along
#|   it; the chord joining the two points is flatter
#|   than the tangent at the start, so a finite order
#|   receives less of asset Y per unit of X than the
#|   marginal rate, which is the slippage.'
library(ggplot2)
library(tibble)
PAL <- c('#2a78d6', '#1baf7a', '#eda100', '#008300',
         '#4a3aa7', '#e34948', '#e87ba4', '#eb6834')
book_theme <- ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(
    panel.grid.minor = ggplot2::element_blank(),
    panel.grid.major = ggplot2::element_line(
      linewidth = 0.25, colour = 'grey88'),
    axis.title = ggplot2::element_text(size = 10),
    legend.position = 'bottom'
  )

k <- 10000
xr <- seq(55, 190, length.out = 200)
curve_df <- tibble(x = xr, y = k / xr)
x0 <- 100; y0 <- k / x0            # before the trade
x1 <- 160; y1 <- k / x1            # after depositing X
slope_tan <- -k / x0^2             # tangent slope at start
tan_df <- tibble(x = c(72, 138),
                 y = y0 + slope_tan * (c(72, 138) - x0))
pts <- tibble(x = c(x0, x1), y = c(y0, y1))

ggplot(curve_df, aes(x, y)) +
  geom_line(colour = PAL[1], linewidth = 0.7) +
  geom_line(data = tan_df, aes(x, y),
            colour = '#9a9a92', linewidth = 0.5, linetype = 2) +
  annotate('segment', x = x0, y = y0, xend = x1, yend = y1,
           colour = PAL[3], linewidth = 0.7) +
  geom_point(data = pts, aes(x, y),
             colour = '#33356b', size = 2.6) +
  annotate('text', x = 104, y = 113, label = 'before',
           hjust = 0, size = 3.1, colour = '#33356b') +
  annotate('text', x = 166, y = 66, label = 'after',
           hjust = 0, size = 3.1, colour = '#33356b') +
  annotate('text', x = 60, y = 150, label = 'tangent: marginal price',
           hjust = 0, size = 3, colour = '#33356b') +
  annotate('text', x = 148, y = 96, label = 'chord: execution price',
           hjust = 0, size = 3, colour = '#33356b') +
  labs(x = 'reserve x of asset X', y = 'reserve y of asset Y') +
  book_theme
```

To make the degradation concrete, the verified
companion code computes the execution price for a
range of trade sizes against a pool that also
charges the standard thirty-basis-point fee, and
plots the result.

```{r ch07-setup}
#| include: false
library(ggplot2)
library(dplyr)
library(tibble)
library(tidyr)
```

```{r ch07-amm}
#| label: fig-cpmm-slippage
#| fig-cap: "Constant-product execution price as a
#|   function of trade size. The dashed line is the
#|   marginal (spot) price at zero size; a larger
#|   order walks further along the hyperbola and
#|   pays strictly more."
cpmm_out <- function(dx, Rx, Ry, fee = 0.003) {
  dx_eff <- dx * (1 - fee)
  Ry * dx_eff / (Rx + dx_eff)
}

Rx <- 1e6; Ry <- 1e6
sizes <- seq(1e3, 5e5, length.out = 100)
exec_price <- sapply(sizes, function(dx) dx / cpmm_out(dx, Rx, Ry))

ggplot(tibble(size = sizes, price = exec_price), aes(size, price)) +
  geom_line() +
  geom_hline(yintercept = Rx / Ry, linetype = 2) +
  labs(x = 'trade size dx (units of X)',
       y = 'execution price (X paid per Y received)',
       title = 'Constant-product slippage: price worsens with size')
```

A trade that consumes a large fraction of the
reserves moves the price far, so large orders
'move the market' mechanically, with no
counterparty required to disagree with them. This
is a protocol fact in the sense of @sec-intro: it
follows from the invariant by algebra and can be
checked by anyone reading the contract.

::: {.callout-note icon=false collapse=true}
## Check your understanding: slippage of a large order

A CPMM pool holds $x = 1{,}000{,}000$ units of a
stablecoin and $y = 1{,}000{,}000$ units of a
token, so the spot price is one stablecoin per
token. Ignoring fees, what average price does a
trader pay for a purchase that spends
$\mathrm{d}x = 250{,}000$ stablecoins, and by what
fraction does that exceed spot?

The trader receives
$\mathrm{d}y = y\,\mathrm{d}x/(x+\mathrm{d}x)
= 10^6 \cdot 2.5\times10^5 / 1.25\times10^6
= 200{,}000$ tokens. The average execution price
is $250{,}000 / 200{,}000 = 1.25$ stablecoins per
token, twenty-five percent above the spot price of
one. The slippage $\mathrm{d}x/y = 0.25$ matches
the algebra exactly. A quarter of the reserve is a
very large order; the point is that the cost is a
deterministic function of depth, not a matter of
who is on the other side.
:::

## Impermanent loss as a convexity cost

A liquidity provider (LP) deposits both assets
into the pool and earns a share of trading fees.
The cost of doing so, relative to simply holding
the two assets, is impermanent loss (IL). It is
the most misunderstood quantity in DeFi, and it is
one line of algebra. We dwell on it because it is
the hidden term in a pitch a public-health reader
will one day be shown: the suggestion that a
preparedness fund or an endowment 'provide
liquidity to earn a yield' on its idle capital. The
advertised yield is visible; the convexity cost
derived below is not, and a fund treasurer who
cannot price it is agreeing to be short volatility
without knowing the premium demanded in return.

Suppose the LP deposits when the price of $Y$ (in
units of $X$) is $p_0$, and later the external
price has moved to $p_1$, a gross ratio
$r = p_1 / p_0$. Arbitrageurs keep the pool's spot
price equal to the external price, so at any price
$p$ the reserves satisfy $x/y = p$ together with
$xy = k$, giving $x = \sqrt{kp}$ and
$y = \sqrt{k/p}$. The value of the pool position,
measured in units of $X$, is

$$
V_{\text{pool}}(p) \;=\; x + p\,y
\;=\; 2\sqrt{k p}.
$$

Had the LP instead held the initial reserves
$x_0 = \sqrt{k p_0}$ and $y_0 = \sqrt{k/p_0}$, the
value would be
$V_{\text{hold}}(p_1) = x_0 + p_1 y_0
= \sqrt{k p_0}\,(1 + r)$. Taking the ratio,

$$
\frac{V_{\text{pool}}}{V_{\text{hold}}}
\;=\;
\frac{2\sqrt{k p_1}}{\sqrt{k p_0}\,(1+r)}
\;=\;
\frac{2\sqrt{r}}{1 + r},
$$

and the impermanent loss is the shortfall

$$
\mathrm{IL}(r) \;=\;
\frac{2\sqrt{r}}{1 + r} - 1 .
$$

The interpretation is immediate once we notice
what the numerator and denominator are:
$\sqrt{r} = \sqrt{1 \cdot r}$ is the geometric mean
of $\{1, r\}$ and $(1+r)/2$ is their arithmetic
mean. Hence

$$
\frac{V_{\text{pool}}}{V_{\text{hold}}}
\;=\;
\frac{\operatorname{GM}(1, r)}{\operatorname{AM}(1, r)}
\;\le\; 1,
$$

with equality if and only if $r = 1$, by the
arithmetic-geometric mean inequality, itself a
special case of Jensen's inequality applied to the
concave function $\sqrt{\cdot}$. The pool value is
a concave (square-root) function of price while the
hold benchmark is linear, and a concave payoff
underperforms its own tangent whenever the argument
departs from the point of tangency. That is the
entire content of impermanent loss.

The mechanism, stated without calculus: the
constant-product rule forces the pool to sell the
asset that appreciates and buy the asset that
depreciates, continuously and in exactly the wrong
direction for a trend follower. The LP is
structurally short volatility. The companion code
gives the closed form and its symmetry.

```{r ch07-il}
#| label: fig-il
#| fig-cap: "Impermanent loss against the price
#|   ratio r on a log axis. The curve is zero at
#|   r = 1, negative on both sides, and symmetric
#|   in log r: a doubling and a halving cost the
#|   provider the same."
il <- function(r) 2 * sqrt(r) / (1 + r) - 1
r <- 10^seq(-1, 1, length.out = 200)

ggplot(tibble(r = r, il = il(r)), aes(r, il)) +
  geom_line() +
  scale_x_log10() +
  labs(x = 'price ratio r (withdraw / deposit), log scale',
       y = 'impermanent loss (fraction vs holding)',
       title = 'Impermanent loss: 2 sqrt(r)/(1+r) - 1')
c(il_2x = il(2), il_4x = il(4), il_half = il(0.5))
```

The loss is zero at $r = 1$, negative on either
side, and symmetric in $\log r$: doubling
($r = 2$) and halving ($r = 1/2$) cost the same,
about two percent, because $\mathrm{IL}(r) =
\mathrm{IL}(1/r)$, a fact the reader can verify
directly from the formula. A four-fold move costs
about six percent. These are not catastrophic
numbers, which is precisely why the loss is
survivable when fee income is adequate and ruinous
when it is not.

::: {.callout-important}
## The sceptic's reflex: 'impermanent loss is impermanent'

The reassurance that the loss is 'impermanent',
recovered if the price returns to where it started,
is a tier-four narrative dressed as a tier-one
fact (@sec-intro). The recovery is conditional on
an event, mean reversion, that the provider does
not control and cannot assume. What is guaranteed,
unconditionally, is the expectation: for any
non-degenerate price process the LP realises a
negative $\mathbb{E}[\mathrm{IL}]$ relative to
holding, because $\mathrm{IL}(r) \le 0$ pointwise
and equals zero only on a measure-zero set. A more
honest name, now common in the literature, is
'loss versus rebalancing', or simply the cost of
being short gamma. Rename it and the fog lifts: the
provider is selling volatility and must be paid the
premium in fees, or the position is losing.
:::

::: {.callout-note icon=false collapse=true}
## Check your understanding: why the AM-GM form matters

Someone claims that impermanent loss can be
eliminated by choosing a pool whose two assets are
'highly correlated', for example two stablecoins.
Using the AM-GM characterisation, explain precisely
what correlation has to do with it and what it does
not change.

Impermanent loss depends only on the realised price
ratio $r = p_1/p_0$ through $2\sqrt{r}/(1+r) - 1$;
it makes no reference to correlation, volatility,
or any distributional assumption. What a
tightly-correlated pair buys the provider is a
distribution of $r$ concentrated near one, where
the curve is
flat (its derivative vanishes at $r = 1$ and the
loss is second-order, $\mathrm{IL} \approx
-\tfrac{1}{8}(r-1)^2$ for small moves). So the
expected loss is small because $r$ rarely strays
from one, not because the mechanism is different.
Correlation shapes the input distribution; the
AM-GM inequality still governs the output for every
realised value. This is why stablecoin pools are
viable LP positions and volatile pairs often are
not.
:::

## Lending and liquidation

The second pillar of DeFi is over-collateralised
lending, exemplified by Aave and Compound. Because
borrowers are pseudonymous and cannot be pursued
for a deficiency, a loan must be secured by
collateral worth more than the debt at all times.
A position is described by its collateral value
$C$, its debt $D$, and a protocol parameter, the
liquidation threshold $\ell \in (0, 1)$. The
position is healthy while

$$
\text{health factor} \;=\;
\frac{\ell \, C}{D} \;\ge\; 1,
$$

and becomes liquidatable the instant the health
factor drops below one, at which point liquidators
may repay part of the debt in exchange for the
borrower's collateral at a discount. Gudgeon and
colleagues analyse the economics of these systems,
and the ways in which they fail under stress
[@gudgeon2020loanable; @gudgeon2020crisis].

Because collateral value $C = q \cdot P$ is the
quantity of collateral $q$ times a price $P$ that
the protocol reads from an oracle, the liquidation
condition is a threshold on price. Solving
$\ell q P = D$ gives the liquidation price

$$
P^\star \;=\; \frac{D}{\ell \, q}.
$$

Below $P^\star$ the position is under water. The
following self-contained illustration computes the
health factor and liquidation price for a concrete
position.

```{r ch07-liq}
health_factor <- function(units, price, debt, liq_threshold) {
  liq_threshold * units * price / debt
}

# SYNTHETIC position: 10 units of collateral at a
# current price of 100, a 500-unit stablecoin debt,
# and an Aave-style liquidation threshold of 0.8.
units <- 10; debt <- 500; ell <- 0.8
liq_price <- debt / (ell * units)

c(current_price = 100,
  current_hf = health_factor(units, 100, debt, ell),
  liquidation_price = liq_price)
```

The position is healthy at a health factor of 1.6
and is liquidated once the collateral price falls
to 62.5, a thirty-seven percent drawdown. The
danger is that liquidation is itself a sale: the
liquidator sells the seized collateral into the
market, which pushes the price down, which drops
the next-nearest position below its own threshold.
Forced selling begets forced selling. The following
compact deterministic model makes the cascade
visible: a synthetic population of leveraged
positions, an initial price shock, and a fixed
price impact per unit liquidated.

```{r ch07-cascade}
simulate_liq_cascade <- function(price0 = 100, shock = -8,
                                 n = 400, leverage = 10,
                                 impact = 0.15) {
  set.seed(7)
  # SYNTHETIC entry prices; longs liquidate at
  # entry * (1 - 1/leverage).
  entry <- rnorm(n, price0, 5)
  liq_price <- entry * (1 - 1 / leverage)
  price <- price0 + shock
  liquidated <- 0L
  repeat {
    newly <- sum(liq_price >= price) - liquidated
    if (newly <= 0) break
    liquidated <- liquidated + newly
    price <- price - impact * newly / n * price0
  }
  list(final_price = price,
       fraction_liquidated = liquidated / n)
}

cascade <- simulate_liq_cascade()
c(final_price = round(cascade$final_price, 2),
  fraction_liquidated = cascade$fraction_liquidated)
```

An eight percent shock is amplified by the
feedback loop into a much larger price decline and
nearly the entire population of leveraged positions
liquidated. The
same avalanche structure recurs across this book:
a small exogenous disturbance amplified by a
structural feedback, as in the mining fork rate of
@sec-mining and the restaking tail of @sec-pos.
The statistician's contribution is to notice that
the cascade's severity is not random but a
computable function of the leverage distribution,
the threshold, and the depth of the market into
which collateral is sold.

## Oracles and flash loans

Every mechanism above consults a price. On a chain
that price must come from somewhere, and the
somewhere is an oracle: a contract or a signed feed
that reports an off-chain quantity to on-chain
code. The oracle is the trust boundary of DeFi.
Everything upstream of it can be trust-minimised by
the consensus machinery of the preceding chapters;
the oracle is where an external number, and with it
an external assumption, re-enters. A protocol that
prices collateral off the current spot price of an
on-chain pool has, in effect, delegated its
solvency to whoever can move that pool.

Flash loans make that delegation dangerous. A flash
loan is an uncollateralised loan of arbitrary size
that must be borrowed and repaid within a single
transaction; atomicity guarantees that if the
repayment fails the entire transaction reverts, so
the lender bears no credit risk. This is a genuine
and elegant primitive with legitimate uses
(arbitrage, collateral swaps, self-liquidation).
It is also the ideal instrument for oracle
manipulation, because it hands an attacker
unlimited capital for the length of one block with
no requirement to already possess it. The
canonical attack: borrow a large sum, use it to
distort a spot-price oracle, transact against a
second protocol that trusts that oracle, and unwind,
all atomically.

The following self-contained numeric example makes
the extraction concrete. A lending market prices a
collateral token off the spot price of a single
CPMM pool. An attacker flash-borrows stablecoins,
buys the collateral token to pump its reported
price, and shows how much the naive valuation
overstates the true value.

```{r ch07-oracle}
swap_x_for_y <- function(dx, Rx, Ry) {
  dy <- Ry * dx / (Rx + dx)
  list(dy = dy, Rx = Rx + dx, Ry = Ry - dy)
}

# SYNTHETIC pool used naively as a price oracle:
# 1e6 stablecoin (X) and 1e6 collateral token (Y),
# so the honest spot price of the token is 1.
Rx0 <- 1e6; Ry0 <- 1e6
spot0 <- Rx0 / Ry0

loan <- 5e5                       # flash-borrowed stablecoins
after <- swap_x_for_y(loan, Rx0, Ry0)
spot1 <- after$Rx / after$Ry      # manipulated spot price

bought <- after$dy                # collateral acquired
ltv <- 0.75
naive_value <- bought * spot1
true_value <- bought * spot0
borrowable <- naive_value * ltv

c(spot_before = spot0, spot_after = spot1,
  collateral_bought = round(bought),
  naive_valuation = round(naive_value),
  borrowable_at_75pct = round(borrowable),
  true_value = round(true_value),
  bad_debt = round(borrowable - true_value))
```

A single half-million swap more than doubles the
reported price. A lending market that valued the
acquired collateral at the manipulated spot and
lent seventy-five percent against it would advance
far more than the collateral is truly worth,
leaving bad debt on its books the moment the pool
price snaps back after the attacker unwinds. The
extraction is deterministic, not probabilistic; it
follows from the invariant.

The defence is a matter of estimator design, which
is why it is the statistician's natural territory.
A time-weighted average price (TWAP) over $N$
blocks replaces the single-block spot with a mean;
to move a TWAP an attacker must hold the pool
displaced for many blocks, paying arbitrage losses
each one, which raises the cost of manipulation by
roughly a factor of $N$ and exposes the position to
being traded against. Median-of-quorum feeds and
signed off-chain prices trade manipulation
resistance for a different trust assumption, namely
the honesty of the reporters. There is no
assumption-free oracle; there is only a choice of
which variance and which adversary one prefers. The
audit question of the statistician's second
contribution, above, is exactly this: which is it,
and what does moving it cost?

The reader will note that nothing in this argument
is special to a price. A parametric health
instrument that reads a case count from a
surveillance feed, an outcome-event tally from the
trial ledger, or a breach count from the
cold-chain record faces the identical question. A
single reporting authority is the equivalent of a
single-block spot price, cheap for an interested
party to distort; a quorum of independent
surveillance sources, or a multi-day average of
reported cases, is the equivalent of a TWAP, more
costly to move but slower to respond. This is
precisely the tension the World Bank's bonds
resolved, in the event, on the side of slowness.
The conservative multi-condition trigger was a
manipulation-resistant estimator purchased at the
price of a delayed and partial payout, and whether
that was the right trade is a question of estimator
design, not of blockchain engineering.

## Stablecoins

A stablecoin is a token engineered to hold a
constant value, almost always one United States
dollar. Moin and colleagues, in their
systematisation of the subject [@moin2020sok],
classify the designs by what backs the peg, and the
taxonomy is worth committing to memory because the
failure modes follow from it.

- **Fiat-collateralised.** Each token is a claim
  on a dollar held by a custodian (Tether, USDC).
  The peg is only as good as the custodian's
  reserves and redemption policy; this is a
  tier-two empirical claim about a balance sheet,
  auditable in principle, and reintroduces exactly
  the trusted intermediary the chain claimed to
  remove.
- **Crypto-collateralised.** Each token is backed
  by an over-collateralised position in volatile
  crypto assets (Dai). Solvency is maintained by
  the same liquidation machinery as a lending
  market, and it inherits the same cascade risk
  when the collateral falls fast.
- **Algorithmic.** The peg is defended not by
  collateral but by a control rule that expands and
  contracts supply, often coupled to a companion
  'governance' or 'seigniorage' token. This is the
  fragile design, and the fragility is analysable.

We collect the taxonomy in one place, since the
failure mode of each design follows directly from
what backs the peg (@tbl-stablecoin-taxonomy).

| Collateral class | Peg mechanism | Failure mode | Example |
|---|---|---|---|
| **Fiat-reserve** | Redeemable one-for-one against custodial dollars | Reserve shortfall or a frozen redemption window | USDC |
| **Crypto over-collateralised** | Over-collateralised vault defended by liquidation | Collateral crash outruns the liquidators | DAI |
| **Algorithmic** | Mint-and-burn against a companion token | Reflexive depeg spiral | TerraUSD |

: Stablecoin designs classified by what backs the peg, with the peg mechanism, the characteristic failure mode, and a canonical example of each {#tbl-stablecoin-taxonomy}

Algorithmic stablecoins are reflexive: the
mechanism that defends the peg depends on the price
of a second token whose value depends, in turn, on
confidence in the peg. Klages-Mundt and Minca
[@klagesmundt2019instability] show how such systems
admit multiple equilibria, a stable peg and a
collapse, with no restoring force between them once
confidence breaks. The canonical realisation was
the collapse of TerraUSD (UST) and its companion
token Luna in May 2022, which Briola and colleagues
[@briola2023anatomy] anatomise in detail. UST
maintained its peg by a mint-and-burn arbitrage:
one UST was always redeemable for one dollar of
newly minted Luna. When UST slipped below the peg
under redemption pressure, defending it minted ever
more Luna, whose price collapsed under the new
supply, which destroyed the very backing the
arbitrage relied on, which in turn accelerated the
redemptions. The feedback loop is the same
positive-feedback deleveraging spiral as the
liquidation cascade above, differing only in that
here the forced selling is of the collateral asset
the protocol itself creates. Tens of billions of
dollars of notional value evaporated in days.

The mechanism is easiest to see as a loop, in
which each step feeds the next and the whole turns
faster the further the peg slips
(@fig-terra-spiral).

```{r}
#| label: fig-terra-spiral
#| fig-cap: 'The TerraUSD collapse as a reflexive
#|   loop: a depeg triggers arbitrage that mints the
#|   companion token LUNA, whose inflating supply
#|   drives its price down, which erodes the
#|   confidence backing the peg and deepens the depeg,
#|   so the loop accelerates the very collapse it was
#|   meant to arrest.'
ang <- 90 - (0:4) * 72
rad <- ang * pi / 180
nodes <- tibble(
  x = cos(rad), y = sin(rad),
  lab = c('UST depegs', 'arbitrage mints LUNA',
          'LUNA supply inflates', 'LUNA price falls',
          'confidence falls')
)
nx <- nodes$x; ny <- nodes$y
bx <- nodes$x[c(2:5, 1)]; by <- nodes$y[c(2:5, 1)]
t <- 0.22                          # trim arrows clear of boxes
edges <- tibble(
  x = nx + t * (bx - nx), y = ny + t * (by - ny),
  xend = nx + (1 - t) * (bx - nx), yend = ny + (1 - t) * (by - ny)
)

ggplot() +
  geom_curve(data = edges,
             aes(x = x, y = y, xend = xend, yend = yend),
             curvature = -0.18, colour = '#e34948', linewidth = 0.7,
             arrow = grid::arrow(length = grid::unit(0.28, 'cm'),
                                 type = 'closed')) +
  geom_label(data = nodes, aes(x, y, label = lab),
             colour = '#33356b', fill = 'white',
             linewidth = 0.3, size = 3,
             label.r = grid::unit(0.15, 'lines')) +
  annotate('text', x = 0, y = 0, label = 'reflexive\nspiral',
           colour = '#e34948', size = 3.4, fontface = 'italic') +
  coord_equal(clip = 'off') +
  xlim(-1.6, 1.6) + ylim(-1.4, 1.4) +
  theme_void()
```


The epidemiologist will recognise the shape of this
dynamic. A reflexive collapse is a positive-feedback
process, and it is tempting to read the UST spiral
as the financial cousin of a panic-driven demand
spike in an outbreak, in which fear of shortage
provokes hoarding, which produces the very shortage
that was feared, which deepens the fear. The
formal resemblance is genuine: in both cases a
quantity that measures confidence, the token price
or the perceived availability of a commodity, feeds
back on the behaviour that determines it, and both
admit a stable branch and a runaway branch
separated by a threshold. We should, however, be
candid about where the analogy ends. An epidemic
spike is driven by human belief and can be arrested
by reassurance, by resupply, or by information,
because the feedback runs through people who can be
addressed; the UST spiral was, once past its
threshold, a mechanical arbitrage executed by code
against a supply schedule, with no reassurance
available to a smart contract and no authority
positioned to inject restoring capital in time. The
analogy is useful for intuition about the existence
of tipping points; it is misleading if it suggests
that the interventions which calm a human panic
would have saved an algorithmic peg. Broadening the
view, Werner and colleagues [@werner2022sok] place
these mechanisms, their oracles, and their attack
surfaces in a single framework worth reading in
full.

::: {.callout-important}
## The sceptic's reflex: a high APY is usually emission

An advertised yield of, say, eighty percent
annually is a tier-four narrative until decomposed.
Write it as $\text{APY} = \text{fees} +
\text{emission}$. The fee component is real income:
counterparties paid it for a service. The emission
component is freshly minted governance tokens
handed to liquidity providers, funded by dilution
of existing holders, and it evaporates when the
emission schedule tapers or the token price falls,
which are correlated events. A yield that is
predominantly emission is a wealth transfer from
later to earlier participants, structurally
identical to the token-unlock overhang studied in
@sec-onchain. Before believing a yield, ask for its
two components separately. Protocols that cannot or
will not provide the split are answering the
question by refusing it.
:::

The decomposition generalises past the two-way
fee-versus-emission split. A headline yield is a
sum over sources, and only some of them are income
a counterparty actually paid (@tbl-yield-source).

| Source | Real cash flow? | Sustainable? | Test |
|---|---|---|---|
| **Trading fees** | Yes | Yes, while volume persists | Trace it to observed pool volume |
| **Lending interest** | Yes | Yes, while borrowers pay | Trace it to borrower demand |
| **Token emission** | No | No, decays as the schedule tapers | Read the emission schedule |
| **New-entrant capital** | No | No, a transfer not an earning | Ask who is paying whom |

: The four sources of an advertised yield, of which only the first two are income a counterparty paid; the last two are transfers that decay or reverse {#tbl-yield-source}

### Public-health connection: parametric escrow

We now return to the instrument with which the
chapter opened and assemble it from the parts we
have derived. A smart contract that pays out
automatically on a verifiable trigger is the escrow
of parametric insurance, in which the payout is
tied not to an assessed loss but to an objective
index: rainfall below a threshold, a wind speed
above a cutoff, or an epidemic case count crossing
a level. The conditional-escrow contract holds the
premium capital and releases it the instant the
index, delivered by an oracle, crosses the trigger.
For public-health financing the intuitive appeal is
real: a preparedness fund could disburse to a
ministry of health the day a notifiable-disease
surveillance feed crosses an outbreak threshold,
with no claims adjudication and no discretionary
delay, and the escrow's solvency and payout logic
would be auditable by anyone. This is, in outline,
what the World Bank's 2017 bonds attempted.

The candour we owe the pitch is this: the contract
does not solve the hard problem, it relocates it.
Everything reduces to the trigger and the oracle
that reports it. Did the outbreak truly occur at
the reported scale? Was the surveillance feed
accurate, timely, and free of manipulation by a
party with a financial stake in the outcome? Was
the threshold set low enough to fire when help was
needed, and high enough not to fire spuriously?
These are the same questions, and the same trust
boundary, as the price oracle of the previous
section, and a blockchain answers none of them. The
mathematics of the escrow is trivial; the
epidemiology and the data integrity of the trigger
are the entire problem, and they live off-chain.

To see how the trigger, rather than the plumbing,
governs the outcome, consider a stylised outbreak.
To illustrate, suppose a preparedness bond promises
a fixed disbursement, and its trigger requires the
reported cumulative case count to exceed a fixed
threshold. The epidemic itself grows roughly
exponentially in its early phase, but the reporting
is neither instantaneous nor complete: cases are
observed with a delay and a detection probability
below one, so the count the oracle sees lags the
count that exists. A conservatively high threshold,
chosen to avoid paying out on a false alarm, then
fires only well after the true burden has mounted.
The following synthetic computation traces the true
and reported curves and marks the day the trigger
fires against the day the funds were first needed.

```{r}
#| label: fig-parametric-trigger
#| fig-cap: "SYNTHETIC outbreak. The true cumulative
#|   case curve (solid) rises ahead of the reported
#|   curve (dashed), because reporting is delayed and
#|   incomplete. A conservatively high payout
#|   threshold (horizontal line) is crossed by the
#|   reported curve only on the trigger day (solid
#|   vertical line), long after the day the response
#|   was first needed (dotted vertical line)."
set.seed(7)
days <- 1:90
# SYNTHETIC early-epidemic incidence: roughly
# exponential growth with Poisson noise.
growth <- 0.09
new_true <- rpois(length(days), lambda = 3 * exp(growth * days))

# Reporting: a detection probability below one and a
# fixed reporting delay in days.
detect <- 0.6
delay <- 10
detected <- rbinom(length(days), size = new_true, prob = detect)
reported_by_day <- numeric(length(days))
for (i in seq_along(days)) {
  j <- i + delay
  if (j <= length(days)) reported_by_day[j] <- detected[i]
}

cum_true <- cumsum(new_true)
cum_reported <- cumsum(reported_by_day)

# 'Need' level: true burden at which the response
# should begin. Conservative payout threshold: set on
# the REPORTED count to avoid a false alarm.
need_level <- 1500
threshold <- 6000
need_day <- days[which(cum_true >= need_level)[1]]
trigger_day <- days[which(cum_reported >= threshold)[1]]

curves <- tibble(day = days, true = cum_true,
                 reported = cum_reported) |>
  pivot_longer(c(true, reported),
               names_to = 'series', values_to = 'cases')

ggplot(curves, aes(day, cases, linetype = series)) +
  geom_line() +
  geom_hline(yintercept = threshold, colour = 'grey50') +
  geom_vline(xintercept = need_day, linetype = 3) +
  geom_vline(xintercept = trigger_day) +
  labs(x = 'day since first case', y = 'cumulative cases',
       linetype = NULL,
       title = 'Trigger fires long after the need arises')

c(need_day = need_day, trigger_day = trigger_day,
  lag_days = trigger_day - need_day,
  true_burden_at_trigger = cum_true[trigger_day])
```

The delay is not a defect of the escrow, which
would pay the instant it was commanded; it is a
property of the trigger and its feed. The reported
curve lags the true curve by the reporting delay and
sits below it by the detection shortfall, so that a
threshold set conservatively high, in order not to
disburse on a false alarm, is crossed only after the
true burden has climbed to many times the level at
which help was first warranted. On the one hand a
lower threshold would fire sooner, and on the other
it would fire on noise and on outbreaks that fizzle,
disbursing scarce capital when no response was
needed. That trade-off, between the sensitivity and
the specificity of the trigger, is the whole design
problem, and it is the epidemiologist's problem, not
the programmer's. This is the lesson of the World
Bank's experience, reproduced here in miniature.

::: {.callout-warning}
## Most 'DeFi for health financing' pitches fail the test

The recurring proposal to fund health programmes
through DeFi yields conflates two things a
statistician should keep apart. The escrow-and-
trigger logic is sound and modest, and where a
donor and a recipient lack a common trusted
intermediary, an auditable conditional payout has
real value. The yield-generation layer, staking the
premium float in a lending market or a liquidity
pool to 'grow the fund', imports impermanent loss,
liquidation cascades, oracle-manipulation risk, and
smart-contract risk onto capital that is supposed
to be there when an outbreak hits, which is exactly
when those risks correlate and fire together. A
preparedness fund is not the place to be short
gamma. As @sec-intro warns, if a trusted custodian
already exists or can be appointed, a well-governed
account with an audit log delivers the auditable
payout without the tail risk. Adopt the escrow
logic; refuse the yield farming.
:::

## Worked example: decomposing a liquidity provider's return

The chapter's threads converge on the practical
question a would-be liquidity provider, and equally
the treasurer of a fund tempted to 'earn a yield on
the float', must answer: does the fee income outrun
the impermanent loss? The net return over a period,
relative to holding, is

$$
\text{net} \;=\;
\underbrace{f \cdot \tfrac{t}{365}}_{\text{fee yield}}
\;-\;
\underbrace{\bigl(-\mathrm{IL}(r)\bigr)}_{\text{IL cost}},
$$

where $f$ is the annualised fee yield the position
earns from trading volume, $t$ is the holding
period in days, and $r$ is the realised price
ratio. The following computes the decomposition
across a grid of price moves and two fee
environments.

```{r ch07-lp-pnl}
scenarios <- expand_grid(
  price_move = c(1.00, 1.25, 1.50, 2.00),
  fee_apr = c(0.05, 0.20)
) |>
  mutate(
    r = price_move,
    il_cost = -il(r),                 # positive = loss vs hold
    fee_yield = fee_apr * 30 / 365,   # 30-day holding period
    net = fee_yield - il_cost
  )

scenarios |>
  select(price_move, fee_apr, il_cost, fee_yield, net)
```

Read the table as the LP's decision problem. When
the price barely moves ($r = 1$) impermanent loss
is zero and every basis point of fee is profit.
As the price move grows the IL cost rises
quadratically at first and overwhelms a thin fee
yield: at a five-percent fee environment a
fifty-percent price move already leaves the
provider behind a simple hold, whereas the same
move is survived comfortably when fees run at
twenty percent. The break-even is the locus where
the two columns cancel, and it is the only number
that matters for the decision. Notice what is
absent: any governance-token emission. Had the
protocol advertised its yield inclusive of
emission, the fee_yield column would be inflated by
a component that is dilution rather than income,
and the break-even analysis would be
correspondingly optimistic. Decomposing the yield
before running this table is the discipline the
chapter has been arguing for throughout.

## Collaborating with an LLM

A language model is a capable assistant for DeFi
mathematics and a dangerous one for DeFi claims.
Use it to derive and to draft; verify the economics
yourself.

- **Prompt:** 'Derive the impermanent loss of a
  constant-product pool as a function of the price
  ratio and show it is symmetric in log price.'
  **Watch for:** sign errors, a missing factor of
  two, or a confident derivation that silently
  assumes zero fees or a fifty-fifty pool without
  saying so. **Verification:** check the closed
  form against `il(2)` and `il(0.5)` in the code
  above; if the model's formula disagrees at
  $r = 2$, it is wrong, and the one-line evaluation
  settles it.

- **Prompt:** 'Explain how a flash loan can
  manipulate this lending protocol's oracle and
  estimate the extractable value.' **Watch for:** a
  fluent narrative that omits the atomic-repayment
  constraint, ignores the fee and slippage the
  attacker pays to move the pool, or fails to note
  that a TWAP oracle would defeat the attack.
  **Verification:** reproduce the extraction as a
  small numeric example, as in the oracle chunk,
  and confirm the attacker's round-trip actually
  nets positive after its own slippage.

- **Prompt:** 'This protocol advertises a
  120-percent APY. Is it sustainable?' **Watch
  for:** the model taking the headline at face
  value or inventing a reserves figure. **Verification:**
  insist on the fee-versus-emission decomposition;
  a sustainable yield is a real cash flow from
  counterparties, and if the model cannot source
  the fee component from actual volume, treat the
  number as emission until shown otherwise.

## Exercises

1. **Slippage bound.** For a no-fee constant-product
   pool with reserves $(x, y)$, prove that the
   average execution price of a purchase of size
   $\mathrm{d}x$ is exactly $(x + \mathrm{d}x)/y$,
   and hence that fractional slippage relative to
   spot equals $\mathrm{d}x/x$. Confirm numerically
   against `cpmm_out()` with the fee set to zero.

2. **Impermanent loss is second order.** Show by
   Taylor expansion that $\mathrm{IL}(r) \approx
   -\tfrac{1}{8}(r - 1)^2$ near $r = 1$, and hence
   that for small price moves the loss scales with
   the square of the move. Verify the approximation
   against `il()` for $r \in \{1.01, 1.05, 1.1\}$
   and report the relative error.

3. **Break-even fee yield.** Using the LP
   decomposition, derive the annualised fee yield
   $f$ that exactly offsets the impermanent loss of
   a given realised price ratio $r$ over a 30-day
   period. Tabulate the break-even $f$ for
   $r \in \{1.25, 1.5, 2, 4\}$ and comment on which
   pools are viable LP positions.

4. **Liquidation price and cascade sensitivity.**
   Extend `simulate_liq_cascade()` to sweep the
   leverage parameter over $\{2, 5, 10, 20\}$ and
   plot the fraction liquidated against leverage
   for a fixed initial shock. Explain the shape in
   terms of the liquidation-price formula
   $P^\star = D/(\ell q)$.

5. **Oracle manipulation cost.** Modify the flash-
   loan example so the lending market prices
   collateral off a TWAP over $N$ blocks rather than
   the single-block spot. Argue, and illustrate
   numerically, why the attacker's cost to move the
   reported price scales with $N$, and state the
   assumption about inter-block arbitrage that the
   argument requires.

6. **Reflexivity of an algorithmic peg.** Write a
   minimal two-state model of a mint-and-burn
   stablecoin in which the backing token's price is
   a decreasing function of its supply. Identify the
   two equilibria of @klagesmundt2019instability and
   the confidence threshold that separates them, and
   relate the collapse branch to the TerraUSD
   account of @briola2023anatomy.

7. **Parametric trigger sensitivity and
   specificity.** Using the synthetic outbreak of
   @fig-parametric-trigger, sweep the payout threshold
   over a range of values and, for each, record the lag
   between the need day and the trigger day.
   Then introduce a second, non-epidemic scenario in
   which reported cases plateau below the true
   outbreak level, and report for each threshold
   whether the trigger fires spuriously. Explain the
   sensitivity-specificity trade-off this exposes,
   and argue why it, and not the escrow contract, is
   the binding constraint on a parametric health
   bond.

## Further reading

The Uniswap v2 [@adams2020uniswapv2] and v3
[@adams2021uniswapv3] specifications are the
primary sources for the constant-product and
concentrated-liquidity market makers; read v2
first, since its invariant is the one derived here
and v3 is a change of variables on the same idea.
The mathematical theory of constant-function
markets, including when prices are well-defined and
how curvature governs slippage and loss, is
developed rigorously by @angeris2020improved and
@angeris2021analysis; these are the papers that
turn AMMs from an engineering trick into an object
with theorems. For lending and its instabilities,
@gudgeon2020loanable models the interest-rate
mechanisms and @gudgeon2020crisis analyses how they
behave under stress. The stablecoin landscape is
mapped by the systematisation of @moin2020sok, and
the specific instability of algorithmic designs by
@klagesmundt2019instability, with the TerraUSD
collapse dissected empirically by
@briola2023anatomy. Finally, the DeFi
systematisation of @werner2022sok is the single
best map of the whole territory, oracles, flash
loans, composability risk, and the attack surface
they jointly create, and it is the reference to
keep at hand when reading any new protocol.

## In conclusion

In conclusion, three points are to be emphasized.
First, the mechanisms of decentralised finance,
once stripped of their promotional language, are
elementary and auditable: a swap is a move along a
level set, impermanent loss is the arithmetic-
geometric mean inequality made financial, a
liquidation is a threshold crossing, and each
admits a closed form a statistician can check in a
line. Second, in every one of these mechanisms the
oracle is the assumption that matters, and the
public-health instrument that motivated the chapter
inherits exactly this: a parametric bond is only as
sound as the surveillance feed that reports its
trigger, and the choice between a fast, manipulable
index and a slow, robust one is a problem of
estimator design, as the World Bank's experience
with its 2017 pandemic bonds makes plain. Third,
the escrow-and-trigger logic merits adoption where
no common trusted custodian exists, but the
yield-generation layer that is so often bolted onto
it, and which imports impermanent loss, liquidation
cascades, and oracle risk onto capital that must be
present precisely when those risks fire together,
should be refused, since a preparedness fund is not
the place to be short gamma, and where a
well-governed account with an audit log will serve,
it is to be preferred.
