# Maximal Extractable Value {#sec-mev}

::: {.bookepigraph}
```{=latex}
\begin{flushright}\small
```
*We additionally show that high fees paid for priority
transaction ordering pose a systemic risk to
consensus-layer security.*

[Daian et al.]{.smallcaps}, *Flash Boys 2.0* (2020)
```{=latex}
\end{flushright}
```
:::

## Learning objectives

By the end of this chapter the reader should be
able to:

- define MEV precisely as the value of
  transaction inclusion and ordering, and locate
  it in the pipeline from mempool to block;
- classify the three canonical strategies
  (arbitrage, liquidation, sandwich) by whether
  they close or manufacture inefficiency;
- formulate the sandwich attack as a constrained
  optimization on the constant-product curve from
  @sec-defi and compute the optimal front-run;
- explain how fee-driven reordering threatens
  consensus stability, following the analysis of
  @daian2020flash;
- describe proposer-builder separation as an
  auction that redistributes rather than abolishes
  MEV;
- audit a claim of 'MEV protection' by quantifying
  the residual extractable value it leaves
  unaddressed;
- transfer the ordering-tax lesson to any
  allocation queue, recognizing that ordering
  power together with advance visibility
  constitutes a tax whether the venue is a mempool
  or a waiting list.

## Orientation

We begin not with a blockchain but with a waiting
list. Whenever a scarce health resource is
allocated by a queue, whether a transplant waiting
list, a vaccine appointment slot during a
shortage, or the procurement of a scarce
therapeutic, the party who controls the order of
that queue, or who can merely see the pending
requests before they settle, holds something of
value. The value is the ability to place a
favored request ahead of others, or to act on the
knowledge of what is coming before it arrives.
History gives us little reason to expect such
value to go unextracted. Wherever ordering power
and advance visibility coincide, they function as
a tax on the participants whose requests are
reordered, and the tax is collected whether or not
anyone troubles to name it.

Maximal extractable value (MEV) is the
cryptocurrency instance of this general
phenomenon. Its setting is the automated market
maker of @sec-defi (a venue that prices each trade
by a fixed formula against a pool of reserves
rather than by matching buyers to sellers), which
priced a swap by a deterministic curve: given the
reserves, the output of any trade is a known
function of its size. That determinism was
presented as a virtue, and for a lone trader it
is. Here we study its
adversarial consequence. If the price a user
receives is a known function of the pool state,
and if the pool state can be altered by inserting
a transaction immediately before the user's, then
the profit an adversary can extract by choosing
what to insert, and in what order, is itself a
deterministic function of the user's trade and the
pool depth. It is not a matter of luck or of rare
exploitation but an optimization problem with a
closed-form-adjacent solution, solved thousands of
times per day.

MEV carries one property that its public-health
analogues lack, and that property is what makes it
worth our close study: on a public chain the
extraction is measurable to the dollar. A waiting
list manipulated for advantage leaves, at best, a
statistical residue open to dispute, whereas a
sandwich attack (an adversary bracketing a
victim's trade with its own transactions in order
to worsen the victim's price) leaves an auditable
trail of transactions from which the extracted
value can be computed exactly. We shall exploit that
measurability to make the general lesson precise,
and then carry the lesson back to the settings in
which it cannot be measured but nonetheless
applies.

The reader arrives equipped for the technical
work. Choosing an inclusion set and an order to
maximize a payoff is a scheduling problem. The
attacker's profit as a function of front-run size
(the size of the trade the attacker inserts ahead
of the victim) is a smooth objective on a compact
set, to be maximized by the same reasoning used
for any one-dimensional optimum. The market that
sells the right to order transactions is an
auction,
analysable with the tools of auction theory. The
work of the chapter is to make these translations
and, in doing so, to remove the mystique MEV is
usually granted and to exhibit it as market
microstructure (the study of how the mechanics of
trading and order handling determine prices).

The stakes are not only pecuniary. Ordering that
is economically valuable and visible before
settlement is an information-integrity failure,
and the same structure recurs wherever a queue of
sensitive operations is exposed before it is
committed. The book's running case study, a
multi-institution clinical-trial data-integrity
ledger, furnishes the health-data instance we
shall develop once the mechanism is precise:
several hospitals maintain a shared, append-only
record of enrollment, randomization, and outcome
events with no single trusted custodian, and a
regulator audits it after the fact. We refer to it
hereafter as the trial ledger. A ledger of that
kind places a queue of pending allocations in
view, and a queue of pending allocations is
precisely the surface this chapter teaches one to
recognize.

The plan of the chapter is as follows. We first
make the notion of extractable value precise and
classify the three canonical strategies by whether
they close or manufacture an inefficiency. We then
treat the sandwich quantitatively, correcting a
common misdescription of its profit function, and
follow @daian2020flash, the study published as
'Flash Boys 2.0', in tracing how competition for
MEV destabilizes the consensus layer beneath it.
Proposer-builder separation is examined as an
auction that redistributes rather than abolishes
the tax. Lastly we return to the trial ledger and
to allocation-order integrity in health systems
generally, where the mechanism recurs but the
remedy is regulatory rather than market-based.

Before proceeding we fix the vocabulary on which
the argument depends, so that the reader need not
carry these terms as mysteries.

::: {.callout-note}
## The vocabulary of this chapter

Several terms recur throughout the chapter. We
define them here and use them consistently
thereafter.

- **Maximal (miner) extractable value (MEV).** The
  value a block producer can capture through its
  power to decide which pending transactions enter
  a block and in what order, over and above the
  ordinary transaction fees and block reward. The
  term arose as 'miner extractable value' under
  proof-of-work and was reread as 'maximal' under
  proof-of-stake.
- **The mempool.** The public holding area of
  pending transactions. In an open network a
  proposed transaction broadcasts to every node
  and waits there, visible to all, until some
  producer includes it in a block.
- **Transaction ordering.** The producer's right
  to choose both which transactions to include and
  the sequence in which they execute. Because a
  trade's price depends on the pool state it meets,
  the sequence is economically consequential, and
  it is ordering, not mere inclusion, that MEV
  monetizes.
- **Front-running and back-running.** Placing
  one's own transaction immediately before a target
  transaction (front-running) or immediately after
  it (back-running) in order to profit from the
  price movement the target causes.
- **The sandwich attack.** The pairing of a
  front-run and a back-run around a victim's trade,
  so that the victim executes at a price the
  front-run has deliberately worsened while the
  attacker unwinds its position immediately after.
- **Arbitrage.** A trade that profits by equalizing
  a price that differs between two venues, or
  between a pool and an external reference, thereby
  restoring a single price.
- **Liquidation.** The forced closing of a
  collateralized loan whose collateral has fallen
  below a required threshold, for which the lending
  protocol pays the closing party a discount.
:::

## The statistician's contribution

Three judgments in this chapter are the analyst's
to make, and no tool makes them automatically.

1. **Quantify extractable value as an optimization
   on a known curve.** MEV is frequently reported
   as an aggregate dollar figure of unclear
   provenance. The disciplined statement is a
   function: given the victim trade, the pool
   reserves, and the fee, the extractable value is
   the maximum of an explicit profit function over
   the attacker's decision variable. Writing that
   function down converts folklore into a
   computation we can check, and, as this chapter
   shows, checking it sometimes contradicts the
   folklore.

2. **Recognize MEV as a structural tax, not a
   bug.** A sandwich is not a protocol defect to be
   patched. It is the equilibrium consequence of
   two design choices: a public queue of pending
   trades and a block producer free to order them.
   The extractable value is therefore a levy on
   users that scales with their trade size and
   their tolerance for slippage, the gap between
   the price a trade is quoted and the price it in
   fact receives. Naming it a tax,
   and estimating its rate, is more useful than
   naming it an attack.

3. **Audit 'MEV protection' claims.** A protocol
   or wallet that advertises protection is making
   an empirical claim: that the residual
   extractable value under its mechanism is small.
   That claim is testable. The statistician's
   contribution is to compute the residual, not to
   accept the adjective.

## What MEV is

A block producer assembles the next block from a
pool of pending transactions. In an open network
that pool, the mempool, is public: proposed
transactions broadcast to every node and wait,
visible, until some producer includes them. The
producer is not obliged to include transactions in
the order they arrived, nor to include any
particular transaction at all. Within the block it
builds, the producer chooses the inclusion set and
the order.

That freedom has value. Under the fee market of
@eip1559 (the 2021 reform that split each
transaction's fee into a base fee and a priority
tip) a transaction carries a priority fee, a tip
paid directly to the producer, so the naive
value of ordering is the sum of tips the producer
can collect. Maximal extractable value is the
larger quantity that includes the profit the
producer can capture by inserting its own
transactions and by ordering the block's contents
to its advantage. The term originated as 'miner
extractable value' in proof-of-work, where the
miner producing a block was the ordering
authority; under proof-of-stake (@sec-pos) the
proposer, the validator whose turn it is to
produce the block, plays that role, and the
initialism was retained by reading it as 'maximal'.

Two roles populate the supply chain. A searcher is
an agent who scans the mempool for profitable
ordering opportunities and submits transactions,
or bundles of transactions, designed to capture
them. A block producer, or an intermediary the
producer delegates to, decides which of these
bundles enters the block. A searcher who is not
the producer must pay the producer for favorable
placement, through a priority fee or an explicit
bribe. The value extracted is thus split between
searcher and producer by a market we examine under
proposer-builder separation below.

::: {.callout-note icon=false collapse=true}
## Check your understanding: why ordering has value

A pending swap will move a pool's price. Two
searchers see it. Searcher A wants its own buy to
execute just before the swap; searcher B wants its
buy just after. Neither transaction is invalid or
dishonest in isolation. Explain, in terms of the
constant-product curve, why the position of a
searcher's transaction relative to the victim's
determines whether it profits, and why this makes
ordering, not merely inclusion, the scarce
resource.

The pool's marginal price of $X$ in units of $Y$
is $R_y / R_x$, increasing in the amount of $X$
already added. A buy placed *before* the victim's
buy executes at a lower price and, crucially,
raises the price the victim then pays, so the
victim's own trade pushes the price further in the
searcher's favor before the searcher sells. A buy
placed *after* the victim's executes at the price
the victim already inflated, with no such feedback.
Identical transactions, opposite outcomes: the
payoff is a function of position in the order, not
of the transaction's content. Inclusion is
necessary but the order is what is monetized, which
is why producers can auction it.
:::

## Three canonical strategies

MEV strategies are not uniform in their welfare
effect. It is worth separating those that correct
a mispricing from those that manufacture one.
Before treating each in turn, we set the three
side by side. @tbl-mev-strategies distinguishes
them by who bears the cost, whether the strategy
performs a function the system needs, what drives
its magnitude, and the health-queue situation each
most resembles.

| Strategy | Who bears the cost | Socially useful | Magnitude driver | Health-queue analogue |
|---|---|---|---|---|
| **Arbitrage** | Liquidity providers | Yes, realigns the pool to the market price | Size of the mispricing | Reordering a queue back to the stated policy |
| **Liquidation** | The undercollateralised borrower | Yes, removes bad debt | Collateral shortfall and the liquidation discount | Enforcing a triage threshold once it is crossed |
| **Sandwich** | The targeted victim trader | No, manufactures then unwinds a mispricing | Victim slippage tolerance times trade size | Advancing a favored patient ahead in the queue |

: MEV strategies compared by welfare effect and health-queue analogue {#tbl-mev-strategies}

### Arbitrage

When the same asset trades at different prices on
two venues, or when a pool's price diverges from an
external reference, a trade that equalizes them
earns the difference. On a constant-product pool
with reserves $R_x, R_y$ (a market maker whose
reserves multiply to a constant $R_x R_y = k$ that
every trade preserves) and no fee, the marginal
price of $X$ is $R_y/R_x$. If an external market
prices $X$ at $p$, the pool is mispriced whenever
$R_y/R_x \neq p$, and the trade that restores
$R_y/R_x = p$ has a closed form. Selling
$\Delta x^\star = \sqrt{R_x R_y / p} - R_x$ units
of $X$ into the pool (valid when the pool
over-prices $X$) brings its marginal price exactly
to $p$.

```{r ch08-setup}
library(ggplot2)
library(dplyr)
library(tibble)

swap_x_for_y <- function(dx, Rx, Ry) {
  dy <- Ry * dx / (Rx + dx)
  list(dy = dy, Rx = Rx + dx, Ry = Ry - dy)
}
swap_y_for_x <- function(dy, Rx, Ry) {
  dx <- Rx * dy / (Ry + dy)
  list(dx = dx, Rx = Rx - dx, Ry = Ry + dy)
}
```

```{r ch08-arb}
Rx <- 1e6; Ry <- 1.1e6            # pool prices X at 1.10
p_ext <- 1.00                     # external price of X

dx_star <- sqrt(Rx * Ry / p_ext) - Rx
after <- swap_x_for_y(dx_star, Rx, Ry)
arb_profit <- after$dy - dx_star * p_ext

c(trade = dx_star,
  pool_price_after = after$Ry / after$Rx,
  profit_Y = arb_profit)
```

The trade drives the pool price to exactly the
external price and books a profit equal to the
mispricing it removed. Arbitrage of this kind is
the benign end of MEV: it is a deterministic
optimization, it is competed for, and its social
effect is to keep pool prices aligned with the
wider market. It taxes liquidity providers (the
participants who fund the pool and stand as
counterparty to the correcting trade) but it does
not target any individual user's transaction.

::: {.callout-important}
## The skeptic's reflex

'MEV is good for users because arbitrage keeps
prices efficient.' Name the tier. That arbitrage
closes price gaps is a protocol-level fact about
the curve, verified by the computation above. That
MEV *as a whole* benefits users is an economic
claim, and it is false as stated: the same ordering
power that enables benign arbitrage enables the
sandwich, which is a direct transfer from a user to
an adversary. The efficiency of arbitrage does not
net against the extraction of sandwiching; they are
different trades with different victims. Conflating
the benign case with the whole category is exactly
the tier-slippage this book trains the reader to
catch.
:::

### Liquidations

Lending protocols hold collateralized loans that
become eligible for liquidation when the collateral
value falls below a threshold. The protocol offers
a liquidator a discount for closing an
undercollateralised position. Because eligibility
is a deterministic function of an on-chain price
and the loan state, the moment a position crosses
the threshold is observable, and searchers compete
to be the transaction that liquidates it. Like
arbitrage, liquidation performs a function the
system needs (removing bad debt) while extracting
value from the ordering right to perform it. The
competition is again a race for position in the
block.

### The sandwich

The sandwich is the extractive case. It targets a
specific pending swap and profits by degrading the
price that swap receives. The attacker places one
transaction before the victim and one after, so the
victim executes 'sandwiched' between them. Unlike
arbitrage, it corrects no mispricing: it creates a
temporary one, forces the victim to trade into it,
and then undoes it. The remainder of the chapter
treats the sandwich quantitatively, because it is
the strategy for which the statistician's framing
is most clarifying and the popular account most
often wrong.

We picture the mechanism before we quantify it.
@fig-sandwich-timeline shows the attacker
occupying the slots immediately before and after
the victim inside one block, the arrangement that
lets a front-run inflate the price the victim then
pays.

```{r}
#| label: fig-sandwich-timeline
#| fig-cap: 'The sandwich places the attacker on both sides of the
#|   victim within a single block, so the victim buys at a price the
#|   front-run has already inflated and the attacker unwinds the
#|   position immediately after.'
#| fig-width: 7
#| fig-height: 3
PAL <- c('#2a78d6', '#1baf7a', '#eda100', '#008300',
         '#4a3aa7', '#e34948', '#e87ba4', '#eb6834')
book_theme <- ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(
    panel.grid.minor = ggplot2::element_blank(),
    panel.grid.major = ggplot2::element_line(
      linewidth = 0.25, colour = 'grey88'),
    axis.title = ggplot2::element_text(size = 10),
    legend.position = 'bottom')
ink <- '#33356b'; hit <- '#e34948'; ctx <- '#9a9a92'

ggplot() +
  annotate('rect', xmin = 0.2, xmax = 2.7, ymin = 2.9, ymax = 4.6,
           fill = 'grey95', colour = ctx, linewidth = 0.5,
           linetype = 2) +
  annotate('text', x = 1.45, y = 4.35, label = 'Public mempool',
           colour = ctx, size = 3.3, fontface = 'bold') +
  annotate('text', x = 1.45, y = 3.65,
           label = 'victim swap\npending, visible', colour = ctx,
           size = 3) +
  annotate('segment', x = 2.75, xend = 3.25, y = 3.75, yend = 3.75,
           colour = ink, arrow = arrow(length = unit(0.16, 'cm'))) +
  annotate('rect', xmin = 3.3, xmax = 9.6, ymin = 2.6, ymax = 4.9,
           fill = NA, colour = ink, linewidth = 0.6) +
  annotate('text', x = 6.45, y = 4.65,
           label = 'Block (ordering fixed by producer)', colour = ink,
           size = 3.2, fontface = 'bold') +
  annotate('rect', xmin = 3.55, xmax = 5.4, ymin = 3.1, ymax = 4.3,
           fill = PAL[1], colour = ink, linewidth = 0.5) +
  annotate('rect', xmin = 5.55, xmax = 7.4, ymin = 3.1, ymax = 4.3,
           fill = hit, colour = ink, linewidth = 0.5) +
  annotate('rect', xmin = 7.55, xmax = 9.4, ymin = 3.1, ymax = 4.3,
           fill = PAL[1], colour = ink, linewidth = 0.5) +
  annotate('text', x = 4.47, y = 3.7,
           label = '1  front-run\nbuy (attacker)', colour = 'white',
           size = 2.9) +
  annotate('text', x = 6.47, y = 3.7,
           label = '2  victim buy\nat worse price', colour = 'white',
           size = 2.9) +
  annotate('text', x = 8.47, y = 3.7,
           label = '3  back-run\nsell (attacker)', colour = 'white',
           size = 2.9) +
  annotate('segment', x = 3.55, xend = 9.4, y = 2.35, yend = 2.35,
           colour = ctx, linewidth = 0.3) +
  annotate('text', x = 4.47, y = 2.05, label = 'price rises\np0 -> p1',
           colour = ink, size = 2.7) +
  annotate('text', x = 6.47, y = 2.05,
           label = 'executes at\ninflated p1', colour = hit,
           size = 2.7) +
  annotate('text', x = 8.47, y = 2.05, label = 'price falls\np1 -> ~p0',
           colour = ink, size = 2.7) +
  coord_cartesian(xlim = c(0, 9.8), ylim = c(1.6, 5)) +
  theme_void()
```

## The sandwich as an optimization

Fix a pool with reserves $R_x, R_y$ and a victim
who will swap $v$ units of $X$ for $Y$. The
attacker front-runs with a buy of size $a$, the
victim then buys at the worsened price, and the
attacker sells back the $Y$ it acquired. Every leg
is a constant-product swap, so the attacker's
profit is an explicit function of the single
decision variable $a$. We reuse the swap primitives
from above and grid-search the objective. The pool
fee is set to zero here so the mechanism is visible
without the fee's confounding; we restore a
binding cost in a moment.

```{r ch08-sandwich}
sandwich_profit <- function(a, v, Rx, Ry) {
  s1 <- swap_x_for_y(a, Rx, Ry)          # attacker front-run buy
  s2 <- swap_x_for_y(v, s1$Rx, s1$Ry)    # victim buy at worse price
  s3 <- swap_y_for_x(s1$dy, s2$Rx, s2$Ry)  # attacker sells its Y back
  s3$dx - a
}

Rx <- 1e6; Ry <- 1e6; v <- 5e4
a_grid <- seq(1, 2e5, length.out = 400)
profit <- sapply(a_grid, sandwich_profit, v = v, Rx = Rx, Ry = Ry)
a_star <- a_grid[which.max(profit)]
c(optimal_a = a_star, max_profit = max(profit))
```

```{r ch08-fig-sandwich}
#| label: fig-sandwich
#| fig-cap: "Zero-fee sandwich profit against front-run size. The
#|   curve rises monotonically over the searched range; the reported
#|   optimum sits at the grid boundary, not at an interior peak."
ggplot(tibble(a = a_grid, profit = profit), aes(a, profit)) +
  geom_line() +
  geom_vline(xintercept = a_star, linetype = 3) +
  labs(x = 'attacker front-run size a', y = 'attacker profit (X)',
       title = 'Zero-fee sandwich profit rises in front-run size')
```

The reported optimum lies at the right edge of the
grid, and @fig-sandwich shows why: with a zero fee
the profit is monotonically increasing in $a$. The
grid search did not find an interior peak because
there is none to find. This is worth dwelling on,
because a widely repeated description of the
sandwich asserts that profit is concave in the
front-run size with an interior optimum, and here
the computation contradicts the description. If we
push the front-run far beyond the grid, the reason
becomes clear.

```{r ch08-ceiling}
c(front_run_1e9 = sandwich_profit(1e9, v, Rx, Ry),
  victim_input = v)
```

With no fee, the attacker's round-trip is costless,
so a larger front-run is always weakly better, and
the profit rises toward a ceiling equal to the
victim's entire input $v$. Intuitively, the
attacker can extract at most the value the victim
loses to slippage, and in the limit of an enormous
front-run the victim receives almost nothing while
the attacker captures very nearly the full $v$.
There is no too-large front-run to punish when the
round-trip is free, hence no interior optimum. The
concavity story is not wrong in spirit, but it
requires a cost that the zero-fee model omits. Two
costs supply it: the pool's trading fee, charged on
every leg including the attacker's own round-trip,
and, more sharply, the victim's slippage limit,
which removes the victim entirely if the front-run
is too aggressive. We develop the second, which is
both the dominant real-world bound and the quantity
a 'protection' claim is really about, in the worked
example.

::: {.callout-note icon=false collapse=true}
## Check your understanding: the extraction ceiling

Using only the constant-product invariant
$R_x R_y = k$, argue that a zero-fee sandwich of a
victim swapping $v$ units of $X$ can extract at
most $v$ units of $X$, and state the limiting
front-run behavior that approaches this bound. Why
does this ceiling, rather than a stationary point
of the profit function, govern the zero-fee case?

The attacker ends holding $X$ only, so its profit
is the net change in the pool's $X$ balance that
accrues to it. Across the three swaps the victim
adds $v$ units of $X$ to the pool and no leg
creates $X$, so the attacker cannot withdraw more
net $X$ than the victim contributed, bounding the
gain by $v$. As $a \to \infty$ the front-run drives
$R_y$ toward zero before the victim trades, the
victim receives almost no $Y$ for its $v$, and the
attacker's sell-back recovers nearly all of that
$v$. The profit function is monotone increasing and
asymptotes to $v$, so the supremum is a boundary
limit, not an interior maximum. An interior optimum
appears only once a per-leg cost makes very large
front-runs self-defeating.
:::

## Flash Boys 2.0: reordering and consensus instability

The foregoing establishes that extractable value is
a deterministic function of public state. The
consequence, drawn out by @daian2020flash in the
paper that named the phenomenon for a general
audience, is that competition to capture it
destabilizes the layer beneath it. Their argument
has two parts, and both are quantitative.

First, front-running competition becomes an
auction conducted in the fee. When several
searchers detect the same opportunity, each wants
its transaction ordered first, and under a
fee-priority regime the producer orders by fee. The
searchers therefore bid each other up in a
'priority gas auction', raising fees until they
approach the value of the opportunity itself. This
is an ascending auction whose proceeds accrue to
the producer, and it explains a puzzling empirical
signature: sudden fee spikes uncorrelated with
ordinary user demand, driven by two bots outbidding
one another over a single arbitrage. The fee market
of @eip1559 reshaped the mechanics, replacing a
first-price free-for-all with a base-fee-plus-tip
structure, but the tip is still an ordering auction
and the competitive dynamic survives.

Second, and more seriously, the extractable value
in *past* blocks can exceed the reward for
producing a *new* block. When it does, a rational
producer's best move is not to extend the chain but
to re-mine a recent block and capture the MEV
inside it for itself, a 'time-bandit'
reorganization that rewrites confirmed history for
profit rather than for a double-spend. This couples
the application layer to consensus in a way the
security analysis of @sec-security did not
contemplate: the cost of rewriting history is
offset by the MEV reward for doing so, lowering the
effective security of finality. @daian2020flash
frame this as a systemic risk, and the framing is
apt. MEV is not only a tax on users; above a
threshold it is a bribe to abandon consensus.

The disciplined reading keeps the two claims in
their tiers. That reordering is profitable and
fee-competition is an auction are protocol-level
facts, verifiable from the mechanism. That
time-bandit reorgs occur at scale is an empirical
claim whose magnitude depends on the ratio of
extractable value to block reward, and it is that
ratio, not the mere possibility, that determines
whether the risk is realized.

## Proposer-builder separation and the MEV supply chain

If ordering is a valuable and contestable right,
the natural institutional response is to organize a
market for it, and that is what proposer-builder
separation (PBS) does. PBS decomposes the producer
into two roles. A builder assembles a full, ordered
block, including whatever MEV bundles it has
sourced, and attaches a bid: the payment it will
make to the proposer if its block is chosen. A
proposer, the validator whose turn it is to add a
block, does not build; it selects the
highest-bidding block and signs it. Between them sit
relays, trusted intermediaries that hold builders'
blocks and reveal them to the proposer only after
the proposer has committed to the bid, preventing
the proposer from stealing the block's contents.

We can lay the supply chain out as a diagram.
@fig-pbs-supply-chain traces a bundle from
searcher through builder and relay to proposer,
marks the sealed-bid auction that settles the
ordering right, and shows private order flow as
the bypass that never touches the public mempool.

```{r}
#| label: fig-pbs-supply-chain
#| fig-cap: 'The proposer-builder separation supply chain routes each
#|   bundle from searcher to builder to relay to proposer, settling the
#|   ordering right in a sealed-bid auction, while private order flow
#|   lets a user reach the builder directly and never enter the public
#|   mempool.'
#| fig-width: 7
#| fig-height: 3.2
node_rect <- function(xmin, xmax, label, fill = 'grey96') {
  list(
    annotate('rect', xmin = xmin, xmax = xmax, ymin = 3.0, ymax = 4.0,
             fill = fill, colour = ink, linewidth = 0.5),
    annotate('text', x = (xmin + xmax) / 2, y = 3.5, label = label,
             colour = ink, size = 3.1, fontface = 'bold'))
}
flow_arr <- function(x, xend) {
  annotate('segment', x = x, xend = xend, y = 3.5, yend = 3.5,
           colour = ink, linewidth = 0.5,
           arrow = arrow(length = unit(0.16, 'cm')))
}
ggplot() +
  node_rect(0.3, 2.0, 'Searcher') +
  node_rect(2.7, 4.4, 'Builder') +
  node_rect(5.1, 6.8, 'Relay') +
  node_rect(7.5, 9.2, 'Proposer') +
  flow_arr(2.02, 2.68) + flow_arr(4.42, 5.08) +
  flow_arr(6.82, 7.48) +
  annotate('text', x = 2.35, y = 3.75, label = 'bundle', colour = ink,
           size = 2.6) +
  annotate('segment', x = 3.55, xend = 8.35, y = 4.55, yend = 4.55,
           colour = PAL[3], linewidth = 0.4) +
  annotate('text', x = 5.95, y = 4.85,
           label = 'sealed-bid auction: highest bid wins the slot',
           colour = PAL[3], size = 2.8, fontface = 'bold') +
  annotate('segment', x = 3.55, xend = 3.55, y = 4.0, yend = 4.5,
           colour = PAL[3], linewidth = 0.3, linetype = 3) +
  annotate('segment', x = 8.35, xend = 8.35, y = 4.0, yend = 4.5,
           colour = PAL[3], linewidth = 0.3, linetype = 3) +
  annotate('rect', xmin = 0.3, xmax = 2.0, ymin = 1.1, ymax = 2.0,
           fill = 'grey96', colour = PAL[2], linewidth = 0.5) +
  annotate('text', x = 1.15, y = 1.55, label = 'User', colour = PAL[2],
           size = 3.1, fontface = 'bold') +
  annotate('curve', x = 2.02, xend = 3.55, y = 1.55, yend = 2.98,
           colour = PAL[2], linewidth = 0.5, curvature = -0.3,
           arrow = arrow(length = unit(0.16, 'cm'))) +
  annotate('text', x = 3.35, y = 1.75,
           label = 'private order flow\n(bypasses mempool)',
           colour = PAL[2], size = 2.7, hjust = 0) +
  coord_cartesian(xlim = c(0, 9.5), ylim = c(0.9, 5.1)) +
  theme_void()
```

The economic content is an auction, and the
statistician should read it as one. Builders compete
in what is effectively a first-price sealed-bid
auction for the right to have their block proposed.
In equilibrium a builder bids up to the MEV it can
extract, less its margin, so the auction transfers
most of the extractable value from the builder to
the proposer. This is a redistribution, not an
elimination. The user whose trade is sandwiched is
no better off; the sandwich still occurs, and its
proceeds now flow up the chain to the proposer via
the builder's bid. PBS improves decentralization (a
small validator can capture MEV revenue without
running a sophisticated searching operation) and it
mitigates the consensus-instability worry by making
MEV a smooth ongoing revenue stream rather than a
lumpy incentive to reorg. What it does not do is
reduce the tax.

A further wrinkle is private order flow. If users
send transactions directly to a builder through a
private channel rather than to the public mempool,
those transactions are never exposed to searchers,
and the sandwich opportunity never forms. Private
order flow is thus a genuine protection, but it
purchases integrity by reintroducing a trusted
intermediary, the builder who sees the order first.
The property the public mempool sacrificed,
integrity of ordering without a custodian, is
recovered only by appointing a custodian.

The posture PBS adopts, to constrain and channel
ordering power rather than to remove it, is
familiar from the governance of allocation queues
in health systems. We do not eliminate ordering
power on a transplant waiting list or in the
appointment queue for a scarce vaccine; we
constrain it with published allocation rules and
we audit it after the fact. The parallel should be
drawn with care, however, for it ends at a
substantive point. MEV is redistributed by a
market, the sealed-bid auction transferring the
tax from builder to proposer, whereas the
integrity of a health-allocation queue is secured
by regulation and audit, not by selling the
ordering right to the highest bidder. Auctioning
the right to order a transplant list would be a
scandal rather than a mechanism. What transfers
between the two settings is diagnostic, the
recognition of an ordering-and-visibility surface,
and not the market remedy.

::: {.callout-important}
## The skeptic's reflex

'PBS solves MEV.' Name the tier. That PBS moves MEV
revenue from builders to proposers and smooths the
reorg incentive is an economic claim that holds
under its stated auction assumptions, and it is
largely true. That PBS *solves* MEV is a narrative,
and it fails on the definition: the extractable
value is a function of the public pending trade and
the pool depth, and PBS changes neither. It
reassigns who collects the tax, not whether the
user pays it. A mechanism that redistributes a levy
is not a mechanism that abolishes it, and marketing
that elides the difference is selling the narrative
in the cadence of the fact.
:::

## Ordering integrity beyond crypto

Strip away the tokens and the sandwich is an
instance of a general failure: a system in which
the order of operations is economically valuable
and is visible before it is committed. Wherever
both conditions hold, whoever controls or observes
the queue can profit by reordering it, and the
participants whose operations are reordered bear
the cost. The public mempool is the crypto-specific
form of the vulnerability, but the structure is not
crypto-specific.

To illustrate, consider the trial ledger in a
concrete shortage scenario. Suppose four hospitals
share the ledger during a trial of a scarce
therapeutic, and that each day a fixed number of
enrollment-and-randomization entries are appended,
say twenty slots against forty eligible patients.
If the pending entries are visible before they are
committed, a coordinator at one site who can see
that the next randomization will assign the active
arm, and who can advance a preferred patient into
that position, extracts value in the precise sense
of this chapter: the value of controlling the order
of a queue whose order is outcome-relevant. No
tokens change hands, yet the integrity of the
allocation is compromised exactly as a swap's
execution price is compromised by a front-run. Here
the analogy both illuminates and reaches its limit.
It illuminates because the diagnosis is identical,
namely order-dependent value together with
pre-settlement visibility. It reaches its limit
because the extractable quantity is not a dollar
figure but a bias in arm assignment, unmeasurable
to the precision the on-chain case affords, and
because the remedy is not an auction that reassigns
the tax but a regulatory prohibition on exposing
the queue at all. The vaccine cold-chain provenance
system, the book's secondary example, admits the
same reading whenever the sequence of pending
custody transfers is visible before it is recorded.

We can tabulate the venues this chapter has
touched. @tbl-ordering-venues compares who holds
the ordering right, how visible the pending queue
is before it settles, and what mechanism
constrains the power in each.

| Venue | Who orders | Visibility before settlement | Constraint mechanism |
|---|---|---|---|
| **Public mempool** | Block producer or builder | Full and public | Market: PBS sealed-bid auction |
| **Private order flow** | The builder that receives it | Hidden from searchers, seen by the builder | Trust in the intermediary, plus PBS |
| **Transplant list** | Allocation policy and the procurement organization | Restricted, not publicly pending | Regulation and after-the-fact audit |
| **Vaccine scheduler** | The scheduling administrator | Restricted, not publicly pending | Regulation and after-the-fact audit |

: Ordering-power venues compared by visibility and constraint {#tbl-ordering-venues}

::: {.callout-warning}
## Public-health connection: order integrity and information leakage

Consider a workflow in which sensitive operations
queue before they settle: pending laboratory orders
visible to a party who also trades on the implied
diagnosis, a formulary system in which the sequence
of pending procurement requests leaks demand before
contracts are set, or the trial ledger's enrollment
pipeline in which the order of pending allocations
is observable to someone with an interest in the
arm assignment.
Each is a front-running surface in the exact sense
of this chapter: order-dependent value plus
pre-settlement visibility. The lesson transfers
cleanly. Making a queue of sensitive operations
public before commitment is an information-integrity
hazard, and the analyst should treat 'the pending
list is visible' as a finding, not a convenience.

The trial ledger's enrollment pipeline is worth
dwelling on, because the hazard it imports is one the
methodological literature named and measured decades
ago under a different word. What a randomized trial
protects, above almost all else, is *allocation
concealment*: the person deciding whether and when to
enroll a participant must not be able to foresee the
arm to which that participant would be assigned, for
if they can, they can enroll the favorable case and
defer the unfavorable one, and the randomization is
quietly subverted. @schulz2002concealment call this
foreknowledge 'deciphering' the sequence, and warn
that investigators will be crafty in obtaining it, so
that designers must be craftier in denying it. This
is front-running in the precise sense of this
chapter: a queue of pending assignments, visible to a
party with an interest in the order, exploited before
the assignments settle. The cost is not hypothetical.
In a study of two hundred and fifty trials,
@schulz1995empirical found that inadequate allocation
concealment exaggerated estimated treatment effects
by around forty per cent, a bias of the same
character as the sandwich tax, extracted here not in
tokens but in the credibility of a result. A public
ledger that renders the enrollment queue observable
does not merely risk a privacy breach; it reintroduces
the exact selection channel that allocation
concealment exists to close.

The lesson also cuts against naive on-chain designs
for health data. A public ledger for clinical
operations would expose precisely the pre-settlement
queue that must stay confidential, importing the
mempool's vulnerability into a setting where the
leaked information is a diagnosis rather than a
trade. The remedies that work on-chain, private
order flow and commit-reveal schemes, amount to
reintroducing a trusted intermediary or delaying
visibility, which a conventional access-controlled
database provides directly and more simply. As
@sec-intro argued, when a custodian is available and
acceptable, the custodial design is usually the
correct one, and ordering integrity is a case where
the public ledger is often the wrong tool.
:::

## Worked example: auditing MEV protection

The dominant real-world bound on a sandwich is the
victim's slippage limit. A well-formed swap
specifies a minimum acceptable output: the
transaction reverts if the received amount falls
below $(1-s)$ times what a trade against the
undisturbed pool would return, for a tolerance $s$
the user sets (a wallet default of one percent is
common). This changes the attacker's problem from
an unbounded optimization into a constrained one.
Front-run too little and value is left unextracted;
front-run enough to push the victim below its
minimum and the victim's transaction reverts,
leaving no victim to sandwich and only the
attacker's own round-trip. The profit function now
rises and then falls, and the interior optimum the
zero-fee model lacked appears, located by the
slippage limit rather than by a fee.

```{r ch08-slippage}
slip <- 0.01
fair_out <- swap_x_for_y(v, Rx, Ry)$dy
min_out <- (1 - slip) * fair_out

victim_reverts <- function(a) {
  after <- swap_x_for_y(a, Rx, Ry)
  swap_x_for_y(v, after$Rx, after$Ry)$dy < min_out
}

bounded_profit <- function(a) {
  if (victim_reverts(a)) 0 else sandwich_profit(a, v, Rx, Ry)
}

b_grid <- seq(1, 3e4, length.out = 500)
b_profit <- vapply(b_grid, bounded_profit, numeric(1))
b_star <- b_grid[which.max(b_profit)]

c(front_run = b_star,
  extracted = max(b_profit),
  tax_fraction = max(b_profit) / v)
```

```{r ch08-fig-slippage}
#| label: fig-slippage
#| fig-cap: "Slippage-bounded sandwich profit. The attacker front-runs
#|   up to the point where the victim would revert, then profit
#|   collapses to zero. The dotted line marks the interior optimum
#|   set by the victim's one-percent slippage tolerance."
ggplot(tibble(a = b_grid, profit = b_profit), aes(a, profit)) +
  geom_line() +
  geom_vline(xintercept = b_star, linetype = 3) +
  labs(x = 'attacker front-run size a', y = 'attacker profit (X)',
       title = 'A slippage cap creates an interior optimum')
```

The extracted value is close to one percent of the
victim's trade, which is to say close to the
slippage the victim authorized. This is the
auditable statement the chapter has been building
toward: under a sandwich, the tax a user pays is
approximately its own slippage tolerance times its
trade size. A wallet advertising 'MEV protection'
through a tighter default slippage is making a
quantitatively correct claim, and the analyst can
state its size rather than accept its adjective. A
protection that halves the tolerance halves the
worst-case extraction.

We can display the identity directly.
@fig-slippage-tax plots the extraction tax against
the tolerance the victim authorized, and the curve
sits almost on the dashed identity line across the
usual range of wallet defaults.

```{r}
#| label: fig-slippage-tax
#| fig-cap: 'Under a binding slippage limit the extraction tax tracks
#|   the victim tolerance almost exactly, sitting just above the
#|   identity line across the common range of wallet defaults.'
#| fig-width: 6
#| fig-height: 3.6
a_search <- seq(1, 3e4, length.out = 500)
tax_frac <- function(s) {
  floor_out <- (1 - s) * swap_x_for_y(v, Rx, Ry)$dy
  prof <- vapply(a_search, function(a) {
    after <- swap_x_for_y(a, Rx, Ry)
    got <- swap_x_for_y(v, after$Rx, after$Ry)$dy
    if (got < floor_out) 0 else sandwich_profit(a, v, Rx, Ry)
  }, numeric(1))
  max(prof) / v
}
tol_grid <- seq(0.0025, 0.03, length.out = 24)
tax_df <- tibble(tol = tol_grid,
                 tax = vapply(tol_grid, tax_frac, numeric(1)))

ggplot(tax_df, aes(tol * 100, tax * 100)) +
  geom_abline(slope = 1, intercept = 0, linetype = 2,
              colour = '#9a9a92') +
  geom_line(colour = PAL[1], linewidth = 0.7) +
  annotate('text', x = 2.55, y = 2.05,
           label = 'identity: tax = tolerance', colour = '#9a9a92',
           size = 3, angle = 32) +
  labs(x = 'victim slippage tolerance (percent)',
       y = 'extraction tax (percent of trade)') +
  book_theme
```

How much does the tax depend on how deep the pool
is? The popular intuition holds that deeper
liquidity protects users, so it is worth checking
against the slippage-bounded model.

```{r ch08-depth}
extract_max <- function(depth) {
  fair <- swap_x_for_y(v, depth, depth)$dy
  floor_out <- (1 - slip) * fair
  prof <- vapply(b_grid, function(a) {
    after <- swap_x_for_y(a, depth, depth)
    got <- swap_x_for_y(v, after$Rx, after$Ry)$dy
    if (got < floor_out) 0 else sandwich_profit(a, v, depth, depth)
  }, numeric(1))
  max(prof)
}

tibble(depth = c(2.5e5, 5e5, 1e6, 2e6, 4e6)) |>
  mutate(max_extract = vapply(depth, extract_max, numeric(1)))
```

The extraction is nearly flat across a sixteen-fold
range of pool depth. This is the correction the
model supplies to the intuition: once the slippage
limit binds, it, and not the pool depth, sets the
extractable value. Deeper liquidity lets the victim
trade with less price impact for a given tolerance,
but the attacker is still permitted to consume the
whole tolerance, so the tax tracks the slippage
allowance almost independently of depth. The
protection lever the user actually controls is the
slippage setting, not the pool it trades against,
and a protection claim should be audited on that
basis. (All pool figures here are SYNTHETIC,
constructed to expose the mechanism rather than to
match any deployed pool.)

## Collaborating with an LLM

An assistant is useful for surveying the MEV
landscape and drafting extraction models, and
hazardous wherever a plausible narrative can pass
for a computed fact. Three exchanges recur.

**Prompt.** 'Write R to compute the optimal
front-run for a sandwich attack on a
constant-product pool and show that profit is
concave in the front-run size.'

*Watch for.* The assistant will very likely produce
code and prose asserting an interior optimum,
because that claim is repeated throughout the
secondary literature it was trained on. As this
chapter showed, the claim is false for the zero-fee
model, where profit is monotone.

*Verification.* Run the grid over a range wide
enough to reveal the shape, as in @fig-sandwich,
and inspect whether the argmax is interior or on the
boundary. Add the cost that actually bounds the
problem (a fee or a slippage limit) before believing
any concavity claim, and confirm the optimum moves
to the interior only once that cost is present.

**Prompt.** 'Summarize how proposer-builder
separation protects users from MEV.'

*Watch for.* Conflation of redistribution with
elimination, the tier-slippage demoted in the
skeptic's reflex above. The assistant may describe
PBS as reducing MEV when it only reassigns who
collects it.

*Verification.* Ask specifically whether the
extractable value, defined as a function of the
pending trade and the pool depth, changes under
PBS. It does not, and forcing the definition into
the answer exposes the redistribution.

**Prompt.** 'Estimate the annual MEV extracted on a
given chain.'

*Watch for.* A confident aggregate with no stated
method. MEV totals circulate as round numbers whose
measurement (which strategies, which pools, gross or
net of costs) is rarely specified.

*Verification.* Demand the estimand and the method
before the number. Treat any figure without a
defined extraction model and a data source as a
tier-four narrative, not a tier-two measurement.

## Conclusion

In conclusion, three points are to be emphasized.
First, MEV is not a mysterious exploit but an
optimization on a known curve: given the victim
trade, the pool reserves, and the fee, the
extractable value is the maximum of an explicit
profit function, and writing that function down
converts folklore into a computation, one that
here contradicted the received account of the
sandwich's profit shape by showing it monotone at
zero fee and interior-optimal only once a per-leg
cost or a slippage bound binds. Second, the
extractable value is a structural tax rather than a
bug, levied by the conjunction of a public pending
queue and a producer free to order it; the
institutional responses we examined,
proposer-builder separation and private order flow,
redistribute or relocate the tax but do not abolish
it, so that a protection claim should be audited by
computing the residual it leaves rather than by
accepting its adjective. Third, the structure is
not confined to cryptocurrency: wherever a scarce
resource is allocated by a queue whose order is
valuable and whose contents are visible before they
settle, from the trial ledger to a transplant
waiting list, ordering power and advance visibility
together constitute a tax. The cryptocurrency case
is distinguished chiefly by its measurability, and
it is that measurability we have used to make
precise a lesson that elsewhere must be enforced by
governance rather than computed to the dollar.

## Exercises

1. **Marginal price and the no-arbitrage
   condition.** For a zero-fee constant-product
   pool with reserves $R_x, R_y$, derive the
   marginal price of $X$ in $Y$ as the limit of the
   average price of an infinitesimal trade, and show
   it equals $R_y/R_x$. Then verify that the
   arbitrage trade
   $\Delta x^\star = \sqrt{R_x R_y / p} - R_x$
   restores the marginal price to an external price
   $p$, and state the condition on $p$ under which
   the trade is a sale of $X$ rather than a
   purchase.

2. **The extraction ceiling.** Prove that the
   zero-fee sandwich profit `sandwich_profit(a, v,
   Rx, Ry)` is monotonically increasing in $a$ and
   bounded above by $v$. Confirm numerically that
   the bound is approached as $a$ grows, and explain
   why introducing a per-leg fee turns the monotone
   function into a concave one with an interior
   maximum.

3. **Fee-bounded optimum.** Modify the swap
   primitives to charge a fee, so that an input
   $\Delta x$ contributes $\gamma \Delta x$ to the
   product with $\gamma = 0.997$ while the full
   $\Delta x$ is added to the reserve. Grid-search
   the resulting sandwich profit and locate the
   interior optimum. Compare its position and the
   extracted value to the slippage-bounded optimum
   of the worked example, and comment on which
   constraint binds first for a realistic pool.

4. **Slippage as a control variable.** Using the
   `bounded_profit` construction, compute the
   maximum extractable value as a function of the
   victim's slippage tolerance $s$ over a grid from
   $0.001$ to $0.05$. Plot extraction against $s$
   and characterize the relationship. What does the
   result imply for a wallet choosing a default
   tolerance, and what does the user trade away by
   tightening it?

5. **Priority-gas auction.** Model two searchers
   competing for a single opportunity worth $\Pi$
   as an ascending auction in the priority fee.
   Argue that in equilibrium the winning fee
   approaches $\Pi$, so that the producer captures
   nearly all of $\Pi$ and the searchers earn nearly
   zero. State the assumptions your argument
   requires, and discuss how sealed-bid
   proposer-builder separation changes the split
   between searcher, builder, and proposer.

6. **The reorg threshold.** Let a block carry
   ordinary reward $\rho$ and let a recent block
   contain extractable value $M$. Write the
   condition under which a rational producer prefers
   to re-mine the recent block rather than extend
   the chain, ignoring for simplicity the
   probability of losing the reorg race. Relate the
   threshold to the security analysis of
   @sec-security, and explain why a rising ratio
   $M/\rho$ erodes the effective cost of rewriting
   history.

## Further reading

@daian2020flash is the primary source for this
chapter and should be read first. It introduced the
vocabulary of the field (front-running, priority
gas auctions, and the coupling of extractable value
to consensus stability) and its argument that MEV
threatens the layer beneath it remains the reference
framing. Read it for the mechanism, and hold its
empirical magnitudes to the tier-two standard the
book applies to all measured claims.

@adams2020uniswapv2, the Uniswap v2 protocol
specification, sets out the constant-product pool
on which every computation in this chapter runs,
including the fee accounting that Exercise 3
asks the reader to implement. It is the ground
truth for the curve, and reading it removes any
ambiguity
about how a swap alters the reserves, which is the
whole basis of the extraction model.

@eip1559 defines the fee market whose priority tip
is the ordering auction of this chapter. Understand
the base-fee-and-tip split to see why fee-driven
reordering survived the reform, and to reason about
where MEV revenue accrues under proposer-builder
separation.

@narayanan2016bitcoin remains the reference text for
the underlying mechanics of transaction propagation
and block assembly, and its treatment of the mempool
and of miner incentives is the foundation on which
the MEV supply chain is built. Read its consensus
chapters alongside @sec-security to place the
time-bandit argument in context.
