# Proof-of-Stake, Finality, and Restaking {#sec-pos}

::: {.bookepigraph}
```{=latex}
\begin{flushright}\small
```
*We present 'Ouroboros', the first blockchain protocol
based on proof of stake with rigorous security
guarantees.*

[Kiayias, Russell, David, and Oliynykov]{.smallcaps},
*Ouroboros: A Provably Secure Proof-of-Stake Blockchain
Protocol* (2017)
```{=latex}
\end{flushright}
```
:::

## Learning objectives

By the end of this chapter the reader should be
able to:

- Explain how proof-of-stake replaces the burned
  energy of proof-of-work with slashable capital
  as the source of Sybil resistance, and describe
  the roles of validators, attestations, and
  proposals.
- State the accountable-safety guarantee of Casper
  FFG and its realization in Gasper, and relate it
  to the provably-secure Ouroboros family.
- Name the behaviors that slashing punishes
  (equivocation and surround voting) and explain
  why the penalty must destroy stake rather than
  merely forfeit reward.
- Read a cost-of-attack calculation as a statement
  about capital at risk, and contrast the recurring
  rental cost of a proof-of-work attack with the
  self-inflicted capital loss of a proof-of-stake
  attack.
- Recognize leader and committee selection as
  weighted sampling, and bound the representativeness
  of a sampled committee with a concentration
  argument.
- Model restaking as a correlated-loss problem with
  a Gaussian copula, and quantify how correlation,
  at fixed marginal risk, inflates the probability
  of a large joint loss.

## Orientation

We return once more to the trial ledger. Several
hospitals maintain a shared, append-only record of
enrollment, randomization, and outcome events, with
no single institution trusted to hold the master
copy, and a regulator who audits the record only
after the fact. In @sec-pow we learned to secure
such a ledger by proof-of-work: to append a block a
participant must burn energy, and to rewrite
finalized history an attacker must out-burn the
honest network (@sec-security). The mechanism is
sound, but it is extravagant, and for a consortium
of hospitals the prospect of defending a
patient-safety record by incinerating electricity
has little intuitive appeal. The matter may be put
as a question. What secures a shared ledger once the
burning of energy is set aside?

Proof-of-stake offers an answer by changing the
collateral. Weight in the protocol is no longer
proportional to the energy a participant burns but
to the capital the participant locks in a bond, and
misbehavior is punished not by wasted electricity
but by destroying part of that bond. The scarce
resource is no longer electricity but capital that
the protocol can confiscate. This single
substitution reorganizes the entire security
argument, and we shall find it best read not as a
change of technology but as a change of collateral.

Before proceeding we fix the vocabulary on which the
rest of the chapter depends, so that the reader need
not carry these terms as mysteries.

::: {.callout-note}
## The vocabulary of this chapter

Several terms recur throughout the chapter. We define
them here and use them consistently thereafter.

- **Proof-of-stake.** A consensus protocol in which a
  participant's influence is proportional to the
  capital it locks in a bond, rather than to the
  computation it performs, and in which misbehavior is
  punished by destroying part of that bonded capital.
- **Validator.** A participant that locks a quantity of
  the chain's native asset in a bond and in return
  takes part in proposing and voting on blocks. Its
  influence is the share of the total staked capital
  that its bond represents.
- **Stake and slashing.** The *stake* is the bonded
  capital a validator places at risk. *Slashing* is the
  protocol's destruction of part of that bond as the
  penalty for a provable violation, so that misbehavior
  costs a validator its own capital rather than a
  forgone reward.
- **Accountable (economic) finality.** A guarantee that
  once a block is finalized, reverting it forces a
  large and identifiable set of validators to have
  provably broken a rule, so that reversal carries a
  price denominated in slashed stake. It is contrasted
  with the merely *probabilistic* finality of
  proof-of-work, under which a reversal grows unlikely
  with depth but is never ruled out.
- **Finality gadget.** A protocol component layered on a
  chain that periodically marks certain blocks as
  finalized, thereby converting the probabilistic
  guarantee into the accountable one.
- **Sortition.** The random selection of validators for
  a consensus duty, drawn in proportion to stake. It is
  a weighted lottery, and in statistical terms it is
  sampling with probability proportional to size.
- **Committee.** The random subset of validators that
  sortition selects to attest to a given block. The
  committee stands in for the whole validator set so
  that not every validator need vote on every block.
- **Restaking and rehypothecation.** *Restaking* reuses
  capital already bonded to secure the base chain to
  secure additional services at the same time.
  *Rehypothecation* is the general term for pledging one
  asset as collateral against several obligations at
  once; restaking is its application to staked capital.
:::

The reader already owns the tools to analyze it. The
claim that 'proof-of-stake is secure' is a
cost-of-attack inequality, and cost of attack is
capital at risk, a quantity a statistician prices
routinely. The claim that a small committee may
stand in for the whole validator set is a statement
about a weighted sample and its concentration around
a population quantity, which is to say a statement of
exactly the kind the reader met in survey-sampling
coursework. The claim that restaking earns extra
yield 'for free' is a claim about a joint tail
probability, and whether it holds depends entirely on
a correlation the marketing does not mention. Each of
these is a translation into a language the reader
speaks fluently, and once translated, each becomes
checkable.

Two protocol facts anchor what follows. First,
finality in the Casper sense (Casper being Ethereum's
finality gadget, treated in detail below) is
*accountable*: should
the chain ever finalize two conflicting histories, a
large and identifiable fraction of the staked capital
has provably broken a rule and can be slashed.
Reverting finalized history therefore carries a price
floor denominated in stake, and we shall let that
price arrive, as a number, through the cost-of-attack
comparison below. Second, restaking rehypothecates
one bond to secure many services at once, which is
leverage, and leverage together with a shared shock
is a correlated-liquidation problem. The
statistician's edge, here as in survival analysis, is
to price the tail that the leverage narrative
ignores.

## The statistician's contribution

Three judgments in this chapter are automated by no
client software and by no audit tool. They are the
work, and each is a translation of a protocol slogan
into a statistical object the reader can interrogate.

**1. Cost of attack is capital at risk, and the
structure of the risk matters more than the headline
number.** A proof-of-work attacker rents hash power
for an attack window and recovers most of the outlay
afterwards, so the cost is an operating expense. A
proof-of-stake attacker must post a bond that
slashing can destroy, so the cost is the attacker's
own capital placed inside the blast radius. The two
figures may be comparable in dollars while being
categorically different in kind, and naming that
difference, rather than quoting a dollar figure, is
the analyst's contribution.

**2. Sortition is sampling, so its guarantees are
concentration inequalities, not promises.** A
committee drawn in proportion to stake is a weighted
sample of the validator set, selected with
probability proportional to size in exactly the sense
the reader met in the design of complex health
surveys. That the committee 'represents' the whole
set is precisely the statement that a sample mean
concentrates around a population mean, and the rate
of concentration is a Hoeffding or Chernoff bound the
reader can write down. The concentration argument
that makes a committee trustworthy is the same
argument that makes a well-designed survey sample
trustworthy, and to treat the guarantee as anything
softer than a tail bound is a category error.

**3. Restaking is correlated risk, and correlation is
the whole story.** At a fixed marginal slashing
probability, whether restaking is prudent or reckless
depends on the correlation between slashing events
across services. Independent risks aggregate near
their mean; correlated risks develop a fat joint
tail. This is the shared-frailty object of survival
analysis and the correlated-censoring object of the
multi-site trial, and it is, at bottom, the
reinsurer's problem: rehypothecating one pool of
capital across many liabilities is structurally the
problem of writing many correlated policies against
one reserve. A statistician who has priced those
tails is well placed to price this one, which the
yield-focused literature systematically underprices.

## From burned energy to slashable capital

In proof-of-work, an identity's influence is its
share $q$ of the global hash rate, and the cost of
acquiring influence is the cost of the corresponding
hardware and electricity. The defense against the
Sybil attack, in which one adversary fabricates many
identities, was broached by @douceur2002sybil, who
named and formalized the attack:
fabricating identities does not multiply influence,
because a thousand node identities backed by one
machine still command only that machine's hash share.
Influence is anchored to a physical, external, burned
resource.

Proof-of-stake anchors influence to an internal
resource instead. A participant becomes a *validator*
by locking a quantity of the chain's own native asset
in a bond, and influence is that bond's share of the
total staked capital. Sybil resistance is then
immediate and by construction: splitting one bond
across a thousand validator identities does not change
its total stake share, so identities are free but
weight is not. The defining move, we note, is that the
scarce resource, capital, is now denominated in the
system's own unit of account and is *confiscable by
the protocol*. That confiscability is what makes the
security argument work, and it is what proof-of-work
cannot offer, since no protocol can reach into the
physical world and burn an attacker's electricity
after the fact.

### Validators, attestations, and proposals

The consensus duties of a validator set reduce to
two primitives.

- A *proposal* is the block a designated validator
  (the *proposer* or leader for a slot) puts forward
  to extend the chain. Time is divided into slots,
  and each slot has one proposer selected by the
  protocol's sortition rule.
- An *attestation* is a signed vote by a validator
  in favor of a particular block as the head of the
  chain and in favor of a particular *checkpoint*
  as a finalization target. Attestations are the raw
  material of both the fork-choice rule (the rule that
  selects the chain's head among competing branches)
  and the finality gadget.

Weight everywhere is stake, not head count. A
block's support, a checkpoint's support, and the
threshold for finality are all measured as a
fraction of total staked capital. A validator with
twice the bond casts an attestation worth twice as
much. This is the single fact that lets us carry the
reader's intuition about weighted samples and
capital-weighted losses directly into the protocol:
everywhere the protocol says 'two thirds of
validators', read 'two thirds of stake'.

The duties resolve into a lifecycle for the bond
itself. @fig-validator-lifecycle traces that bond
from deposit, through the repeated per-slot duty of
proposing and attesting, to one of two exits: a
voluntary withdrawal, or the involuntary destruction
of the bond by slashing should the validator
equivocate.

```{r}
#| label: fig-validator-lifecycle
#| fig-cap: 'A validator bond moves from deposit, through the repeated per-slot duty of proposing and attesting, to one of two exits: a voluntary withdrawal, or the involuntary destruction of the bond by slashing when the validator equivocates.'
#| fig-width: 7
#| fig-height: 3.4
library(ggplot2)
library(tibble)

PAL <- c('#2a78d6', '#1baf7a', '#eda100', '#008300',
         '#4a3aa7', '#e34948', '#e87ba4', '#eb6834')
book_theme <- ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(
    panel.grid.minor = ggplot2::element_blank(),
    panel.grid.major = ggplot2::element_line(
      linewidth = 0.25, colour = 'grey88'),
    axis.title = ggplot2::element_text(size = 10),
    legend.position = 'bottom'
  )

nodes <- tibble::tribble(
  ~id,        ~x, ~y, ~label,                        ~fill,
  'deposit',  1,  3,  'Deposit\nstake',              '#33356b',
  'active',   3,  3,  'Active:\npropose / attest',   '#2a78d6',
  'exit',     5,  3,  'Exit\nqueue',                 '#33356b',
  'withdraw', 7,  3,  'Withdraw\nstake',             '#33356b',
  'slashed',  3,  1,  'Slashed:\nbond destroyed',    '#e34948'
)
edges <- tibble::tribble(
  ~x,    ~y,   ~xend, ~yend,
  1.75,  3,    2.25,  3,
  3.75,  3,    4.25,  3,
  5.75,  3,    6.25,  3,
  3,     2.55, 3,     1.45
)
ggplot() +
  geom_tile(data = nodes, aes(x, y, fill = fill),
            width = 1.5, height = 0.9, colour = NA) +
  geom_text(data = nodes, aes(x, y, label = label),
            colour = 'white', size = 3.1, lineheight = 0.95) +
  geom_segment(data = edges,
               aes(x = x, y = y, xend = xend, yend = yend),
               arrow = arrow(length = unit(0.16, 'cm'),
                             type = 'closed'),
               colour = '#33356b', linewidth = 0.5) +
  geom_curve(aes(x = 2.55, y = 3.45, xend = 3.45, yend = 3.45),
             curvature = -1.1,
             arrow = arrow(length = unit(0.14, 'cm'),
                           type = 'closed'),
             colour = '#2a78d6', linewidth = 0.4) +
  annotate('text', x = 3, y = 4.05, label = 'each slot',
           colour = '#2a78d6', size = 2.8) +
  annotate('text', x = 3.35, y = 2, label = 'equivocation',
           colour = '#e34948', size = 2.8, hjust = 0) +
  scale_fill_identity() +
  coord_equal(xlim = c(0, 8), ylim = c(0.3, 4.3)) +
  theme_void()
```

::: {.callout-note icon=false collapse=true}
## Check your understanding: identities are free but weight is not

In proof-of-work, why does running a thousand node
identities on one machine not increase an attacker's
consensus weight, and what is the exact analogue in
proof-of-stake?

In proof-of-work, consensus weight is the fraction
of global hash rate a participant controls. A
thousand identities backed by one machine share that
machine's single hash rate, so the total weight is
unchanged; identities are costless but weight is
bought only with hashing. In proof-of-stake the
analogue is exact with 'stake' substituted for 'hash
rate': a thousand validator identities splitting one
bond share that bond's stake, so total weight equals
the total stake regardless of how many identities it
is spread across. In both systems the Sybil defense
is that weight is tied to an external scarce resource
that fabricating identities does not multiply.
:::

## Finality gadgets: Casper, Gasper, and Ouroboros

Nakamoto consensus offers only *probabilistic*
finality: a block $z$ deep can still be reverted, but
the probability of a successful reorganization decays
geometrically in $z$ (@sec-security). A *finality
gadget* replaces this asymptotic guarantee with a
sharp economic one. After a block is finalized,
reverting it is not merely unlikely; it is provably
expensive, with a price denominated in slashable
stake.

### Casper FFG and accountable safety

Casper the Friendly Finality Gadget, proposed by
Buterin and Griffith [@buterin2017casper], operates
on *checkpoints*, periodic blocks (one per epoch, an
epoch being a fixed span of consecutive slots)
that validators vote to justify and then finalize. A validator's vote is a
*link* from a source checkpoint $s$ to a target
checkpoint $t$, written $s \to t$. A checkpoint is
*justified* when links from more than two thirds of
total stake point to it from a justified source, and
a justified checkpoint is *finalized* when its direct
child is justified in the immediately following
epoch.

The two-thirds threshold is not arbitrary. It is the
smallest supermajority for which two conflicting
finalizations force a large, identifiable set of
validators to have broken a rule. Casper imposes two
slashing conditions on every validator:

1. **No double vote (no equivocation).** A validator
   must not publish two distinct attestations for the
   same target epoch. Formally, no two links
   $s_1 \to t$ and $s_2 \to t$ with $s_1 \ne s_2$.
2. **No surround vote.** A validator must not publish
   a link $s_1 \to t_1$ that strictly surrounds
   another of its links $s_2 \to t_2$, meaning
   $s_1 < s_2$ and $t_2 < t_1$.

From these two conditions follows the central result.

::: {.callout-tip}
## Accountable safety (informal)

If two conflicting checkpoints are both finalized,
then validators controlling at least one third of
the total stake have provably violated a slashing
condition, and the violation is a publicly verifiable
piece of evidence (two of their own signed,
contradictory attestations).
:::

The proof is a counting argument the reader can
reconstruct. Each of the two conflicting
finalizations required a two-thirds supermajority.
Two sets each holding more than $2/3$ of the stake
must overlap in more than $1/3$ of the stake, and the
overlapping validators, having voted for both
histories, must have either double-voted or
surround-voted. Their contradictory signatures are
the evidence. This is *economic* or *accountable*
finality: safety is not unconditional, but any
violation of it burns at least a third of all staked
capital and leaves a signed confession. Reverting
finalized history therefore has a hard price floor,
and that floor is the $0.34$ coefficient that will
reappear in the cost-of-attack calculation below.

### Gasper: fork choice plus finality

A finality gadget must be married to a fork-choice
rule that decides the chain head between
finalizations. Gasper, set out by Buterin and
colleagues [@buterin2020gasper], is that marriage: it
combines a stake-weighted variant of the GHOST
fork-choice rule (LMD-GHOST, which weights subtrees by
validators' latest messages) with Casper FFG finality.
GHOST descends from the proposal of Sompolinsky and
Zohar [@sompolinsky2015ghost]; Gasper's contribution
is to prove that the composite protocol is safe and
live in a partially synchronous network under an
honest-majority-of-stake assumption.
It is the consensus core of Ethereum after its
transition to proof-of-stake. The division of labor
is worth naming: LMD-GHOST chooses a head every slot
so the chain makes progress, while Casper FFG
periodically stamps checkpoints as finalized so that
progress becomes irreversible at a known cost.

### Ouroboros: provable security from the ground up

Where Casper and Gasper add finality to an existing
chain, Kiayias and colleagues designed the Ouroboros
family [@kiayias2017ouroboros] from the outset as a
proof-of-stake protocol with a formal security
reduction. Its leader election is a stake-weighted
lottery seeded by a jointly-generated randomness
beacon, and its principal theorem shows that the
protocol realizes a robust transaction ledger (with
persistence and liveness) as long as a majority of
stake is controlled by honest parties, under a
rigorously specified network and adversary model. The
significance for a quantitative reader is the style of
the guarantee: it is a theorem with stated
assumptions, in the tradition of the 'backbone'
analyses that model a blockchain as a formal protocol
and prove its properties, due to Garay, Kiayias and
Leonardos [@garay2015backbone] and Pass, Seeman and
Shelat [@pass2017asynchronous], not an appeal to
observed behavior. When assessing any proof-of-stake
system,
the first question is which of these two regimes it
lives in, a provable-security reduction with explicit
assumptions, or an economic argument whose assumptions
must be surfaced and checked.

::: {.callout-warning}
## 'Proof-of-stake is more secure' needs a qualifier

The comparison is not scalar. Proof-of-stake offers
accountable finality, which proof-of-work does not,
and it removes a large recurring energy cost. It also
introduces failure modes proof-of-work lacks:
stake can be highly concentrated (a security-relevant
Gini, revisited in @sec-onchain), long-range and
weak-subjectivity attacks (in which an adversary
rewrites history from far in the past, exploiting that
staked capital, unlike burned energy, leaves no
external trace) require out-of-band checkpoints to
defend, and the same capital can be levered across
services through restaking. 'More secure' is only
meaningful once one names the threat model, the
assumption, and the cost of violating it.
Demote the unqualified claim to the economic tier and
ask for its assumptions.
:::

## The cost of attack: capital at risk

The accountable-safety result gives a lower bound on
the cost of reverting finalized history: at least one
third of total stake is slashed. We can now make the
proof-of-work versus proof-of-stake contrast concrete.
The following helper is reused verbatim from the
verified companion; the numbers are illustrative but
the *structure* is the lesson.

```{r ch09-cost}
pow_attack_cost <- function(network_hashrate_ths, hours,
                            rent_per_ths_hour) {
  0.51 * network_hashrate_ths * hours * rent_per_ths_hour
}
pos_attack_cost <- function(total_staked_usd, slash_fraction = 1) {
  0.34 * total_staked_usd * slash_fraction
}

c(pow = pow_attack_cost(4e8, 6, 0.02),
  pos = pos_attack_cost(1e11, 1))
```

Read the two functions as models, not as quotes. The
proof-of-work cost is a *flow*: rent $51\%$ of the
hash rate for the duration of the attack window, an
operating expense the attacker recovers most of once
the rented hardware is returned. The proof-of-stake
cost is a *stock*: acquire roughly a third of the
staked capital and then lose it to slashing when the
equivocation is detected. The coefficient $0.34$ is
not decoration; it is the one-third overlap from the
accountable-safety proof. The categorical difference
is the point. In proof-of-work the attacker rents a
weapon; in proof-of-stake the attacker's own capital
*is* the weapon, and firing it destroys it. This is
the real content of the slogan 'security is capital at
risk', and it is a statement a statistician should
insist on stating precisely rather than accepting as
folklore.

@tbl-pow-pos sets the two mechanisms side by side
along the dimensions a consortium actually weighs:
what is scarce, whether the cost is a flow or a
stock, how much of it the attacker recovers, the
character of finality, and the monitoring analogue a
hospital consortium would deploy in place of a
posted bond.

: Proof-of-work and proof-of-stake compared along the dimensions a consortium weighs. {#tbl-pow-pos}

| Dimension | Proof-of-work | Proof-of-stake |
|-----------|---------------|----------------|
| **Scarce resource** | External hash power: energy and hardware | Internal bonded capital, in the chain's own unit |
| **Cost structure** | Flow: rent hash power for the attack window | Stock: post a bond the protocol can destroy |
| **Attack cost recovery** | Mostly recovered; the rented hardware is returned | Self-inflicted; the bond is slashed and gone |
| **Finality character** | Probabilistic; reversal risk decays with depth | Accountable; conflicting finality slashes a third of stake |
| **Energy** | Large and recurring | Negligible |
| **Consortium analogue** | No natural counterpart | Contractual and reputational penalty on a named party |

To illustrate, consider the trial ledger operated, as
some proposals would have it, not as a permissioned
consortium but as a public proof-of-stake chain, so
that any party may validate by bonding the chain's
asset. Suppose the total staked capital securing the
ledger is one hundred million dollars. An adversary who
wishes to revert a finalized randomization record, so
as to reassign a patient's arm after an adverse event
has been observed, must, by the accountable-safety
bound, place at least a third of that capital inside the
blast radius and lose it to slashing once the
equivocation is detected. The cost of tampering with a
single finalized entry is therefore on the order of
thirty-four million dollars of the attacker's own
capital, which is what `pos_attack_cost(1e8)` records.
We should say plainly, however, where the argument
ends. The trial ledger of @sec-intro is a consortium of
named, contractually accountable hospitals, and for
such a body the analogue of slashing is not the
destruction of a posted bond but a contractual and
reputational penalty, which is proof-of-stake's logic
without its machinery. Where a regulator can compel
honest behavior by contract, a thirty-four-million-dollar
bond is an expensive way to purchase what a signed
audit obligation already provides.

::: {.callout-important}
## The skeptic's reflex: 'security is capital at risk' as a tier-three claim

'Capital at risk' is an *economic* claim (tier
three), true only under stated assumptions: that
slashing evidence is actually detected and included,
that the attacker cannot exit before the penalty
lands, that the staked asset does not collapse in
value the instant an attack succeeds (which would let
an attacker who has shorted the asset profit despite
the slash), and that stake is not so concentrated
that one entity already holds the threshold. Each
assumption is a place where the headline dollar figure
can be wrong by an order of magnitude. Name the tier,
then interrogate the assumptions before quoting the
number.
:::

## Committee sampling as weighted sortition

A validator set may number in the hundreds of
thousands. Having every validator attest to every
block does not scale, so protocols delegate each slot
to a *committee*, a random subset drawn in proportion
to stake. This is sortition, and sortition is weighted
sampling of a kind the reader has already met.
Selecting a validator with probability proportional to
its stake is precisely probability-proportional-to-size
sampling, the design a survey statistician reaches for
when the sampling units are hospitals of unequal
census or clusters of unequal size. The protocol is,
in effect, drawing a PPS sample of the validator
population every slot, and the security of the scheme
rests on the representativeness of that sample. But how
faithfully can a sample of a few hundred seats stand in
for a population of hundreds of thousands? The question
is a concentration question, and it has a textbook
answer.

Before we write the bound down it helps to fix the
correspondence in one place. @tbl-sortition-pps sets
the survey-sampling vocabulary the reader already
owns beside its proof-of-stake counterpart, with the
caveat that keeps each row honest.

: The committee-sortition and probability-proportional-to-size correspondence, with the caveat attached to each row. {#tbl-sortition-pps}

| Concept | Survey sampling | Proof-of-stake | Caveat |
|---------|-----------------|----------------|--------|
| **Population units** | Sites or clusters of unequal size | Validators of unequal bond | Validators may collude; survey units do not |
| **Size measure** | Enrollment or census count | Staked capital | Stake concentrates far more than a census |
| **PPS draw** | Sites drawn proportional to size | Committee seats drawn proportional to stake | The committee is redrawn every slot |
| **Estimator reliability** | Sample mean concentrates around the population | Honest committee share concentrates (Hoeffding) | The bound assumes independent seats and a fixed honest fraction |

Suppose a fraction $\pi$ of total stake is honest.
Draw a committee of $m$ seats, each seat filled
independently in proportion to stake. Let
$\bar{X}_m$ be the fraction of committee seats held
by honest stake. Then $\mathbb{E}[\bar{X}_m] = \pi$,
and because the seats are independent Bernoulli($\pi$)
draws, Hoeffding's inequality gives

$$
\Pr\bigl(|\bar{X}_m - \pi| \ge t\bigr)
  \le 2\exp(-2 m t^2).
$$

The probability that a sampled committee
misrepresents the honest stake share by more than $t$
decays exponentially in the committee size $m$. A
committee of a few hundred is, with overwhelming
probability, within a few percentage points of the
population honest fraction. The following short
simulation makes the bound tangible on a SYNTHETIC
validator set with heavy-tailed stakes.

```{r ch09-sortition}
library(dplyr)
library(tibble)

set.seed(23)
n_validators <- 5000
# SYNTHETIC: heterogeneous stakes, heavy-tailed as on-chain.
stake <- rlnorm(n_validators, meanlog = 0, sdlog = 1)
stake <- stake / sum(stake)

# Label a random set of validators 'honest' holding ~80% of stake.
target_honest <- 0.80
ord <- sample(n_validators)
cum <- cumsum(stake[ord])
is_honest <- logical(n_validators)
is_honest[ord[cum <= target_honest]] <- TRUE
pi_honest <- sum(stake[is_honest])

committee_size <- 400
draw_committee <- function() {
  seats <- sample(n_validators, committee_size,
                  replace = TRUE, prob = stake)
  mean(is_honest[seats])
}

reps <- replicate(3000, draw_committee())
hoeffding_bound <- 2 * exp(-2 * committee_size * 0.10^2)
tibble(
  pop_honest_share = pi_honest,
  mean_committee_honest = mean(reps),
  sd_committee_honest = sd(reps),
  p_committee_below_two_thirds = mean(reps < 2 / 3),
  hoeffding_bound_at_t_0.10 = hoeffding_bound
)
```

The committee's honest share concentrates tightly
around the population share $\pi \approx 0.8$, with a
standard deviation close to the binomial value
$\sqrt{\pi(1-\pi)/m}$, and the empirical probability
that a committee falls below the two-thirds honesty
threshold is negligible, comfortably under the
Hoeffding bound at $t = 0.10$. Committee-based
consensus is safe precisely because a weighted sample
of a few hundred seats is an excellent estimator of
the population stake distribution. This is the same
assurance that licenses a well-designed survey: a
regulator who audits a probability-proportional-to-size
sample of the trial ledger's sites, drawn with
probability proportional to enrollment, trusts the audit
for exactly the reason the protocol trusts the
committee, namely that a weighted sample of adequate
size concentrates around the population from which it
is drawn. When a protocol claims that a small committee
'represents' the validator set, it is asserting a
concentration inequality, and the reader can check the
rate.

The rate is worth seeing rather than only asserting.
@fig-committee-concentration draws the sampling
distribution of the honest seat share for two
committee sizes on a SYNTHETIC validator set. The
larger committee concentrates visibly harder, and the
mass below the two-thirds threshold, the region in
which consensus is unsafe, shrinks sharply as the
committee grows.

```{r}
#| label: fig-committee-concentration
#| fig-cap: 'A larger committee is a tighter estimator of the honest stake share, so the chance of drawing a committee below the two-thirds safety threshold falls sharply as the size grows from 128 to 512 seats. The honest fraction is held near seventy-two percent, close to the threshold, to keep the shrinking unsafe tail visible; under a realistic honest majority it is far smaller.'
#| fig-width: 6
#| fig-height: 3.6
set.seed(512)
n_val <- 5000
# SYNTHETIC: heavy-tailed stakes; honest fraction set near threshold.
stake_c <- rlnorm(n_val, meanlog = 0, sdlog = 1)
stake_c <- stake_c / sum(stake_c)
pi_h <- 0.72
ord_c <- sample(n_val)
cum_c <- cumsum(stake_c[ord_c])
is_h <- logical(n_val)
is_h[ord_c[cum_c <= pi_h]] <- TRUE

draw_share <- function(m) {
  mean(is_h[sample(n_val, m, replace = TRUE, prob = stake_c)])
}
conc <- tibble::tibble(
  share = c(replicate(4000, draw_share(128)),
            replicate(4000, draw_share(512))),
  committee = rep(c('m = 128', 'm = 512'), each = 4000)
)
ggplot(conc, aes(share, colour = committee, fill = committee)) +
  geom_density(alpha = 0.14, linewidth = 0.7) +
  geom_vline(xintercept = 2 / 3, linetype = 'dashed',
             colour = '#9a9a92', linewidth = 0.4) +
  annotate('text', x = 2 / 3 - 0.006, y = 15,
           label = '2/3 threshold', colour = '#6f6f68',
           size = 2.9, hjust = 1, vjust = -0.3, angle = 90) +
  scale_colour_manual(values = PAL[1:2]) +
  scale_fill_manual(values = PAL[1:2]) +
  labs(x = 'honest fraction of committee seats',
       y = 'density', colour = NULL, fill = NULL) +
  book_theme
```

::: {.callout-note icon=false collapse=true}
## Check your understanding: why does committee size, not validator-set size, drive the guarantee?

The Hoeffding bound
$2\exp(-2 m t^2)$ depends on the committee size $m$
but not on the total number of validators $N$. Why is
the representativeness of a sampled committee governed
by $m$ alone, and what does this imply for scaling the
validator set?

Each committee seat is an independent draw from the
population stake distribution, so the committee's
honest share is an average of $m$ independent
Bernoulli($\pi$) variables whose concentration rate is
set by $m$. The population size $N$ affects the
granularity of the distribution being sampled but not
the sampling error of an $m$-seat draw (with
replacement, or with negligible finite-population
correction when $N \gg m$). The practical implication
is favorable: a protocol can admit an arbitrarily
large validator set for decentralization and Sybil
resistance while keeping per-slot communication fixed,
because a committee of a few hundred already delivers
exponentially small misrepresentation probability
regardless of how many validators exist in total.
:::

### Metastable consensus: Avalanche as repeated sampling

We include the present subsection on the strength
of its statistical fit rather than its deployment
prominence, and the reader should know as much at
the outset. The committee protocols of the
preceding section, whether the stake-weighted
lottery of Ouroboros [@kiayias2017ouroboros] or the
cryptographic sortition of Algorand, the
proof-of-stake protocol of Gilad and colleagues
[@gilad2017algorand] that draws its committee by a
verifiable random function, draw one representative
sample and let it vote. The Avalanche family,
proposed by Rocket and colleagues
[@rocket2019avalanche], reaches agreement by a
different device entirely, and it is a device a
sampling statistician will find familiar. But how
does a node decide what the network prefers without
a committee and without a tally? It samples,
repeatedly. Each node polls a small random sample
of its peers for their current preference, adopts
the majority answer of that sample, and repeats,
accumulating confidence over many such rounds until
it commits. There is no leader and no global vote;
there is only a node, a small poll, and a rule for
updating on what the poll returns.

The statistical content is worth making explicit,
and it is worth being precise about which urn model
applies, since the obvious one is the wrong one. The
process is a reinforcement scheme in which each
round's majority makes that majority more likely to
be drawn in the next, so an initial lead is amplified
rather than averaged away. It is tempting to call
this a Polya urn, but the classical linear Polya urn
does the opposite of what we need: its proportion is
a bounded martingale that converges to a random,
strictly interior limit, and it never locks to zero
or one. The amplification to a near-unanimous
absorbing state (a configuration the process cannot
leave once it has entered it) is instead the
behavior of a
nonlinear, super-linear reinforcement urn, or
equivalently of a biased majority (voter) dynamic,
in which feedback stronger than linear creates a
tipping point and two absorbing states rather than a
continuum of random limits. Equivalently, and more
familiarly to an epidemiologist, it is a contagion
process on the peer graph, in which a preference
spreads from node to node by contact and a majority
past a threshold sweeps the network. A
biostatistician will recognize the super-linear
reinforcement and an epidemiologist the contagion
framing at once. The contrast with the preceding
section is clean: both are sampling designs, but
sortition draws a single committee meant to
represent the whole, whereas Avalanche iterates many
small polls to drive a reinforcement process to a
metastable, near-unanimous absorbing state (a state
that, once reached, persists indefinitely absent a
large perturbation) with high probability. @fig-avalanche-metastable shows the
reinforcement at work on a SYNTHETIC peer network,
with several trajectories started from different
initial fractions.

```{r}
#| label: fig-avalanche-metastable
#| fig-cap: 'SYNTHETIC. Several repeated-subsampling trajectories of the network fraction preferring one option, over polling rounds, on a small peer network. Preferences starting below the one-half tipping fraction die out toward zero, while those above it sweep to near-unanimity, the two metastable absorbing states. The same threshold behavior is the epidemic threshold of infectious-disease models in another costume.'
#| fig-width: 6
#| fig-height: 3.6
library(ggplot2)
library(tibble)

PAL <- c('#2a78d6', '#1baf7a', '#eda100', '#008300',
         '#4a3aa7', '#e34948', '#e87ba4', '#eb6834')
book_theme <- ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(
    panel.grid.minor = ggplot2::element_blank(),
    panel.grid.major = ggplot2::element_line(
      linewidth = 0.25, colour = 'grey88'),
    axis.title = ggplot2::element_text(size = 10),
    legend.position = 'bottom'
  )

set.seed(206)
# SYNTHETIC: repeated-subsampling reinforcement on a peer network.
n_nodes <- 800
sample_k <- 15
n_rounds <- 20
starts <- c(0.30, 0.42, 0.50, 0.58, 0.70)

run_trajectory <- function(p0) {
  pref <- as.integer(runif(n_nodes) < p0)
  frac <- numeric(n_rounds + 1)
  frac[1] <- mean(pref)
  for (r in seq_len(n_rounds)) {
    new_pref <- pref
    for (i in seq_len(n_nodes)) {
      peers <- sample.int(n_nodes, sample_k, replace = TRUE)
      new_pref[i] <- as.integer(mean(pref[peers]) > 0.5)
    }
    pref <- new_pref
    frac[r + 1] <- mean(pref)
  }
  frac
}

traj <- lapply(starts, run_trajectory)
df <- do.call(rbind, lapply(seq_along(starts), function(j) {
  tibble::tibble(
    round = 0:n_rounds,
    frac = traj[[j]],
    start = sprintf('p0 = %.2f', starts[j])
  )
}))

ggplot(df, aes(round, frac, colour = start)) +
  geom_hline(yintercept = 0.5, linetype = 'dashed',
             colour = '#9a9a92', linewidth = 0.4) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 1.0) +
  scale_colour_manual(values = PAL[seq_along(starts)]) +
  labs(x = 'polling round',
       y = 'network fraction preferring option 1',
       colour = NULL) +
  book_theme
```

We should be even-handed about what this device
buys and what it costs. On the plus side, repeated
subsampling gives fast, leaderless agreement with
high probability, and it does so without the
communication burden of collecting every validator's
vote. On the minus side, the guarantee is
probabilistic rather than classically final: the
process settles into its absorbing state with high
probability, not with the accountable certainty of a
slashing bound, and the whole security argument
rests on sampling assumptions, an honest
supermajority among the polled and an unbiased draw
of peers, that merit exactly the scrutiny this
chapter has applied elsewhere. A biased sample or a
dishonest supermajority breaks the amplification in
the adversary's favor, and neither assumption
verifies itself.

The contagion framing is not a loose metaphor. The
mathematics of preference spread here is the
mathematics of epidemic spread the reader may know
from infectious-disease modeling, and the threshold
behavior is the same in both: a preference held
below a tipping fraction dies out, while one above it
sweeps the network, which is the epidemic threshold
in another costume, below the critical reproduction
number an outbreak fizzles and above it takes off.
The analogy has a limit worth marking, however. In a
natural epidemic the transmission rule is fixed by
biology and the modeller only observes it, whereas in
Avalanche the designer chooses the sampling rule, the
sample size, and the decision threshold, and so can
engineer the tipping fraction and the speed of
convergence rather than merely estimate them.

::: {.callout-note icon=false collapse=true}
## Check your understanding: why does an initial majority sweep the network?

In Avalanche's repeated subsampling each node adopts
the majority preference of a small random poll every
round. If a preference already holds a modest
majority of the network, why does repeated polling
drive it toward near-unanimity rather than back
toward the fifty-fifty split, and what feature of the
dynamics makes the two extremes absorbing?

The poll is drawn from the current population, so a
preference held by a majority is more likely than not
to be the majority of any given sample, and each node
that adopts it enlarges the majority faced by the
next. This is super-linear reinforcement, the
majority-amplifying feedback of a biased voter
dynamic rather than the mean-reverting draw of a
classical linear urn: success breeds success, and the
lead compounds rather than reverting to the mean. The
fifty-fifty split is
an unstable equilibrium, since any chance departure
from it is amplified, while near-unanimity is stable,
because a sample from a near-unanimous population
almost surely returns the same answer and no node
changes its mind. The two extremes are therefore
absorbing, and which one the process reaches is
decided by which side of the tipping fraction the
initial preference lies on, exactly as an epidemic's
fate is decided by whether transmission sits above or
below its threshold.
:::

## Worked example: restaking as correlated liquidation

Restaking, set out in the whitepaper of EigenLayer,
the protocol that introduced the practice to Ethereum
[@eigenlayer2023], lets capital already bonded to
secure the base chain be *rehypothecated* to secure
additional services, called actively validated
services (AVSs). Each additional service carries its
own slashing conditions, so the same bond is now
exposed to slashing from several sources at once.
Marketed as extra yield on idle-looking collateral, it
is, in the language the reader already owns,
*leverage*: one unit of capital backing several
liabilities. It is also, and more usefully for us, the
reinsurer's situation exactly, one reserve written
against many policies that may or may not fail
together, and leverage together with a shared shock is
a correlated-liquidation problem.

The shape of the exposure is worth drawing.
@fig-restaking-rehypothecation places one bonded pool
at the center, rehypothecated outward to secure
several services at once, and a single correlated
shock that can strike all of them together. The
arrows outward are the leverage; the red arc is the
dependence the yield calculation never mentions.

```{r}
#| label: fig-restaking-rehypothecation
#| fig-cap: 'Restaking rehypothecates one bonded pool to secure several services at once, so a single correlated shock can slash the same capital from several directions together. It is the situation of a reinsurer: one reserve written against many liabilities that may fail together.'
#| fig-width: 6.2
#| fig-height: 4
pool <- tibble::tibble(x = 0, y = 0, label = 'One\nstaked\nbond')
svc <- tibble::tibble(
  x = c(2.4, 2.4, 2.4),
  y = c(1.4, 0, -1.4),
  label = c('Service A', 'Service B', 'Service C')
)
spokes <- tibble::tibble(
  x = 0.55, y = 0,
  xend = svc$x - 0.55, yend = svc$y
)
ggplot() +
  geom_segment(data = spokes,
               aes(x = x, y = y, xend = xend, yend = yend),
               arrow = arrow(length = unit(0.16, 'cm'),
                             type = 'closed'),
               colour = '#33356b', linewidth = 0.5) +
  annotate('curve', x = 3.6, y = 2.1, xend = 3.6, yend = -2.1,
           curvature = -0.9, colour = '#e34948', linewidth = 0.7,
           arrow = arrow(length = unit(0.16, 'cm'),
                         ends = 'both', type = 'closed')) +
  annotate('text', x = 4.35, y = 0, label = 'correlated\nshock',
           colour = '#e34948', size = 2.9, lineheight = 0.95) +
  geom_point(data = svc, aes(x, y), size = 22, shape = 21,
             fill = '#2a78d6', colour = '#2a78d6') +
  geom_text(data = svc, aes(x, y, label = label),
            colour = 'white', size = 3.0) +
  geom_point(data = pool, aes(x, y), size = 40, shape = 21,
             fill = '#33356b', colour = '#33356b') +
  geom_text(data = pool, aes(x, y, label = label),
            colour = 'white', size = 3.2, lineheight = 0.95) +
  coord_equal(xlim = c(-1.1, 5.2), ylim = c(-2.4, 2.4)) +
  theme_void()
```

The systemic risk is not the marginal slashing
probability of any one service; we may fix that at a
modest level and set it aside. The risk lives in the
*joint* tail, the probability that many services slash
the same bond in the same event. Whether that tail is
thin or fat is governed entirely by the correlation
between slashing events, a parameter the yield
calculation omits. We model it with a Gaussian copula,
the same instrument by which a reinsurer binds a set of
marginal failure probabilities to a dependence
structure, drawing correlated latent normals via a
Cholesky factorization (and so avoiding a `MASS`
dependency), and we read a service as slashing when its
latent variable falls below the threshold implied by
its marginal slashing probability. The code is reused
verbatim from the verified companion.

```{r ch09-restaking}
library(tibble)

# Cholesky-based correlated normal draws, so no MASS dependency.
rmvn <- function(n, Sigma) {
  matrix(rnorm(n * ncol(Sigma)), n) %*% chol(Sigma)
}

set.seed(9)
simulate_restaking_loss <- function(n_services = 10, rho = 0.0,
                                    p_slash = 0.05, sims = 5000) {
  Sigma <- matrix(rho, n_services, n_services)
  diag(Sigma) <- 1
  z <- rmvn(sims, Sigma)
  thresh <- qnorm(p_slash)
  slashed <- z < thresh
  rowSums(slashed) / n_services
}

loss_indep <- simulate_restaking_loss(rho = 0.0)
loss_corr  <- simulate_restaking_loss(rho = 0.7)
tibble(
  scenario = c('independent', 'correlated (rho=0.7)'),
  mean_loss = c(mean(loss_indep), mean(loss_corr)),
  p_loss_over_30pct = c(mean(loss_indep > 0.3), mean(loss_corr > 0.3))
)
```

The two scenarios are calibrated to the *same*
marginal slashing probability and therefore the same
mean loss. They differ only in correlation, and that
difference is the entire story of the tail. Under
independence a joint loss exceeding thirty percent of
the bond is vanishingly rare, because it requires many
independent low-probability events to coincide. At
correlation $\rho = 0.7$ the same large loss becomes
orders of magnitude more likely, because a single
common shock can trip many services together. Two
portfolios with identical expected loss can have
tail-loss probabilities that differ by an order of
magnitude; correlation, not the marginal, is what
prices the disaster. This is precisely what every
reinsurer knows and what the yield brochure omits: at
a fixed mean, correlation moves risk into the tail.

The point is worth seeing as a curve rather than as
two scenarios. @fig-restaking-rho traces the
probability of a joint loss exceeding thirty percent of
the bond as the cross-service correlation is swept from
zero to nine tenths, with the marginal slashing
probability held fixed throughout.

```{r}
#| label: fig-restaking-rho
#| fig-cap: "SYNTHETIC. Probability that restaking's joint slashing loss exceeds thirty percent of the bond, as the cross-service slashing correlation rises, at a fixed marginal slashing probability of 0.05. Because the marginal is held fixed, the mean loss is unchanged along the curve; only the tail moves. A multi-site trial exhibits the same shape when a shared frailty couples site-level dropout: the marginal dropout rate is unmoved, but the variance of any aggregate estimate thickens with correlation."
library(ggplot2)

set.seed(94)
rho_grid <- seq(0, 0.9, by = 0.05)
p_tail <- vapply(rho_grid, function(r) {
  mean(simulate_restaking_loss(rho = r, sims = 8000) > 0.3)
}, numeric(1))
tail_curve <- tibble::tibble(rho = rho_grid, p_tail = p_tail)

ggplot(tail_curve, aes(rho, p_tail)) +
  geom_line(linewidth = 0.8, colour = '#2c3e50') +
  geom_point(size = 1.6, colour = '#2c3e50') +
  labs(x = 'cross-service slashing correlation (rho)',
       y = 'P(joint loss > 30% of bond)') +
  theme_minimal(base_size = 12)
```

::: {.callout-important}
## The skeptic's reflex: 'restaking is free extra yield'

The phrase 'extra yield with no additional risk'
delivers a *narrative* (tier four) in the cadence of a
protocol fact. Restaking adds risk; it merely moves
that risk into the joint tail, where a mean-and-yield
summary cannot see it. The correct restatement is a
tier-three claim with an explicit parameter: the extra
yield compensates for a correlated slashing exposure
whose severity is governed by the cross-service
correlation $\rho$, and the yield is fairly priced only
if it exceeds the expected tail loss at the *true*
$\rho$, which is almost never measured. Demote 'free
yield' to 'a short position in a correlated tail', and
ask who is estimating the correlation.
:::

### The public-health connection

The correlated-loss object is not blockchain-specific.
It is the same statistical structure we confront as
*correlated censoring* or *dropout* in a clinical
trial, and as *shared-frailty* failure in a survival
model. In a multi-site trial, an analysis that treats
participant dropout as independent across sites will
badly understate the variance of any aggregate estimate
when a common cause, a protocol amendment, a supply
interruption, a regional event, makes many participants
leave together. The failure mode is identical to
restaking's: a summary calibrated on the marginals is
blind to a tail generated by a shared shock. A
shared-frailty model, in which a latent common factor
couples individual hazards, is structurally the same
device as the Gaussian copula above, with the frailty
playing the role of the common latent normal, and both
are the reinsurer's device deployed for the same reason.
The correspondence is not merely rhetorical.
@clayton1978bivariate introduced association into
bivariate survival precisely by letting a shared latent
term act multiplicatively on otherwise independent
hazards, which is the copula construction we used above
written in the language of failure times, and
@hougaard1995frailty gives the general treatment,
including the point most relevant here: the choice of
mixing distribution governs *where in the tail* the
dependence concentrates, so that two models agreeing on
the marginal and the average correlation can still
disagree sharply on the extreme joint loss. That is the
restaking problem exactly. Fixing the marginal slashing
probability and the nominal correlation does not fix the
tail, and the yield-focused literature, which rarely
looks past the marginal, is systematically blind to the
object a frailty model is built to expose.

The transferable lesson, we would suggest, is that a
statistician trained to distrust independence
assumptions in survival and longitudinal data is
precisely the person who should price restaking's
systemic risk. The tail that the yield-focused
literature ignores is the tail that correlated-dropout
methodology exists to quantify. Where a restaked
service secures on-chain records on which a health
system relies, the provenance of a vaccine cold chain,
say, or the enrollment log of the trial ledger, the
correlated-liquidation risk of the securing capital
becomes a data-integrity risk for the health
application, and the same correlation analysis governs
both.

We should be candid, finally, about the negative result
of @sec-intro. If a single custodian can be trusted to
keep the record, none of this machinery is needed, and
a database with an audit log is the correct tool. For
the trial ledger, which is a consortium of named and
contractually accountable hospitals rather than an open
set of anonymous bonders, the analogue of slashing is
not the destruction of a posted bond but a contractual
and reputational penalty. That is proof-of-stake's
logic, the deterrence of misbehavior by a credible
loss, without proof-of-stake's machinery, and where a
regulator can impose the penalty by contract it is the
cheaper instrument.

## Collaborating with an LLM

An assistant is useful for drafting protocol summaries
and simulation scaffolding, but the load-bearing
judgments here are exactly the ones an assistant will
paper over. We offer three prompt-and-verification
triples.

**Prompt.** 'Explain why reverting a finalized
Ethereum block costs at least one third of total
stake.'

- *Watch for*: a fluent restatement that omits the
  overlap argument, or that conflates 'one third
  slashed' with 'one third needed to attack liveness'.
  The safety bound (one-third overlap of two
  two-thirds supermajorities) and the liveness bound
  are different quantities.
- *Verification*: the reader should reconstruct the
  counting argument independently. Two sets each
  exceeding $2/3$ of stake overlap in more than $1/3$;
  confirm the assistant's
  explanation reduces to this and cites the slashing
  conditions [@buterin2017casper].

**Prompt.** 'Write R to show that restaking increases
risk as correlation rises.'

- *Watch for*: code that changes the marginal slashing
  probability along with the correlation, which
  confounds the two effects and makes the demonstration
  meaningless. Also watch for a multivariate-normal
  call pulling in an unnecessary package dependency.
- *Verification*: confirm the mean loss is held fixed
  across correlation levels (as in `ch09-restaking`),
  so that only the tail moves. Check that the copula is
  correctly specified: the threshold must be
  $\Phi^{-1}(p_\text{slash})$ for the target marginal.

**Prompt.** 'Is a small validator committee safe as a
stand-in for the whole set?'

- *Watch for*: an appeal to 'randomness' or 'large
  numbers' with no rate. Safety here is a concentration
  bound, and a claim without a bound is not an answer.
- *Verification*: demand the Hoeffding or Chernoff
  inequality and its dependence on committee size $m$,
  and sanity-check it against a simulation like
  `ch09-sortition`.

## In conclusion

In conclusion, three points are to be emphasized.
First, the security of a proof-of-stake ledger is a
cost-of-attack inequality and nothing more mysterious:
the cost is capital placed at risk of slashing, and the
accountable-safety bound fixes its floor at roughly one
third of the total stake, a figure the analyst should
state precisely, with its assumptions, rather than
borrow as folklore. Second, committee sortition is
probability-proportional-to-size sampling, so a
committee's fidelity to the validator set is a
concentration inequality of exactly the kind that
licenses a well-designed survey sample; the governing
quantity is the committee size, not the size of the
validator set, and the guarantee should be quoted as a
tail bound and never as a promise. Third, restaking is
leverage, and its danger lives entirely in the joint
tail: at a fixed marginal slashing probability,
correlation moves the risk into the tail exactly as it
does for a reinsurer, or for a multi-site trial with a
shared frailty, and the statistician trained to
distrust independence assumptions is the right person
to price it. For the trial ledger itself we would add,
in the spirit of @sec-intro, that the honest conclusion
is frequently that a permissioned consortium bound by
contract achieves the same integrity as a bonded chain
at a fraction of the cost, and the cost-of-attack
calculation developed here is the instrument by which
that judgment is made.

## Exercises

1. **Reconstruct accountable safety.** Prove that if
   two conflicting checkpoints are each finalized under
   Casper FFG's two-thirds rule, then validators
   holding at least one third of total stake have
   violated a slashing condition. State precisely where
   each of the two slashing conditions (no double vote,
   no surround vote) is used.

2. **Committee concentration (quantitative).** For an
   honest stake fraction $\pi = 0.7$, use Hoeffding's
   inequality to find the smallest committee size $m$
   guaranteeing that the probability a committee falls
   below the two-thirds honesty threshold is at most
   $10^{-6}$. Then verify your $m$ by simulation,
   adapting `ch09-sortition`, and comment on the gap
   between the bound and the empirical probability.

3. **Cost-of-attack sensitivity (quantitative).**
   Extend `ch09-cost` so that the proof-of-stake
   attacker can recover a fraction $\gamma$ of the
   staked asset's value by shorting it before the
   attack, and plot net attack cost against $\gamma$.
   At what $\gamma$ does the proof-of-stake attack
   become cheaper than the proof-of-work attack under
   the chapter's illustrative parameters? Interpret the
   result as a statement about which assumption in the
   skeptic's-reflex callout is doing the work.

4. **Correlated-loss tail (quantitative).** Using
   `simulate_restaking_loss`, plot the probability of a
   joint loss exceeding thirty percent as a function of
   correlation $\rho \in [0, 0.9]$, holding the marginal
   slashing probability fixed at $0.05$. Fit or describe
   the shape of the curve and identify the correlation
   at which the tail probability first exceeds one
   percent. Relate the curve to the shared-frailty
   analogy.

5. **Provable versus economic security.** Contrast the
   security guarantee of Ouroboros [@kiayias2017ouroboros]
   with the economic cost-of-attack argument for a
   Casper-finalized chain. For each, state the threat
   model, the assumption, and what an adversary must
   control to break it. Which guarantee is falsifiable
   by which kind of evidence?

6. **When restaking is the wrong tool.** Identify a
   public-health data-integrity problem for which a
   restaked service could in principle provide
   trust-minimized verification, then argue whether the
   added correlated-liquidation risk is justified
   relative to a governed database with an audit log.
   Make the argument quantitative where you can.

## Further reading

- @buterin2017casper is the primary source for Casper
  FFG. Read it for the two slashing conditions and the
  accountable-safety theorem; the counting argument
  behind the one-third bound is worth reconstructing on
  paper.
- @buterin2020gasper specifies Gasper, the composition
  of LMD-GHOST fork choice with Casper finality, and
  states its safety and liveness properties in a
  partially synchronous model. It is the consensus core
  of proof-of-stake Ethereum.
- @kiayias2017ouroboros is the foundational paper of the
  provably-secure proof-of-stake family. Read it for the
  style of guarantee: a formal reduction to
  honest-majority-of-stake under an explicit network and
  adversary model, in the tradition of
  @garay2015backbone.
- @eigenlayer2023 is the restaking whitepaper. Read it
  for the mechanism (rehypothecation of stake to secure
  AVSs) and read it *adversarially* for the correlation
  assumption its risk discussion leaves implicit.
- @narayanan2016bitcoin remains the reference textbook
  for the surrounding consensus and security material,
  and its treatment of cost-of-attack reasoning is the
  baseline this chapter extends from proof-of-work to
  proof-of-stake.
