# On-Chain Data Analysis and Tokenomics {#sec-onchain}

## Learning objectives

By the end of this chapter the reader should be
able to:

- Frame the analysis of a public ledger as an
  observational study, and name the selection,
  confounding, and measurement threats that
  attach to it.
- Treat address clustering as an entity-
  resolution problem, apply the common-input-
  ownership heuristic, and state why
  pseudonymity is not anonymity.
- Quantify wealth concentration with a Lorenz
  curve and a Gini coefficient, and explain why
  concentration is security-relevant under
  proof-of-stake.
- Fit and honestly test a block-interval model,
  and locate where the exponential assumption is
  expected to fail on real data.
- Read a token's economics through market
  capitalisation versus fully diluted valuation,
  build an unlock schedule, and give an
  operational test for 'Ponzinomics'.
- Keep a scored forecasting log, compute a Brier
  score, and read a calibration curve, closing
  the falsifiability programme of @sec-intro.

## Orientation

We open, and shall in due course close the book,
with a single auditing question. Given an
append-only data store, whether a public
blockchain or a health registry that no custodian
may quietly overwrite, what can we honestly infer
from it, and how do we keep our own claims honest
while we do so? The question is older than
blockchains. It is the question an epidemiologist
asks of any record assembled without a controlled
design, and this closing chapter is, in the end,
an epidemiology of a ledger.

The preceding chapters treated blockchains as
mechanisms: a consensus protocol to be analysed,
a cryptographic primitive to be measured, a
market to be modelled. This chapter treats a
blockchain as something else, a dataset. Every
confirmed transaction on a public chain is a
permanent, world-readable record, and the
aggregate is one of the largest openly available
behavioural datasets in existence. For a
statistician this is the payoff of everything
that came before: the guarantees established in
@sec-consensus and @sec-crypto are exactly what
make the data trustworthy enough to analyse.

It will help to fix a concrete public-health
instance of the same auditing problem, and we
shall carry it through the chapter. Consider
several hospitals that agree to share one
append-only record of a multi-centre clinical
trial. Each enrolment, each randomisation, and
each outcome event is written to a ledger that no
single institution owns and none may silently
revise, and a regulator audits the record after
the fact. We shall refer to this arrangement
throughout as the trial ledger. A vaccine
cold-chain provenance system, in which each
custody transfer of a refrigerated lot is
committed to a shared log, is a second instance we
shall reach for where it illuminates a point. Both
are append-only stores that a distrustful party
must audit after the fact, which is precisely the
situation the on-chain analyst faces.

The central discipline of the chapter is to
remember what kind of dataset it is. On-chain
data is observational, not experimental. Nobody
randomised addresses into treatment arms. The
population of participants is self-selected, the
measured quantities are proxies for the latent
quantities of interest, and the generating
process is adversarial, for some of the entities
in the data are actively trying to be
misclassified. Every caution a biostatistician
brings to an electronic health record or an
insurance claims database applies here, and a few
more besides.

The reader already owns the tools, under other
names. Wealth concentration is a Lorenz curve and
a Gini coefficient (@kondor2014rich, an empirical
study of the concentration of Bitcoin holdings),
the very instruments of health-inequity
measurement. Block arrivals are a renewal process
(a sequence of independent waiting times between
successive events) whose fit can be tested
(@sec-mining). De-anonymisation is entity
resolution, the record-linkage problem of registry
deduplication (@meiklejohn2013fistful, the study
of Meiklejohn and colleagues that clustered the
Bitcoin address graph into economic actors). Token
release is a deterministic schedule whose future
supply is a piece of cohort bookkeeping,
computable today. The work of the chapter is to
apply these familiar instruments to an unfamiliar
substrate, and to report the results with the
honesty that observational data demands. We take
up each in turn: the observational-study framing
in the next section, then clustering,
concentration, the block-interval model,
tokenomics, and finally the forecasting log with
which the book closes.

Several terms recur through the chapter, and we
gather the least familiar of them here so that the
reader need not hold them as mysteries while the
argument proceeds.

::: {.callout-note}
## The vocabulary of this chapter

- **On-chain data.** The permanent, world-readable
  record of confirmed transactions on a public
  blockchain, read here as an observational dataset
  rather than as a designed experiment.
- **Address clustering (entity resolution).** The
  grouping of many pseudonymous addresses into the
  single economic actor that controls them. The
  workhorse rule is the common-input-ownership
  heuristic: when one transaction spends several
  inputs together, all those input addresses are
  presumed to belong to one owner, since the spender
  had to sign for each. It is the record-linkage
  problem of a patient registry applied to a ledger.
- **Pseudonymity versus anonymity.** Pseudonymity
  hides only the map from addresses to identities
  while leaving every transaction permanently public;
  anonymity would hide the activity itself. A chain
  is pseudonymous, and pseudonymity degrades under
  analysis.
- **Mixer.** A service that pools funds from many
  users and redistributes them, deliberately breaking
  the input-output linkage on which clustering
  depends.
- **Market capitalisation versus fully diluted
  valuation (FDV).** Market capitalisation is the
  circulating supply times price; FDV is the total
  eventual supply times price. The gap between them is
  the value of the supply not yet released.
- **Total value locked (TVL).** The token value
  deposited in a protocol, denominated in the very
  tokens whose prices it aggregates, and therefore not
  a direct measure of adoption or of safety.
- **Vesting and unlock schedule.** The deterministic
  timetable, fixed in a token's contract, by which
  allocated but locked tokens (to a team, to early
  investors) become circulating, each unlock adding to
  the effective supply.
- **Ponzinomics.** A token structure in which the
  returns paid to existing participants are funded
  from the capital of new entrants rather than from
  any revenue exogenous to the token.
:::

Because each instrument arrives under a name the
reader already owns, we set the correspondence out
in one place before we begin. @tbl-tool-twin pairs
each on-chain tool with its public-health twin and
the caveat both inherit, so that the rest of the
chapter can be read as a single method applied to
two substrates.

| On-chain tool | Public-health twin | Shared caveat |
|---|---|---|
| **Address clustering** (common-input-ownership) | Record linkage and capture-recapture | A heuristic classifier; false and missed matches must be reported, not hidden |
| **Gini and Lorenz** concentration | Health-inequity measurement | The unit of analysis (address versus entity) sets the number before any estimator does |
| **Block-interval fit** | Survival and arrival-time diagnostics | A pooled fit can pass while the stress regimes one cares about fail |
| **Unlock schedule** | Cohort follow-up accrual | Deterministic today, but only as far as the vesting contract is honoured |
| **Brier calibration** | Epidemic-forecast scoring | A low score is not calibration; the log must be kept prospectively |

: Each on-chain instrument is a public-health instrument under another name, and both inherit the same caveat. {#tbl-tool-twin}

## The statistician's contribution

Before turning to the instruments, we should name
three judgements that are the analyst's own, for
no tool performs them for us.

1. **Reading a chain as an observational study,
   not a ledger of facts.** A transaction graph
   is a sample, not a census of economic
   behaviour, and it is a biased sample: the
   entities that transact on-chain differ
   systematically from those that do not, custodial
   addresses aggregate many users behind one
   pseudonym, and wash trading, the practice of
   trading with oneself to fabricate the appearance
   of activity, manufactures volume. Naming the
   estimand, and the gap between it and
   what the chain actually measures, is prior to any
   number.

2. **Operationalising a narrative into a test.**
   The field runs on unfalsifiable claims:
   'decentralised', 'sound money', 'sustainable
   yield'. The statistician's contribution is to
   convert each into a measurable quantity with a
   rejection region. 'Decentralised' becomes a Gini
   coefficient on stake. 'Sustainable yield' becomes
   a question about whether returns are paid from
   exogenous cash flow or from new entrants. Once a
   claim has been given a test it can be shown false,
   which is the entire purpose of formulating it.

3. **Scoring one's own forecasts.** Scepticism
   that is never recorded and never graded is merely a
   posture. The calibration log introduced in
   @sec-intro and closed here holds the analyst's own
   predictions to the same standard demanded of the
   promoters: a probability, an outcome, and a Brier
   score. A well-calibrated critic is of more value
   than a loud one.

The second judgement is the engine of the whole
chapter, so we make its machinery explicit.
@tbl-falsifiability sorts a claim into one of the
four falsifiability tiers of @sec-intro and names,
for each tier, the test that would settle it. The
same ladder governs a crypto slogan and a
public-health assertion, and the analyst's task is
to keep pushing a claim down the ladder until it
names an outcome that could arrive and prove it
wrong.

| Tier | Crypto claim | Health claim | Test that settles it |
|---|---|---|---|
| **1 Protocol fact** | The contract releases 30% of supply at month 12 | The registry recorded 400 enrolments | Read it off the ledger or registry; deterministic |
| **2 Empirical** | Active addresses rose 40% this quarter | Coverage reached 80% in the district | Measure it from records, with the error stated |
| **3 Model-dependent** | This staking yield is sustainable | This vaccine prevents 60% of cases | Fix the model and estimator; report assumptions and fit |
| **4 Narrative** | The chain is decentralised | The programme is a success | Not falsifiable until forced down to a tier-2 measurement |

: The four falsifiability tiers operationalised, with a crypto and a health claim at each and the test that would settle it. {#tbl-falsifiability}

```{r ch10-setup}
library(digest)
library(dplyr)
library(tibble)
library(tidyr)
library(purrr)
library(ggplot2)
```

```{r ch10-visuals-setup}
PAL <- c('#2a78d6', '#1baf7a', '#eda100', '#008300',
         '#4a3aa7', '#e34948', '#e87ba4', '#eb6834')

book_theme <- ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(
    panel.grid.minor = ggplot2::element_blank(),
    panel.grid.major = ggplot2::element_line(
      linewidth = 0.25, colour = 'grey88'),
    axis.title = ggplot2::element_text(size = 10),
    legend.position = 'bottom'
  )
```

## A chain is an observational dataset

We consider the estimand before the estimator.
Suppose we want the distribution of wealth among
the economic participants of a chain. What the
ledger gives us is the distribution of balances
across addresses. These are not the same object,
and the map between them is exactly the source of
every serious error in on-chain analysis. The
statistician who has audited a disease registry
will recognise the difficulty at once, for it is
the difficulty of any observational study: the
quantity we can measure is not, in general, the
quantity we should like to have measured.

An address is a public key, not a person. One
person may control thousands of addresses, and
one address, an exchange's hot wallet, that is, the
online account it uses for day-to-day settlement,
may hold the pooled balances of millions of people.
The
unit of observation the chain hands us is
therefore not the unit of analysis we want. This
is a measurement problem in the precise sense of
the term, for the observed variable is a noisy,
biased, and non-injective function of the latent
variable.

To name the hazards is to recite the epidemiologic
catechism unchanged. Three classical threats
organise the caution.

- **Selection.** The set of addresses that ever
  transact is self-selected and unrepresentative
  of any wider population. Inferences about
  'people' from a chain are inferences about a
  peculiar, wealthy, technically sophisticated,
  and partly automated subpopulation.
- **Confounding.** Apparent relationships in the
  transaction graph are routinely driven by
  unobserved common causes: a single market maker,
  a single exploit, a single exchange's internal
  rebalancing. Correlation of address activity is
  weak evidence of a relationship between distinct
  economic actors.
- **Measurement.** Balances, volumes, and active-
  address counts are proxies. Volume can be
  fabricated by wash trading at near-zero cost.
  Active-address counts are inflated by dust, that
  is, economically negligible amounts scattered
  across many addresses, and by wallet software that
  splits change across new addresses. The proxy and
  the target diverge, and
  the divergence is not random.

The methodological posture that follows is the
one a biostatistician already holds toward
registry data, and it transfers to the trial
ledger without amendment. The data is real and
valuable, the access is unusually complete, and
precisely for that reason the temptation to treat
a proxy as its target is unusually strong. On the
one hand the completeness of an append-only record
is a genuine analytic gift; on the other,
completeness is not representativeness, and a
census of addresses is still not a census of
people.

::: {.callout-note icon=false collapse=true}
## Check your understanding: estimand versus ledger

A widely cited dashboard reports that a chain
has '1.2 million daily active addresses' and
concludes that adoption is growing. Name one
selection threat and one measurement threat to
the claim that this number tracks the count of
distinct human users.

Selection: addresses that transact on-chain are
a self-selected minority; custodial users who
trade on an exchange never appear as distinct
addresses, so the measure omits an unknown and
probably large population and over-represents
automated and professional actors. Measurement:
one human can generate many addresses (wallet
software creates a fresh change address per
transaction) and one address can front millions
of humans (an exchange wallet), so the count is a
non-injective proxy for users, inflated by
address churn and deflated by custodial pooling.
The number can rise while the human population
falls.
:::

## Address clustering and de-anonymisation

Bitcoin is pseudonymous, not anonymous, and the
distinction is the entire content of
@meiklejohn2013fistful. Every transaction is
public forever, and what is hidden is only the map
from addresses to identities. Recovering part of
that map is a statistical problem the reader will
recognise immediately, for it is entity
resolution, the record-linkage task that
deduplicates a patient registry or links records
across health databases that share no common key.
The epidemiologist who has estimated the size of a
hidden population by capture-recapture, matching
individuals across incomplete lists, has already
practised the core move: we accumulate partial
identifying evidence until distinct records
resolve into one entity.

The workhorse is the common-input-ownership
heuristic. When a transaction spends several
inputs at once, the spender had to sign for all of
them, so under normal wallet behaviour all those
input addresses are controlled by one entity. Each
multi-input transaction is therefore an assertion
that its input addresses belong to the same
cluster. Accumulating these assertions is a
union-find (disjoint-set) computation: addresses
are nodes, co-spending is an edge, and the
connected components are candidate entities.

@fig-entity-resolution draws the move in the small.
Six pseudonymous addresses carry co-spending edges
that merge them into two entities, while one
exchange address, though a single node, fronts many
users behind one pseudonym and so is the measurement
error the section warns against, drawn in red.

```{r}
#| label: fig-entity-resolution
#| fig-cap: 'Entity resolution on six addresses. Common-input-ownership edges merge them into two entities (circled), while the exchange address fronts many users behind one pseudonym.'
#| fig-height: 3.4
ellipse <- function(cx, cy, rx, ry, grp, n = 80) {
  t <- seq(0, 2 * pi, length.out = n)
  tibble(x = cx + rx * cos(t), y = cy + ry * sin(t), grp = grp)
}

# SYNTHETIC layout: six addresses, three co-spending
# edges, one exchange fronting several users.
nodes <- tibble(
  id = c('a1', 'a2', 'a3', 'a4', 'a5', 'exchange'),
  x  = c(1.1, 2.1, 1.7, 4.9, 5.9, 8.4),
  y  = c(3.7, 3.9, 2.6, 3.8, 2.9, 3.2),
  kind = c('addr', 'addr', 'addr', 'addr', 'addr', 'exchange')
)
edges <- tibble(
  x    = c(1.1, 1.7, 4.9),
  y    = c(3.7, 2.6, 3.8),
  xend = c(2.1, 2.1, 5.9),
  yend = c(3.9, 3.9, 2.9)
)
hulls <- bind_rows(
  ellipse(1.6, 3.4, 1.0, 1.0, 'entity 1'),
  ellipse(5.4, 3.35, 0.85, 0.9, 'entity 2')
)
users <- tibble(x = c(9.0, 9.25, 9.0), y = c(3.9, 3.2, 2.5))

ggplot() +
  geom_path(data = hulls, aes(x, y, group = grp, colour = grp),
            linewidth = 0.7) +
  geom_segment(data = edges, aes(x, y, xend = xend, yend = yend),
               colour = '#33356b', linewidth = 0.6) +
  geom_point(data = filter(nodes, kind == 'addr'), aes(x, y),
             colour = '#33356b', size = 5) +
  geom_point(data = users, aes(x, y), colour = '#9a9a92', size = 2.4) +
  geom_segment(data = users, aes(x = 8.4, y = 3.2, xend = x, yend = y),
               colour = '#9a9a92', linewidth = 0.3) +
  geom_point(data = filter(nodes, kind == 'exchange'), aes(x, y),
             colour = '#e34948', size = 6) +
  geom_text(data = filter(nodes, kind == 'addr'), aes(x, y, label = id),
            colour = 'white', size = 3) +
  annotate('text', x = 1.6, y = 4.65, label = 'entity 1',
           colour = PAL[1], size = 3.3) +
  annotate('text', x = 5.4, y = 4.55, label = 'entity 2',
           colour = PAL[2], size = 3.3) +
  annotate('text', x = 8.4, y = 1.7,
           label = 'exchange: many users,\none pseudonym',
           colour = '#e34948', size = 3.1) +
  scale_colour_manual(values = c('entity 1' = PAL[1],
                                 'entity 2' = PAL[2])) +
  expand_limits(y = c(1.3, 4.9)) +
  coord_equal(clip = 'off') +
  guides(colour = 'none') +
  theme_void()
```

```{r ch10-clustering}
set.seed(21)
# SYNTHETIC: 800 transactions co-spending among
# 300 pseudonymous addresses. Real wallets reuse
# addresses, so co-spending links accumulate.
n_addr <- 300
n_tx <- 800
tx_inputs <- replicate(
  n_tx,
  sample(n_addr, size = sample(1:3, 1)),
  simplify = FALSE
)

parent <- seq_len(n_addr)
uf_find <- function(i) {
  while (parent[i] != i) i <- parent[i]
  i
}
uf_union <- function(a, b) {
  ra <- uf_find(a)
  rb <- uf_find(b)
  if (ra != rb) parent[rb] <<- ra
}

for (ins in tx_inputs) {
  if (length(ins) > 1L) {
    for (k in 2:length(ins)) uf_union(ins[1], ins[k])
  }
}

roots <- vapply(seq_len(n_addr), uf_find, integer(1))
c(addresses = n_addr,
  entities = length(unique(roots)),
  largest_cluster = max(tabulate(roots)))
```

We see the 300 pseudonymous addresses collapse
into far fewer entities, and a single super-cluster
emerges, the familiar giant component of a random
graph past its percolation threshold. This is
SYNTHETIC data, but the qualitative result is the
one @meiklejohn2013fistful obtained on the real
Bitcoin graph: a modest set of heuristics, applied
at scale, reduces a pseudonymous ledger to a
tractable number of economic actors, many of which
can then be tagged by their interactions with
known services. The same collapse is what a
registry statistician calls deduplication, and it
is neither more nor less than that here.

The heuristic is a classifier, and like any
classifier it makes errors. CoinJoin transactions,
which several independent users construct jointly,
deliberately violate common-input-ownership by
having many independent signers co-spend, injecting
false-positive links, and wallets that avoid
address reuse suppress true-positive links. We
should therefore report the heuristic's assumptions
and its failure modes rather than present the
clustering as ground truth, exactly as one would
report the false-match and missed-match rates of a
probabilistic record-linkage model in a registry
study.

::: {.callout-important}
## The sceptic's reflex

'The blockchain is anonymous' is a tier-four
narrative delivered in the cadence of a tier-one
protocol fact. The protocol fact is the opposite:
the ledger is permanently and globally public, and
only the identity map is hidden. Pseudonymity
degrades under analysis, and it degrades
retroactively, because the data never expires. A
transaction that is unlinkable today may be
de-anonymised by a heuristic or a subpoena years
from now. Any claim of on-chain privacy must name
the specific technique that provides it (a mixer,
a zero-knowledge scheme) and the adversary it is
proven against, or it is marketing.
:::

### Public-health relevance

The same machinery that de-anonymises a payment
ledger de-anonymises a health-data store published
under pseudonyms, and here the trial ledger
demands a word of caution. If the consortium
commits an append-only log of patient events under
per-patient pseudonyms, the linkage structure of
the events, that is, which pseudonyms co-occur in
which records, is itself identifying, exactly as
co-spending is identifying on Bitcoin. The lesson
transfers without modification: pseudonymised
health data on a public ledger is a
re-identification dataset waiting for a heuristic,
and the permanence of the ledger means the risk
only grows rather than decays. We state the
warning plainly, for it is the privacy corollary
of the whole section: pseudonymity is not
anonymity, on a chain or in a registry. This is a
decisive argument against writing patient-level
records, even pseudonymised ones, to any public
chain, and it is why the trial ledger of our
running example commits only integrity hashes and
access-controlled pointers, never the clinical
payload itself.

### The de-anonymisation arms race

The heuristics of the previous section are not the
end of the story but a move in a continuing game.
Each advance in clustering has provoked a
countermeasure, and each countermeasure has
provoked a counter-countermeasure, so that the
practical privacy of a chain is best read not as a
settled fact but as the current state of an arms
race. Meiklejohn and colleagues showed that a
modest set of heuristics could collapse a
pseudonymous ledger into a tractable set of
economic actors (@meiklejohn2013fistful), and the
response was a family of privacy technologies
engineered precisely to defeat that collapse.

Two of these responses mark out the design space.
The first is the mixing service, which pools funds
from many users and redistributes them, so that the
input-output linkage on which common-input-ownership
depends is deliberately broken: the several inputs
of a joint transaction are no longer one entity, and
the heuristic that read co-spending as co-ownership
now records a false positive. The second is the
privacy-by-design chain, which builds concealment
into the protocol rather than bolting it on
afterward, and here two contrasting constructions
merit mention. Monero hides the true spender among a
set of decoys using ring signatures (a scheme in
which a signature is valid for any member of a group,
so that the true signer cannot be singled out),
whereas Zerocash, the construction of Ben-Sasson and
colleagues that underlies the Zcash chain, conceals
the transaction graph altogether behind
zero-knowledge shielded transactions
(@bensasson2014zerocash). The
latter is the zero-knowledge machinery of @sec-crypto
put to work: a shielded transfer proves that a valid
spend occurred without revealing the sender, the
receiver, or the amount.

But does privacy-by-design settle the question in
the user's favour? It does not, and here the honest
finding is two-sided. On the one hand the clustering
literature shows that pseudonymity leaks, for the
heuristics recover far more structure than the naive
user expects. On the other hand the privacy tools
are themselves fallible. Moser and colleagues, in a
traceability study of Monero, traced a substantial
fraction of its inputs by exploiting weaknesses in
the early decoy sampling (@moser2018monero), which
is to say that a
concealment scheme is only as strong as its
implementation and its parameters. Neither anonymity
nor de-anonymisation is therefore absolute. The
practical privacy of a given user is an empirical
and moving quantity, revised with each new heuristic
and each new countermeasure, and it is emphatically
not a theorem. @tbl-arms-race sets the exchange of
moves out in one place, with the health-data
analogue of each.

| Clustering move | On-chain countermeasure | Health-data analogue |
|---|---|---|
| Common-input-ownership heuristic | A mixing service pools and redistributes funds, breaking the input-output linkage | Aggregation or k-anonymity over quasi-identifiers |
| Tagging clusters by known services | A fresh address per payment; privacy-by-design chains | Pseudonym rotation and cell suppression |
| Decoy and ring-signature analysis | Larger, better-sampled decoy sets (Monero) | Adding noise or synthetic records to a release |
| Following the transaction graph | Zero-knowledge shielded transactions (Zcash) | Formal privacy guarantees inside a governed enclave |

: Each clustering move provokes an on-chain countermeasure, and each countermeasure has a health-data analogue; the arms race is one game played on two substrates. {#tbl-arms-race}

The reader who works with health data has met this
arms race before, under another name. The
re-identification literature established long ago
that a small number of quasi-identifiers, a
postcode, a date of birth, a sex, can single out an
individual in a release published as de-identified.
A biostatistician who has absorbed that result
already understands the mixer problem without
further instruction, for it is the same problem:
pseudonymity is not anonymity, on a chain or in a
registry, and the privacy of a released dataset is a
property to be tested empirically against a
motivated adversary, not asserted by the custodian
who released it. The mixing service and the
k-anonymity transform, which coarsens a release
until every record is indistinguishable from at
least k-1 others, are two answers to one
question, namely how much identifying structure
survives an adversary who is actively trying to
recover it.

Where does the parallel end? It ends at permanence
and reach. A quasi-identifier attack on a health
release is mounted by whoever holds the release,
within a data enclave whose access can in principle
be controlled, revoked, and audited. An on-chain
pseudonym, by contrast, is permanent and
world-readable: the data never expires and the
adversary set is everyone, indefinitely. A linkage
that is infeasible today may become feasible
tomorrow with a better heuristic, or with an
auxiliary dataset that does not yet exist, and the
ledger will still be there to receive it. The arms
race is thus the same game, but played on a chain
for considerably higher stakes, which is one further
reason the trial ledger of our running example
commits only integrity proofs and keeps every
identifying datum in a store that can still be
locked.

::: {.callout-note icon=false collapse=true}
## Check your understanding: does a mixer make you anonymous?

A user runs their coins through a mixing service and
concludes that their transactions are now anonymous.
Name one reason this conclusion is premature, and
state the health-data analogue.

The mixer breaks the common-input-ownership linkage
at the moment of mixing, but it does not erase the
permanent public record on either side of the mix,
and analysis of amounts, timing, and downstream
address reuse can partially undo the concealment;
moreover the guarantee depends on a large and honest
anonymity set, which a thinly used mixer or a
compromised operator may not provide. The health-data
analogue is a de-identified release that strips
direct identifiers but leaves quasi-identifiers and
linkage structure intact: in both cases privacy is
an empirical property of the release against a
motivated adversary, to be measured rather than a
label the custodian may simply attach. On a public
chain the record is permanent, so the measurement can
only worsen with time.
:::

## Wealth concentration: Lorenz and Gini

The most-cited empirical fact about on-chain
wealth is its concentration, and the instrument is
one every reader has already used to measure
health inequity, whether across incomes, insurance
coverage, or the reach of a vaccination campaign.
The Lorenz curve plots the cumulative share of
total wealth held by the poorest fraction $p$ of
addresses, and the Gini coefficient is twice the
area between that curve and the line of perfect
equality. For a population $x_1 \le \dots \le x_n$
the Gini has the convenient closed form

$$
G = \frac{2 \sum_{i=1}^{n} i\, x_i}
         {n \sum_{i=1}^{n} x_i}
    - \frac{n+1}{n},
$$

which we implement directly rather than reaching
for a package. We test it on a heavy-tailed
balance distribution, the SYNTHETIC stand-in for
the log-normal-to-power-law shape that on-chain
balances reliably exhibit.

```{r ch10-gini}
gini_coef <- function(x) {
  x <- sort(x)
  n <- length(x)
  (2 * sum(seq_len(n) * x) / (n * sum(x))) - (n + 1) / n
}

set.seed(14)
# SYNTHETIC: heavy-tailed balances (log-normal),
# typical of on-chain balance distributions.
balances <- rlnorm(10000, meanlog = 0, sdlog = 2)
c(gini = gini_coef(balances),
  top1pct_share = sum(sort(balances, decreasing = TRUE)[1:100]) /
    sum(balances))
```

The Gini sits high and the top one percent of
addresses hold a large share of the total, the
qualitative signature @kondor2014rich documented
for Bitcoin's real balance distribution.

To illustrate that the instrument is indifferent
to its substrate, consider the same coefficient
applied not to token balances but to vaccine
coverage. Suppose an immunisation programme reports
the number of fully immunised children in each of
eight districts as 950, 910, 880, 840, 300, 120,
60 and 20. A national average would look tolerable,
yet the coefficient exposes the two districts
carrying almost no coverage.

```{r ch10-vaccine-gini}
# SYNTHETIC: fully immunised children in eight
# districts. The same gini_coef, now measuring
# health inequity rather than token wealth.
coverage <- c(950, 910, 880, 840, 300, 120, 60, 20)
c(gini = gini_coef(coverage),
  bottom2_share = sum(sort(coverage)[1:2]) / sum(coverage))
```

The Gini here carries exactly the meaning it
carries on balances, namely that nearer zero is
nearer parity and nearer one is nearer to a few
districts holding all the coverage. The instrument
does not change when the substrate changes from
stake to sera; only the estimand does. A student
who has quantified inequity in vaccination has
therefore already met every idea in this section.
To see the concentration rather than summarise it,
we draw the Lorenz curve. The few lines below sort
the balances, form the cumulative wealth share
against the cumulative population share, and
compare the result to the diagonal of perfect
equality.

```{r}
#| label: fig-lorenz
#| fig-cap: 'Lorenz curve for a SYNTHETIC heavy-tailed balance
#|   distribution. The gap to the diagonal is the Gini area.'
lorenz_curve <- function(x) {
  x <- sort(x)
  n <- length(x)
  tibble(
    pop_share = c(0, seq_len(n) / n),
    wealth_share = c(0, cumsum(x) / sum(x))
  )
}

lc <- lorenz_curve(balances)

ggplot(lc, aes(pop_share, wealth_share)) +
  geom_abline(linetype = 2) +
  geom_line(linewidth = 0.8) +
  coord_equal() +
  labs(x = 'cumulative share of addresses',
       y = 'cumulative share of wealth',
       title = 'Lorenz curve of on-chain balances (SYNTHETIC)')
```

The curve bows far below the diagonal: the poorest
ninety percent of addresses hold a negligible
fraction of the wealth, while the top sliver holds
almost all of it. For a proof-of-work chain this
is a distributional curiosity. For a proof-of-stake
chain it is a security parameter. Sybil resistance,
the property that an actor cannot gain
disproportionate influence merely by creating many
identities, under proof-of-stake (@sec-pos,
@douceur2002sybil) presumes that influence is
proportional to stake
and that stake is not trivially concentrated in a
few hands. A high stake Gini means a small number
of entities approach the thresholds that matter for
liveness and finality, and the clustering of
@meiklejohn2013fistful can make the effective
concentration worse than the address-level Gini
suggests, because addresses that look independent
may resolve to one staking entity. Concentration
measurement is thus not sociological commentary; it
is an input to the safety analysis.

::: {.callout-note icon=false collapse=true}
## Check your understanding: address Gini versus entity Gini

An analyst computes a stake Gini of 0.82 across
addresses and concludes the chain is dangerously
concentrated.
A defender replies that many of those addresses are
retail users of one large exchange, so the true
concentration is lower. Who is right, and which
direction does address clustering push the estimate?

Both effects are real and they point in opposite
directions, so the address-level Gini is neither an
upper nor a lower bound without further work. Pooling
custodial users behind one exchange address
overstates that entity's independent balance and can
inflate the Gini relative to the underlying human
distribution. Conversely, one entity that spreads
stake across many addresses to look decentralised
deflates the address Gini relative to the true
entity concentration. The correct move is to cluster
addresses into entities first (the union-find of the
previous section) and compute the Gini on entities,
reporting the clustering assumptions. For the
security question, the entity Gini is the relevant
one, because an attacker who controls an exchange
controls its pooled stake regardless of how many
addresses front it.
:::

## Block-interval modelling on data

@sec-mining derived the block interval as an
exponential random variable, the memoryless waiting
time of a Poisson process whose rate the difficulty
controller holds near a set-point. Here we treat a
series of intervals as data and ask whether the
model survives contact with it. The workflow is the
one any analyst applies to a distributional
hypothesis: we estimate the rate, run a
goodness-of-fit test, and inspect a QQ plot. It is
the same workflow an epidemiologist brings to the
spacing of events in a surveillance stream, and on
the trial ledger it is the workflow that would test
whether enrolments arrive as a steady Poisson
process or in suspicious bursts.

```{r}
#| label: fig-qq
#| fig-cap: 'QQ plot of SYNTHETIC block intervals against the
#|   fitted exponential. Points on the diagonal indicate a
#|   good fit; systematic departure is where the science is.'
set.seed(13)
# SYNTHETIC: intervals from Exp(mean 600s), the very
# model under test. Replace with a real CoinMetrics
# CSV to make this a genuine test of a real chain.
intervals <- rexp(2016, rate = 1 / 600)

fit_rate <- 1 / mean(intervals)
ks <- ks.test(intervals, 'pexp', rate = fit_rate)
c(mean_interval = mean(intervals), fitted_rate = fit_rate,
  ks_p_value = ks$p.value)

ggplot(tibble(x = sort(intervals),
              theo = qexp(ppoints(length(intervals)), fit_rate)),
       aes(theo, x)) +
  geom_point(size = 0.6) + geom_abline(linetype = 2) +
  labs(x = 'exponential quantiles', y = 'observed intervals',
       title = 'Block intervals vs fitted exponential (SYNTHETIC)')
```

The QQ points lie on the diagonal and the
Kolmogorov-Smirnov test does not reject, because we
drew from the model. That is the point of the
SYNTHETIC exercise: it establishes what a passing
result looks like, so that a failure on real data is
legible. On a real chain the interesting science is
precisely where the exponential frays. Around a
difficulty retarget, the periodic adjustment that
resets the mining difficulty toward the target
block time, the rate parameter is stale, so
intervals cluster short or long until the controller
catches up. During a hash-rate shock, the exogenous
migration of miners, the interval distribution
develops a heavy right tail that the single-rate
exponential cannot represent. A memoryless model
also cannot express the serial dependence that these
episodes induce. Reporting where the fit breaks, and
resisting the urge to declare victory because the
bulk of the distribution looks exponential, is the
deliverable. A model that fits the easy 95 percent
and fails exactly during the stress events one cares
about is worse than useless if its failure is not
disclosed.

## Tokenomics and the sceptic's rubric

Tokenomics is the study of a token's supply
schedule and the claims made about its value.
Almost all of it reduces to arithmetic that the
promotional literature is structured to obscure.
The first and most consequential distinction is
between market capitalisation and fully diluted
valuation.

Market capitalisation is the circulating supply
times price. Fully diluted valuation (FDV) is the
total eventual supply times price. When a large
fraction of the supply has not yet been released,
these two numbers diverge by an order of magnitude,
and the released supply is the numerator that keeps
the price up while the unreleased supply is a
schedule of future sell pressure. A token can look
cheap on market cap and be enormously expensive on
FDV, and the gap is not an opinion; it is on the
vesting contract.

Three headline valuations circulate, and each
misleads in its own way, so we set their definitions
against what inflates them and when a number may be
trusted. @tbl-valuation-metrics is the reference we
return to whenever a dashboard quotes one figure and
suppresses the other two.

| Metric | Definition | What inflates it | When to trust it |
|---|---|---|---|
| **Market cap** | Circulating supply times price | A low float that suppresses the circulating count | Only when paired with FDV and the unlock schedule |
| **Fully diluted valuation** | Total eventual supply times price | Pricing tokens that are not yet released at today price | As a ceiling on dilution, never as a price target |
| **Total value locked** | Token value deposited in a protocol | Token-price appreciation and value rehypothecated across protocols | As a denominated flow, not as adoption or safety |

: Market cap, fully diluted valuation, and total value locked compared: what each is, what inflates it, and when it can be trusted. {#tbl-valuation-metrics}

The exercise that follows is, in its structure,
plain cohort bookkeeping. Each allocation tranche
is a cohort with an entry time, its cliff, and a
follow-up schedule, its vesting period, over which
its members become circulating. Summing the
released fractions across cohorts to obtain the
circulating supply at each month is exactly the
accounting an epidemiologist performs to track how
many members of successive enrolment cohorts have
accrued follow-up by a given date. To make the
overhang concrete we build a synthetic allocation
with cliffs and linear vesting and plot circulating
supply over time, marking where insider tranches
unlock.

```{r}
#| label: fig-unlock
#| fig-cap: 'Circulating supply under a SYNTHETIC allocation.
#|   Dotted lines mark the investor (month 6) and team
#|   (month 12) unlock cliffs, each a step of future supply.'
months <- 0:48
vest <- function(alloc, cliff, duration, t) {
  if (t < cliff) return(0)
  frac <- min(1, (t - cliff) / duration)
  alloc * frac
}

schedule <- tibble(month = months) |>
  rowwise() |>
  mutate(
    public   = vest(20, 0, 1, month),
    team     = vest(20, 12, 36, month),
    investor = vest(30, 6, 24, month),
    treasury = vest(30, 0, 48, month)
  ) |>
  ungroup() |>
  mutate(circulating = public + team + investor + treasury)

ggplot(schedule, aes(month, circulating)) +
  geom_line(linewidth = 0.8) +
  geom_vline(xintercept = c(6, 12), linetype = 3) +
  labs(x = 'month', y = 'circulating supply (% of total)',
       title = 'Circulating supply: cliffs at months 6 and 12')
tail(schedule |> select(month, circulating), 3)
```

At launch only the public tranche and a sliver of
the linear allocations circulate, so market cap is a
small fraction of FDV. As the investor cliff at
month 6 and the team cliff at month 12 pass, supply
climbs toward its full value. A price chart that
looks healthy while a large team allocation
approaches its cliff is carrying an overhang that
market-cap-only reporting hides. The unlock schedule
is a tier-one protocol fact, computable today from
the vesting contract, and any valuation that ignores
it is a tier-four narrative.

::: {.callout-important}
## The sceptic's reflex

Two headline metrics deserve immediate demotion.
Market cap, quoted without FDV, hides the unlock
overhang and should never be read alone; the
honest pair is (market cap, FDV) with the release
schedule attached. Total value locked (TVL), the
banner statistic of decentralised finance, is a
tier-two empirical quantity masquerading as a
measure of adoption or safety. TVL is denominated
in the very tokens whose prices it aggregates, so
it inflates during a bubble by construction; it
double-counts value that is rehypothecated, that is,
pledged as collateral in one protocol and then
re-pledged in another, across protocols; and it says
nothing about whether the
locked capital can actually exit. A protocol whose
TVL doubled because its own token tripled has not
become twice as used or twice as safe. Demote both
numbers to their tier before quoting them.
:::

### Ponzinomics, operationalised

'Ponzinomics' is thrown around as an epithet, but
it names a precise and testable structure. A scheme
is Ponzi-like when the returns paid to existing
participants are funded from the capital
contributed by new entrants rather than from any
exogenous cash flow. But how do we tell a genuine
yield from a disguised transfer between cohorts?
The operational test follows directly from the
definition: we trace the source of the yield. If the protocol produces revenue from
outside the token system (trading fees paid by
users who want a service, interest paid by
borrowers with outside income) the yield has an
exogenous source and the scheme is not Ponzi by
this test. If the yield is paid in newly minted
tokens whose only demand is the expectation of
future yield, the source is endogenous and the
structure is Ponzi-like: it requires monotonically
increasing inflows to sustain the promised return,
and it fails when inflows stall.

The test is a flow accounting question, not a moral
one, and it is answerable from on-chain data. A
statistician can decompose the yield into its
funding sources and report the fraction that
originates outside the token. The 'sustainable
yield' narrative becomes a measurable ratio with a
value between zero and one, and the closer that
ratio sits to zero, the closer the structure sits to
a pyramid.

### Stablecoin fragility, revisited

The empirical counterpart to this analysis is the
stablecoin, a token engineered to hold a fixed
value against a reference such as the dollar,
studied in detail by @briola2023anatomy, an
on-chain and market-microstructure account of a
stablecoin collapse. An algorithmic stablecoin that
maintains its peg through a reflexive relationship
with a companion token, minting one to defend the
other, is a Ponzi-like structure with a fixed point
that is stable only while inflows grow. The
collapse of such a design in May 2022 was a
liquidation cascade of exactly the positive-feedback
form met in the study of perpetual funding: a small
loss of peg triggered redemptions, redemptions
minted companion tokens, the new supply drove the
companion price down, and the falling companion
undermined the very collateral meant to defend the
peg. @briola2023anatomy anatomise the on-chain and
market microstructure of the episode, and the
lesson for the analyst is that the fragility was
visible in the mechanism design before it was
visible in the price. A reflexive peg has no
exogenous anchor, and the operational Ponzi test
would have flagged it.

## Forecasting and calibration

The chapter, and the book, closes by turning the
critical lens on the critic. Scepticism is cheap
and unfalsifiable unless it commits to predictions
and grades them. The instrument, promised in
@sec-intro, is a forecasting log scored by the
Brier score, the mean squared error between a
stated probability and the realised binary outcome.
The reader who has followed the COVID-19
forecast-hub literature will recognise both the
score and the discipline at once, for scoring a log
of crypto predictions is the very exercise by which
epidemic forecasters are held to account: a
probability, an outcome, and a score, kept
prospectively so that the record cannot be edited
after the fact.

$$
\text{Brier} = \frac{1}{n} \sum_{i=1}^{n}
  (p_i - o_i)^2, \qquad o_i \in \{0, 1\}.
$$

A forecaster who always says 0.5 scores 0.25, so
any score below that is informative and the ideal
is zero. But is a low score the same as an honest
one? It is not, for a low Brier score alone is not
calibration. A forecaster is calibrated when, among
all the events assigned probability near $p$, the
long-run frequency of occurrence is $p$. Calibration
is checked by binning predictions and plotting
observed frequency against predicted probability: a
calibrated forecaster hugs the diagonal.

We exercise the machinery on a SYNTHETIC log of a
well-calibrated forecaster, drawing each outcome as
a Bernoulli variable at the stated probability, so
the calibration curve should track the diagonal up
to sampling noise. Replacing this simulated log with
a real `predictions.csv` turns the same code into an
assessment of one's own judgement.

```{r}
#| label: fig-calibration
#| fig-cap: 'Calibration curve for a SYNTHETIC well-calibrated
#|   forecaster. Points hug the diagonal; the Brier score sits
#|   well below the uninformed 0.25.'
brier_score <- function(prob, outcome) mean((prob - outcome)^2)

set.seed(7)
n <- 200
preds <- tibble(
  prob = runif(n),
  outcome = rbinom(n, 1, runif(n))  # SYNTHETIC: mildly miscalibrated
)
# Force approximate calibration for the demonstration:
preds$outcome <- rbinom(n, 1, preds$prob)

calibration_report <- function(preds, n_bins = 5) {
  overall <- brier_score(preds$prob, preds$outcome)
  binned <- preds |>
    mutate(bin = cut(prob, breaks = seq(0, 1, length.out = n_bins + 1),
                     include.lowest = TRUE)) |>
    group_by(bin) |>
    summarise(n = n(), mean_prob = mean(prob),
              observed = mean(outcome), .groups = 'drop')
  list(brier = overall, binned = binned)
}

rep_out <- calibration_report(preds)
cat('Brier score:', round(rep_out$brier, 4),
    '(0.25 = uninformed coin flip)\n')

ggplot(rep_out$binned, aes(mean_prob, observed)) +
  geom_abline(linetype = 2) +
  geom_point(aes(size = n)) +
  coord_equal(xlim = c(0, 1), ylim = c(0, 1)) +
  labs(x = 'Predicted probability', y = 'Observed frequency',
       title = 'Calibration of a well-calibrated forecaster')
```

The points hug the diagonal and the Brier score
sits well below 0.25, as expected for a forecaster
whose probabilities match reality. A real log of
crypto predictions will rarely look this clean,
which is exactly why one keeps it. The log connects
directly to the four falsifiability tiers of
@sec-intro. A tier-four narrative cannot be scored,
because it names no outcome and no date; forcing a
claim into the log is what demotes it to a
tier-two empirical statement with a probability, a
horizon, and eventually a grade. The discipline is
symmetric: it holds the analyst's scepticism to the
same standard it demands of the promoter's
enthusiasm.

A calibration curve earns its keep only by contrast,
so we place a second forecaster beside the first.
@fig-calibration-compare draws two SYNTHETIC logs on
one reliability plot: a well-calibrated forecaster
whose curve tracks the diagonal, and an overconfident
one whose curve is too flat, stating extreme
probabilities that the outcomes do not earn. The
overconfident forecaster is the one to fear, for a
strong-sounding prediction is exactly the one a
promoter rewards and a careful analyst distrusts.

```{r}
#| label: fig-calibration-compare
#| fig-cap: 'Two SYNTHETIC forecasters on one reliability plot. The well-calibrated series hugs the diagonal; the overconfident series is too flat, its extreme probabilities not earned by outcomes.'
set.seed(11)
n_cal <- 2000
mk_forecaster <- function(slope) {
  # SYNTHETIC: slope 1 is calibrated; slope < 1 pulls
  # the realised rate toward 0.5, the overconfidence.
  p <- runif(n_cal)
  true_p <- pmin(pmax(0.5 + (p - 0.5) * slope, 0), 1)
  tibble(prob = p, outcome = rbinom(n_cal, 1, true_p))
}
bin_reliability <- function(d, lab, n_bins = 8) {
  d |>
    mutate(bin = cut(prob, seq(0, 1, length.out = n_bins + 1),
                     include.lowest = TRUE)) |>
    group_by(bin) |>
    summarise(mean_prob = mean(prob), observed = mean(outcome),
              .groups = 'drop') |>
    mutate(forecaster = lab)
}
cal_compare <- bind_rows(
  bin_reliability(mk_forecaster(1.0), 'well calibrated'),
  bin_reliability(mk_forecaster(0.5), 'overconfident')
)

ggplot(cal_compare, aes(mean_prob, observed, colour = forecaster)) +
  geom_abline(linetype = 2, colour = 'grey60') +
  geom_line(linewidth = 0.7) +
  geom_point(size = 1.8) +
  scale_colour_manual(values = c('well calibrated' = PAL[1],
                                 'overconfident' = PAL[2])) +
  coord_equal(xlim = c(0, 1), ylim = c(0, 1)) +
  labs(x = 'predicted probability', y = 'observed frequency',
       colour = NULL) +
  book_theme
```

## Worked example: an audit in six numbers

We now assemble the chapter's instruments into a
single pass over one SYNTHETIC chain snapshot, the
shape an actual audit would take against a real
export. The point is that a handful of numbers,
each the output of a familiar statistical tool,
summarises the health of a chain more honestly than
any dashboard.

```{r ch10-audit}
set.seed(2026)
# SYNTHETIC snapshot: balances, block intervals, and
# a small forecast log for one chain.
audit_balances <- rlnorm(5000, meanlog = 0, sdlog = 1.9)
audit_intervals <- rexp(1000, rate = 1 / 600)
audit_preds <- tibble(
  prob = runif(50),
  outcome = rbinom(50, 1, runif(50))
)

audit <- tibble(
  metric = c('gini', 'top1pct_share', 'mean_interval_s',
             'ks_p_exponential', 'n_forecasts', 'brier'),
  value = c(
    gini_coef(audit_balances),
    sum(sort(audit_balances, decreasing = TRUE)[1:50]) /
      sum(audit_balances),
    mean(audit_intervals),
    ks.test(audit_intervals, 'pexp',
            rate = 1 / mean(audit_intervals))$p.value,
    nrow(audit_preds),
    brier_score(audit_preds$prob, audit_preds$outcome)
  )
)
audit |> mutate(value = round(value, 4))
```

Read the report honestly. The Gini and top-share
quantify concentration, and would be recomputed on
clustered entities before any security claim. The
mean interval and the KS p-value report whether the
block-arrival model holds on this snapshot. The
forecast count and Brier score keep the analyst
accountable. No single number is a verdict; together
they are a defensible, reproducible summary that a
reader can rerun and challenge. That reproducibility,
every figure regenerated from the code beside it, is
the standard the book has tried to hold throughout.

### The public-health template

This audit is, line for line, the audit a
regulator would run against the trial ledger, or
against the vaccine cold-chain provenance log. We
verify the integrity commitments, the hash chain of
@sec-intro and @sec-crypto, so that the record is
known to be tamper-evident. We measure concentration
and anomalies, asking whether a handful of sites or
actors are responsible for a suspicious share of
events. We model the arrival process and report
where it departs from expectation, for an
unexpected burst of enrolments may be data-entry
error or fraud, and an interrupted cold-chain
handoff may be a spoiled lot. And we score any
predictions the monitoring committee makes, so that
oversight is itself calibrated. The statistical
toolkit is identical because the underlying question
is identical: can this append-only record be
trusted, and by how much? A blockchain and a
well-kept audit log are two implementations of the
same integrity guarantee, and, as @sec-intro argued,
the simpler implementation is usually the right one.

## The right to be forgotten meets the append-only ledger

One tension deserves a section of its own, because it is
the point at which this book's subject matter collides
most directly with the law that governs health data. The
whole enterprise of the preceding chapters has been to
build a record that cannot be altered or deleted. Yet
data-protection law presumes the opposite. The European
Union's General Data Protection Regulation grants a right
to erasure, and the spirit of health-privacy regimes such
as the United States' HIPAA presumes throughout that
personal data can be corrected, restricted, and in the end
destroyed. An immutable ledger is engineered precisely so
that it cannot honour such a request. Finck states the
collision plainly [@finck2018gdpr]: a technology whose
value proposition is that nothing can be removed sits
uneasily with a body of law whose premise is that the
individual may demand removal.

But surely, one might object, a hash or a ciphertext is
not personal data, so why not commit those to the chain
and keep the plaintext elsewhere? The objection does not
survive the machinery of @sec-crypto. A hash of a
low-cardinality identifier, a national health number or a
date of birth, is not anonymisation: it is a commitment
that yields to precisely the dictionary attack we ran
against the avalanche property, since an adversary who can
enumerate the possible inputs can simply hash them all and
match. Encryption is stronger, but a key is a secret that
can leak, be compelled, or be broken by an advance in
capability, and a ledger designed to outlive its
participants will still be holding the ciphertext when
that day comes. The prudent reading is that anything
committed to an immutable ledger should be treated as
permanently and potentially publicly disclosed.

The defensible pattern that follows is the one this book
has stated wherever the question has arisen: personal
health data stays off-chain, in a governed store that can
honour an erasure request, and the ledger commits only to
non-identifying proofs, a Merkle root (a single hash that
commits to an entire dataset) of a locked dataset,
a signature over an access grant, a timestamp of an event.
The chain then attests that something happened without
recording what, to whom, or containing any datum a
regulator could later order destroyed.

::: {.callout-warning}
## What may go on the ledger

Under a health-privacy regime, treat the on-chain and
off-chain boundary as a hard rule rather than a
preference. On-chain: hashes and Merkle roots of locked
records, commitments, signatures, and pointers.
Off-chain, in an erasable governed store: every datum that
identifies a person, and every field a subject could ask
to correct or delete. A hash of an identifier is not a
safe harbour; it belongs off-chain with the identifier it
commits to.
:::

We do not claim the tension is resolved. Reconciling
immutability with a right to erasure is an unsettled
research question, and the proposed answers, redactable
chains built on chameleon hashes that permit an authorised
edit, or off-chain storage with on-chain pointers that can
be severed, each carry costs and each weaken the very
immutability that motivated the ledger. The honest
position for a public-health practitioner is that the law
here is a binding design constraint and not an
afterthought, and that a design which cannot answer 'how
would this honour an erasure request?' is not yet
finished.

## Collaborating with an LLM

On-chain analysis is a domain where a language model
accelerates the routine and endangers the
inferential. Use it for the first and guard the
second.

**Prompt.** 'Write R to compute the Gini
coefficient and the top-1% share of a vector of
account balances, and draw the Lorenz curve, using
only base R and ggplot2.'

*Watch for:* a solution that pulls in `ineq` or
`DescTools` when the closed form is four lines; an
off-by-one in the rank weighting of the Gini
formula; a Lorenz curve that omits the origin point
and so understates the area.

*Verification:* check the Gini against a known case
(perfect equality gives 0, one holder gives near 1)
and confirm the Lorenz curve starts at (0, 0) and
ends at (1, 1).

**Prompt.** 'Here is a description of a token's
staking rewards. Is the yield sustainable?'

*Watch for:* a confident narrative verdict with no
flow accounting. The model will happily echo the
project's own framing.

*Verification:* insist on the operational test.
Trace the yield to an exogenous or endogenous
source, and demand the on-chain quantity that would
distinguish them. A verdict without a funding
decomposition is a tier-four opinion.

**Prompt.** 'Fit an exponential to these block
intervals and tell me if it is a good model.'

*Watch for:* a single KS p-value reported as a
verdict, with no attention to serial dependence or
to the retarget and hash-shock regimes where the
model is expected to fail.

*Verification:* require the QQ plot and an explicit
statement of where the fit is weakest. A high
p-value on the pooled data is consistent with severe
misfit in exactly the stress episodes that matter.

## Exercises

1. **Estimand audit.** Take a published on-chain
   metric (daily active addresses, TVL, or realised
   capitalisation) and write one paragraph naming its
   estimand, the ledger quantity actually measured,
   and the selection, confounding, and measurement
   gaps between them.

2. **Gini sanity checks (quantitative).** Prove that
   the closed-form `gini_coef` returns 0 for a
   perfectly equal population and approaches
   $(n-1)/n$ when one holder owns everything. Verify
   both limits numerically, and confirm the estimator
   is scale-invariant.

3. **Lorenz and concentration (quantitative).**
   Simulate balances from a Pareto distribution with
   tail index $\alpha \in \{1.5, 2.0, 2.5\}$. For each,
   compute the Gini and the top-1% share and overlay
   the three Lorenz curves. Relate the direction of
   the change in Gini to the change in $\alpha$, and
   explain the mechanism.

4. **Entity versus address Gini (quantitative).**
   Extend the union-find clustering to attach a
   balance to each address, then compute the Gini on
   addresses and on the resulting entities for the
   same synthetic chain. Construct one scenario where
   clustering raises the Gini and one where it lowers
   it, and state which is the security-relevant number.

5. **Where the exponential breaks (quantitative).**
   Simulate block intervals with a rate that steps up
   by 50% halfway through the series (a hash-rate
   shock). Fit a single exponential to the pooled
   data, report the KS p-value, and show via the QQ
   plot and a plot of intervals over time that the
   pooled test can fail to detect a regime it visibly
   contains.

6. **Ponzi flow test.** Design a schema for the
   minimal on-chain quantities you would need to
   compute the fraction of a protocol's yield that
   originates from exogenous revenue. State the
   decision rule that would classify the yield as
   Ponzi-like, and name one way the schema could be
   gamed.

## Further reading

@meiklejohn2013fistful is the foundational
de-anonymisation paper: read it for the heuristics
and for the demonstration that pseudonymity degrades
under patient analysis. @kondor2014rich gives the
empirical wealth-concentration study whose Lorenz and
Gini analysis this chapter reproduces on synthetic
data. @briola2023anatomy anatomise a stablecoin
collapse with on-chain and microstructure data, the
empirical case for the reflexive-peg fragility
discussed here.

For quantitative treatments of the wider field,
@goffard2023blockastics collects probabilistic and
statistical models of blockchain phenomena, and the
two systematisation-of-knowledge papers,
@bonneau2015sok on Bitcoin and cryptocurrencies and
@werner2022sok on decentralised finance, map the
research landscape and its open problems with more
rigour than any promotional source. @narayanan2016bitcoin
remains the reference textbook, and its final
chapters on anonymity and community are the natural
sequel to this one.

## Synthesis: achievements, theatre, and calibration

We began the book with a claim that a blockchain is
an append-only ledger that mutually distrusting
parties agree on without a trusted intermediary, and
that this one sentence contains the entire subject.
Ten weeks later the sentence has been made
quantitative at every clause. The consensus is a
random walk and a cost-of-attack calculation. The
cryptography is a measurable set of statistical
properties. The economics is a collection of control
systems and feedback loops. And the data, as this
chapter has shown, is an observational dataset to be
analysed with the ordinary and powerful instruments
of the trade, the same instruments we bring to a
registry or a trial.

What remains is judgement, and judgement divides the
genuine from the theatrical. The genuine achievement
is real and difficult, namely open-membership
Byzantine agreement without a trusted party, a
problem @lamport1982byzantine posed and
@nakamoto2008bitcoin answered, delivering a
tamper-evident record whose integrity a health
scientist has good reason to want. The theatre is
also real, in the unfalsifiable valuations, the
metrics engineered to impress, and the reflexive
schemes that dress a pyramid in the vocabulary of
finance. On the one hand a credulous enthusiasm
mistakes the theatre for the achievement; on the
other, a reflexive dismissal throws out the
achievement with the theatre. A calibrated analyst
holds both without collapsing into either, and the
instruments that keep the two apart are the ones on
which this chapter and this book have ended.

In conclusion, three points are to be emphasized.
First, a public chain is an observational dataset
and must be read as one, with the estimand named
before the estimator and the selection, confounding,
and measurement gaps between the ledger and the
latent quantity stated in the open; every instrument
in this chapter, from the Gini coefficient to the
capture-recapture logic of address clustering, has a
public-health twin, and the honesty demanded of a
registry analysis is demanded here without
relaxation. Second, pseudonymity is not anonymity,
neither on a chain nor in a registry, and because an
append-only store never forgets, the re-identification
risk of pseudonymised health data written to a
public ledger grows rather than decays with time;
this is a decisive argument for keeping patient-level
records off any public chain and for confining the
trial ledger to integrity commitments alone. Third,
scepticism that is never recorded is merely a
posture, and the remedy is to name the falsifiability
tier of every claim, to operationalise each narrative
into a test with a rejection region, and to keep a
prospectively scored forecasting log so that the
analyst's own judgement is calibrated against outcomes
by the same Brier score demanded of the epidemic
forecaster and the promoter alike. A blockchain and a
well-kept audit log are, in the end, two
implementations of one integrity guarantee, and, as
we argued in @sec-intro, the simpler implementation is
usually the right one. That principle, and not any
enthusiasm for the technology, is what we hope the
reader carries forward.
